### 2017-09-20 ms.Rmd
init commit

---
title: "Museum specimens as baseline for assessing a potential global pollinator crisis"
author: "I. Bartomeus & J. Stavert"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#req: 6000 words max. 3 figures
#deadline: 1st April
```

##Anstract

## Intro

What we know, geographical bias, ...

## Case study one: Spain

## Case study two: NZ

## Joint efforts needed at the global scale

highlight digitalyzing initiatives including AMNH, LondonMNH, ¿? Highlight gaps that will never be filled probably.

### 2017-09-20 ms.Rmd
Add quick abstract

@@
 
+Building on previous studies showing the use of historical museum specimens to compare historical bee occurrences in northeast USA, Central Europe, and South Africa with current occurrence patterns, we show the potential for Museum specimens to play a key role in elucidating the extent of the pollinator crisis and compare the responses obtained along different environmental situations.
+Interestingly, we will add data from New Zealand and Spain increasing the geographical coverage and including bee diversity hotspots like Spain, for which bee declines have never been assessed.
+
 ## Intro

### 2017-10-05 NZexploration_bees.R
Add first pass at NZ data. Probably outdated

@@ -1,0 +1,286 @@
+#load data
+
+d <- read.csv("data/NZbees_raw.csv", header = TRUE, sep = ",")
+str(d)
+
+levels(d$Species) #need to correct space
+#d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
+d$Gen_sp <- d$Species
+
+sort(table(d$Gen_sp))
+sum(table(d$Gen_sp))
+head(d)
+
+#date
+date2 <- strptime(d$Date, "%d/%m/%Y") #several formats mixed in... will be painfull
+d$jday <- date2$yday
+d$year <- 1900 + date2$year
+d$month <- date2$mon + 1
+head(d$month)
+
+#select fields to be independent
+uni <- as.data.frame(cbind(d$Gen_sp, d$Locality, d$Region, d$Date))
+str(uni)
+dup <- duplicated(uni)
+
+# remove duplicates
+d2 <- d[-which(dup == TRUE),]
+str(d2)
+sort(table(d2$Gen_sp)) #14 species with more than 30 datapoints
+
+#visualize the data: #not with bees
+#plot(d2$Long, d2$Lat)
+#library(maps)
+#map("world", col="red", add=TRUE)
+
+#plot(d2$Long, d2$Lat, ylim = c(-50,-32), xlim = c(163,183))
+#map("world", col="red", add=TRUE) #fix wrong lat/long
+
+
+#create time periods
+d4 <- d2
+#for now
+d4$year <- d4$year+1900
+d4$time_period50  <- cut(d4$year, breaks = seq(1875,2025,50), labels = c("1875-1925","1925-1975","1975-2011"))
+
+d4$time_period25  <- cut(d4$year, breaks = seq(1875,2025,25), labels = c("1875-1900","1900-1925","1925-1950","1950-1975","1975-2000","2000-2011"))
+levels(d4$time_period25)
+
+#d4$time_period_custom  <- cut(d4$year, breaks = c(1875,1945,1985,2025), labels = c("1875-1945","1950-1985","1985-2012"))
+
+i = 3
+break1 <- quantile(d4$year, probs = seq(0,1,1/i), na.rm = TRUE)
+lab <- c(1:i)
+for(j in 1:i){
+    lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
+}
+d4$time_period_custom  <- cut(d4$year, breaks = break1, labels = lab)
[…]

### 2017-10-05 NZexploration_flies.R

@@ -1,0 +1,298 @@
+#load data
+
+d <- read.csv(file = "NZ_fly_pollinators.csv", header = TRUE, sep = ",")
+str(d)
+
+
+d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
+
+sort(table(d$Gen_sp))
+sum(table(d$Gen_sp))
+head(d)
+
+#date
+date2<- strptime(d$CollectionDateISOFrom, "%Y\\%m\\%d")
+d$jday <- date2$yday
+d$year <- 1900 + date2$year
+d$month <- date2$mon + 1
+head(month)
+
+#select fields to be independent
+#two approaches lat = 0 and others
+lat <- subset(d, Lat != 0 & is.na(year) == F)
+str(lat)
+uni <- as.data.frame(cbind(lat$Gen_sp, lat$Lat, lat$Long, lat$jday, lat$year))
+str(uni)
+dup <- duplicated(uni)
+# lat/long = 0
+#lat0 <- subset(d, Dlat == 0 & is.na(year) == F)
+#str(lat0)
+#uni2 <- as.data.frame(cbind(lat0$Gen, lat0$sp, lat0$LocalityStr, lat0$SubDivStr, lat0$StateProv, lat0$Country, lat0$jday, lat0$year))
+#str(uni2)
+#dup2 <- duplicated(uni2)
+
+# remove duplicates
+lat <- lat[-which(dup == TRUE),]
+#lat0 <- lat0[-which(dup2 == TRUE),]
+#d2 <- rbind(lat,lat0)
+d2 <- lat
+str(d2)
+sort(table(d2$Gen_sp)) #14 species with more than 30 datapoints
+
+#visualize the data:
+plot(d2$Long, d2$Lat)
+library(maps)
+map("world", col="red", add=TRUE)
+
+plot(d2$Long, d2$Lat, ylim = c(-50,-32), xlim = c(163,183))
+map("world", col="red", add=TRUE) #fix wrong lat/long
+
+
+#create time periods
+d4 <- d2
+d4$time_period50  <- cut(d4$year, breaks = seq(1875,2025,50), labels = c("1875-1925","1925-1975","1975-2011"))
+
+d4$time_period25  <- cut(d4$year, breaks = seq(1875,2025,25), labels = c("1875-1900","1900-1925","1925-1950","1950-1975","1975-2000","2000-2011"))
+levels(d4$time_period25)
+
[…]

### 2017-10-05 SP_data_cleaning.Rmd
Add spanish data and initial exploration

---
title: "SP"
author: "I. Bartomeus"
date: "05/10/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Notas:

- 'Localidades?': Info per species. Recorded only at the first record of each species. Refers to geograpfic known records. Usefull for abejassilvestres.es ?
- 'obrera' needs to be merged with 'sex'
- ID and 'cuaderno' need to remove duplicates, which are not in cuaderno!!
- IF several IT's Oscar Aguade duplicated the netry as many times as need. Those sp are in red in original excel.

Old data:

```{r}
sp <- read.csv(file = "raw_data/Spain.csv", h = T)
head(sp)

```

New data:

- Not entered yet.

### 2017-10-05 ms.Rmd
Add scheleton to ms with Jamie

@@
 title: "Museum specimens as baseline for assessing a potential global pollinator crisis"
-author: "I. Bartomeus & J. Stavert"
+author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
 output: html_document
@@
 knitr::opts_chunk$set(echo = TRUE)
-#req: 6000 words max. 3 figures
+#req: 6000 words max. inc ref's 3 figures
 #deadline: 1st April
@@
 
-##Anstract
+##Abstract
 
 Building on previous studies showing the use of historical museum specimens to compare historical bee occurrences in northeast USA, Central Europe, and South Africa with current occurrence patterns, we show the potential for Museum specimens to play a key role in elucidating the extent of the pollinator crisis and compare the responses obtained along different environmental situations.
+Highlight knowledge gaps and oportunities (e.g.
+London museum, zooniverse...).
 Interestingly, we will add data from New Zealand and Spain increasing the geographical coverage and including bee diversity hotspots like Spain, for which bee declines have never been assessed.
@@
 
-What we know, geographical bias, ...
+Paragraph on pollinator importance and claims of a crisis.
+initial claims driving to NAS 2007.
+Goulson Science, Gonzalez -varo TREE, Potts TREE.
+BUT highlight not global assessment and doubts (Gazoul "buzzines as usual", Forrest or Rafferty Oikos paper on geograpfical bias).
+Taxonomic bias?
+
+Define trends: Decline, turnover (changes in evennes winners loosers), exotics...
+Review current methods to assess 1) space for time (recent reviews rae, ...), 2) monitoring schemes /citicen science (iNatralis) (UK) and 3) museums type data.
+Highlight its importance (cite Primak).
+Pros and cons with an enphasis on Museums.
+
+##Current evidence:
+- US: Bartomeus 2013 and Burkle 201?, ... NAS book Critizize Burkle dessign and compare to mine.
+- EU: Beijsmier, Carvalho, Schepper, ... UK? Peter Keil flies? Moths? Buterflies!
+- South Africa: Pauw, maybe with plants...
+- Specific examples e.g. Bombus dalbhomi (Aizen), ... Thomson ecology letters
+
+##Other uses of museums in pollinators. Identifying mechanisms of decline
+- Oliveira plos one 
+- Osorio Ecol Lettres
+- Phenoloy (Bartomeus)
+
+## Future needs /knowledge gaps:
+- Digitalization of MORE museums (coordinated)
+- Integration of DATA and analysis:
+- Communication Museums- Sci (biderictional. Importance of voucering specimens fused for science).
+- Compare global differences geographically (Eu Z US, for example)
 
@@
 
-## Joint efforts needed at the global scale
+## Conclusions: Joint efforts needed at the global scale, scit science, ...
 
-highlight digitalyzing initiatives including AMNH, LondonMNH, ¿? Highlight gaps that will never be filled probably.
+highlight digitalyzing initiatives including AMNH, LondonMNH, ¿? Highlight gaps that will never be filled probably (central africa).

### 2018-02-02 NZexploration_bees.R
new commit

@@ -1,6 +1,7 @@
 #load data
 
 d <- read.csv("data/NZbees_raw.csv", header = TRUE, sep = ",")
+NZAC <- read.csv("data/NZbees_NZAC.csv", header = TRUE, sep = ",")
 str(d)
 
 levels(d$Species) #need to correct space

### 2018-02-02 ms.Rmd

@@
 
+Statements about pollinator crisis, or bee declines are widespread, but information is currenti limiting....
 Building on previous studies showing the use of historical museum specimens to compare historical bee occurrences in northeast USA, Central Europe, and South Africa with current occurrence patterns, we show the potential for Museum specimens to play a key role in elucidating the extent of the pollinator crisis and compare the responses obtained along different environmental situations.
@@
 - Specific examples e.g. Bombus dalbhomi (Aizen), ... Thomson ecology letters
+ Fig 1: a global map showing evidence (% decline in richness) and Data deficient areas would be great.
 
@@
 ## Future needs /knowledge gaps:
-- Digitalization of MORE museums (coordinated)
+- Digitalization of MORE museums (coordinated) Just common protocols? RopenSci? What about an interactive bee decline map? Predicts?
 - Integration of DATA and analysis:
@@
 highlight digitalyzing initiatives including AMNH, LondonMNH, ¿? Highlight gaps that will never be filled probably (central africa).
+
+#Acknowledgements
+
+E. Asensio (or maybe coauthor?). Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum, ITACyL.

### 2018-02-02 NZexploration_bees.R
updated pollinator decline references

@@ -1,0 +1,287 @@
+#load data
+
+d <- read.csv("data/NZbees_raw.csv", header = TRUE, sep = ",")
+NZAC <- read.csv("data/NZbees_NZAC.csv", header = TRUE, sep = ",")
+str(d)
+
+levels(d$Species) #need to correct space
+#d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
+d$Gen_sp <- d$Species
+
+sort(table(d$Gen_sp))
+sum(table(d$Gen_sp))
+head(d)
+
+#date
+date2 <- strptime(d$Date, "%d/%m/%Y") #several formats mixed in... will be painfull
+d$jday <- date2$yday
+d$year <- 1900 + date2$year
+d$month <- date2$mon + 1
+head(d$month)
+
+#select fields to be independent
+uni <- as.data.frame(cbind(d$Gen_sp, d$Locality, d$Region, d$Date))
+str(uni)
+dup <- duplicated(uni)
+
+# remove duplicates
+d2 <- d[-which(dup == TRUE),]
+str(d2)
+sort(table(d2$Gen_sp)) #14 species with more than 30 datapoints
+
+#visualize the data: #not with bees
+#plot(d2$Long, d2$Lat)
+#library(maps)
+#map("world", col="red", add=TRUE)
+
+#plot(d2$Long, d2$Lat, ylim = c(-50,-32), xlim = c(163,183))
+#map("world", col="red", add=TRUE) #fix wrong lat/long
+
+
+#create time periods
+d4 <- d2
+#for now
+d4$year <- d4$year+1900
+d4$time_period50  <- cut(d4$year, breaks = seq(1875,2025,50), labels = c("1875-1925","1925-1975","1975-2011"))
+
+d4$time_period25  <- cut(d4$year, breaks = seq(1875,2025,25), labels = c("1875-1900","1900-1925","1925-1950","1950-1975","1975-2000","2000-2011"))
+levels(d4$time_period25)
+
+#d4$time_period_custom  <- cut(d4$year, breaks = c(1875,1945,1985,2025), labels = c("1875-1945","1950-1985","1985-2012"))
+
+i = 3
+break1 <- quantile(d4$year, probs = seq(0,1,1/i), na.rm = TRUE)
+lab <- c(1:i)
+for(j in 1:i){
+    lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
+}
[…]

### 2018-02-05 ms.Rmd
Add more ideas and figures

@@
 Define trends: Decline, turnover (changes in evennes winners loosers), exotics...
-Review current methods to assess 1) space for time (recent reviews rae, ...), 2) monitoring schemes /citicen science (iNatralis) (UK) and 3) museums type data.
+Review current methods to assess 1) space for time (recent reviews rae, ...), 2) monitoring schemes /citicen science (iNaturalis) (UK) and 3) museums type data.
 Highlight its importance (cite Primak).
@@
 - Specific examples e.g. Bombus dalbhomi (Aizen), ... Thomson ecology letters
+
+We compiled current evidence worldwide and show that ... (short methods included in text)
+
+```{r map, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
+library(rworldmap)
+map <- data.frame(iso2 = c("ES", "NZ"), richness = c(88, 75))
+#FAKE data, we need to compile a bbdd with the data at hand.
+map2 <- joinCountryData2Map(map, joinCode="ISO2", nameJoinColumn = "iso2", nameCountryColumn = "distribution")
+mapCountryData(mapToPlot = map2, numCats = 2, nameColumnToPlot = "richness"
+               , mapTitle = "Bee richness decline", addLegend = FALSE)
+
+```
+
 Fig 1: a global map showing evidence (% decline in richness) and Data deficient areas would be great.
 
-##Other uses of museums in pollinators. Identifying mechanisms of decline
+## Beyond species occurrences
+Other uses of museums in pollinators. Identifying mechanisms of decline
 - Oliveira plos one 
@@
 - Phenoloy (Bartomeus)
+
+## Prioritizing the low hanging fruit.
+
+Analyze trade off between number of records in Gbif for > 1980 and pre <1980. Each quadrant needs an analysis. 
++ we provide code to run it, and maybe add it to the repo per country? distributed database style?
+We do it for two regions:
+
+```{r low hanging fruits, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
+library(rgbif)
+#spain_code <- isocodes[grep("Spain", isocodes$name), "code"]
+#portugal_code <- isocodes[grep("Portugal", isocodes$name), "code"]
+apidae_key <- name_backbone(name="Apidae", rank = "family")$usageKey
+andrenidae_key <- name_backbone(name="Andrenidae", rank = "family")$usageKey
+halictidae_key <- name_backbone(name="Halictidae", rank = "family")$usageKey
+colletidae_key <- name_backbone(name="Colletidae", rank = "family")$usageKey
+megachilidae_key <- name_backbone(name="Megachilidae", rank = "family")$usageKey
+stenotritidae_key <- name_backbone(name="Stenotritidae", rank = "family")$usageKey
+melittidae_key <- name_backbone(name="Melittidae", rank = "family")$usageKey
+
+keys <- c(apidae_key, andrenidae_key, halictidae_key, halictidae_key, colletidae_key,
+          megachilidae_key, stenotritidae_key, melittidae_key)
+#loop through all families, through all countries and store in datarame
+country_codes <- c("ES", "NZ", "PT", "GE") #need to fetch them all EASY using rgbif
+lhf <- matrix(ncol = 1+(7*2), nrow = length(country_codes))
+lhf <- as.data.frame(lhf)
+colnames(lhf) <- c("country", "apidaepre1980", "andrenidaepre1980", "halictidaepre1980", "colletidaepre1980", 
+                    "megachilidaepre1980", "stenotritidaepre1980", "melittidaepre1980",
+                   "apidaepost1980", "andrenidaepost1980", "halictidaepost1980", "colletidaepost1980", 
+                    "megachilidaepost1980", "stenotritidaepost1980", "melittidaepost1980")
+for(i in 1:length(country_codes)){
+    for(j in 1:length(keys)){
+        lhf[i,1] <- country_codes[i]
+        lhf[i,1+j] <- as.numeric(occ_count(taxonKey= keys[j],
+                            from = 1980, to = 2018,
+                            georeferenced=TRUE,
+                            country = country_codes[i]))
+        lhf[i,1+j+7] <- as.numeric(occ_count(taxonKey= keys[j],
+                            from = 1850, to = 1980,
+                            georeferenced=TRUE,
+                            country = country_codes[i]))
+    }
+}
+#lhf
+plot(rowSums(lhf[,2:8]), rowSums(lhf[,9:15]), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
+abline(h = 3000, v = 5000)
+#plot needs legend by country, and to be ggplotted
+
+```
+
+Fig 2: number of bee occurrences pre 1980 and post 1980 in gbif.
+Upper rigth cuadrat are well covered countries, upper left cuadrat are countries that need resurvey of old sites.
+Lower right are sites which needs to digg museum records and lower left are countries hard to get data.
+
+We ilustrate how this can be done using two countries and providing code to reproduce our analysis, so that more countries can contribute.
+
+## Case study one: Spain
+
+- Describe data and analysis.
+- Present main results
+
+## Case study two: NZ
+
+- Describe data and analysis.
+- Present main results
 
@@
 - Compare global differences geographically (Eu Z US, for example)
-
-## Case study one: Spain
-
-## Case study two: NZ
 

### 2018-02-05 ms.Rmd
Refine potential figures

--- 2018-02-02 ms.Rmd
@@
 - Specific examples e.g. Bombus dalbhomi (Aizen), ... Thomson ecology letters
+
+We compiled current evidence worldwide and show that ... (short methods included in text)
+
+```{r map, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
+library(rworldmap)
+map <- read.csv("data/bee_declines.csv", sep = ";")
+#Jamie, if you want, we could complete this data set based on the papers.
+# If two estimates available for the same country, I will use two rows. 
+# We can compute the mean or the max using reshape here.
+# For studies reporting large regions within country this is ok, I think, but
+# for studies reporting locations (like Burkle Science paper) the % richness loss would be inflated when scaling up... not sure how to add those up. Add a column maybe saying coverage
+# country, region, individual sites.
+#Bee richness can be gathered from Discover life webpage
+#head(map)
+map2 <- joinCountryData2Map(map, joinCode="NAME", nameJoinColumn = "country",
+                            nameCountryColumn = "distribution")
+#need to use ISO's to make sure we don't carry over NA's due to country misspellings
+mapCountryData(mapToPlot = map2, nameColumnToPlot = "bee_richness"
+               , mapTitle = "Bee richness", addLegend = FALSE)
+
+```
+
 Fig 1: a global map showing evidence (% decline in richness) and Data deficient areas would be great.
 
-##Other uses of museums in pollinators. Identifying mechanisms of decline
+## Beyond species occurrences
+Other uses of museums in pollinators. Identifying mechanisms of decline
 - Oliveira plos one 
@@
 - Phenoloy (Bartomeus)
+
+## Prioritizing the low hanging fruit.
+
+Analyze trade off between number of records in Gbif for > 1980 and pre <1980. Each quadrant needs an analysis. 
++ we provide code to run it, and maybe add it to the repo per country? distributed database style?
+We do it for two regions:
+
+```{r low hanging fruits, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
+library(rgbif)
+apidae_key <- name_backbone(name="Apidae", rank = "family")$usageKey
+andrenidae_key <- name_backbone(name="Andrenidae", rank = "family")$usageKey
+halictidae_key <- name_backbone(name="Halictidae", rank = "family")$usageKey
+colletidae_key <- name_backbone(name="Colletidae", rank = "family")$usageKey
+megachilidae_key <- name_backbone(name="Megachilidae", rank = "family")$usageKey
+stenotritidae_key <- name_backbone(name="Stenotritidae", rank = "family")$usageKey
+melittidae_key <- name_backbone(name="Melittidae", rank = "family")$usageKey
+
+keys <- c(apidae_key, andrenidae_key, halictidae_key, halictidae_key, colletidae_key,
+          megachilidae_key, stenotritidae_key, melittidae_key)
+#loop through all families, through all countries and store in datarame
+country_codes <- isocodes$code 
+lhf <- matrix(ncol = 1+(7*2), nrow = length(country_codes))
+lhf <- as.data.frame(lhf)
+colnames(lhf) <- c("country", "apidaepre1980", "andrenidaepre1980", "halictidaepre1980", "colletidaepre1980", 
+                    "megachilidaepre1980", "stenotritidaepre1980", "melittidaepre1980",
+                   "apidaepost1980", "andrenidaepost1980", "halictidaepost1980", "colletidaepost1980", 
+                    "megachilidaepost1980", "stenotritidaepost1980", "melittidaepost1980")
+for(i in 1:length(country_codes)){
+    for(j in 1:length(keys)){
+        lhf[i,1] <- country_codes[i]
+        lhf[i,1+j] <- as.numeric(occ_count(taxonKey= keys[j],
+                            from = 1980, to = 2018,
+                            georeferenced=TRUE,
+                            country = country_codes[i]))
+        lhf[i,1+j+7] <- as.numeric(occ_count(taxonKey= keys[j],
+                            from = 1850, to = 1980,
+                            georeferenced=TRUE,
+                            country = country_codes[i]))
+    }
+}
+#very slow, we should save the data and load here when needed.
+lhf$post1980 <- rowSums(lhf[,2:8])
+lhf$pre1980 <- rowSums(lhf[,9:15])
+write.csv(lhf, "data/rgbif_output.csv")
+```
+
+```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
+lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
+#lhf
+#head(lhf)
+#str(lhf)
+lhf[order(lhf$pre1980),c(1,17,18)]
+plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
+abline(h = 150000, v = 150000)
+#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.
+
+```
+
+Fig 2: number of bee occurrences pre 1980 and post 1980 in gbif.
+Upper rigth cuadrat are well covered countries, upper left cuadrat are countries that need resurvey of old sites.
+Lower right are sites which needs to digg museum records and lower left are countries hard to get data.
+
+I am not sure the 1980 cut is apropiate. Looks like it is too correlated. Maybe we should correct per country size? Or per total records? % post vs % pre.
+
+```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
+lhf$percent_post1980 <- lhf$post1980 / (lhf$pre1980+lhf$post1980)
+lhf$percent_pre1980 <- lhf$pre1980 / (lhf$pre1980+lhf$post1980)
+
+plot(lhf$percent_post1980, lhf$percent_pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,1), ylim = c(0, 1))
+abline(h = 0.50, v = 0.50)
+```
+This is even worst... correcting for size would be the best BUT I don't think richness and size have a linear relationship, so we should account for that then??
+
+We ilustrate how this can be done using two countries and providing code to reproduce our analysis, so that more countries can contribute.
+
+## Case study one: Spain
+
+- Describe data and analysis.
+- Present main results
+
+## Case study two: NZ
+
+- Describe data and analysis.
+- Present main results
 

### 2018-02-05 ms.Rmd
Try logs on Fig 2

@@
 abline(h = 0.50, v = 0.50)
+
+plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
+abline(h = 8, v = 8)
+
 ```
 This is even worst... correcting for size would be the best BUT I don't think richness and size have a linear relationship, so we should account for that then??
+Simply logging the data may work, we can point out % of countries with > 1000 records in either axes, for example, etc...
 

### 2018-02-06 ms.Rmd
Wrtoe the firs draft of the paper... probably is very weak, but its a start.

@@
 
-Statements about pollinator crisis, or bee declines are widespread, but information is currenti limiting....
-Building on previous studies showing the use of historical museum specimens to compare historical bee occurrences in northeast USA, Central Europe, and South Africa with current occurrence patterns, we show the potential for Museum specimens to play a key role in elucidating the extent of the pollinator crisis and compare the responses obtained along different environmental situations.
-Highlight knowledge gaps and oportunities (e.g.
-London museum, zooniverse...).
-Interestingly, we will add data from New Zealand and Spain increasing the geographical coverage and including bee diversity hotspots like Spain, for which bee declines have never been assessed.
+There is an ever-increasing concern about the current decline of pollinators, specifically bees.
+Despite pollinator decline reports are widespread, the information available is scarce and biased geographically and taxonomically, limiting the inference about the a potential pollination crisis.
+The most direct form for understanding species decline patterns is to analyze historical time series, however, those are rare or often non-existing for most pollinator groups.
+However, museum collections can provide the only source of historical data serving as a baseline to understand the reach of pollinator declines.
+Specimens historically collected by naturalists and preserved in museums can provide information on when and where species were collected, as well as other ecological information like historical species interactions, or morphological traits.
+We show here how researchers have used historical museum specimens to compare historical occurrences with current occurrence patterns to show changes over time in biodiversity levels, species relative abundances, body size and even pollination services delivery.
+Despite this advances, we show that we still miss information for the status and trends of most pollinators and that often a bottle neck is digitizing the historical data.
+We analyze the opportunities and limitations to do so worldwide and provide two worked examples on how to analyze this type of data for two distinct regions for which historical bee declines have never been assessed, New Zealand and Spain.
+There is a great potential for Museum specimens to play a key role in elucidating the extent of the pollinator crisis and compare the responses obtained along different environmental situations if we succeed in unleashing the power of historical data.
 
-## Intro
+## Introduction
 
-Paragraph on pollinator importance and claims of a crisis.
-initial claims driving to NAS 2007.
-Goulson Science, Gonzalez -varo TREE, Potts TREE.
-BUT highlight not global assessment and doubts (Gazoul "buzzines as usual", Forrest or Rafferty Oikos paper on geograpfical bias).
-Taxonomic bias?
+Pollinators are of paramount importance given its prominent role in plant reproduction (@ollerton2012) and food security (@kleinetal2007).
+As many other taxa, pollinators are also exposed to a plethora of human induced rapid environmental changes potentially threatening its populations.
+This conundrum was realized early on by researchers, and was popularized by the influential book "the forgotten pollinators" (1996).
+However, at that time there was almost no data on the trends and status of pollinators and this initial claims trigger the first efforts to elucidate the problem, including a commission of a US National Academy of Science (NAS) panel to assess the reach of pollinators declines (2006).
+The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult".
+Since then, the studies on pollinator responses to global change has multiplied and we reached a strong consensus that pressures like habitat destruction, land use change, chemical exposure, exotic species and global warming are the leading causes of pollinator declines and act sinergistically (@goulson2016, @potts2010).
+However, the status and trends of most pollinator species remains unknown.
+For example, for a well studied region like Europe, 60% [check number]() of the bee species assessed by a recent IUCN report fall into the category "Data Deficient" (@Nieto2014).
 
-Define trends: Decline, turnover (changes in evennes winners loosers), exotics...
-Review current methods to assess 1) space for time (recent reviews rae, ...), 2) monitoring schemes /citicen science (iNaturalis) (UK) and 3) museums type data.
-Highlight its importance (cite Primak).
-Pros and cons with an enphasis on Museums.
+One of the main barriers to profile pollinator population trends is that pollinators conform an heterogeneous group including insects such as bees, flies, butterflies and beetles, but also birds, bats and even lizards (@Winfreeetal2011).
+In particular bees, usually regarded as the more important in terms of abundance, efficiency and widespread distribution (@ollerton2018), are a diverse group with more than 20 000 species worldwide (@PickeringAscher) and most of them require expert taxonomic expertise for its identification.
+This has led not only to a biased geographic patterns due to the uneven distribution of researchers (@Martinetal2012), but also a taxonomic bias to species easier to identify, like bumblebees (@ArcherOikos).
 
-##Current evidence:
-- US: Bartomeus 2013 and Burkle 201?, ... NAS book Critizize Burkle dessign and compare to mine.
-- EU: Beijsmier, Carvalho, Schepper, ... UK? Peter Keil flies? Moths? Buterflies!
-- South Africa: Pauw, maybe with plants...
-- Specific examples e.g. Bombus dalbhomi (Aizen), ... Thomson ecology letters
+The second barrier is methodological and involves how to assess population trends for highly mobile organisms, which in addition are often  short-lived and small.
+A practical approach has been to use space for time substitutions, where the researchers compares natural areas with areas affected by different environmental conditions.
+Despite the critiques on the robustness of this approach (@ref), this is the best source of information, specially for Europe, where recent joint efforts had estimated bee richness declines at the country level using the predictions from such models (@DePalma).
+A second promising method is pollinator monitoring programs.
+Inspired in the successful butterfly monitoring programs (@ref), this protocols, often carried by citizen scientists, are being implemented on other taxa (@ref).
+The main limitation is that those programs need time to build up long term data series and do not reveal anything about the original conditions.
+Finally, the best source of historical data can be found in museums.
+The core aim of Museums is precisely to conserve and curate this kind of collections, and serve not only of a precious voucher of specimens, but often also assures its taxonomic identification quality (@primak).
+The major bottleneck to use this information, is however, digitizing the specimens.
+Efforts to do so have provide great benefits and allow the characterization of the status and trends of bee pollinators is the US and EU (e.g.
+Bartomeus et al 2013, Scheper et al 2014) and despite data deposited in museums is often collected opportunistically and have its own biases (@Keil), is the only data we have to reconstruct the past.
 
-We compiled current evidence worldwide and show that ... (short methods included in text)
+Regardless of the method, it's important to consider that biodiversity has many aspects beyond species richness, and we should put more effort in differentiating declines not only in local (alpha) diversity or regional (gamma) diversity, but also in the turnover over sites (beta), the abundance of key species, and the evenness of the communities (@winfree2016ELE).
+In fact, environmental change often give us few winners, and lots of loosers (@Bartomeusetal2013), and characterizing those is extremely important as the few winners are often exotic species (@Stavertetal2018).
+[Add something about traits of winners and loosers?]()
+
+## The use of museum specimens
+
+Digitizing old specimens is not a trivial task and it needs expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geolocate the coordinates of the collection event (@ref and software) and (iii) store the data in a proper curated database (@ref on database management).
+Doing this process one by one with each of the thousands of specimens located in a museum is a daunting task and needs specialized personnel.
+While some tasks can be only done by experts taxonomists, new technologies and citizen science can speed up the process of digitizing old collections.
+High resolution photos and scanning of the associated labels have been uploaded to the internet to get the power of distributed work where thousands of volunteers can transcript the images to text.
+This process is validated, etc...
+Examples include zooniverse, ...
+[Is there anything on AI to automate the process?
+e.g.
+there is something on wing venation from photos at Netherlands.
+Can be mentioned?]
+[how much do we want to extend on this? Is interesting info, but boring o read?]()
+
+##Current evidence on pollinatior declines
+
+Is not surprising the the best evidence of pollinator declines comes from the more disturbed countries.
+A pioneer study using citizen science and museum specimens in UK and Netherlands, two of the most populated and transformed countries showed a worrisome trend for both bees, flies and flowering plants, but specially for bees (@Beisjmieretal2006).
+However, it's important to note that even within those countries, the local estimates are all nearby big cities and agricultural areas, and data is lacking for better conserved areas.
+Further exploration of this data set shows that the declining trend has attenuated in recent decades (@carvalheiroetal).
+Detailed trends for common species in the Netherlands using museum specimens confirm the link between pollinator and plant declines, as bees wich preferred host plants have declined the most are the more threatened (@Schepper2014).
+[- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies!]()
+
+In comparison, regional estimates for the eastern US using museum specimens show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions as most species can still persist (@Colla2012?).
+This is in harsh contrast with the high local extinctions reported by Burkle (@Burkle) when comparing historical observations of a large forested ecosystem with the few remnants that still exists, but note that this comparison do not takes into account the opportunistic bees using other land use changes, and that the local richness may have changes, but not declined as drastically.
+In any case, there is a great concordance between species locally extinct and regionally declining (@Bartomeus_winfree2013).
+
+Outside of Europe and the US there is surprisingly little evidence of pollinator declines using historical records.
+[South Africa: Pauw, maybe with plants...]() .
+However, there are specific examples showing the historical decline of certain species e.g.
+Bombus dalbhomi (Aizen)
+
+[add % in text also]()
+
+We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species declines (b), syrphid species declines (c) and ??
+to highlight the little coverage we have...
 
@@
 mapCountryData(mapToPlot = map2, nameColumnToPlot = "bee_richness"
-               , mapTitle = "Bee richness", addLegend = FALSE)
+               , mapTitle = "Bee richness", addLegend = TRUE)
 
@@
 
-Fig 1: a global map showing evidence (% decline in richness) and Data deficient areas would be great.
+Fig 1: A global map showing evidence (% decline in richness) and Data deficient areas would be great.
 
 ## Beyond species occurrences
-Other uses of museums in pollinators. Identifying mechanisms of decline
-- Oliveira plos one 
-- Osorio Ecol Lettres
-- Phenoloy (Bartomeus)
 
-## Prioritizing the low hanging fruit.
+Museum can provide valuable information besides species occurrence records, and this is important to track when digitizing the collections.
+This is specially relevant to identify mechanisms of decline and adaptation.
+One particularly important source of information is the date of capture, which has allowed to track phenological advances congruent with contemporary climate change (@Bartomeusetal2011).
+In addition, as mentioned above, insect pollinator labels often include as information the host plant where collected.
+This information is precious to understand past and present species interactions (@tylianakis2010ELE).
+Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events (@kleinj_raemakers).
+Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length (@refAlaskaScience) and body size (@Oliveira) shrinkage with climate and land use change.
 
-Analyze trade off between number of records in Gbif for > 1980 and pre <1980. Each quadrant needs an analysis. 
-+ we provide code to run it, and maybe add it to the repo per country? distributed database style?
-We do it for two regions:
+## The way forward: Prioritizing the low hanging fruit.
+
+As illustrated in Figure 1, there is a paucity of countries for which museum data is available, and hence can be used as baseline for assessing pollinator population declines.
+While ideally, one would aim to digitize all museum records in the world, in the practical world this is an unlikely to happen due mainly to funding constrains.
+Here we illustrate how this can be prioritized and optimized to encourage independent researchers to assess declines in its own country.
+
+Gbif (@ref) is a central repository centralizing occurrence data.
+Much of this data comes from museum, but it integrates several other sources.
+Alongside the popular statistical language R (@ref), gbif can be directly queried into your computer (@rgbif) and check data availability for your region of interest.
+In Figure 2 we show the number of modern and historical bee records already available per country.
+While for some countries like UK and US the numbers are high in both axes, and hence, data is directly analyzable, most countries fall short in one or both of the axes.
+Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis.
+On the contrary, for countries with large number of historical records, organized re-surveys should be organized (e.g.
+@cameron2013).
+Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X...
+Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.
 
@@
 Fig 2: number of bee occurrences pre 1980 and post 1980 in gbif.
-Upper rigth cuadrat are well covered countries, upper left cuadrat are countries that need resurvey of old sites.
-Lower right are sites which needs to digg museum records and lower left are countries hard to get data.
+Upper right cuadrat are well covered countries, upper left cuadrat are countries that need resurvey of old sites.
+Lower right are sites which needs to dig museum records and lower left are countries hard to get data.
 
-I am not sure the 1980 cut is apropiate. Looks like it is too correlated. Maybe we should correct per country size? Or per total records? % post vs % pre.
-
+Alternative plots:  
 ```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
@@
 
-plot(lhf$percent_post1980, lhf$percent_pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,1), ylim = c(0, 1))
-abline(h = 0.50, v = 0.50)
+plot(sort(lhf$percent_post1980), xlab = "post1980", ylab = "%", col = as.factor(lhf$country))
+#Add country labels in x axes.
 
@@
 ```
-This is even worst... correcting for size would be the best BUT I don't think richness and size have a linear relationship, so we should account for that then??
+% makes it unidimensional.
 Simply logging the data may work, we can point out % of countries with > 1000 records in either axes, for example, etc...
+Correcting for country size would be the best BUT I don't think richness and size have a linear relationship, so we should account for that then?
 
-We ilustrate how this can be done using two countries and providing code to reproduce our analysis, so that more countries can contribute.
+Once the data sets are made available, the researcher needs to gauge very carefully the potential bias in data.
+We illustrate this process with two examples for Spain and New Zealand.
+In the first one, most historical data comes from a few locations and is collected by the same team of people.
+In this case, we contacted the original authors to understand their sampling protocol, and resurveyed the same sites 20 years later.
+The second case shows a diversity of collectors and sites.
+In this case, we should ensure that geographical and taxonomic coverage do not change systematically with time.
+For example, some collectors may target only certain taxa, biasing the sample.
+We provide annotated R scripts to treat this kind of data and show two relevant case studies adding data for two unique regions with contrasting histories and hope this resource will cheer people to add more data and analysis (include a repository where people can contribute?)
 
@@
 
-- Describe data and analysis.
-- Present main results
+Spain is an interesting study system because is has been transformed by human heavely, but is not as intensified as other EU countries.
+In addition, Spain is a hotspot of bee diversity and maintains an heterogeneous landscape.
+As illustrated in Figure 2, Spain has already digitalized a decent amount of occurrence data both historical, and recent.
+However, a visual inspection reveals that data points are clustered and historical and recent records do not match spatially, making them hard to compare.
+Among the old records, it is striking that most of them are located around Valladolid and collected by Enrique Asensio and his team.
+Unfortunately, there is no recent sampling in the area.
+After some investigations, we discover that Enrique sampled systematically 8 localities and that more data is available at the "Museo de historia Nacional" and other minor collections.
+An effort to digitize the remaining material, and a re-survey of the area provides the perfect data set to make a comparison before-after of this 8 sites.
+
+In brief, we check sampling completeness at both time periods, and compare the richness level of each site before 1980 and after 1980 with a paired t-test.
+[Maybe more analysis arise...]().
+We show that...
+
+Fig 3: Mutipanel figure
 
@@
 
-## Future needs /knowledge gaps:
-- Digitalization of MORE museums (coordinated) Just common protocols? RopenSci? What about an interactive bee decline map? Predicts?
-- Integration of DATA and analysis:
-- Communication Museums- Sci (biderictional. Importance of voucering specimens fused for science).
-- Compare global differences geographically (Eu Z US, for example)
+Fig 3: Mutipanel figure
 
-## Conclusions: Joint efforts needed at the global scale, scit science, ...
+## The way forward
 
-highlight digitalyzing initiatives including AMNH, LondonMNH, ¿? Highlight gaps that will never be filled probably (central africa).
+As illustrated in this examples, unleashing the power of museum data to answer pressing questions is at our hands, and need the coordinated effort of many actors.
+We envision that rather than a central repository of museum records, we should aim to make the digitalized data available and easy to share, and if possible ultimately centralized in existing platforms like Gbif.
+Even when the data lives in different museum webpages or databases, if it's easily accessible and has programmatic access (API e.g.
+@heltmint) it can be retried and joined with other data sets using open science tools (@ropensci).
+
+We also need to revalorize museums and foster a birectional relationship between researchers and museums.
+Researchers collect massive amounts of specimens, not always appropriately vouched.
+Depositing this in museums should ensure its long term preservation.
+However, this require appropriately funded museums and recognizing data (and not only papers) as an important research output.
+
+[What about an interactive bee decline map? or Integration of DATA and analysis:]()
+
+Ultimately, we need more data to compare global differences geographically and see which are the generalized trends.
+As far as we know there have been few bee regional extinctions (@Ollerton, @Bartomeus2013), but bee declines are widespread in disturbed ecosystems of north america and Europe (@Bartomeus, @dePalma, @Carvalheiro).
+We know almost nothing about the rest...
+for those areas, the first step is to know what pollinators exists and name the species before we loss them.
 

### 2018-02-07 ms.Rmd
added global bee richness data and altered code to create map

@@
 library(rworldmap)
-map <- read.csv("data/bee_declines.csv", sep = ";")
+map <- read.csv("data/bee_richness.csv")
 #Jamie, if you want, we could complete this data set based on the papers.
@@
 #need to use ISO's to make sure we don't carry over NA's due to country misspellings
-mapCountryData(mapToPlot = map2, nameColumnToPlot = "bee_richness"
-               , mapTitle = "Bee richness", addLegend = TRUE)
+#have plotted countries with missing values as black (shows we have richness for pretty much everywhere)
+map3 <- mapCountryData(map2, nameColumnToPlot = "species", missingCountryCol = "black",
+               mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)
+
+#add a modified legend using the same initial parameters as mapCountryData
+do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5))
 
 ```
-
 Fig 1: A global map showing evidence (% decline in richness) and Data deficient areas would be great.

### 2018-02-07 ms.Rmd
small edits

@@
 There is an ever-increasing concern about the current decline of pollinators, specifically bees.
-Despite pollinator decline reports are widespread, the information available is scarce and biased geographically and taxonomically, limiting the inference about the a potential pollination crisis.
+Despite pollinator decline reports are widespread, the information available is scarce and biased geographically and taxonomically, limiting the inference about a potential pollination crisis.
 The most direct form for understanding species decline patterns is to analyze historical time series, however, those are rare or often non-existing for most pollinator groups.
@@
 Despite this advances, we show that we still miss information for the status and trends of most pollinators and that often a bottle neck is digitizing the historical data.
-We analyze the opportunities and limitations to do so worldwide and provide two worked examples on how to analyze this type of data for two distinct regions for which historical bee declines have never been assessed, New Zealand and Spain.
+We analyze the opportunities and limitations to do so worldwide and provide two contrasting worked examples on how to analyze this type of data for two distinct regions for which historical bee declines have never been assessed, New Zealand and Spain.
 There is a great potential for Museum specimens to play a key role in elucidating the extent of the pollinator crisis and compare the responses obtained along different environmental situations if we succeed in unleashing the power of historical data.
@@
 A practical approach has been to use space for time substitutions, where the researchers compares natural areas with areas affected by different environmental conditions.
-Despite the critiques on the robustness of this approach (@ref), this is the best source of information, specially for Europe, where recent joint efforts had estimated bee richness declines at the country level using the predictions from such models (@DePalma).
+Despite the critiques on the robustness of this approach (@ref), this is the larger source of information we have, specially for Europe, where recent joint efforts had estimated bee richness declines at the country level using the predictions from such models (@DePalma).
 A second promising method is pollinator monitoring programs.
@@
 [Add something about traits of winners and loosers?]()
+https://www.sciencedirect.com/science/article/pii/S0065250417300296
+[move this and expand to the discussion on the map?]()
 
@@
 Doing this process one by one with each of the thousands of specimens located in a museum is a daunting task and needs specialized personnel.
-While some tasks can be only done by experts taxonomists, new technologies and citizen science can speed up the process of digitizing old collections.
+While some tasks can only be done by expert taxonomists, new technologies and citizen science can speed up the process of digitizing old collections.
 High resolution photos and scanning of the associated labels have been uploaded to the internet to get the power of distributed work where thousands of volunteers can transcript the images to text.
-This process is validated, etc...
-Examples include zooniverse, ...
+New algorithms have emerged to georeference localities based on vernacular names (@refs geonames, geoparser and opencage).
+Explain zooniverse, NZ initiative and how data can be validated...
 [Is there anything on AI to automate the process?
@@
 
+Start with buterflies?
+
 Is not surprising the the best evidence of pollinator declines comes from the more disturbed countries.
 A pioneer study using citizen science and museum specimens in UK and Netherlands, two of the most populated and transformed countries showed a worrisome trend for both bees, flies and flowering plants, but specially for bees (@Beisjmieretal2006).
-However, it's important to note that even within those countries, the local estimates are all nearby big cities and agricultural areas, and data is lacking for better conserved areas.
+However, it's important to note that even within those countries, the local estimates are all nearby big cities and agricultural areas, and data is lacking for the better conserved areas.
 Further exploration of this data set shows that the declining trend has attenuated in recent decades (@carvalheiroetal).
-Detailed trends for common species in the Netherlands using museum specimens confirm the link between pollinator and plant declines, as bees wich preferred host plants have declined the most are the more threatened (@Schepper2014).
-[- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies!]()
+Detailed trends for common bee species in the Netherlands using museum specimens confirm the link between pollinator and plant declines, as bees wich preferred host plants have declined the most are the more threatened (@Schepper2014).
+[- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies! Add Ollerton paper!]()
 
@@
 This is in harsh contrast with the high local extinctions reported by Burkle (@Burkle) when comparing historical observations of a large forested ecosystem with the few remnants that still exists, but note that this comparison do not takes into account the opportunistic bees using other land use changes, and that the local richness may have changes, but not declined as drastically.
-In any case, there is a great concordance between species locally extinct and regionally declining (@Bartomeus_winfree2013).
+In any case, there is a great concordance between species locally extinct and regionally declining (@Bartomeus_winfree2013) indicating that local extinctions are part of the downward regional population trends.
 
@@
 
+Justify focus on bees?
 We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species declines (b), syrphid species declines (c) and ??
@@
 
-Museum can provide valuable information besides species occurrence records, and this is important to track when digitizing the collections.
+Museum can provide valuable information besides species occurrence records, and this is important to take into account when digitizing the collections.
 This is specially relevant to identify mechanisms of decline and adaptation.
@@
 Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length (@refAlaskaScience) and body size (@Oliveira) shrinkage with climate and land use change.
+[move this section below]()
 
@@
 In Figure 2 we show the number of modern and historical bee records already available per country.
-While for some countries like UK and US the numbers are high in both axes, and hence, data is directly analyzable, most countries fall short in one or both of the axes.
+While for some countries like UK and US the numbers are high in both axes, and hence, data can be potentially analyzed without further data collection efforts, most countries fall short in one or both of the axes.
 Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis.
@@
 
-Once the data sets are made available, the researcher needs to gauge very carefully the potential bias in data.
-We illustrate this process with two examples for Spain and New Zealand.
+Once the data sets are made available, the researcher needs to gauge very carefully the potential bias in occurrence data.
+We illustrate this process with two contrasted examples for Spain and New Zealand.
 In the first one, most historical data comes from a few locations and is collected by the same team of people.
@@
 For example, some collectors may target only certain taxa, biasing the sample.
-We provide annotated R scripts to treat this kind of data and show two relevant case studies adding data for two unique regions with contrasting histories and hope this resource will cheer people to add more data and analysis (include a repository where people can contribute?)
+We provide annotated R scripts to treat both kind of data while revealing trends for two relevant regions with contrasting histories.
+We hope this resource will encourage people to add more data and analysis...
+(include a repository where people can contribute?)
 
@@
 
-Spain is an interesting study system because is has been transformed by human heavely, but is not as intensified as other EU countries.
+Spain is an interesting study system because is has been transformed by humans heavely, but is not as intensified as other EU countries.
 In addition, Spain is a hotspot of bee diversity and maintains an heterogeneous landscape.
@@
 Unfortunately, there is no recent sampling in the area.
-After some investigations, we discover that Enrique sampled systematically 8 localities and that more data is available at the "Museo de historia Nacional" and other minor collections.
-An effort to digitize the remaining material, and a re-survey of the area provides the perfect data set to make a comparison before-after of this 8 sites.
+After some investigations, we discover that Enrique sampled systematically 7 localities and that more data is available at the "Museo de historia Nacional" and other minor collections.
+An effort to digitize the remaining material, and a re-survey of the area provides the perfect data set to make a comparison before-after of this 7 sites.
 
@@
 [Maybe more analysis arise...]().
-We show that...
+We show that ...
 
@@
 As illustrated in this examples, unleashing the power of museum data to answer pressing questions is at our hands, and need the coordinated effort of many actors.
+As evidenceed by the case studies, collaborating with museum curators and original data collectors is key to understand the data and treat it appropiatelly.
 We envision that rather than a central repository of museum records, we should aim to make the digitalized data available and easy to share, and if possible ultimately centralized in existing platforms like Gbif.

### 2018-02-07 SP_data_cleaning.Rmd
Start preparing Spain data

@@
 - ID and 'cuaderno' need to remove duplicates, which are not in cuaderno!!
-- IF several IT's Oscar Aguade duplicated the netry as many times as need. Those sp are in red in original excel.
+- IF several IT's Oscar Aguado duplicated the enetry as many times as need. Those sp are in red in original excel.
 
@@
 head(sp)
+head(sort(table(sp$LOCALIDAD), decreasing = TRUE))
+write.csv(sort(table(sp$LOCALIDAD), decreasing = TRUE),
+          "raw_data/localidades.csv")
+
+sp$gen_sp <- paste(sp$G.NERO, sp$ESPECIE, sep = "_")
+unique(sp$gen_sp) #721 species! #clean, but only for the subset of used data. The 7 localities.
+head(sort(table(sp$gen_sp), decreasing = TRUE), 50)
+
+#subset only the 7 localities.
+
+#check sampling completness
+library(vegan)
+#need to fix first locality
+comm <- table(sp2$LOCALIDAD, sp2$gen_sp)
+temp <- specaccum(comm, "random")
+summary(temp)
+plot(temp)
+rarecurve(comm)
 
@@
 
-- Not entered yet.
+```{r}
+now <- read.csv(file = "raw_data/Spain_2016.csv", h = T)
+head(now)
+table(now$LOCALIDAD)
+levels(now$LOCALIDAD) <- c("auditorio_Delibes",
+                           "auditorio_Delibes",
+                           "hornillos",
+                           "hornillos",
+                           "olmedo",
+                           "sanbernardo",
+                           "toro",
+                           "vilalba",
+                           "zamaduenas")
+#get lat long
+library(spatial)
+#biogeomancer(country = "Spain", adm1 = "Castilla y Leon", adm2 = "Valladolid", 
+#             locality = levels(now$LOCALIDAD), 
+#             singleRecord=TRUE, progress="text")
+#not working, but Ropen sci has alternative packages.
+now$gen_sp <- paste(now$GÉNERO, now$ESPECIE, sep = "_")
+sort(table(now$gen_sp))
+#clean species
+unique(now$gen_sp) #197!!! but still some duplicates
+
+#check sampling completness
+library(vegan)
+comm <- table(now$LOCALIDAD, now$gen_sp)
+temp <- specaccum(comm, "random")
+summary(temp)
+plot(temp)
+rarecurve(comm) #uf... not really well sampled... but ~80 species per site it's quite good!
+```
+
+Data Gbif:
+```{r}
+gbif <- read.csv(file = "raw_data/data_gbif.csv", h = T)
+head(gbif)
+hist(gbif$year) #nice, bimodal!
+sort(table(gbif$recordedBy), decreasing = TRUE)
+#Engel, M, Hinojosa, I, Bennett, D, & Davis, S
+#G.E. Bohart
+#Torchio, Asensio, etc...
+#Baker, D. & Baker, M.
+# see
+set <- which(gbif$recordedBy == "Engel, M, Hinojosa, I, Bennett, D, & Davis, S")
+library(mapr)
+colnames(gbif)[2:3] <- c("latitude", "longitude") 
+map_leaflet(gbif[set,]) #interesting, 
+# Engel collected in 2007 in the south! And taxonomy is bad!
+# Baker in Estartit #del 62! 
+# Bohart Madrid, islands, south... in 78's
+
+#Overall, most pre 1980 collections around valladolid.
+head(sort(table(gbif$species), decreasing = TRUE), 50)
+unique(gbif$species) #443! BUT here lots of duplicates?
+```
+
+#Analysis
+
+1) only for 7 localities, join now and then and paired t-test on rarefied richness per site
+
+2) For common species we can do the same but with standardized frequency.

### 2018-03-08 ms.Rmd
added a section on the NZ study and have edited the introduction and abstract

@@
 
-There is an ever-increasing concern about the current decline of pollinators, specifically bees.
-Despite pollinator decline reports are widespread, the information available is scarce and biased geographically and taxonomically, limiting the inference about a potential pollination crisis.
-The most direct form for understanding species decline patterns is to analyze historical time series, however, those are rare or often non-existing for most pollinator groups.
-However, museum collections can provide the only source of historical data serving as a baseline to understand the reach of pollinator declines.
-Specimens historically collected by naturalists and preserved in museums can provide information on when and where species were collected, as well as other ecological information like historical species interactions, or morphological traits.
-We show here how researchers have used historical museum specimens to compare historical occurrences with current occurrence patterns to show changes over time in biodiversity levels, species relative abundances, body size and even pollination services delivery.
-Despite this advances, we show that we still miss information for the status and trends of most pollinators and that often a bottle neck is digitizing the historical data.
-We analyze the opportunities and limitations to do so worldwide and provide two contrasting worked examples on how to analyze this type of data for two distinct regions for which historical bee declines have never been assessed, New Zealand and Spain.
-There is a great potential for Museum specimens to play a key role in elucidating the extent of the pollinator crisis and compare the responses obtained along different environmental situations if we succeed in unleashing the power of historical data.
+There is ever-increasing concern about the decline of pollinators, especially for bees.
+However, despite reports that pollinator declines are widespread, data is scarce and geographically and taxonomically biased.
+This limits any robust inference about a potential pollination crisis.
+The most direct approach for identifying patterns in species declines is analysis of historical time series data.
+However, these data are often rare or non-existent for most pollinator groups, but provide the only source of historical information serving as a baseline to understand the extent of pollinator declines.
+Specimens historically collected by naturalists and preserved in museums can provide information on when and where species were collected, as well as other ecological information (e.g., historical species interactions and morphological traits).
+Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current occurrence patterns to identify changes in biodiversity, relative species abundances, body size and even pollination services over time.
+Despite recent advances, we show that information on the status and trends of most pollinators is missing and a key bottle neck slowing progress in this field is the digitization of historical data.
+We identify opportunities and limitations to progress understanding of pollinator declines worldwide.
+Further, we demonstrate how to analyse historical museum collection data, by providing two contrasting case studies from distinct geographical regions, New Zealand and Spain, for which historical pollinator declines have never been assessed.
+We show that there is a great potential for museum specimens to play a key role in identifying the extent of the pollination crisis if focus on unleashing the power of historical collection data.
 
@@
 
-Pollinators are of paramount importance given its prominent role in plant reproduction (@ollerton2012) and food security (@kleinetal2007).
-As many other taxa, pollinators are also exposed to a plethora of human induced rapid environmental changes potentially threatening its populations.
-This conundrum was realized early on by researchers, and was popularized by the influential book "the forgotten pollinators" (1996).
-However, at that time there was almost no data on the trends and status of pollinators and this initial claims trigger the first efforts to elucidate the problem, including a commission of a US National Academy of Science (NAS) panel to assess the reach of pollinators declines (2006).
+Animal pollinators are acritical component of ecosystems, given their role in plant reproduction (@ollerton2012) and food security worldwide (@kleinetal2007).
+As many other taxa, pollinators are vulnerable to multiple human-mediated environmental changes, which potentially threatening their populations.
+This problem was identified by researchers several decades ago, and was popularized by the influential book "the forgotten pollinators" (1996).
+However, these accounts were relatively anecdotal as there were almost no data on pollinator trends and statuses at the time.
+These initial claims triggered the first efforts to address the problem, including a US National Academy of Science (NAS) panel commissioned to assess the extent of pollinators declines (2006).
 The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult".
-Since then, the studies on pollinator responses to global change has multiplied and we reached a strong consensus that pressures like habitat destruction, land use change, chemical exposure, exotic species and global warming are the leading causes of pollinator declines and act sinergistically (@goulson2016, @potts2010).
-However, the status and trends of most pollinator species remains unknown.
-For example, for a well studied region like Europe, 60% [check number]() of the bee species assessed by a recent IUCN report fall into the category "Data Deficient" (@Nieto2014).
+Since then, studies on pollinator responses to global change have multiplied and researchers have reached strong consensus that pressures such as habitat destruction, land use change, chemical exposure, exotic species and climate change are the primary causes of pollinator declines, and often act synergistically (@goulson2016, @potts2010).
+Yet the status and trends of most pollinator species worldwide remain unknown.
+For example, even for a well-studied region such as Europe, 60% [check number]() of bee species assessed by in recent IUCN report fall into the category "Data Deficient" (@Nieto2014).
+For countries outside of Europe and the US, data on pollinator population is almost non-existent.
 
-One of the main barriers to profile pollinator population trends is that pollinators conform an heterogeneous group including insects such as bees, flies, butterflies and beetles, but also birds, bats and even lizards (@Winfreeetal2011).
-In particular bees, usually regarded as the more important in terms of abundance, efficiency and widespread distribution (@ollerton2018), are a diverse group with more than 20 000 species worldwide (@PickeringAscher) and most of them require expert taxonomic expertise for its identification.
-This has led not only to a biased geographic patterns due to the uneven distribution of researchers (@Martinetal2012), but also a taxonomic bias to species easier to identify, like bumblebees (@ArcherOikos).
+One of the main barriers to identifying pollinator population trends is that pollinators are a diverse group of taxa including bees, flies, butterflies and beetles, and also birds, bats and even lizards (@Winfreeetal2011).
+In particular, bees are usually regarded as the more important pollinaotrs due to their abundance, pollination efficiency and widespread distribution (@ollerton2018).
+However, bees are an incredibly diverse group, with more than 20,000 species worldwide (@PickeringAscher), and most require taxonomic expertise for identification.
+This has led to a biased geographic patterns in bee decline research due to the uneven distribution of researchers (@Martinetal2012), but also taxonomic biases toward species that are easier to identify, such bumblebees (@ArcherOikos).
+Another barrier is methodological and involves challenges for assessing population trends for highly mobile organisms, which are also often short-lived and small.
+One solution to overcoming these barriers has been to use space for time substitutions, where researchers compare pollinator populations across environmental gradients.
+Despite critiques on the robustness of this approach (@ref), this is the most extensive source of data currently available and in Europe, researchers have recently estimated bee richness declines for each country using predictions generated from models (@DePalma).
+A second promising method is pollinator monitoring programs, with data often collected by citizen scientists.
+This approach was inspired by successful butterfly monitoring programs and is currently being extended to other pollinator taxa (@ref).
+The main limitation is that these programs need significant time to generate long-term datasets and do not reveal anything about historic pollinator communities.
+Finally, the most robust approach for identifying historical pollinator tends is to use museum collection data.
+The core aim of museums is precisely to conserve and curate collections, thus they not only serve as a precious repository specimens, but often also ensure high quality taxonomic identification (@primak).
+However, the major bottleneck for researchers wanting to use this information is the digitization of specimens.
+Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have facilitated identification of current status and trends in pollinator populations (e.g.
+Bartomeus et al 2013, Scheper et al 2014).
+This is despite the fact that museum collections often have a number of biases (@Keil).
+In reality, museum collections are the only data we have to reconstruct the past.
 
-The second barrier is methodological and involves how to assess population trends for highly mobile organisms, which in addition are often short-lived and small.
-A practical approach has been to use space for time substitutions, where the researchers compares natural areas with areas affected by different environmental conditions.
-Despite the critiques on the robustness of this approach (@ref), this is the larger source of information we have, specially for Europe, where recent joint efforts had estimated bee richness declines at the country level using the predictions from such models (@DePalma).
-A second promising method is pollinator monitoring programs.
-Inspired in the successful butterfly monitoring programs (@ref), this protocols, often carried by citizen scientists, are being implemented on other taxa (@ref).
-The main limitation is that those programs need time to build up long term data series and do not reveal anything about the original conditions.
-Finally, the best source of historical data can be found in museums.
-The core aim of Museums is precisely to conserve and curate this kind of collections, and serve not only of a precious voucher of specimens, but often also assures its taxonomic identification quality (@primak).
-The major bottleneck to use this information, is however, digitizing the specimens.
-Efforts to do so have provide great benefits and allow the characterization of the status and trends of bee pollinators is the US and EU (e.g.
-Bartomeus et al 2013, Scheper et al 2014) and despite data deposited in museums is often collected opportunistically and have its own biases (@Keil), is the only data we have to reconstruct the past.
-
-Regardless of the method, it's important to consider that biodiversity has many aspects beyond species richness, and we should put more effort in differentiating declines not only in local (alpha) diversity or regional (gamma) diversity, but also in the turnover over sites (beta), the abundance of key species, and the evenness of the communities (@winfree2016ELE).
-In fact, environmental change often give us few winners, and lots of loosers (@Bartomeusetal2013), and characterizing those is extremely important as the few winners are often exotic species (@Stavertetal2018).
-[Add something about traits of winners and loosers?]()
+Regardless of the method, it is important to consider that biodiversity has many aspects beyond species richness, and we should into identifying changes not only in local (alpha) diversity or regional (gamma) diversity, but also in turnover between sites (beta diversity), abundance of key species, and evenness of communities (@winfree2016ELE).
+Environmental changes often results in few winners and many of losers (@Bartomeusetal2013).
+Identifying winners and losers is critical as the few winners are often exotic species, and represent a subset of trait that facilitate survival in highly modified environments (@Stavertetal2018).
+These changes can have important flow on effects for pollination of native species and crops.
+[Add something about traits of winners and losers?]()
 https://www.sciencedirect.com/science/article/pii/S0065250417300296
@@
 
-Digitizing old specimens is not a trivial task and it needs expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geolocate the coordinates of the collection event (@ref and software) and (iii) store the data in a proper curated database (@ref on database management).
-Doing this process one by one with each of the thousands of specimens located in a museum is a daunting task and needs specialized personnel.
-While some tasks can only be done by expert taxonomists, new technologies and citizen science can speed up the process of digitizing old collections.
-High resolution photos and scanning of the associated labels have been uploaded to the internet to get the power of distributed work where thousands of volunteers can transcript the images to text.
-New algorithms have emerged to georeference localities based on vernacular names (@refs geonames, geoparser and opencage).
+Digitizing old specimens is not a trivial task and it needs expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geo-locate the coordinates of the collection event (@ref and software) and (iii) store the data in a properly curated database (@ref on database management).
+Undertaking this process for the thousands of specimens a museum collection is a daunting task and requires specialized personnel.
+While some tasks can only be undertaken by those with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the process of digitizing collections.
+High resolution photos of specimens and the associated labels can be uploaded to the internet, where the work of transcribing the images can be distributed across hundreds or thousands of volunteers.
+Furthermore, new algorithms have been created to geo-reference localities based on vernacular names (@refs geonames, geoparser and opencage).
 Explain zooniverse, NZ initiative and how data can be validated...
@@
 
-Is not surprising the the best evidence of pollinator declines comes from the more disturbed countries.
-A pioneer study using citizen science and museum specimens in UK and Netherlands, two of the most populated and transformed countries showed a worrisome trend for both bees, flies and flowering plants, but specially for bees (@Beisjmieretal2006).
-However, it's important to note that even within those countries, the local estimates are all nearby big cities and agricultural areas, and data is lacking for the better conserved areas.
-Further exploration of this data set shows that the declining trend has attenuated in recent decades (@carvalheiroetal).
-Detailed trends for common bee species in the Netherlands using museum specimens confirm the link between pollinator and plant declines, as bees wich preferred host plants have declined the most are the more threatened (@Schepper2014).
+Is not surprising the best evidence of pollinator declines comes from the more disturbed countries western countries.
+For the UK and the Netherlands, which have been subject to extensive environmental changes, a study using citizen science and museum specimens found declines for bees, flies and flowering plants (@Beisjmieretal2006).
+However, it is important to note that even for these two countries, the local estimates for pollinator richness are nearby all for large cities and agricultural areas, and data is lacking for conserved natural areas.
+Further exploration of this dataset showed that the declining trend has attenuated in recent decades (@carvalheiroetal).
+Using museum specimens in the Netherlands, detailed trends for common bee species confirm the link between pollinator and plant declines, as bees which preferred host plants have declined the most are the more threatened (@Schepper2014).??confusing.
 [- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies! Add Ollerton paper!]()
 
-In comparison, regional estimates for the eastern US using museum specimens show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions as most species can still persist (@Colla2012?).
-This is in harsh contrast with the high local extinctions reported by Burkle (@Burkle) when comparing historical observations of a large forested ecosystem with the few remnants that still exists, but note that this comparison do not takes into account the opportunistic bees using other land use changes, and that the local richness may have changes, but not declined as drastically.
-In any case, there is a great concordance between species locally extinct and regionally declining (@Bartomeus_winfree2013) indicating that local extinctions are part of the downward regional population trends.
+In comparison, regional estimates in species richness changes for the eastern US show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions as most species tend to persist (@Colla2012?).
+These findings are in stark contrast with the widespread local extinctions reported by Burkle (@Burkle), who compared historical observations of bees from a large forested ecosystem with the few remnants.
+However, it is important to note that this comparison does not take into account opportunistic bee species using novel land use types in the region.
+Thus, local richness may have declines but regional richness may not have declined as drastically.
+In any case, there is strong concordance between local extinctions and regional declines (@Bartomeus_winfree2013), suggesting that local extinctions are indicators of regional population declines.
 
-Outside of Europe and the US there is surprisingly little evidence of pollinator declines using historical records.
+Outside of Europe and the US there is surprisingly studies on pollinator declines using historical records.
 [South Africa: Pauw, maybe with plants...]() .
@@
 Justify focus on bees?
-We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species declines (b), syrphid species declines (c) and ??
-to highlight the little coverage we have...
+We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes to highlight the little coverage that we have globally.
 
@@
 
-## Case study two: NZ
+##Case study two: NZ    
+New Zealand is an isolated oceanic archipelago, with a distinctive pollinator biota and a unique history of human occupation.
+Much of New Zealand’s pollinator fauna is also relatively depauperate.
+For example, New Zealand has only 27 species of native bees (@Donovan, 2007), which is a fraction of nearby Australia’s c.
+1600 species (REF).
+However, New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems (@Newstrom & Robertson, 2005).
+Therefore, New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.
 
-- Describe data and analysis.   
-- Present main results
+Human colonisation of New Zealand was relatively recent in global terms (c.
+740 y) (@Wilmshurst et al.
+2008).
+Before human arrival, New Zealand was mostly forested, but has since been dramatically altered by people.
+Early Māori settlers cleared forests by burning and more recently, European colonists cleared much of remaining forests for agriculture, mostly before 1900 (@Perry et al, 2014).
+Therefore, human activity likely affected New Zealand pollinator communities long before collections began.
+Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.
 
-Fig 3: Mutipanel figure
+New Zealand pollinator collections
+We used New Zealand bee collection records gathered from multiple sources, including university, research institute, museum and private collections.
+Fly pollinator data was obtained from six participating New Zealand museums and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators.
+Collections for the bee and fly datasets span over 100 years (early 1900s to late 2000s).
+
+Data analyses
+First, we filtered our original datasets so that data used for analyses only included unique collection events.
+To do this, we removed specimens collected at the same location, on the same data and by the same collector.
+We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980.
+We accounted for differences in collection effort through time by binning collection records, so that each bin had a similar number of records but a different number of years (see @Bartomeusetal2013 for further details).
+We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin.
+Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness.
+Thus reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.
+
+To test whether there were relative differences in occurrences of species over time (should we analyse natives and exotics together or separately?).
+To determine if the probability of finding a species in the collection changed over time we used a general linear model with a binomial distribution and a logit link.
+For species that showed overdispersion, we used a quasi-binomial distribution.
+We only included species in this analysis that had more than 30 records.
+To account for differences in sampling effort between years, we weighted each year by the total number of samples collected in that year.
+
+Changes in species richness over time 
+We found that rarefied richness for both native and exotic bees increased overtime, but these trends were non-significant (P-values for both groups > 0.05).
+In contrast, native fly richness declined, whereas exotic fly richness increased, although results for these groups were also non-significant (P-values for both groups > 0.05).
+
+Changes in species relative abundances over time 
+We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic).
+Three bee species declined in relative abundance (one native and two exotic).
+However, only two out of 14 fly species increased in relative abundance, both of which were exotic, whereas four species decreased in abundance (three native and one exotic).
+
+Fig 3: Multi-panel figure
+
+-	Do we need some sort of separate discussion or integrate into the results section?
+… Although this is surprising, it could simply indicate a change in the evenness of species collected over time as the method we used actually measures both richness and evenness.
 
@@
 
-E. Asensio (or maybe coauthor?). Patrick McQuinn, and Crona McMonagle for data entry. Madrid museum, ITACyL.
+E.
+Asensio (or maybe coauthor?).
+Patrick McQuinn, and Crona McMonagle for data entry.
+Madrid museum, ITACyL.
+Barry Donovan for providing New Zealand bee collection records.
+Carola Warner for entering the New Zealand data.

### 2018-03-08 ms.Rmd
updated ms

@@
 Before human arrival, New Zealand was mostly forested, but has since been dramatically altered by people.
-Early Māori settlers cleared forests by burning and more recently, European colonists cleared much of remaining forests for agriculture, mostly before 1900 (@Perry et al, 2014).
-Therefore, human activity likely affected New Zealand pollinator communities long before collections began.
+Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 (@Perry et al, 2014).
+Therefore, human activity likely affected New Zealand pollinator communities long before surveys and collections of pollinators began.
 Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.
@@
 Data analyses
-First, we filtered our original datasets so that data used for analyses only included unique collection events.
-To do this, we removed specimens collected at the same location, on the same data and by the same collector.
+First, we filtered our original datasets so that data for analyses only included unique collection events.
+To do this, we removed specimens collected at the same location, on the same date, and by the same collector.
 We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980.
@@
 
-To test whether there were relative differences in occurrences of species over time (should we analyse natives and exotics together or separately?).
-To determine if the probability of finding a species in the collection changed over time we used a general linear model with a binomial distribution and a logit link.
+To determine if the probability of finding a species in the collection changed over time, we used a general linear model with a binomial distribution and a logit link.
+(should we analyse natives and exotics together or separately?).
 For species that showed overdispersion, we used a quasi-binomial distribution.

### 2018-03-15 ms.Rmd
ms edited again

@@
 
-There is ever-increasing concern about the decline of pollinators, especially for bees.
-However, despite reports that pollinator declines are widespread, data is scarce and geographically and taxonomically biased.
-This limits any robust inference about a potential pollination crisis.
-The most direct approach for identifying patterns in species declines is analysis of historical time series data.
-However, these data are often rare or non-existent for most pollinator groups, but provide the only source of historical information serving as a baseline to understand the extent of pollinator declines.
-Specimens historically collected by naturalists and preserved in museums can provide information on when and where species were collected, as well as other ecological information (e.g., historical species interactions and morphological traits).
-Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current occurrence patterns to identify changes in biodiversity, relative species abundances, body size and even pollination services over time.
-Despite recent advances, we show that information on the status and trends of most pollinators is missing and a key bottle neck slowing progress in this field is the digitization of historical data.
-We identify opportunities and limitations to progress understanding of pollinator declines worldwide.
-Further, we demonstrate how to analyse historical museum collection data, by providing two contrasting case studies from distinct geographical regions, New Zealand and Spain, for which historical pollinator declines have never been assessed.
-We show that there is a great potential for museum specimens to play a key role in identifying the extent of the pollination crisis if focus on unleashing the power of historical collection data.
+There is increasing concern about the decline of pollinators worldwide.
+However, despite reports that pollinator declines are widespread, data is scarce and often geographically and taxonomically biased.
+This limits robust inference about any potential “pollination crisis”.
+[something about how to address the problem].
+Analysis of historical time series data is the most direct approach for assessing species declines.
+In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines.
+Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits).
+However, data of this kind are typically rare or non-existent for most pollinator taxa.
+Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and even pollination services.
+Despite recent advances, we show that information on the status and trends of most pollinators is absent, and that the lack of collection record data digitization is a key bottleneck, impeding progress in this field.
+We identify opportunities and limitations to progress in assessing pollinator declines at a global scale.
+Further, we demonstrate various approaches to analysing museum collection data by providing two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
+Finally, we suggest that there is immense potential for museum specimens to play a central role in assessing the extent of the pollination crisis globally.
 
@@
 
-Animal pollinators are acritical component of ecosystems, given their role in plant reproduction (@ollerton2012) and food security worldwide (@kleinetal2007).
-As many other taxa, pollinators are vulnerable to multiple human-mediated environmental changes, which potentially threatening their populations.
-This problem was identified by researchers several decades ago, and was popularized by the influential book "the forgotten pollinators" (1996).
-However, these accounts were relatively anecdotal as there were almost no data on pollinator trends and statuses at the time.
-These initial claims triggered the first efforts to address the problem, including a US National Academy of Science (NAS) panel commissioned to assess the extent of pollinators declines (2006).
+Animal pollinators are a critical component of both natural and production ecosystems worldwide, given their role in plant reproduction (@ollerton2012) and food security (@kleinetal2007).
+As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions.
+This issue was identified several decades ago, and was popularized in 1996 by the influential book "the forgotten pollinators".
+However, these early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time.
+These initial claims triggered the first efforts to asses the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines.
 The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult".
-Since then, studies on pollinator responses to global change have multiplied and researchers have reached strong consensus that pressures such as habitat destruction, land use change, chemical exposure, exotic species and climate change are the primary causes of pollinator declines, and often act synergistically (@goulson2016, @potts2010).
-Yet the status and trends of most pollinator species worldwide remain unknown.
-For example, even for a well-studied region such as Europe, 60% [check number]() of bee species assessed by in recent IUCN report fall into the category "Data Deficient" (@Nieto2014).
-For countries outside of Europe and the US, data on pollinator population is almost non-existent.
+Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically (@goulson2016, @potts2010).
+Yet the current status and population trends of most pollinator species worldwide remain unknown.
+For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category (@Nieto2014).
+For countries outside of Europe and the US, data on pollinator populations is almost non-existent.
 
-One of the main barriers to identifying pollinator population trends is that pollinators are a diverse group of taxa including bees, flies, butterflies and beetles, and also birds, bats and even lizards (@Winfreeetal2011).
-In particular, bees are usually regarded as the more important pollinaotrs due to their abundance, pollination efficiency and widespread distribution (@ollerton2018).
-However, bees are an incredibly diverse group, with more than 20,000 species worldwide (@PickeringAscher), and most require taxonomic expertise for identification.
-This has led to a biased geographic patterns in bee decline research due to the uneven distribution of researchers (@Martinetal2012), but also taxonomic biases toward species that are easier to identify, such bumblebees (@ArcherOikos).
-Another barrier is methodological and involves challenges for assessing population trends for highly mobile organisms, which are also often short-lived and small.
-One solution to overcoming these barriers has been to use space for time substitutions, where researchers compare pollinator populations across environmental gradients.
-Despite critiques on the robustness of this approach (@ref), this is the most extensive source of data currently available and in Europe, researchers have recently estimated bee richness declines for each country using predictions generated from models (@DePalma).
-A second promising method is pollinator monitoring programs, with data often collected by citizen scientists.
-This approach was inspired by successful butterfly monitoring programs and is currently being extended to other pollinator taxa (@ref).
-The main limitation is that these programs need significant time to generate long-term datasets and do not reveal anything about historic pollinator communities.
-Finally, the most robust approach for identifying historical pollinator tends is to use museum collection data.
-The core aim of museums is precisely to conserve and curate collections, thus they not only serve as a precious repository specimens, but often also ensure high quality taxonomic identification (@primak).
-However, the major bottleneck for researchers wanting to use this information is the digitization of specimens.
-Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have facilitated identification of current status and trends in pollinator populations (e.g.
+One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and including bees, flies, butterflies, beetles, birds, bats and even lizards (@Winfreeetal2011).
+Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution (@ollerton2018).
+However, bees are also extremely diverse, with more than 20,000 species currently described worldwide (@PickeringAscher), and require often require expert taxonomists for identification.
+The uneven distribution of bee researchers has resulted in geographical biases in bee decline research (@Martinetal2012) and there are also taxonomic biases toward species that are easier to identify, such bumblebees (@ArcherOikos).
+Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.
+One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients.
+Despite critiques on the robustness of this approach (@ref), this is the most extensive source of data on pollinator populations currently available.
+In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models [using what parameters?] (@DePalma).
+A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists.
+This approach was inspired by successful butterfly monitoring programs in countries such as the UK and is currently being extended to other pollinator taxa (@ref).
+However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations.
+Finally, the most robust approach for assessing long-term pollinator population trends is to use museum specimen collection data.
+The core aim of museums is to conserve and curate collections.
+Thus, they serve as a precious repository for specimens and also ensure high quality taxonomic identification (@primak).
+Yet currently, the major bottleneck for researchers wanting to use contained within museum collections is the digitization of collection data.
+Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g.
 Bartomeus et al 2013, Scheper et al 2014).
-This is despite the fact that museum collections often have a number of biases (@Keil).
-In reality, museum collections are the only data we have to reconstruct the past.
+This is despite the fact that museum collections often have a number of biases [including...] (@Keil).
+But in many cases, museum collections are the only data we have to reconstruct the past.
 
-Regardless of the method, it is important to consider that biodiversity has many aspects beyond species richness, and we should into identifying changes not only in local (alpha) diversity or regional (gamma) diversity, but also in turnover between sites (beta diversity), abundance of key species, and evenness of communities (@winfree2016ELE).
-Environmental changes often results in few winners and many of losers (@Bartomeusetal2013).
-Identifying winners and losers is critical as the few winners are often exotic species, and represent a subset of trait that facilitate survival in highly modified environments (@Stavertetal2018).
-These changes can have important flow on effects for pollination of native species and crops.
+Regardless of the method, it is important for researchers to consider aspects of biodiversity beyond species richness.
+In addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities (@winfree2016ELE).
+Environmental changes often results in few “winners” and many “losers” (@Bartomeusetal2013).
+Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments (@Stavertetal2018).
+These changes can have important flow on effects for pollination of native plant species and crops.
 [Add something about traits of winners and losers?]()
@@
 
-Digitizing old specimens is not a trivial task and it needs expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geo-locate the coordinates of the collection event (@ref and software) and (iii) store the data in a properly curated database (@ref on database management).
-Undertaking this process for the thousands of specimens a museum collection is a daunting task and requires specialized personnel.
-While some tasks can only be undertaken by those with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the process of digitizing collections.
-High resolution photos of specimens and the associated labels can be uploaded to the internet, where the work of transcribing the images can be distributed across hundreds or thousands of volunteers.
-Furthermore, new algorithms have been created to geo-reference localities based on vernacular names (@refs geonames, geoparser and opencage).
-Explain zooniverse, NZ initiative and how data can be validated...
-[Is there anything on AI to automate the process?
+Digitizing old specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geo-locate the coordinates of collection events (@ref and software) and (iii) store the data in a properly curated database (@ref on database management).
+Undertaking this process for the thousands of specimens in museum collections can be a daunting task and requires specialized personnel.
+While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process.
+High resolution specimen photos and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers.
+In addition to this, new algorithms have been created that allow location geo-referencing based on vernacular names (@refs geonames, geoparser and opencage).
+[Explain zooniverse, NZ initiative and how data can be validated...
+Is there anything on AI to automate the process?
 e.g.
@@
 Can be mentioned?]
+
 [how much do we want to extend on this? Is interesting info, but boring o read?]()
@@
 
-As illustrated in this examples, unleashing the power of museum data to answer pressing questions is at our hands, and need the coordinated effort of many actors.
-As evidenceed by the case studies, collaborating with museum curators and original data collectors is key to understand the data and treat it appropiatelly.
-We envision that rather than a central repository of museum records, we should aim to make the digitalized data available and easy to share, and if possible ultimately centralized in existing platforms like Gbif.
-Even when the data lives in different museum webpages or databases, if it's easily accessible and has programmatic access (API e.g.
-@heltmint) it can be retried and joined with other data sets using open science tools (@ropensci).
+As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors.
+We show in our case studies that collaboration with both museum curators and data collectors is key to understanding the data and treating it appropriately.
+We envision that rather than a central repository of museum collection records, we should aim to make digitalized data readily available and easy to share.
+Ultimately, centralisation of these data in existing platforms such as Gbif would facilitate free and widespread access.
+However, even when data is stored in different webpages or databases, it can be retried and joined with other data sets using open science tools (@ropensci) if it is easily accessible and has programmatic access (API e.g.
+@heltmint).
 
-We also need to revalorize museums and foster a birectional relationship between researchers and museums.
-Researchers collect massive amounts of specimens, not always appropriately vouched.
-Depositing this in museums should ensure its long term preservation.
-However, this require appropriately funded museums and recognizing data (and not only papers) as an important research output.
-
+We also need to revolutionize the way that researchers collaborate with museums by foresting healthy bidirectional relationships.
+For instance, researchers collect massive numbers of specimens, but these are often inappropriately vouched, rendering them useless for research purposes.
+To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is adequate funding and recognition that accurate data recording and long term preservation are critical for research.
 [What about an interactive bee decline map? or Integration of DATA and analysis:]()
 
-Ultimately, we need more data to compare global differences geographically and see which are the generalized trends.
-As far as we know there have been few bee regional extinctions (@Ollerton, @Bartomeus2013), but bee declines are widespread in disturbed ecosystems of north america and Europe (@Bartomeus, @dePalma, @Carvalheiro).
-We know almost nothing about the rest...
-for those areas, the first step is to know what pollinators exists and name the species before we loss them.
+Ultimately, to identify global trends in pollinator declines we require data from diverse geographic regions.
+In particular, we need to put increased emphasis on quantifying pollinator declines in regions outside of the US and Europe.
+Furthermore, we need to investigate whether there have been declines in pollinator groups other than bees.
+As far as we know, there have been few regional bee extinctions (@Ollerton, @Bartomeus2013), but declines are widespread in disturbed ecosystems of the US and Europe (@Bartomeus, @dePalma, @Carvalheiro).
+For most other pollinator taxa we know almost nothing.
+For many taxa, the first step is to identify and describe them and only then can we begin to document declines.
 
@@
 
-E.
-Asensio (or maybe coauthor?).
-Patrick McQuinn, and Crona McMonagle for data entry.
+We thank E.
+Asensio (or maybe coauthor?), Patrick McQuinn, and Crona McMonagle for data entry.
 Madrid museum, ITACyL.
-Barry Donovan for providing New Zealand bee collection records.
-Carola Warner for entering the New Zealand data.
+We also thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.

### 2018-03-16 ms.Rmd
ammended manuscript and added script to produce frequency estimate graph

@@
 
-Start with buterflies?
+Start with butterflies?
 
-Is not surprising the best evidence of pollinator declines comes from the more disturbed countries western countries.
-For the UK and the Netherlands, which have been subject to extensive environmental changes, a study using citizen science and museum specimens found declines for bees, flies and flowering plants (@Beisjmieretal2006).
-However, it is important to note that even for these two countries, the local estimates for pollinator richness are nearby all for large cities and agricultural areas, and data is lacking for conserved natural areas.
-Further exploration of this dataset showed that the declining trend has attenuated in recent decades (@carvalheiroetal).
-Using museum specimens in the Netherlands, detailed trends for common bee species confirm the link between pollinator and plant declines, as bees which preferred host plants have declined the most are the more threatened (@Schepper2014).??confusing.
-[- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies! Add Ollerton paper!]()
+It is unsurprising that most evidence of pollinator declines is biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance.
+For example, in the UK and the Netherlands, which have experienced a multitude of widespread anthropogenic changes, a citizen science based study using and museum collections detected declines for bees, flies and flowering plants (@Beisjmieretal2006).
+However, it is important to note that even for these two countries, pollinator richness local estimates are biased toward large cities and regions dominated by agriculture, and data is lacking for natural areas.
+Further exploration of this dataset revealed that the declining trend for pollinator taxa has attenuated in recent decades (@carvalheiroetal).
+More detailed trends for common bee species in the Netherlands have also been explored using museum specimens and confirm the link between simultaneous plant and pollinator declines (@Schepper2014).
+This study found that bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus were most threatened with extinction.
+[check this] [- EU: UK?
+Ollerton Science, Peter Keil flies?
+Moths?
+Buterflies!
+Add Ollerton paper!]()
 
-In comparison, regional estimates in species richness changes for the eastern US show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions as most species tend to persist (@Colla2012?).
-These findings are in stark contrast with the widespread local extinctions reported by Burkle (@Burkle), who compared historical observations of bees from a large forested ecosystem with the few remnants.
-However, it is important to note that this comparison does not take into account opportunistic bee species using novel land use types in the region.
-Thus, local richness may have declines but regional richness may not have declined as drastically.
+In comparison to local decline estimates, regional estimates for changes in species richness in the eastern US show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions (@Colla2012?).
+These findings are in stark contrast with the widespread local extinctions reported by Burkle (@Burkle), who compared historical observations of bee occurrence in a large forested ecosystem with the present-day remaining forest remnants.
+However, it is important to note that this comparison does not take into account opportunistic bee species that use novel land-use types.
+Thus, studies may detect local richness declines, while regional richness remains relatively stable.
 In any case, there is strong concordance between local extinctions and regional declines (@Bartomeus_winfree2013), suggesting that local extinctions are indicators of regional population declines.
 
-Outside of Europe and the US there is surprisingly studies on pollinator declines using historical records.
+Outside of Europe and the US there are no studies on pollinator declines using historical museum collection records.
 [South Africa: Pauw, maybe with plants...]() .
-However, there are specific examples showing the historical decline of certain species e.g.
-Bombus dalbhomi (Aizen)
+However, there are species-specific examples of historical (e.g., Bombus dalbhomi; Aizen).
 
@@
 
-Once the data sets are made available, the researcher needs to gauge very carefully the potential bias in occurrence data.
-We illustrate this process with two contrasted examples for Spain and New Zealand.
-In the first one, most historical data comes from a few locations and is collected by the same team of people.
-In this case, we contacted the original authors to understand their sampling protocol, and resurveyed the same sites 20 years later.
-The second case shows a diversity of collectors and sites.
-In this case, we should ensure that geographical and taxonomic coverage do not change systematically with time.
-For example, some collectors may target only certain taxa, biasing the sample.
-We provide annotated R scripts to treat both kind of data while revealing trends for two relevant regions with contrasting histories.
-We hope this resource will encourage people to add more data and analysis...
-(include a repository where people can contribute?)
+Once museum collection datasets are made available, researchers needs to gauge any potential biases in the data.
+We explore this process with two contrasting examples from Spain and New Zealand.
+In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people.
+In this case, we contacted the original collectors to define their sampling protocols.
+We then resurveyed the same sites [using the same protocols?] 20 years later.
+The New Zealand dataset includes a wide suite of collectors and collection locations.
+For this dataset, we ensured that there no clear changes to geographical [I haven’t actually looked at this as there weren’t and GPS coordinates in the data] and taxonomic collection patterns through time.
+For example, some collectors may have targeted particular taxa, at specific locations or during certain time periods, which would represent a bias in the data.
+For these case studies, we provide annotated R scripts to analyse the different kinds of data (as described above).
+Essentially, these different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories.
+We hope this resource will encourage researchers to analyses data for regions where little is currently know about pollinator declines.
+[include a repository where people can contribute?]
 
@@
 
-Spain is an interesting study system because is has been transformed by humans heavely, but is not as intensified as other EU countries.
-In addition, Spain is a hotspot of bee diversity and maintains an heterogeneous landscape.
-As illustrated in Figure 2, Spain has already digitalized a decent amount of occurrence data both historical, and recent.
-However, a visual inspection reveals that data points are clustered and historical and recent records do not match spatially, making them hard to compare.
-Among the old records, it is striking that most of them are located around Valladolid and collected by Enrique Asensio and his team.
-Unfortunately, there is no recent sampling in the area.
-After some investigations, we discover that Enrique sampled systematically 7 localities and that more data is available at the "Museo de historia Nacional" and other minor collections.
-An effort to digitize the remaining material, and a re-survey of the area provides the perfect data set to make a comparison before-after of this 7 sites.
+Spain provides an interesting study system because its natural habitats have been transformed extensively by humans, but land-use is not as intensive as many other European countries.
+In addition, Spain is a bee diversity hotspot and maintains a relatively heterogeneous landscape.
+Spain has already digitalized a decent amount of occurrence data both historical, and recent (Figure 2).
+However, visual inspection of the data reveals that the data are clustered [in what way?].
+Further, historic records do not spatially match recent records, making them difficult to compare.
+For this dataset, most of the old records are located around Valladolid and were collected by Enrique Asensio and his team.
+There has been no recent sampling of bees in this area.
+However we found that Enrique systematically sampled seven locations and that there was additional data available at the "Museo de historia Nacional" and other minor collections.
+Digitization of these records, along with a re-survey of the locations that Enrique and his team collected from, provided us with an excellent dataset for before and after comparisons of bee communities at these sites.
 
 In brief, we check sampling completeness at both time periods, and compare the richness level of each site before 1980 and after 1980 with a paired t-test.
-[Maybe more analysis arise...]().
+[Maybe more analyses arise...]().
 We show that ...
 
-Fig 3: Mutipanel figure
+Fig 3: Multi-panel figure
 
 ##Case study two: NZ    
-New Zealand is an isolated oceanic archipelago, with a distinctive pollinator biota and a unique history of human occupation.
-Much of New Zealand’s pollinator fauna is also relatively depauperate.
-For example, New Zealand has only 27 species of native bees (@Donovan, 2007), which is a fraction of nearby Australia’s c.
+In contrast to Spain, Aotearoa-New Zealand is an isolated oceanic archipelago, with a distinctive pollinator biota and a unique history of human occupation.
+Much of Aotearoa-New Zealand’s pollinator fauna is also relatively depauperate.
+For example, Aotearoa-New Zealand has only 27 species of native bees (@Donovan, 2007), which is a fraction of nearby Australia’s c.
 1600 species (REF).
-However, New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems (@Newstrom & Robertson, 2005).
-Therefore, New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.
+However, Aotearoa-New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems (@Newstrom & Robertson, 2005).
+Thus, Aotearoa-New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.
 
-Human colonisation of New Zealand was relatively recent in global terms (c.
+In global terms, human colonisation of Aotearoa-New Zealand was relatively recent (c.
 740 y) (@Wilmshurst et al.
 2008).
-Before human arrival, New Zealand was mostly forested, but has since been dramatically altered by people.
+Before human arrival, Aotearoa-New Zealand was predominately forested, but has since been dramatically altered by people.
 Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 (@Perry et al, 2014).
-Therefore, human activity likely affected New Zealand pollinator communities long before surveys and collections of pollinators began.
-Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.
+Therefore, human activity likely affected pollinator communities in Aotearoa-New Zealand long before surveys and collections began.
+Nevertheless, we can use museum records to identify trends in pollinator communities during Aotearoa-New Zealand’s more recent history.
 
@@
 Data analyses
-First, we filtered our original datasets so that data for analyses only included unique collection events.
+First, we filtered our original datasets so that data used for analyses only included unique collection events.
 To do this, we removed specimens collected at the same location, on the same date, and by the same collector.
 We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980.
-We accounted for differences in collection effort through time by binning collection records, so that each bin had a similar number of records but a different number of years (see @Bartomeusetal2013 for further details).
+We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years (see @Bartomeusetal2013 for further details).
 We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin.
@@
 To determine if the probability of finding a species in the collection changed over time, we used a general linear model with a binomial distribution and a logit link.
-(should we analyse natives and exotics together or separately?).
+[should we analyse natives and exotics together or separately?].
 For species that showed overdispersion, we used a quasi-binomial distribution.
-We only included species in this analysis that had more than 30 records.
+Further, we only included species in this analysis for which we had 30 or more records.
 To account for differences in sampling effort between years, we weighted each year by the total number of samples collected in that year.
@@
 Changes in species relative abundances over time 
-We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic).
-Three bee species declined in relative abundance (one native and two exotic).
+We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Figure 3).
 However, only two out of 14 fly species increased in relative abundance, both of which were exotic, whereas four species decreased in abundance (three native and one exotic).
 
-Fig 3: Multi-panel figure
+```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
+#load packages
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(ggplot2)
+library(maps)
+library(vegan)
+
+d <- read.csv("data/NZBees_allrecords.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+
+#select required variables
+d <- select(d, one_of(c("species", "date", "locality", "collector", "collection", "method", "exotic_native")))
+d <- na.omit(d) #remove NAs
+
+#remove honeybees
+d$species <- as.factor(d$species)
+d <- d[ ! d$species %in% c("Apis mellifera"), ]
+d$species <- droplevels(d$species)
+
+#need to correct space between genus and species
+#levels(d$species)
+#d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
+d$Gen_sp <- d$species
+#sort(table(d$Gen_sp))
+#sum(table(d$Gen_sp))
+#head(d)
+
+#standardise dates
+date2 <- strptime(d$date, "%d/%m/%Y")
+d$jday <- date2$yday
+d$year <- 1900 + date2$year
+d$month <- date2$mon + 1
+#head(d$month)
+
+#select fields to be independent
+#completeFun <- function(data, desiredCols) {
+#  completeVec <- complete.cases(data[, desiredCols])
+#  return(data[completeVec, ])
+#}
+
+#d <- completeFun(d, c("day", "month", "year"))
+uni <- as.data.frame(cbind(d$Gen_sp, d$locality, d$collector, d$jday, d$year))
+#str(uni)
+dup <- duplicated(uni)
+
+# remove duplicates
+d2 <- d[-which(dup == TRUE),]
+
+#retain only sweep netting and no data
+#keep <- c("sweep", "no_data") #method is missing for a number of records
+#d2 <- d2[d2$method %in% keep, ]
+
+#create dataframes to remove species with less that 30 records
+no.ind <- as.data.frame(sort(table(d2$Gen_sp))) #14 species with more than 30 datapoints
+species.remove <- filter(no.ind, Freq < 30)#create df with species that have less than 30 records
+#visualize the data: #not with bees
+#plot(d2$Long, d2$Lat)
+#library(maps)
+#map("world", col="red", add=TRUE)
+
+#plot(d2$Long, d2$Lat, ylim = c(-50,-32), xlim = c(163,183))
+#map("world", col="red", add=TRUE) #fix wrong lat/long
+
+All <- d2
+#############################
+# species models
+#############################
+#collection history per year
+B <- All
+frq <- table(factor(B$Gen_sp),factor(B$year))
+bin <- ifelse(frq > 0, 1, 0)
+
+i=1
+last_year2 <- c(1:length(bin[,1]))
+#for (i in 1:length(bin[,1])){
+  #  bin2 <- bin[i,-which(bin[i,] == 0)]
+  #  last_year2[i] <- names(tail(bin2,1))
+#}
+l_yBB <- data.frame(Gen_sp = rownames(bin), last_year = last_year2)
+#l_yBB #broken at i = 5 check.
+
+#year vector
+y <- as.numeric(colnames(bin))
+
+#weights per year (specimens collected/max specimens collected)
+
+effort <- c(1:length(y))
+wei <- c(c(1:length(y)))
+for (i in 1:length(y)){effort[i] <- sum(frq[,i])}
+for (i in 1:length(y)){wei[i] <- sum(frq[,i])/max(effort)}
+
+#### estimates for all sp
+#mB <- frq[,c(1:11)]
+#colnames(mB) <- c("estimate_bin", "SE_bin", "p_bin","estimate_frq", "SE_frq", "p_frq", "pred1880", "pred 2010","pred1880SE", "pred 2010SE","Logit")
+mB <- frq[,c(1:10)]
+colnames(mB) <- c("estimate_bin", "SE_bin", "p_bin","estimate_frq", "SE_frq", "p_frq", "pred1880", "pred 2015","pred1880SE", "pred 2015SE")
+try(for (i in 1:length(frq[,1])){
+    mB[i,1] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[2]
+    mB[i,2] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[4]
+    mB[i,3] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[8]
+    if (deviance(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))/df.residual(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit)) > 1.5) {
+        mB[i,4] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[2]
+        mB[i,5] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[4]
+        mB[i,6] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[8]
+        mB[i,7] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[1]
+        mB[i,8] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[2]
+        mB[i,9] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[1]
+        mB[i,10] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[2]
+    }
+    #mB[i,11] <- plogis(summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[2])-0.5
+    else {
+        mB[i,4] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[2]
+        mB[i,5] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[4]
+        mB[i,6] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[8]
+        mB[i,7] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[1]
+        mB[i,8] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[2]
+        mB[i,9] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[1]
+        mB[i,10] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[2]
+    }
+})
+
+############################################
+#plot model estimates and predictions
+############################################
+
+#prepare dataframe
+xx <- as.data.frame(mB)
+xx.cast <- dcast(xx, Var1 ~ Var2, value.var="Freq")
+xx.cast <- xx.cast[!(xx.cast$Var1 %in% species.remove$Var1),]
+
+#add direction change based on estimate and p-value
+xx.cast$frq_direction <- with(xx.cast, ifelse(
+    estimate_frq < 0 & p_frq < 0.05, 'negative', ifelse(
+        p_frq >= 0.05, 'stable', ifelse(
+            estimate_frq > 0 & p_frq >= 0.05, 'whoops', 'positive'))))
+
+####################################################
+#plot frequency estimates
+####################################################
+
+#need to add fly data and analyse bees and flies separately. Include all analyses in a multi-panel figure.
+p <- ggplot(xx.cast, aes(Var1, estimate_frq, group=Var1))
+p <- p + xlab(NULL) + ylab("Frequency estimate")
+p <- p + theme(text = element_text(size=18))
+p <- p + geom_point(aes(colour=frq_direction), size = 3)
+p <- p + geom_errorbar(aes(ymin=estimate_frq-SE_frq, ymax=estimate_frq+SE_frq, colour=frq_direction), width = 0)
+p <- p + geom_hline(aes(yintercept=0), linetype="dashed")
+p <- p + theme(panel.grid.minor = element_blank(),
+               panel.background = element_blank(),
+               axis.line = element_line(colour = "black")) +
+    theme(panel.border=element_rect(colour = "black", fill = "NA", size = 1)) +
+    theme(axis.text.y=element_text(angle= 360, hjust = 0.5, vjust = 0.5, size =8),
+          axis.title.y=element_text(size=30, vjust = 1),
+          axis.text.x=element_text(angle= 90, hjust = 0.5, vjust = 0.5, size =8),
+          axis.title.x=element_text(size=30, vjust = 1),
+          axis.text=element_text(colour = "black"))+
+    theme(strip.background = element_rect(colour="NA", fill=NA),
+          strip.text = element_text(size=10))
+p <- p + theme(axis.title.y=element_text(margin=margin(0,20,0,0)))
+p <- p + scale_colour_brewer(palette = "Dark2")
+p
+
+####################################################
+
+```
+
+Figure 3. Model estimated changes in the realtive occurance frequencey for diffent New Zealand bee species in collections (± 1 SE) over time.
 
@@
 As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors.
-We show in our case studies that collaboration with both museum curators and data collectors is key to understanding the data and treating it appropriately.
-We envision that rather than a central repository of museum collection records, we should aim to make digitalized data readily available and easy to share.
+Using case studies, we show that collaboration with both museum curators and data collectors is key to adequately understanding the data and treating it appropriately.
+We envision that instead of a central museum collection record repository, researchers and curators should simply aim to make digitalized data readily available and easy to share.
 Ultimately, centralisation of these data in existing platforms such as Gbif would facilitate free and widespread access.
-However, even when data is stored in different webpages or databases, it can be retried and joined with other data sets using open science tools (@ropensci) if it is easily accessible and has programmatic access (API e.g.
+However, even when data are stored in different webpages or databases, it can be retried and combined with other datasets using open science tools (@ropensci), providing these tools are accessible (API e.g.
 @heltmint).
 
-We also need to revolutionize the way that researchers collaborate with museums by foresting healthy bidirectional relationships.
-For instance, researchers collect massive numbers of specimens, but these are often inappropriately vouched, rendering them useless for research purposes.
-To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is adequate funding and recognition that accurate data recording and long term preservation are critical for research.
+We also need to revolutionize the way that researchers collaborate with museums, which should start with foresting healthy bidirectional relationships.
+For instance, researchers collect massive amounts of specimens, but these are often inappropriately vouchered, rendering them useless for research purposes.
+To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is adequate funding, and recognition that accurate data recording and long-term preservation are critical for research.
 [What about an interactive bee decline map? or Integration of DATA and analysis:]()
 
-Ultimately, to identify global trends in pollinator declines we require data from diverse geographic regions.
-In particular, we need to put increased emphasis on quantifying pollinator declines in regions outside of the US and Europe.
-Furthermore, we need to investigate whether there have been declines in pollinator groups other than bees.
-As far as we know, there have been few regional bee extinctions (@Ollerton, @Bartomeus2013), but declines are widespread in disturbed ecosystems of the US and Europe (@Bartomeus, @dePalma, @Carvalheiro).
-For most other pollinator taxa we know almost nothing.
-For many taxa, the first step is to identify and describe them and only then can we begin to document declines.
+Ultimately, to identify global trends in pollinator declines we require robust data, collected from diverse geographic regions.
+Further, it is crucial that data are analysed in the appropriate way.
+This requires researches identifying biases and filling in gaps where possible.
+We need to put increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees.
+For the US and Europe there have been few regional bee extinctions (@Ollerton, @Bartomeus2013), but declines are widespread in disturbed ecosystems (@Bartomeus, @dePalma, @Carvalheiro).
+For most other pollinator taxa and regions throughout the world we know almost nothing.
+Moving forward, the first step for many taxa will be to identify and describe species.
+Only then can we begin to document declines.
 

### 2018-03-20 ms.Rmd
added NZ data graphs to the manuscript

@@
 ```
-Fig 1: A global map showing evidence (% decline in richness) and Data deficient areas would be great.
+
+Figure 1: Global map showing bee species richness. Countries without available data are coloured in black. Data were downloaded from the Discover Life website.
 
@@
 #str(lhf)
-lhf[order(lhf$pre1980),c(1,17,18)]
+lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
 plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
@@
 Changes in species richness over time
+
 We found that rarefied richness for both native and exotic bees increased overtime, but these trends were non-significant (P-values for both groups > 0.05).
@@
 
-Changes in species relative abundances over time 
-We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Figure 3).
-However, only two out of 14 fly species increased in relative abundance, both of which were exotic, whereas four species decreased in abundance (three native and one exotic).
-
 ```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
-#load packages
 library(plyr)
@@
 library(vegan)
+
+########################################
+#read in NZ bee data
+########################################
+
+d <- read.csv("data/NZBees_allrecords.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+
+#select required variables
+d <- select(d, one_of(c("species", "date", "locality", "collector", "collection", "method", "exotic_native")))
+d <- na.omit(d) #remove NAs
+
+#remove honeybees
+d$species <- as.factor(d$species)
+d <- d[ ! d$species %in% c("Apis mellifera"), ]
+d$species <- droplevels(d$species)
+
+#need to correct space between genus and species
+#levels(d$species)
+#d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
+d$Gen_sp <- d$species
+#sort(table(d$Gen_sp))
+#sum(table(d$Gen_sp))
+#head(d)
+
+#standardise dates
+date2 <- strptime(d$date, "%d/%m/%Y")
+d$jday <- date2$yday
+d$year <- 1900 + date2$year
+d$month <- date2$mon + 1
+#head(d$month)
+
+#select fields to be independent
+#completeFun <- function(data, desiredCols) {
+#  completeVec <- complete.cases(data[, desiredCols])
+#  return(data[completeVec, ])
+#}
+
+#d <- completeFun(d, c("day", "month", "year"))
+uni <- as.data.frame(cbind(d$Gen_sp, d$locality, d$collector, d$jday, d$year))
+dup <- duplicated(uni)
+
+# remove duplicates
+d2 <- d[-which(dup == TRUE),]
+
+#retain only sweep netting and no data
+#keep <- c("sweep", "no_data") #method is missing for a number of records
+#d2 <- d2[d2$method %in% keep, ]
+
+native <- filter(d2, exotic_native == "native") %>% droplevels()
+exotic <- filter(d2, exotic_native == "exotic") %>% droplevels()
+
+#############################################
+#Resampling Analysis for natives species 
+#############################################
+
+par(mfrow = c(2,2), mar = c(4,4,2,1))
+
+#fin year = 0 quick and dirty
+#length(which(native$year < 1900))
+native <- subset(native, year > 1900)
+for (i in 3:10){
+    break1 <- quantile(native$year, probs = seq(0,1,1/i), na.rm = TRUE)
+    lab <- c(1:i)
+    for(j in 1:i){
+        lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
+    }
+    native$time_period_quant  <- cut(native$year, breaks = break1, labels = lab)
+    comm2 <- table(native$time_period_quant, factor(native$Gen_sp))
+    (x2 <- rarefy(comm2,150, se = TRUE)) #same result as my resampling...
+    #rar[[i-2+10]] <- x2
+    #str(x2)
+    time <- c(1:i)
+    for(j in 1:i){
+        time[j] <- (as.numeric(break1)[j] + as.numeric(break1)[j+1])/2
+    }
+    xx2 <- as.data.frame(x2)[1,]
+    se <- as.data.frame(x2)[2,]
+    #if(i == 3){
+        #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
+        #abline(v = 150)}
+    m <- lm(t(xx2)~ time)
+    summary(m)
+    corr <- cor(t(xx2), time, use = "complete.obs")
+    cor_dis <- c(1:1000)
+    for (k in 1:1000){
+        cor_dis[k] <- cor(sample(t(xx2),i), time, use = "complete.obs")
+    }
+    #hist(cor_dis)
+    #lines(c(corr, corr), c(0,200), col = "red")
+    (p <- pnorm(corr, mean = mean(cor_dis), sd = sd(cor_dis)))
+    #with lm
+    #if(summary(m)$coef[8]> 0.05 | is.na(summary(m)$coef[8]) == TRUE){
+    #abline(m, lty = 2)}
+    #else{
+    #abline(m, lty = 1)}    
+    #with corrs
+    #if(p > 0.05){
+        #abline(m, lty = 2)}
+    #else{
+        #abline(m, lty = 1)}    
+    #if(i == 10){
+        #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
+        #abline(v = 150)}    
+    #rarecurve(comm2)
+    #abline(v = 1000)
+
+}
+    plot(time,t(xx2), xlab = "", las = 2, ylab = "Number of native bee species", main = "", xaxt = "n", ylim = c(min(xx2, na.rm = TRUE)- max(se), max(xx2, na.rm = TRUE)+ max(se)))
+    axis(1, at = time ,labels = lab, las = 2, cex.axis = 0.7)
+    arrows(time,as.numeric(xx2)+as.numeric(se), time, as.numeric(xx2)-as.numeric(se), angle=90, length = 0)
+    abline(m, lty = 2)
+
+#############################################
+#Resampling Analysis for exotic species 
+#############################################
+
+#length(which(exotic$year < 1900))
+exotic <- subset(exotic, year > 1900)
+for (i in 3:6){ #error saying break are not unique - how to fix?
+    break1 <- quantile(exotic$year, probs = seq(0,1,1/i), na.rm = TRUE)
+    lab <- c(1:i)
+    for(j in 1:i){
+        lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
+    }
+    exotic$time_period_quant  <- cut(exotic$year, breaks = break1, labels = lab)
+    comm2 <- table(exotic$time_period_quant, factor(exotic$Gen_sp))
+    (x2 <- rarefy(comm2,25, se = TRUE)) #same result as my resampling...
+    #rar[[i-2+10]] <- x2
+    #str(x2)
+    time <- c(1:i)
+    for(j in 1:i){
+        time[j] <- (as.numeric(break1)[j] + as.numeric(break1)[j+1])/2
+    }
+    xx2 <- as.data.frame(x2)[1,]
+    se <- as.data.frame(x2)[2,]
+    #if(i == 3){
+        #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
+        #abline(v = 25)}
+    m <- lm(t(xx2)~ time)
+    summary(m)
+    corr <- cor(t(xx2), time, use = "complete.obs")
+    cor_dis <- c(1:1000)
+    for (k in 1:1000){
+        cor_dis[k] <- cor(sample(t(xx2),i), time, use = "complete.obs")
+    }
+    #hist(cor_dis)
+    #lines(c(corr, corr), c(0,200), col = "red")
+    (p <- pnorm(corr, mean = mean(cor_dis), sd = sd(cor_dis)))
+    #with lm
+    #if(summary(m)$coef[8]> 0.05 | is.na(summary(m)$coef[8]) == TRUE){
+    #abline(m, lty = 2)}
+    #else{
+    #abline(m, lty = 1)}    
+    #with corrs
+    #if(p > 0.05){
+      # abline(m, lty = 2)}
+   # else{
+       # abline(m, lty = 1)}    
+   # if(i == 10){
+        #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
+        #abline(v = 25)}    
+    #rarecurve(comm2)
+    #abline(v = 1000)
+}
+    plot(time,t(xx2), xlab = "", las = 2, ylab = "Number of exotic bee species", main = "", xaxt = "n", ylim = c(min(xx2, na.rm = TRUE)- max(se), max(xx2, na.rm = TRUE)+ max(se)))
+    axis(1, at = time ,labels = lab, las = 2, cex.axis = 0.7)
+    arrows(time,as.numeric(xx2)+as.numeric(se), time, as.numeric(xx2)-as.numeric(se), angle=90, length = 0)
+    abline(m, lty = 2)
+
+############################################
+#resampling analysis for flies
+############################################
+
+    #load data
+    d <- read.csv(file = "data/NZ_fly_pollinators.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+
+    #remove unidentified species
+    completeFun <- function(data, desiredCols) {
+        completeVec <- complete.cases(data[, desiredCols])
+        return(data[completeVec, ])
+    }
+
+    d <- completeFun(d, "exotic_native")
+
+    d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
+
+    #date
+    date2 <- strptime(d$CollectionDateISOFrom, "%Y\\%m\\%d")
+    d$jday <- date2$yday
+    d$year <- 1900 + date2$year
+    d$month <- date2$mon + 1
+
+    #select fields to be independent
+    #two approaches lat = 0 and others
+    lat <- subset(d, Lat != 0 & is.na(year) == F)#remove all records with no latitude recording
+    #unique fields by lata, long, day of the year and year
+    uni <- as.data.frame(cbind(lat$Gen_sp, lat$Lat, lat$Long, lat$jday, lat$year))
+    dup <- duplicated(uni)
+    #d2 <- d[-which(dup == TRUE),]
+    # lat/long = 0
+    #lat0 <- subset(d, Dlat == 0 & is.na(year) == F)
+    #str(lat0)
+    #uni2 <- as.data.frame(cbind(lat0$Gen, lat0$sp, lat0$LocalityStr, lat0$SubDivStr, lat0$StateProv, lat0$Country, lat0$jday, lat0$year))
+    #str(uni2)
+    #dup2 <- duplicated(uni2)
+
+    # remove duplicates
+    lat <- lat[-which(dup == TRUE),]
+    #lat0 <- lat0[-which(dup2 == TRUE),]
+    #d2 <- rbind(lat,lat0)
+    d2 <- lat
+    native <- filter(d2, exotic_native == "native") %>% droplevels()
+    exotic <- filter(d2, exotic_native == "exotic") %>% droplevels()
+
+    #length(which(native$year < 1900))
+    native <- subset(native, year > 1900)
+    for (i in 3:10){
+        break1 <- quantile(native$year, probs = seq(0,1,1/i))
+        lab <- c(1:i)
+        for(j in 1:i){
+            lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
+        }
+        native$time_period_quant  <- cut(native$year, breaks = break1, labels = lab)
+        comm2 <- table(native$time_period_quant, factor(native$Gen_sp))
+        (x2 <- rarefy(comm2,100, se = TRUE)) #same result as my resampling...
+        #rar[[i-2+10]] <- x2
+        #str(x2)
+        time <- c(1:i)
+        for(j in 1:i){
+            time[j] <- (as.numeric(break1)[j] + as.numeric(break1)[j+1])/2
+        }
+        xx2 <- as.data.frame(x2)[1,]
+        se <- as.data.frame(x2)[2,]
+        #if(i == 3){
+            #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
+            #abline(v = 100)}
+        m <- lm(t(xx2)~ time)
+        summary(m)
+        corr <- cor(t(xx2), time, use = "complete.obs")
+        cor_dis <- c(1:1000)
+        for (k in 1:1000){
+            cor_dis[k] <- cor(sample(t(xx2),i), time, use = "complete.obs")
+        }
+        #hist(cor_dis)
+        #lines(c(corr, corr), c(0,200), col = "red")
+        (p <- pnorm(corr, mean = mean(cor_dis), sd = sd(cor_dis)))
+        #with lm
+        #if(summary(m)$coef[8]> 0.05 | is.na(summary(m)$coef[8]) == TRUE){
+        #abline(m, lty = 2)}
+        #else{
+        #abline(m, lty = 1)}    
+        #with corrs
+        #if(p > 0.05){
+           # abline(m, lty = 2)}
+        #else{
+            #abline(m, lty = 1)}    
+        #if(i == 10){
+            #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
+            #abline(v = 40)}	
+        #rarecurve(comm2)
+        #abline(v = 1000)
+    }
+
+    plot(time,t(xx2), xlab = "", las = 2, ylab = "Number of native fly species", main = "", xaxt = "n", ylim = c(min(xx2, na.rm = TRUE)- max(se), max(xx2, na.rm = TRUE)+ max(se)))
+    axis(1, at = time ,labels = lab, las = 2, cex.axis = 0.7)
+    arrows(time,as.numeric(xx2)+as.numeric(se), time, as.numeric(xx2)-as.numeric(se), angle=90, length = 0)
+    abline(m, lty = 2)
+
+    ##########################################################################
+    #Resampling Analysis for exotic flies
+    ##########################################################################
+
+    #fin year = 0 quick and dirty
+    #length(which(exotic$year < 1900))
+    exotic <- subset(exotic, year > 1900)
+    for (i in 3:10){
+        break1 <- quantile(exotic$year, probs = seq(0,1,1/i))
+        lab <- c(1:i)
+        for(j in 1:i){
+            lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
+        }
+        exotic$time_period_quant  <- cut(exotic$year, breaks = break1, labels = lab)
+        comm2 <- table(exotic$time_period_quant, factor(exotic$Gen_sp))
+        (x2 <- rarefy(comm2,20, se = TRUE)) #same result as my resampling...
+        #rar[[i-2+10]] <- x2
+        #str(x2)
+        time <- c(1:i)
+        for(j in 1:i){
+            time[j] <- (as.numeric(break1)[j] + as.numeric(break1)[j+1])/2
+        }
+        xx2 <- as.data.frame(x2)[1,]
+        se <- as.data.frame(x2)[2,]
+        #if(i == 3){
+            #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
+            #abline(v = 20)}
+        m <- lm(t(xx2)~ time)
+        summary(m)
+        corr <- cor(t(xx2), time, use = "complete.obs")
+        cor_dis <- c(1:1000)
+        for (k in 1:1000){
+            cor_dis[k] <- cor(sample(t(xx2),i), time, use = "complete.obs")
+        }
+        #hist(cor_dis)
+        #lines(c(corr, corr), c(0,200), col = "red")
+        (p <- pnorm(corr, mean = mean(cor_dis), sd = sd(cor_dis)))
+        #with lm
+        #if(summary(m)$coef[8]> 0.05 | is.na(summary(m)$coef[8]) == TRUE){
+        #abline(m, lty = 2)}
+        #else{
+        #abline(m, lty = 1)}    
+        #with corrs
+        #if(p > 0.05){
+            #abline(m, lty = 2)}
+        #else{
+            #abline(m, lty = 1)}    
+        #if(i == 10){
+            #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
+            #abline(v = 40)}	
+        #rarecurve(comm2)
+        #abline(v = 1000)
+    }
+
+    plot(time,t(xx2), xlab = "", las = 2, ylab = "Number of exotic fly species", main = "", xaxt = "n", ylim = c(min(xx2, na.rm = TRUE)- max(se), max(xx2, na.rm = TRUE)+ max(se)))
+    axis(1, at = time ,labels = lab, las = 2, cex.axis = 0.7)
+    arrows(time,as.numeric(xx2)+as.numeric(se), time, as.numeric(xx2)-as.numeric(se), angle=90, length = 0)
+    abline(m, lty = 2)
+
+    ##########################################################################
+    #END
+    ##########################################################################
+
+```
+
+Figure 4. Changes in species richness for different pollinator groups in New Zealand over time. All trends were non-signifficant.
+
+Changes in species relative abundances over time
+
+We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Figure 3).
+Interestingly, the two exotic bee species that declined in relative abundance were both Bombus.
+Native bees that increased in relative abundance were mostly from the genus Leioproctus.
+Only two out of 14 fly species increased in relative abundance, both of which were exotic, whereas four species decreased in abundance (three native and one exotic).
+Native flies that decreased in relative abundance were all Syrphidae in the genus Helophilus.
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
+
+############################################
+#analyse the bee dataset
+############################################
 
@@
 
-All <- d2
 #############################
@@
 #collection history per year
-B <- All
+B <- d2
 frq <- table(factor(B$Gen_sp),factor(B$year))
@@
 
-i=1
-last_year2 <- c(1:length(bin[,1]))
+#i=1
+#last_year2 <- c(1:length(bin[,1]))
 #for (i in 1:length(bin[,1])){
@@
 #}
-l_yBB <- data.frame(Gen_sp = rownames(bin), last_year = last_year2)
-#l_yBB #broken at i = 5 check.
+#l_yBB <- data.frame(Gen_sp = rownames(bin), last_year = last_year2)
+##l_yBB #broken at i = 5 check.
 
@@
 #prepare dataframe
-xx <- as.data.frame(mB)
-xx.cast <- dcast(xx, Var1 ~ Var2, value.var="Freq")
-xx.cast <- xx.cast[!(xx.cast$Var1 %in% species.remove$Var1),]
+bee.est <- as.data.frame(mB)
+bee.est.cast <- dcast(bee.est, Var1 ~ Var2, value.var="Freq")
+bee.est.cast <- bee.est.cast[!(bee.est.cast$Var1 %in% species.remove$Var1),]
 
 #add direction change based on estimate and p-value
-xx.cast$frq_direction <- with(xx.cast, ifelse(
-    estimate_frq < 0 & p_frq < 0.05, 'negative', ifelse(
-        p_frq >= 0.05, 'stable', ifelse(
-            estimate_frq > 0 & p_frq >= 0.05, 'whoops', 'positive'))))
+bee.est.cast$change_in_occurance <- with(bee.est.cast, ifelse(
+    estimate_frq < 0 & p_frq < 0.05, 'Negative', ifelse(
+        p_frq >= 0.05, 'Stable', ifelse(
+            estimate_frq > 0 & p_frq >= 0.05, 'whoops', 'Positive'))))
+bee.est.cast$taxa <- "Bee"
+
+############################################
+#analyse fly dataset
+############################################
+
+#load data
+d <- read.csv(file = "data/NZ_fly_pollinators.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+
+#remove unidentified species
+completeFun <- function(data, desiredCols) {
+    completeVec <- complete.cases(data[, desiredCols])
+    return(data[completeVec, ])
+}
+
+d <- completeFun(d, "exotic_native")
+
+d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
+
+#date
+date2 <- strptime(d$CollectionDateISOFrom, "%Y\\%m\\%d")
+d$jday <- date2$yday
+d$year <- 1900 + date2$year
+d$month <- date2$mon + 1
+
+#select fields to be independent
+#two approaches lat = 0 and others
+lat <- subset(d, Lat != 0 & is.na(year) == F)#remove all records with no latitude recording
+#unique fields by lata, long, day of the year and year
+uni <- as.data.frame(cbind(lat$Gen_sp, lat$Lat, lat$Long, lat$jday, lat$year))
+dup <- duplicated(uni)
+#d2 <- d[-which(dup == TRUE),]
+# lat/long = 0
+#lat0 <- subset(d, Dlat == 0 & is.na(year) == F)
+#str(lat0)
+#uni2 <- as.data.frame(cbind(lat0$Gen, lat0$sp, lat0$LocalityStr, lat0$SubDivStr, lat0$StateProv, lat0$Country, lat0$jday, lat0$year))
+#str(uni2)
+#dup2 <- duplicated(uni2)
+
+# remove duplicates
+lat <- lat[-which(dup == TRUE),]
+#lat0 <- lat0[-which(dup2 == TRUE),]
+#d2 <- rbind(lat,lat0)
+d2 <- lat
+no.ind <- as.data.frame(sort(table(d2$Gen_sp))) #14 species with more than 30 datapoints
+species.remove <- filter(no.ind, Freq < 30)#create df with species that have less than 30 records
+
+no.ind.fly <- as.data.frame(sort(table(d2$Gen_sp))) #14 species with more than 30 datapoints
+species.remove.fly <- filter(no.ind.fly, Freq < 30)#create df with species that have less than 30 records
+colnames(species.remove.fly)[1] <- "Gen_sp"
+
+##########################################################################
+#species models
+##########################################################################
+
+#collection history per year
+B <- d2
+frq <- table(factor(B$Gen_sp),factor(B$year))
+bin <- ifelse(frq > 0, 1, 0)
+
+#i=1
+#last_year2 <- c(1:length(bin[,1]))
+#for (i in 1:length(bin[,1])){
+   #bin2 <- bin[i,-which(bin[i,] == 0)]
+    #last_year2[i] <- names(tail(bin2,1))
+#}
+#l_yBB <- data.frame(Gen_sp = rownames(bin), last_year = last_year2)
+#l_yBB #broken at i = 5 check.
+
+#year vector
+y <- as.numeric(colnames(bin))
+
+#weights per year (specimens collected/max specimens collected)
+
+effort <- c(1:length(y))
+wei <- c(c(1:length(y)))
+for (i in 1:length(y)){effort[i] <- sum(frq[,i])}
+for (i in 1:length(y)){wei[i] <- sum(frq[,i])/max(effort)}
+
+##########################################################################
+#estimates for all sp
+##########################################################################
+
+#mB <- frq[,c(1:11)]
+#colnames(mB) <- c("estimate_bin", "SE_bin", "p_bin","estimate_frq", "SE_frq", "p_frq", "pred1880", "pred 2010","pred1880SE", "pred 2010SE","Logit")
+mB <- frq[,c(1:10)]
+colnames(mB) <- c("estimate_bin", "SE_bin", "p_bin","estimate_frq", "SE_frq", "p_frq", "pred1880", "pred 2015","pred1880SE", "pred 2015SE")
+try(for (i in 1:length(frq[,1])){
+    mB[i,1] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[2]
+    mB[i,2] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[4]
+    mB[i,3] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[8]
+    if (deviance(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))/df.residual(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit)) > 1.5) {
+        mB[i,4] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[2]
+        mB[i,5] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[4]
+        mB[i,6] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[8]
+        mB[i,7] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[1]
+        mB[i,8] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[2]
+        mB[i,9] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[1]
+        mB[i,10] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[2]
+    }
+    #mB[i,11] <- plogis(summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[2])-0.5
+    else {
+        mB[i,4] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[2]
+        mB[i,5] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[4]
+        mB[i,6] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[8]
+        mB[i,7] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[1]
+        mB[i,8] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[2]
+        mB[i,9] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[1]
+        mB[i,10] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[2]
+    }
+})
+
+##########################################################################
+#plot model estimates and predictions
+##########################################################################
+
+#prepare dataframe
+fly.est <- as.data.frame(mB)
+fly.est.cast <- dcast(fly.est, Var1 ~ Var2, value.var="Freq")
+fly.est.cast <- fly.est.cast[!(fly.est.cast$Var1 %in% species.remove$Var1),]
+
+#add direction change based on estimate and p-value
+fly.est.cast$change_in_occurance <- with(fly.est.cast, ifelse(
+    estimate_frq < 0 & p_frq < 0.05, 'Negative', ifelse(
+        p_frq >= 0.05, 'Stable', ifelse(
+            estimate_frq > 0 & p_frq >= 0.05, 'whoops', 'Positive'))))
+fly.est.cast$taxa <- "Fly"
+
+#merge bee and fly estimate datasets
+est.all <- rbind(fly.est.cast,bee.est.cast)
+est.all$Var1 <- gsub("_", " ", est.all$Var1)
 
@@
 
-#need to add fly data and analyse bees and flies separately. Include all analyses in a multi-panel figure.
-p <- ggplot(xx.cast, aes(Var1, estimate_frq, group=Var1))
+p <- ggplot(est.all, aes(Var1, estimate_frq, group=taxa))
 p <- p + xlab(NULL) + ylab("Frequency estimate")
-p <- p + theme(text = element_text(size=18))
-p <- p + geom_point(aes(colour=frq_direction), size = 3)
-p <- p + geom_errorbar(aes(ymin=estimate_frq-SE_frq, ymax=estimate_frq+SE_frq, colour=frq_direction), width = 0)
+p <- p + theme(text = element_text(size=16))
+p <- p + facet_wrap(~taxa,scales = "free_x")
+p <- p + geom_point(aes(colour=change_in_occurance), size = 2)
+p <- p + geom_errorbar(aes(ymin=estimate_frq-SE_frq, ymax=estimate_frq+SE_frq, colour=change_in_occurance), width = 0)
 p <- p + geom_hline(aes(yintercept=0), linetype="dashed")
@@
     theme(panel.border=element_rect(colour = "black", fill = "NA", size = 1)) +
-    theme(axis.text.y=element_text(angle= 360, hjust = 0.5, vjust = 0.5, size =8),
-          axis.title.y=element_text(size=30, vjust = 1),
-          axis.text.x=element_text(angle= 90, hjust = 0.5, vjust = 0.5, size =8),
-          axis.title.x=element_text(size=30, vjust = 1),
+    theme(axis.text.y=element_text(angle= 360, hjust = 0.5, vjust = 0.5, size =14),
+          axis.title.y=element_text(size=22, vjust = 1),
+          axis.text.x=element_text(face = "italic", angle= 90, hjust = 0, vjust = 0.5, size =7),
+          axis.title.x=element_text(size=22, vjust = 1),
           axis.text=element_text(colour = "black"))+
     theme(strip.background = element_rect(colour="NA", fill=NA),
-          strip.text = element_text(size=10))
+          strip.text = element_text(size=16))
 p <- p + theme(axis.title.y=element_text(margin=margin(0,20,0,0)))
 p <- p + scale_colour_brewer(palette = "Dark2")
+p$labels$colour <- "Change in occurance"
 p
 
+####################################################
+#END
 ####################################################
@@
 
-Figure 3. Model estimated changes in the realtive occurance frequencey for diffent New Zealand bee species in collections (± 1 SE) over time.
+Figure 5. Model estimated changes (± 1 SE) in the relative occurance frequencey for diffent New Zealand bee and fly species in museum collections over time.
 

### 2018-03-20 script/new_script.R

@@ -74,6 +74,7 @@ p
 
 
 library(reshape2)
+#need to give unique dataframe names to each group of taxa estimates
 xx2_melt <- melt(xx2)
 colnames(xx2_melt)[2] <- "estimate"
 se_melt <- melt(se)
@@ -81,17 +82,18 @@ colnames(se_melt)[2] <- "se"
 xx2_join <- merge(xx2_melt, se_melt, by="variable")
 time <- as.data.frame(time)
 xx <- cbind(xx2_join,time)
-m <- lm(estimate~time, data=xx)
-summary(m)
+p <- predict(m)
+xxx <- cbind(xx, p)
 
-predict(m)
-
+#need to plot each plot separately and combine plots at end (can't facet them)
 #plot chnages in fly richness across 10 time periods
-p <- ggplot(xx2_join, aes(time, estimate))
-p <- p + xlab(NULL) + ylab("Number of fly species")
+p <- ggplot(xxx, aes(variable, estimate, group=1))
+p <- p + xlab("Year") + ylab("Number of fly species")
+p <- p + scale_x_discrete(labels=c("1909-1924","1924-1940","1940-1949","1949-1962","1962-1974","1974-1977","1977-1980","1980-1986","1986-1996","1996-2007"))
 p <- p + theme(text = element_text(size=18))
 p <- p + geom_point(size = 3)
-p <- p + geom_abline(intercept = 24.90350, slope = -0.05035,linetype="dashed")#intercpt seems too high??
+p <- p + geom_abline(intercept = 4.5868425, slope = 0.139648, linetype="dashed")#intercept = first year*coef-intercept
+                                                                            #slope = coef *10
 p <- p + geom_errorbar(aes(ymin=estimate-se, ymax=estimate+se), width = 0)
 p <- p + theme(panel.grid.minor = element_blank(),
                panel.background = element_blank(),
@@ -108,4 +110,10 @@ p <- p + theme(legend.position="none")
 p <- p + theme(axis.title.y=element_text(margin=margin(0,20,0,0)))
 p
 
+par(mfrow = c(2,2), mar = c(4,3,1,1))
+
+plot(time,t(xx2), xlab = "", las = 2, ylab = "Number of species", main = "Exotic flies", xaxt = "n", ylim = c(min(xx2, na.rm = TRUE)- max(se), max(xx2, na.rm = TRUE)+ max(se)))
[…]

### 2018-03-23 ms.Rmd
updtaed in text citations. Also added reference file and citation style file (Phil B).

@@
 output: html_document
+bibliography: data/references.xml
+csl: data/philb_style.xml
 ---
@@
 
-Animal pollinators are a critical component of both natural and production ecosystems worldwide, given their role in plant reproduction (@ollerton2012) and food security (@kleinetal2007).
+Animal pollinators are a critical component of both natural and production ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007].
 As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions.
@@
 The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult".
-Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically (@goulson2016, @potts2010).
+Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010].
 Yet the current status and population trends of most pollinator species worldwide remain unknown.
-For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category (@Nieto2014).
+For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category [@Nieto2014].
 For countries outside of Europe and the US, data on pollinator populations is almost non-existent.
 
-One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and including bees, flies, butterflies, beetles, birds, bats and even lizards (@Winfreeetal2011).
-Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution (@ollerton2018).
-However, bees are also extremely diverse, with more than 20,000 species currently described worldwide (@PickeringAscher), and require often require expert taxonomists for identification.
-The uneven distribution of bee researchers has resulted in geographical biases in bee decline research (@Martinetal2012) and there are also taxonomic biases toward species that are easier to identify, such bumblebees (@ArcherOikos).
+One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and including bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011].
+Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017].
+However, bees are also extremely diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and require often require expert taxonomists for identification.
+The uneven distribution of bee researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are also taxonomic biases toward species that are easier to identify, such bumblebees [#@ArcherOikos].
 Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.
 One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients.
-Despite critiques on the robustness of this approach (@ref), this is the most extensive source of data on pollinator populations currently available.
-In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models [using what parameters?] (@DePalma).
+Despite critiques on the robustness of this approach [@ref], this is the most extensive source of data on pollinator populations currently available.
+In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models (using what parameters?) [@DePalma2017].
 A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists.
-This approach was inspired by successful butterfly monitoring programs in countries such as the UK and is currently being extended to other pollinator taxa (@ref).
+This approach was inspired by successful butterfly monitoring programs in countries such as the UK and is currently being extended to other pollinator taxa [@ref].
 However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations.
@@
 The core aim of museums is to conserve and curate collections.
-Thus, they serve as a precious repository for specimens and also ensure high quality taxonomic identification (@primak).
+Thus, they serve as a precious repository for specimens and also ensure high quality taxonomic identification [@primak].
 Yet currently, the major bottleneck for researchers wanting to use contained within museum collections is the digitization of collection data.
-Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g.
-Bartomeus et al 2013, Scheper et al 2014).
-This is despite the fact that museum collections often have a number of biases [including...] (@Keil).
+Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., @Bartomeus2013a; #@Scheper2014).
+This is despite the fact that museum collections often have a number of biases [including...] [@Keil].
 But in many cases, museum collections are the only data we have to reconstruct the past.
@@
 Regardless of the method, it is important for researchers to consider aspects of biodiversity beyond species richness.
-In addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities (@winfree2016ELE).
-Environmental changes often results in few “winners” and many “losers” (@Bartomeusetal2013).
-Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments (@Stavertetal2018).
+In addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities [@Winfree2015].
+Environmental changes often results in few “winners” and many “losers” [@Bartomeus2013b].
+Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018].
 These changes can have important flow on effects for pollination of native plant species and crops.
@@
 
-Digitizing old specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geo-locate the coordinates of collection events (@ref and software) and (iii) store the data in a properly curated database (@ref on database management).
+Digitizing old specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@ref cory paper on taxonomy needs], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@ref on database management].
 Undertaking this process for the thousands of specimens in museum collections can be a daunting task and requires specialized personnel.
@@
 High resolution specimen photos and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers.
-In addition to this, new algorithms have been created that allow location geo-referencing based on vernacular names (@refs geonames, geoparser and opencage).
+In addition to this, new algorithms have been created that allow location geo-referencing based on vernacular names [@refs geonames, geoparser and opencage].
 [Explain zooniverse, NZ initiative and how data can be validated...
@@
 It is unsurprising that most evidence of pollinator declines is biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance.
-For example, in the UK and the Netherlands, which have experienced a multitude of widespread anthropogenic changes, a citizen science based study using and museum collections detected declines for bees, flies and flowering plants (@Beisjmieretal2006).
+For example, in the UK and the Netherlands, which have experienced a multitude of widespread anthropogenic changes, a citizen science based study using and museum collections detected declines for bees, flies and flowering plants [@Biesmeijer2006].
 However, it is important to note that even for these two countries, pollinator richness local estimates are biased toward large cities and regions dominated by agriculture, and data is lacking for natural areas.
-Further exploration of this dataset revealed that the declining trend for pollinator taxa has attenuated in recent decades (@carvalheiroetal).
-More detailed trends for common bee species in the Netherlands have also been explored using museum specimens and confirm the link between simultaneous plant and pollinator declines (@Schepper2014).
+Further exploration of this dataset revealed that the declining trend for pollinator taxa has attenuated in recent decades [@Carvalheiro2013].
+More detailed trends for common bee species in the Netherlands have also been explored using museum specimens and confirm the link between simultaneous plant and pollinator declines [#@Scheper2014].
 This study found that bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus were most threatened with extinction.
@@
 
-In comparison to local decline estimates, regional estimates for changes in species richness in the eastern US show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions (@Colla2012?).
-These findings are in stark contrast with the widespread local extinctions reported by Burkle (@Burkle), who compared historical observations of bee occurrence in a large forested ecosystem with the present-day remaining forest remnants.
+In comparison to local decline estimates, regional estimates for changes in species richness in the eastern US show a moderate decline in bee diversity [@Bartomeus2013a] and very few regional extinctions [@Colla2012].
+These findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with the present-day remaining forest remnants.
 However, it is important to note that this comparison does not take into account opportunistic bee species that use novel land-use types.
 Thus, studies may detect local richness declines, while regional richness remains relatively stable.
-In any case, there is strong concordance between local extinctions and regional declines (@Bartomeus_winfree2013), suggesting that local extinctions are indicators of regional population declines.
+In any case, there is strong concordance between local extinctions and regional declines [#@Bartomeus_Winfree2013], suggesting that local extinctions are indicators of regional population declines.
 
@@
 This is specially relevant to identify mechanisms of decline and adaptation.
-One particularly important source of information is the date of capture, which has allowed to track phenological advances congruent with contemporary climate change (@Bartomeusetal2011).
+One particularly important source of information is the date of capture, which has allowed to track phenological advances congruent with contemporary climate change [#@Bartomeus2011].
 In addition, as mentioned above, insect pollinator labels often include as information the host plant where collected.
-This information is precious to understand past and present species interactions (@tylianakis2010ELE).
-Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events (@kleinj_raemakers).
-Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length (@refAlaskaScience) and body size (@Oliveira) shrinkage with climate and land use change.
+This information is precious to understand past and present species interactions [#@Tylianakis2010ELE].
+Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [#@kleinj_raemakers].
+Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.
 [move this section below]()
@@
 
-Gbif (@ref) is a central repository centralizing occurrence data.
+Gbif [@ref] is a central repository centralizing occurrence data.
 Much of this data comes from museum, but it integrates several other sources.
-Alongside the popular statistical language R (@ref), gbif can be directly queried into your computer (@rgbif) and check data availability for your region of interest.
+Alongside the popular statistical language R [@ref], gbif can be directly queried into your computer [#@rgbif] and check data availability for your region of interest.
 In Figure 2 we show the number of modern and historical bee records already available per country.
@@
 Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis.
-On the contrary, for countries with large number of historical records, organized re-surveys should be organized (e.g.
-@cameron2013).
+On the contrary, for countries with large number of historical records, organized re-surveys should be organized [e.g.
+#@cameron2013].
 Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X...
@@
 Much of Aotearoa-New Zealand’s pollinator fauna is also relatively depauperate.
-For example, Aotearoa-New Zealand has only 27 species of native bees (@Donovan, 2007), which is a fraction of nearby Australia’s c.
-1600 species (REF).
-However, Aotearoa-New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems (@Newstrom & Robertson, 2005).
+For example, Aotearoa-New Zealand has only 27 species of native bees [@Donovan2007], which is a fraction of nearby Australia’s c.
+1600 species [@REF].
+However, Aotearoa-New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems [@Newstrom2005].
 Thus, Aotearoa-New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.
@@
 In global terms, human colonisation of Aotearoa-New Zealand was relatively recent (c.
-740 y) (@Wilmshurst et al.
-2008).
+740 y) [@Wilmshurst2008].
 Before human arrival, Aotearoa-New Zealand was predominately forested, but has since been dramatically altered by people.
-Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 (@Perry et al, 2014).
+Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 [@Perry2014].
 Therefore, human activity likely affected pollinator communities in Aotearoa-New Zealand long before surveys and collections began.
@@
 We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980.
-We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years (see @Bartomeusetal2013 for further details).
+We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years [see @Bartomeus2013a for further details].
 We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin.
@@
 Ultimately, centralisation of these data in existing platforms such as Gbif would facilitate free and widespread access.
-However, even when data are stored in different webpages or databases, it can be retried and combined with other datasets using open science tools (@ropensci), providing these tools are accessible (API e.g.
-@heltmint).
+However, even when data are stored in different webpages or databases, it can be retried and combined with other datasets using open science tools (#@ropensci), providing these tools are accessible [API e.g., #@heltmint].
 
@@
 We need to put increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees.
-For the US and Europe there have been few regional bee extinctions (@Ollerton, @Bartomeus2013), but declines are widespread in disturbed ecosystems (@Bartomeus, @dePalma, @Carvalheiro).
+For the US and Europe there have been few regional bee extinctions [@Ollerton; @Bartomeus2013c] (#check this Bartomeus reference), but declines are widespread in disturbed ecosystems [@Bartomeus2013a; @DePalma2017; @Carvalheiro2013].
 For most other pollinator taxa and regions throughout the world we know almost nothing.
@@
 We also thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.
+
+#References

### 2018-03-26 ToDo.Rmd
cleaning repo

---
title: "Untitled"
author: "I. Bartomeus"
date: "26/03/2018"
output: html_document
---

To Do:

- check formating rules (words in abstract, main body, etc...)
- correct fig1 for country size?
- Species declines in % for different taxa... what to do?
- think on gbif plot...
- Spain.

#refs clarifications

@Ollerton2012 
@Winfree2011 - >

### 2018-03-26 SP_data_cleaning.Rmd
Spanish data

@@
 output: html_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 head(sort(table(sp$LOCALIDAD), decreasing = TRUE))
-write.csv(sort(table(sp$LOCALIDAD), decreasing = TRUE),
-          "raw_data/localidades.csv")
+#write.csv(sort(table(sp$LOCALIDAD), decreasing = TRUE),
+ #         "raw_data/localidades.csv")
 
@@
 
+sp2 <- subset(sp, LOCALIDAD %in% c('"Finca Zamadue?as"',
+                            "Toro",
+                            "Granja Escuela Jose Antonio",
+                            "Villab??ez",
+                            '""Finca Zamadue?as""',
+                            "Hornillos",
+                            '"Camino Hondo" Granja Escuela Jose Antonio',
+                            "Parc. 16 Granja Escuela Jose Antonio",
+                            "Mojados",
+                            "Coto Bajo Matallana",
+                            "San llorente",
+                            "San Llorente",
+                            "_Finca Zamadue?as_",
+                            "La Senovilla Olmedo",
+                            "Coto Bajo de Matallana Villalba de los Alcores",
+                            "Mojados",
+                            "Olmedo",
+                            "Coto Bajo Matallana Villalba de los Alcores",
+                            "Las Fuentes Mojados",
+                            "Paramo de Matallana",
+                            "Villalba de los Alcores",
+                            "_Coto Bajo de Matallana_",
+                            "_Coto Bajo de Matallana_",
+                            "Coto Bajo de Matallana Montealegre",
+                            "Coto Bajo Matallana",
+                            "Las Fuentes  Mojados",
+                            "P?ramo de Matallana Villanubla",
+                            "Olmedo"))
+
+head(sp2)
+
+sp2 <- droplevels(sp2)
+levels(sp2$LOCALIDAD) <- c("zamaduenas",
+                           "auditorio_Delibes", 
+                           "zamaduenas",
+                           "auditorio_Delibes",
+                           "hornillos",
+                           "olmedo",
+                           "olmedo",
+                           "auditorio_Delibes",
+                           "sanbernardo",
+                           "sanbernardo",
+                           "toro",
+                           "vilalba",
+                           "vilalba")
+
 #check sampling completness
@@
 #need to fix first locality
+head(sp2)
 comm <- table(sp2$LOCALIDAD, sp2$gen_sp)
@@
 
+```{r}
+#make final dataset
+sp3 <- sp2[,c("LOCALIDAD", "gen_sp", "NUM_EJEMPLARES")]
+now3 <- now[, c("LOCALIDAD", "gen_sp")]
+now3$NUM_EJEMPLARES <- 1
+now3$when <- "now"
+sp3$when <- "asensio"
+
+library(lettercase)
+str_ucfirst(str_lower_case(sp3$gen_sp[1:10]))
+
+sp3$gen_sp <- str_ucfirst(str_lower_case(sp3$gen_sp))
+
+dat <- rbind(sp3, now3)
+#check stuff
+head(dat)
+unique(dat$gen_sp) #400
+#this can be cleaned, but looks decent!
+
+colnames(dat) <- c("locality", "gen_sp", "freq", "when")
+unique(dat$locality) #400
+
+write.csv(dat, "data/spain_clean.csv", row.names = F)
+```
+
 1) only for 7 localities, join now and then and paired t-test on rarefied richness per site
 
+1.1) Show raref per site
+
 2) For common species we can do the same but with standardized frequency.

### 2018-03-26 ms.Rmd
Edit ms

@@
 However, despite reports that pollinator declines are widespread, data is scarce and often geographically and taxonomically biased.
-This limits robust inference about any potential “pollination crisis”.
-[something about how to address the problem].
-Analysis of historical time series data is the most direct approach for assessing species declines.
+This limits robust inference about any potential “pollinator crisis”.
+Analysis of historical time series data is the most direct approach for assessing species declines, but long term monitoring programms targetting pollinators are scarce.
+However, non-structured and often opportunistic historical data collections are more frequent.
 In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines.
 Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits).
-However, data of this kind are typically rare or non-existent for most pollinator taxa.
 Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and even pollination services.
@@
 Further, we demonstrate various approaches to analysing museum collection data by providing two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
-Finally, we suggest that there is immense potential for museum specimens to play a central role in assessing the extent of the pollination crisis globally.
+There is immense potential for museum specimens to play a central role in assessing the extent of the pollination crisis globally if we are able to make this data accesible and open.
 
@@
 
-Animal pollinators are a critical component of both natural and production ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007].
+Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007].
 As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions.
@@
 Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017].
-However, bees are also extremely diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and require often require expert taxonomists for identification.
-The uneven distribution of bee researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are also taxonomic biases toward species that are easier to identify, such bumblebees [#@ArcherOikos].
+However, bees are also extremely diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification.
+The uneven distribution of bee researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are also taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos].
 Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.
+
 One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients.
 Despite critiques on the robustness of this approach [@ref], this is the most extensive source of data on pollinator populations currently available.
-In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models (using what parameters?) [@DePalma2017].
+In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [@DePalma2017].
 A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists.
@@
 However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations.
-Finally, the most robust approach for assessing long-term pollinator population trends is to use museum specimen collection data.
+Finally, the most practical approach for assessing long-term pollinator population trends is to use museum specimen collection data.
 The core aim of museums is to conserve and curate collections.
-Thus, they serve as a precious repository for specimens and also ensure high quality taxonomic identification [@primak].
-Yet currently, the major bottleneck for researchers wanting to use contained within museum collections is the digitization of collection data.
+Thus, they serve as a precious repository for specimens at the same time that they ensure high quality taxonomic identification [@primak].
+Yet, the major bottleneck for researchers wanting to use data contained within museum collections is the digitization of collection data.
 Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., @Bartomeus2013a; #@Scheper2014).
@@
 
-Regardless of the method, it is important for researchers to consider aspects of biodiversity beyond species richness.
+Regardless of the method, it is important for researchers to go beyond aspects of biodiversity focused solely in species richness.
 In addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities [@Winfree2015].
@@
 It is unsurprising that most evidence of pollinator declines is biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance.
-For example, in the UK and the Netherlands, which have experienced a multitude of widespread anthropogenic changes, a citizen science based study using and museum collections detected declines for bees, flies and flowering plants [@Biesmeijer2006].
+For example, in the UK and the Netherlands, which have experienced a multitude of widespread anthropogenic changes, a citizen science based study using both observations and museum collections detected strong declines for bee, fly and flowering plant richness [@Biesmeijer2006].
 However, it is important to note that even for these two countries, pollinator richness local estimates are biased toward large cities and regions dominated by agriculture, and data is lacking for natural areas.
-Further exploration of this dataset revealed that the declining trend for pollinator taxa has attenuated in recent decades [@Carvalheiro2013].
+Further exploration of this dataset revealed that the declining trend for those pollinator taxa has attenuated in recent decades [@Carvalheiro2013].
 More detailed trends for common bee species in the Netherlands have also been explored using museum specimens and confirm the link between simultaneous plant and pollinator declines [#@Scheper2014].
@@
 These findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with the present-day remaining forest remnants.
-However, it is important to note that this comparison does not take into account opportunistic bee species that use novel land-use types.
-Thus, studies may detect local richness declines, while regional richness remains relatively stable.
+However, it is important to note that this comparison focus only on the woodland remnants and does not take into account the new opportunistic bee species that use novel land-use types (@collado2018).
+Thus, local studies may detect richness declines, while regional richness remains relatively stable.
 In any case, there is strong concordance between local extinctions and regional declines [#@Bartomeus_Winfree2013], suggesting that local extinctions are indicators of regional population declines.
@@
 #have plotted countries with missing values as black (shows we have richness for pretty much everywhere)
-map3 <- mapCountryData(map2, nameColumnToPlot = "species", missingCountryCol = "black",
+map3 <- mapCountryData(map2, nameColumnToPlot = "species", missingCountryCol = "light grey",
                mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)
+#change it to grey, blak was too contrasting!
 
@@
 
+#Nacho needs to correct for country size
+
 ```
 
-Figure 1: Global map showing bee species richness. Countries without available data are coloured in black. Data were downloaded from the Discover Life website.
+Figure 1: Global map showing bee species richness. Countries without available data are coloured in grey. Data were downloaded from the Discover Life website.
+
+```{r}
+#plot for species declines??
+
+```
 
@@
 As illustrated in Figure 1, there is a paucity of countries for which museum data is available, and hence can be used as baseline for assessing pollinator population declines.
-While ideally, one would aim to digitize all museum records in the world, in the practical world this is an unlikely to happen due mainly to funding constrains.
+While ideally, one would aim to digitize all museum records in the world, in the practical world this is unlikely to happen due mainly to funding constrains.
 Here we illustrate how this can be prioritized and optimized to encourage independent researchers to assess declines in its own country.
@@
 Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis.
-On the contrary, for countries with large number of historical records, organized re-surveys should be organized [e.g.
+On the contrary, for countries with large number of historical records, re-surveys should be organized [e.g.
 #@cameron2013].
@@
 
-plot(sort(lhf$percent_post1980), xlab = "post1980", ylab = "%", col = as.factor(lhf$country))
+#plot(sort(lhf$percent_post1980), xlab = "post1980", ylab = "%", col = as.factor(lhf$country))
 #Add country labels in x axes.
@@
 In this case, we contacted the original collectors to define their sampling protocols.
-We then resurveyed the same sites [using the same protocols?] 20 years later.
+We then resurveyed the same sites using the same protocols 20 years later.
 The New Zealand dataset includes a wide suite of collectors and collection locations.
-For this dataset, we ensured that there no clear changes to geographical [I haven’t actually looked at this as there weren’t and GPS coordinates in the data] and taxonomic collection patterns through time.
+For this dataset, we ensured that there is no clear change to geographical [I haven’t actually looked at this as there weren’t and GPS coordinates in the data]() and taxonomic collection patterns through time.
 For example, some collectors may have targeted particular taxa, at specific locations or during certain time periods, which would represent a bias in the data.
@@
 Essentially, these different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories.
-We hope this resource will encourage researchers to analyses data for regions where little is currently know about pollinator declines.
+We hope this resource will encourage researchers to analyze data for regions where there is little current information on pollinator declines.
 [include a repository where people can contribute?]
@@
 Spain has already digitalized a decent amount of occurrence data both historical, and recent (Figure 2).
-However, visual inspection of the data reveals that the data are clustered [in what way?].
+However, visual inspection of the data reveals that the data is clustered around a few localities.
 Further, historic records do not spatially match recent records, making them difficult to compare.
-For this dataset, most of the old records are located around Valladolid and were collected by Enrique Asensio and his team.
-There has been no recent sampling of bees in this area.
-However we found that Enrique systematically sampled seven locations and that there was additional data available at the "Museo de historia Nacional" and other minor collections.
-Digitization of these records, along with a re-survey of the locations that Enrique and his team collected from, provided us with an excellent dataset for before and after comparisons of bee communities at these sites.
+For this dataset, most of the old records are located around Valladolid and were collected by Enrique Asensio and his colaborators There has been no recent sampling of bees in this area.
+However we found that Enrique systematically sampled seven locations and that there was additional data available at the "Museo de Historia Nacional" and other minor collections.
+Digitization of these records, along with a re-survey of the locations that Enrique and his team collected from, provided us with an excellent dataset for a before and after comparison of bee communities at these sites.
 
@@
 
-New Zealand pollinator collections
 We used New Zealand bee collection records gathered from multiple sources, including university, research institute, museum and private collections.
@@
 
-Data analyses
 First, we filtered our original datasets so that data used for analyses only included unique collection events.
@@
 
-Changes in species richness over time
-
 We found that rarefied richness for both native and exotic bees increased overtime, but these trends were non-significant (P-values for both groups > 0.05).
 In contrast, native fly richness declined, whereas exotic fly richness increased, although results for these groups were also non-significant (P-values for both groups > 0.05).
+Note that rarefied richness is sensitive to species evenness...
 
@@
     for (i in 3:10){
-        break1 <- quantile(exotic$year, probs = seq(0,1,1/i))
+        break1 <- quantile(exotic$year, probs = seq(0,1,1/i)) #breaks can be equal, un tie those
+        if(any(duplicated(break1))){
+            break1[duplicated(break1)] <- break1[duplicated(break1)]+1
+        }
+        if(any(duplicated(break1))){
+            break1[duplicated(break1)] <- break1[duplicated(break1)]+1
+        }
         lab <- c(1:i)
@@
 
--	Do we need some sort of separate discussion or integrate into the results section?
-… Although this is surprising, it could simply indicate a change in the evenness of species collected over time as the method we used actually measures both richness and evenness.
+- Do we need some sort of separate discussion or integrate into the results section? i would say integrate!
+… Although this is surprising, it could simply indicate a change in the evenness of species collected over time as the method we used actually measures both richness and evenness. Yes.
 
@@
 As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors.
-Using case studies, we show that collaboration with both museum curators and data collectors is key to adequately understanding the data and treating it appropriately.
+Using two case studies, we show that collaboration with both museum curators and data collectors is key to adequately understanding the data and treating it appropriately.
 We envision that instead of a central museum collection record repository, researchers and curators should simply aim to make digitalized data readily available and easy to share.
 Ultimately, centralisation of these data in existing platforms such as Gbif would facilitate free and widespread access.
-However, even when data are stored in different webpages or databases, it can be retried and combined with other datasets using open science tools (#@ropensci), providing these tools are accessible [API e.g., #@heltmint].
+However, even when data are stored in different webpages or databases, it can be retrieved and combined with other datasets using open science tools (#@ropensci), providing these tools are accessible [API e.g., #@heltmint].
 
-We also need to revolutionize the way that researchers collaborate with museums, which should start with foresting healthy bidirectional relationships.
+We also need to revolutionize the way that researchers collaborate with museums, which should start with fostering healthy bidirectional relationships.
 For instance, researchers collect massive amounts of specimens, but these are often inappropriately vouchered, rendering them useless for research purposes.
-To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is adequate funding, and recognition that accurate data recording and long-term preservation are critical for research.
-[What about an interactive bee decline map? or Integration of DATA and analysis:]()
+To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is not adequate funding, and recognition that accurate data recording and long-term preservation are critical for research.
 
@@
 We thank E.
-Asensio (or maybe coauthor?), Patrick McQuinn, and Crona McMonagle for data entry.
+Asensio (or maybe coauthor?), Curro Molina,  Patrick McQuinn, and Crona McMonagle for data entry.
 Madrid museum, ITACyL.

### 2018-03-26 ms.Rmd
More edits, specially to Spain analysis

@@
 Further, we demonstrate various approaches to analysing museum collection data by providing two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
+We show for both regions that despite richness levels do not change significantly, winners and looser species emerge.
 There is immense potential for museum specimens to play a central role in assessing the extent of the pollination crisis globally if we are able to make this data accesible and open.
@@
 Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010].
-Yet the current status and population trends of most pollinator species worldwide remain unknown.
+Yet, the current status and population trends of most pollinator species worldwide remain unknown.
 For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category [@Nieto2014].
@@
 
-One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and including bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011].
+One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and includes different taxa sush as bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011].
 Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017].
 However, bees are also extremely diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification.
-The uneven distribution of bee researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are also taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos].
+The uneven distribution of pollinator researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are also taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos].
 Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.
@@
 
-Regardless of the method, it is important for researchers to go beyond aspects of biodiversity focused solely in species richness.
-In addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities [@Winfree2015].
-Environmental changes often results in few “winners” and many “losers” [@Bartomeus2013b].
-Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018].
-These changes can have important flow on effects for pollination of native plant species and crops.
-[Add something about traits of winners and losers?]()
-https://www.sciencedirect.com/science/article/pii/S0065250417300296
-[move this and expand to the discussion on the map?]()
+In this paper we first highligth the main technical dificulties of using museum specimens, we show current evidence of diversity decline and highligth the need to go beyond diversity despriptors.
+Finally we draw a road map on how to exploit museum data to fill the knowledge gaps and ilustrate with two ase studies how to snalyze this kind of data.
 
@@
 
-[how much do we want to extend on this? Is interesting info, but boring o read?]()
-
 ##Current evidence on pollinatior declines
@@
 Start with butterflies?
+[Jamie, if you can complete this section with the refernces you found about declines would be super!]()
 
@@
 
+#Plots look mainly white... I wonder if they are useful...
+
 ```
@@
 ## Beyond species occurrences
+
+[need integrate this two paragraphs!]()
+Regardless of the method, it is important for researchers to go beyond aspects of biodiversity focused solely in species richness.
+A recent survey highlighted that more than 90% of the papers studying pollintar responses to land use change focused on richness and abundance descriptors.
+But in addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities [@Winfree2015].
+Environmental changes often results in few “winners” and many “losers” [@Bartomeus2013b].
+Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018].
+These changes can have important flow on effects for pollination of native plant species and crops.
+[Add something about traits of winners and losers?]()
+https://www.sciencedirect.com/science/article/pii/S0065250417300296
 
@@
 Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.
-[move this section below]()
 
@@
 lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
-plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
-abline(h = 150000, v = 150000)
+#plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
+#abline(h = 150000, v = 150000)
 #plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.
@@
 
+```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
+lhf$percent_post1980 <- lhf$post1980 / (lhf$pre1980+lhf$post1980)
+lhf$percent_pre1980 <- lhf$pre1980 / (lhf$pre1980+lhf$post1980)
+
+#plot(sort(lhf$percent_post1980), xlab = "post1980", ylab = "%", col = as.factor(lhf$country))
+#Add country labels in x axes.
+
+plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
+abline(h = 8, v = 8)
+
+```
 Fig 2: number of bee occurrences pre 1980 and post 1980 in gbif.
@@
 Lower right are sites which needs to dig museum records and lower left are countries hard to get data.
-
-Alternative plots:  
-```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
-lhf$percent_post1980 <- lhf$post1980 / (lhf$pre1980+lhf$post1980)
-lhf$percent_pre1980 <- lhf$pre1980 / (lhf$pre1980+lhf$post1980)
-
-#plot(sort(lhf$percent_post1980), xlab = "post1980", ylab = "%", col = as.factor(lhf$country))
-#Add country labels in x axes.
-
-plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
-abline(h = 8, v = 8)
-
-```
-% makes it unidimensional.
-Simply logging the data may work, we can point out % of countries with > 1000 records in either axes, for example, etc...
-Correcting for country size would be the best BUT I don't think richness and size have a linear relationship, so we should account for that then?
+[we can point out % of countries with > 1000 records in either axes, for example, etc...
+would be cool to ggplot this and point where NZ and SP fall, for example.
+As well as the green and red dots in the lower right cuadrat]()
 
@@
 We hope this resource will encourage researchers to analyze data for regions where there is little current information on pollinator declines.
-[include a repository where people can contribute?]
 
@@
 In brief, we check sampling completeness at both time periods, and compare the richness level of each site before 1980 and after 1980 with a paired t-test.
-[Maybe more analyses arise...]().
-We show that ...
-
-Fig 3: Multi-panel figure
+We show that overall there is no richness loss (p > 0.05), as 3 sites present lower number of species, but other three sites present higher numbers.
+However, raw numbers show a steep deline...
+and groups like nomadas, tyreus, prosopis and teralonias are ocmpletely extirpated (appendix?).
+
+```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
+dat <- read.csv("data/spain_clean.csv")
+#head(dat)
+library(reshape2)
+dat2 <- melt(dat, id.vars = c("locality", "gen_sp", "when"), 
+             measure.vars = "freq")
+#head(dat2, 100)
+#sort(unique(dat2$gen_sp))
+#clean with clean_sp
+#remove sp, NA, " ",
+#NA_NA   
+#Hylaeus_Variegatus
+
+#table(dat2$gen_sp, dat2$when)
+#Andrena_pilipes = carbonaria
+#remove A. mellifera
+
+#losts easely spotted: nomadas & Tyreus, prosopis, teralonias
+
+#calculate richness per site and time period
+#head(dat)
+rich <- dcast(dat, locality ~ when, value.var = "gen_sp", 
+              fun.aggregate = function(x) length(unique(x)))
+#rich
+#t.test(rich$asensio, rich$now, paired = TRUE)
+
+library(vegan)
+temp <- subset(dat, when == "now")
+comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
+              fun.aggregate = sum, na.rm = TRUE)
+rownames(comm)[-8] <- as.character(comm$locality[-8])
+comm2 <- comm[,-1]
+rowSums(comm2)
+temp <- specaccum(comm2, "random")
+#plot(temp)
+#rarecurve(comm2)
+now <- rarefy(x = comm2, sample = 120)
+
+temp <- subset(dat, when == "asensio")
+comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
+              fun.aggregate = sum, na.rm = TRUE)
+rownames(comm)[-8] <- as.character(comm$locality[-8])
+comm2 <- comm[,-1]
+rowSums(comm2)
+temp <- specaccum(comm2, "random")
+#plot(temp)
+#rarecurve(comm2)
+asensio <- rarefy(x = comm2, sample = 120)
+
+ttest <- t.test(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)], paired = TRUE)
+
+boxplot(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)], names = c("1980's", "2016"), ylab = "rarefies species richness", 
+            las = 1, border = "light grey")
+points(x = c(rep(1,5), rep (2,5)), y = c(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)]), col = rep(1:5,2), pch = 19)
+for(i in c(1,2,5,6,7)){
+    lines(x = 1:2, y = c(asensio[i], now[i]), col = i)
+}
+
+```
+Fig 3: Rarefied richness at the 5 plots... need to work more on the raw data.
 
@@
 
+IB was funded by FBBVA project.
 We thank E.
-Asensio (or maybe coauthor?), Curro Molina,  Patrick McQuinn, and Crona McMonagle for data entry.
-Madrid museum, ITACyL.
+Asensio for sharing his histrical data and knowledge.
+We thank Curro Molina, Patrick McQuinn, and Crona McMonagle for data entry.
+Madrid museum of Historia Natural and ITACyL for access to historical collections.
 We also thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.

### 2018-03-27 ms.Rmd
integrated comments/changes from Darren and myself. Have added segments to section on global pollinaotr declines (what we aready know)

@@
 author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
-output: html_document
+csl: data/philb_style.xml
+output:
+  html_document: default
+  word_document: default
 bibliography: data/references.xml
-csl: data/philb_style.xml
 ---
@@
 
-##Abstract
+#Abstract
 
 There is increasing concern about the decline of pollinators worldwide.
-However, despite reports that pollinator declines are widespread, data is scarce and often geographically and taxonomically biased.
-This limits robust inference about any potential “pollinator crisis”.
-Analysis of historical time series data is the most direct approach for assessing species declines, but long term monitoring programms targetting pollinators are scarce.
-However, non-structured and often opportunistic historical data collections are more frequent.
+However, despite reports that pollinator declines are widespread, data are scarce and often geographically and taxonomically biased.
+These biases limit robust inference about any potential “pollinator crisis”.
+Analysis of historical time series data is the most direct approach for assessing species declines, but long term monitoring programs targeting pollinators are scarce.
+However, non-structured and often opportunistic historical data collections are more frequently available.
 In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines.
@@
 We identify opportunities and limitations to progress in assessing pollinator declines at a global scale.
-Further, we demonstrate various approaches to analysing museum collection data by providing two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
-We show for both regions that despite richness levels do not change significantly, winners and looser species emerge.
-There is immense potential for museum specimens to play a central role in assessing the extent of the pollination crisis globally if we are able to make this data accesible and open.
-
-## Introduction
-
-Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007].
+Further, we demonstrate various approaches to analyzing museum collection data by providing two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
+We show that for both regions, winners and loser species emerge despite a lack of significant richness changes over time.
+There is immense potential for museum specimens to play a central role in assessing the global pollination crisis extent, but we must make these data feely accessible to researchers.
+
+#Introduction
+
+Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [# ???] and food security [@Klein2007].
 As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions.
-This issue was identified several decades ago, and was popularized in 1996 by the influential book "the forgotten pollinators".
-However, these early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time.
-These initial claims triggered the first efforts to asses the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines.
-The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult".
-Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010].
+The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@ref].
+However, early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time.
+These initial claims triggered the first efforts to assess the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines.
+The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@ref].
+Since then, studies on pollinator responses to various global change drivers have multiplied rapidly.
+Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @ Potts2010].
 Yet, the current status and population trends of most pollinator species worldwide remain unknown.
-For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category [@Nieto2014].
+For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species check number fell into the “Data Deficient” category [@Nieto2014].
 For countries outside of Europe and the US, data on pollinator populations is almost non-existent.
 
-One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and includes different taxa sush as bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011].
+One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011].
 Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017].
-However, bees are also extremely diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification.
-The uneven distribution of pollinator researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are also taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos].
-Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.
+However, bees are also diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification.
+Furthermore, the uneven distribution of pollinator researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos].
+Another barrier to assessing pollinator declines is that pollinators are often highly mobile, short-lived and small.
 
@@
 Despite critiques on the robustness of this approach [@ref], this is the most extensive source of data on pollinator populations currently available.
-In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [@DePalma2017].
+In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [DePalma2017].
 A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists.
@@
 Finally, the most practical approach for assessing long-term pollinator population trends is to use museum specimen collection data.
+
 The core aim of museums is to conserve and curate collections.
-Thus, they serve as a precious repository for specimens at the same time that they ensure high quality taxonomic identification [@primak].
-Yet, the major bottleneck for researchers wanting to use data contained within museum collections is the digitization of collection data.
-Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., @Bartomeus2013a; #@Scheper2014).
-This is despite the fact that museum collections often have a number of biases [including...] [@Keil].
-But in many cases, museum collections are the only data we have to reconstruct the past.
-
-In this paper we first highligth the main technical dificulties of using museum specimens, we show current evidence of diversity decline and highligth the need to go beyond diversity despriptors.
-Finally we draw a road map on how to exploit museum data to fill the knowledge gaps and ilustrate with two ase studies how to snalyze this kind of data.
-
-## The use of museum specimens
-
-Digitizing old specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@ref cory paper on taxonomy needs], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@ref on database management].
-Undertaking this process for the thousands of specimens in museum collections can be a daunting task and requires specialized personnel.
+Thus, they serve as a precious repository for specimens, and at the same time often ensure higher quality taxonomic identification [@primak].
+Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization.
+Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014).
+This is despite the fact that museum collections often have a number of biases including: the unknown sampling effort that was employed; the personal interests and curatorial techniques of collectors (e.g.
+discarding damaged individuals, only accessioning a certain number of individuals, targeting rare or unusual over common taxa); the spatial biases where areas have been undersampled, or where samples are biased towards easily collected localities (e.g.
+near towns/cities and/or along roadsides);that information is often restricted only to the presence of a species (i.e.
+there is no information on where a species is absent);and the difficulty of getting information on other taxa from the same location (e.g.
+NHCs are organised taxonomically, not geographically [#@Keil; other @refs?].
+Nevertheless, in many cases, museum collections are the only data we have to reconstruct the past.
+
+In this review we first highlight the major technical difficulties involved in using museum collection data.
+We then assess current evidence for diversity declines and highlight the need to move beyond simple diversity descriptors [this is not very clear – are we talking about richness?].
+Finally we present a road map outlining a strategy for using museum data to fill current knowledge gaps.
+We demonstrate how to approach analysis of different types for museum collection data to identify long-term trends in pollinator populations.
+
+#The use of museum specimens
+
+Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@ref cory paper on taxonomy needs], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@ref on database management].
+Undertaking this process for tens or hundreds of thousands of museum collections specimens can be a daunting task and requires specialized personnel.
 While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process.
-High resolution specimen photos and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers.
-In addition to this, new algorithms have been created that allow location geo-referencing based on vernacular names [@refs geonames, geoparser and opencage].
-[Explain zooniverse, NZ initiative and how data can be validated...
-Is there anything on AI to automate the process?
+High resolution photos of specimens and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers.
+In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage].
+[Explain zooniverse, NZ initiative and how data can be validated… Is there anything on AI to automate the process?
 e.g.
@@
 
-##Current evidence on pollinatior declines
-
-Start with butterflies?
-[Jamie, if you can complete this section with the refernces you found about declines would be super!]()
-
+#Current evidence on pollinatior declines
+
+At a global scale, current evidence of pollinator declines highly limited with most data restricted to the US and countries in Europe.
 It is unsurprising that most evidence of pollinator declines is biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance.
-For example, in the UK and the Netherlands, which have experienced a multitude of widespread anthropogenic changes, a citizen science based study using both observations and museum collections detected strong declines for bee, fly and flowering plant richness [@Biesmeijer2006].
-However, it is important to note that even for these two countries, pollinator richness local estimates are biased toward large cities and regions dominated by agriculture, and data is lacking for natural areas.
-Further exploration of this dataset revealed that the declining trend for those pollinator taxa has attenuated in recent decades [@Carvalheiro2013].
-More detailed trends for common bee species in the Netherlands have also been explored using museum specimens and confirm the link between simultaneous plant and pollinator declines [#@Scheper2014].
-This study found that bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus were most threatened with extinction.
+For example, in the UK and the Netherlands, a citizen science based study using both observations and museum collections detected strong declines for bee, hoverfly and flowering plant richness [@Biesmeijer2006].
+However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and data is lacking for natural areas.
+Further exploration of this dataset revealed that for declining pollinator taxa, the trend has attenuated in recent decades [@Carvalheiro2013].
+
+More detailed patterns for common bee species in the Netherlands have shown simultaneous plant and pollinator declines [#@Scheper2014].
+Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction.
 [check this] [- EU: UK?
@@
 Moths?
-Buterflies!
-Add Ollerton paper!]()
-
-In comparison to local decline estimates, regional estimates for changes in species richness in the eastern US show a moderate decline in bee diversity [@Bartomeus2013a] and very few regional extinctions [@Colla2012].
-These findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with the present-day remaining forest remnants.
-However, it is important to note that this comparison focus only on the woodland remnants and does not take into account the new opportunistic bee species that use novel land-use types (@collado2018).
-Thus, local studies may detect richness declines, while regional richness remains relatively stable.
+Butterflies!
+Add Ollerton paper!].
+Although studies of local pollinator communities often detect richness declines, regional richness may remain relatively stable.
+For example, regional estimates for species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012].
+These findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with remaining forest remnants.
+However, it is important to note that this study focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (#@collado2018).
 In any case, there is strong concordance between local extinctions and regional declines [#@Bartomeus_Winfree2013], suggesting that local extinctions are indicators of regional population declines.
 
+As with bees, studies on changes in species richness for other pollinator taxa is scarce and geographically restricted.
+Generally speaking, reported declines for bumblebees have been more severe.
+For example, declines of up to 18% in local bumble richness have been reported for Belgium and the Netherlands [@Carvalheiro2013].
+In other parts of Europe local richness declines range from 5% for Great Britain [@Carvalheiro2013] to 42% for Denmark [@Dupont].
+In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti] and 30% [@Bartomeus2013a].
+For butterflies, the only evidence of richness declines come from Europe.
+For example, butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although in Great Britain declines have been less severe [@Carvalheiro].
+Another study in Belgium showed that richness declines have been more severe (approximately 30%), although this study exampled changes in richness over a longer time period (early 1900’s to 2000) [@Maes].
+In parts of Germany, declines in local butterfly richness have been even more severe, with declines of up to 70% reported [@Wenzel].
+There are very few studies on changes in hoverfly species richness and are once again restricted to Europe.
+In Belgium, Great Britain and the Netherlands, changes in hoverfly richness have been modest [@Carvalheiro].
+In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer].
+Furthermore, directionality (richness increases or decreases) varies depending on the time period that richness changes have been assessed.
+For example, hoverfly richness decreased in Belgium decreased by approximately 6% from 1950-69 compared with 1970-80, but increased by approximately 10% between 1970-89 and 1990-2009 [@Carvalheiro2013].
+
 Outside of Europe and the US there are no studies on pollinator declines using historical museum collection records.
-[South Africa: Pauw, maybe with plants...]() .
-However, there are species-specific examples of historical (e.g., Bombus dalbhomi; Aizen).
-
-[add % in text also]()
-
+South Africa: Pauw, maybe with plants… .
+However, there are species-specific examples of historical (e.g., Bombus dalbhomi; #@Aizen).
+add % in text also
 Justify focus on bees?
@@
 
-## Beyond species occurrences
+#Beyond species occurrences
 
 [need integrate this two paragraphs!]()
+need integrate this two paragraphs!
 Regardless of the method, it is important for researchers to go beyond aspects of biodiversity focused solely in species richness.
-A recent survey highlighted that more than 90% of the papers studying pollintar responses to land use change focused on richness and abundance descriptors.
+A recent survey highlighted that more than 90% of the papers studying pollinator responses to land use change focused on richness and abundance descriptors.
 But in addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities [@Winfree2015].
-Environmental changes often results in few “winners” and many “losers” [@Bartomeus2013b].
+Environmental changes often results in a few species as “winners” and many species becoming “losers” [@Bartomeus2013b].
 Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018].
 These changes can have important flow on effects for pollination of native plant species and crops.
-[Add something about traits of winners and losers?]()
+Add something about traits of winners and losers?
 https://www.sciencedirect.com/science/article/pii/S0065250417300296
 
-Museum can provide valuable information besides species occurrence records, and this is important to take into account when digitizing the collections.
-This is specially relevant to identify mechanisms of decline and adaptation.
-One particularly important source of information is the date of capture, which has allowed to track phenological advances congruent with contemporary climate change [#@Bartomeus2011].
-In addition, as mentioned above, insect pollinator labels often include as information the host plant where collected.
+Museum can provide valuable information besides species occurrence records, and such information is important to capture when digitizing collections.
+This is specifically relevant to identify mechanisms of decline and adaptation.
+For example, one particularly important source of information is the date the specimen was collected, which allows the tracking of phenological advances congruent with contemporary climate change [#@Bartomeus2011].
+In addition, museum labels on pollinator specimens often include information about the host plant on which the specimen was collected.
 This information is precious to understand past and present species interactions [#@Tylianakis2010ELE].
@@
 
-## The way forward: Prioritizing the low hanging fruit.
-
-As illustrated in Figure 1, there is a paucity of countries for which museum data is available, and hence can be used as baseline for assessing pollinator population declines.
-While ideally, one would aim to digitize all museum records in the world, in the practical world this is unlikely to happen due mainly to funding constrains.
+#The way forward: Prioritizing the low hanging fruit.
+
+There is a paucity of countries for which museum data is available (Figure 1), and hence can be used as baseline for assessing pollinator population declines.
+While ideally, one would aim to digitize all museum records in the world, this is unlikely to happen in the near future due mainly to funding constraints.
+[from Darren: Not just about lack of funding, it is also about time.
+It could happen faster IF other things were not done but museums have competing priorities].
 Here we illustrate how this can be prioritized and optimized to encourage independent researchers to assess declines in its own country.
 
-Gbif [@ref] is a central repository centralizing occurrence data.
-Much of this data comes from museum, but it integrates several other sources.
-Alongside the popular statistical language R [@ref], gbif can be directly queried into your computer [#@rgbif] and check data availability for your region of interest.
+GBIF [#@ref] is a central repository centralizing occurrence data from around the globe.
+Much of this data comes from museums and taxonomic collections [from Darren: Wider than just a ‘museum’ in the traditional sense; also a lot goes on at govt research institutes that have a collection (eg like NZAC), and also sometimes at Universities,], but it integrates several other sources.
+Alongside the popular statistical language R [#@ref], GBIF can be directly queried into your computer [#@rgbif] and data availability checked for your region of interest.
 In Figure 2 we show the number of modern and historical bee records already available per country.
@@
 #@cameron2013].
-Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X...
-Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.
+Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X… Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.
 
@@
 ```
-Fig 2: number of bee occurrences pre 1980 and post 1980 in gbif.
-Upper right cuadrat are well covered countries, upper left cuadrat are countries that need resurvey of old sites.
+
+Figure 2.
+The number of bee occurrences pre 1980 and post 1980 in GBIF.
+Upper right quadrat are well covered countries, upper left quadrat are countries that need resurvey of old sites.
 Lower right are sites which needs to dig museum records and lower left are countries hard to get data.
+
 [we can point out % of countries with > 1000 records in either axes, for example, etc...
@@
 
-Once museum collection datasets are made available, researchers needs to gauge any potential biases in the data.
+Once museum collection datasets are made available, researchers need to gauge any potential biases in the data.
 We explore this process with two contrasting examples from Spain and New Zealand.
@@
 We then resurveyed the same sites using the same protocols 20 years later.
-The New Zealand dataset includes a wide suite of collectors and collection locations.
-For this dataset, we ensured that there is no clear change to geographical [I haven’t actually looked at this as there weren’t and GPS coordinates in the data]() and taxonomic collection patterns through time.
+Conversely, the New Zealand dataset includes a wide suite of collectors and collection locations.
+For this dataset, we ensured that there is no clear differences to geographical I haven’t actually looked at this as there weren’t and GPS coordinates in the data and taxonomic collection patterns through time.
 For example, some collectors may have targeted particular taxa, at specific locations or during certain time periods, which would represent a bias in the data.
-For these case studies, we provide annotated R scripts to analyse the different kinds of data (as described above).
+For these case studies, we provide annotated R scripts to analyse the different kinds of data (e.g.
+for Spain and New Zealand, as described above).
 Essentially, these different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories.
@@
 
-## Case study one: Spain
+#Case study one: Spain
 
@@
 In addition, Spain is a bee diversity hotspot and maintains a relatively heterogeneous landscape.
-Spain has already digitalized a decent amount of occurrence data both historical, and recent (Figure 2).
+Spain has already digitalized a large amount of occurrence data for both historical and recent periods (Figure 2).
 However, visual inspection of the data reveals that the data is clustered around a few localities.
 Further, historic records do not spatially match recent records, making them difficult to compare.
-For this dataset, most of the old records are located around Valladolid and were collected by Enrique Asensio and his colaborators There has been no recent sampling of bees in this area.
-However we found that Enrique systematically sampled seven locations and that there was additional data available at the "Museo de Historia Nacional" and other minor collections.
-Digitization of these records, along with a re-survey of the locations that Enrique and his team collected from, provided us with an excellent dataset for a before and after comparison of bee communities at these sites.
-
-In brief, we check sampling completeness at both time periods, and compare the richness level of each site before 1980 and after 1980 with a paired t-test.
-We show that overall there is no richness loss (p > 0.05), as 3 sites present lower number of species, but other three sites present higher numbers.
-However, raw numbers show a steep deline...
-and groups like nomadas, tyreus, prosopis and teralonias are ocmpletely extirpated (appendix?).
+For this dataset, most of the historical records are located around Valladolid and were collected by Enrique Asensio and his collaborators.
+There has been no recent sampling of bees in this area.
+However, we found that Enrique systematically sampled seven locations and that there was additional data available at the “Museo de Historia Nacional” and other minor collections.
+Digitization of these records, along with a re-survey of these locations provided an excellent dataset for a before and after comparison of bee communities at these locations.
+[from Darren: Has there been any changes in the land use between the two periods?]
+
+In brief, we check sampling completeness at both time periods, and compare the species richness level of each site before 1980 and after 1980 with a paired t-test.
+We show that overall there is no richness loss (p > 0.05) [Should it be “no difference in species richness”].
+However, raw abundance [is this what you mean by numbers?] show a steep decline… and groups like nomadas, tyreus, prosopis and teralonias are ocmpletely extirpated (appendix?).
+[from Darren: Any information on these groups.
+Are the specialists species, or more sensitive to land change?].
 
@@
 ```
-Fig 3: Rarefied richness at the 5 plots... need to work more on the raw data.
-
-##Case study two: NZ    
-In contrast to Spain, Aotearoa-New Zealand is an isolated oceanic archipelago, with a distinctive pollinator biota and a unique history of human occupation.
-Much of Aotearoa-New Zealand’s pollinator fauna is also relatively depauperate.
-For example, Aotearoa-New Zealand has only 27 species of native bees [@Donovan2007], which is a fraction of nearby Australia’s c.
-1600 species [@REF].
-However, Aotearoa-New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems [@Newstrom2005].
-Thus, Aotearoa-New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.
-
-In global terms, human colonisation of Aotearoa-New Zealand was relatively recent (c.
+
+Figure 3. Rarefied richness at the 5 plots... need to work more on the raw data.
+
+#Case study two: New Zealand
+
+In contrast to Spain, New Zealand is an isolated oceanic archipelago, with a distinctive pollinator biota and a unique history of human occupation.
+Much of New Zealand’s pollinator fauna is also relatively depauperate.
+For example, New Zealand has only 27 species of native bees [@Donovan2007], which is a fraction of nearby Australia’s c.
+1600 species [#@ref].
+However, New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems [@Newstrom2005].
+Thus, New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.
+
+In global terms, human colonisation of a-New Zealand was relatively recent (c.
 740 y) [@Wilmshurst2008].
-Before human arrival, Aotearoa-New Zealand was predominately forested, but has since been dramatically altered by people.
+Before human arrival, New Zealand was predominately forested, but has since been dramatically altered by people.
 Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 [@Perry2014].
-Therefore, human activity likely affected pollinator communities in Aotearoa-New Zealand long before surveys and collections began.
-Nevertheless, we can use museum records to identify trends in pollinator communities during Aotearoa-New Zealand’s more recent history.
+Therefore, human activity likely affected pollinator communities in New Zealand long before surveys and collections began.
+Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.
 
 We used New Zealand bee collection records gathered from multiple sources, including university, research institute, museum and private collections.
-Fly pollinator data was obtained from six participating New Zealand museums and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators.
+Collection records from the New Zealand Arthropod Collection (NZAC) are freely available online (https://scd.landcareresearch.co.nz/).
+Fly pollinator data was obtained from six participating New Zealand museums [need to list in the acknowledgments] and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators.
 Collections for the bee and fly datasets span over 100 years (early 1900s to late 2000s).
@@
 We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980.
-We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years [see @Bartomeus2013a for further details].
+We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years [@Bartomeus2013a].
 We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin.
 Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness.
-Thus reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.
+Thus, reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.
 
@@
 In contrast, native fly richness declined, whereas exotic fly richness increased, although results for these groups were also non-significant (P-values for both groups > 0.05).
-Note that rarefied richness is sensitive to species evenness...
+Note that rarefied richness is sensitive to species evenness, so increases in rarefied richness over time may actually indicate increased species evenness.
 
@@
 
-Figure 4. Changes in species richness for different pollinator groups in New Zealand over time. All trends were non-signifficant.
-
-Changes in species relative abundances over time
+Figure 4. Changes in rarefied species richness for different pollinator groups in New Zealand over time. All trends were non-significant (α = 0.05).
+
+##Changes in species relative abundances over time
 
 We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Figure 3).
-Interestingly, the two exotic bee species that declined in relative abundance were both Bombus.
+Interestingly, the two exotic bee species that declined in relative abundance were both in the genus Bombus, intentionally introduced into New Zealand for the pollination of crops.
 Native bees that increased in relative abundance were mostly from the genus Leioproctus.
@@
 
-Figure 5. Model estimated changes (± 1 SE) in the relative occurance frequencey for diffent New Zealand bee and fly species in museum collections over time.
-
-- Do we need some sort of separate discussion or integrate into the results section? i would say integrate!
-… Although this is surprising, it could simply indicate a change in the evenness of species collected over time as the method we used actually measures both richness and evenness. Yes.
-
-## The way forward
+Figure 5. Model estimated changes (± 1 SE) in the relative occurrence frequency for different New Zealand bee and fly species in museum collections over time.
+
+- Do we need some sort of separate discussion or integrate into the results section?
+i would say integrate!
+… Although this is surprising, it could simply indicate a change in the evenness of species collected over time as the method we used actually measures both richness and evenness.
+Yes.
+
+#The way forward
 
 As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors.
-Using two case studies, we show that collaboration with both museum curators and data collectors is key to adequately understanding the data and treating it appropriately.
+Using two case studies, we show that collaboration with both museum curators and ecologists is key to adequately understanding the data and treating it appropriately.
 We envision that instead of a central museum collection record repository, researchers and curators should simply aim to make digitalized data readily available and easy to share.
-Ultimately, centralisation of these data in existing platforms such as Gbif would facilitate free and widespread access.
-However, even when data are stored in different webpages or databases, it can be retrieved and combined with other datasets using open science tools (#@ropensci), providing these tools are accessible [API e.g., #@heltmint].
+[from Darren: I don’t think this is a good statement.
+It can be deleted and the discussion is still good.
+Saying “instead of a central museum..” is an argument against the main part of this paper (early in introduction) and also is against the discussion about collaboration].
+Ultimately, centralization of regional and national museum data in existing global platforms such as GBIF would facilitate free and widespread access.
+However, ecological [from Darren: Adding “ecological” means the data is different from the museum records in a central repository.
+Ecological datasets also do not fit well into museums databases unless there are voucher specimens from the ecology project] datasets can also be stored in different webpages or database repositories (e.g.
+Dryad), and can be retrieved and combined with other datasets using open science tools (#@ropensci), providing these tools are accessible [API e.g., # @heltmint].
 
 We also need to revolutionize the way that researchers collaborate with museums, which should start with fostering healthy bidirectional relationships.
-For instance, researchers collect massive amounts of specimens, but these are often inappropriately vouchered, rendering them useless for research purposes.
-To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is not adequate funding, and recognition that accurate data recording and long-term preservation are critical for research.
+For instance, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered, rendering them less useful for future research.
+To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is not adequate funding, and recognition that accurate data recording and long-term preservation are critical for research [@WardTEE].
 
 Ultimately, to identify global trends in pollinator declines we require robust data, collected from diverse geographic regions.
-Further, it is crucial that data are analysed in the appropriate way.
-This requires researches identifying biases and filling in gaps where possible.
-We need to put increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees.
-For the US and Europe there have been few regional bee extinctions [@Ollerton; @Bartomeus2013c] (#check this Bartomeus reference), but declines are widespread in disturbed ecosystems [@Bartomeus2013a; @DePalma2017; @Carvalheiro2013].
+Further, it is crucial that data are analysed appropriately.
+This requires researches to identify biases in the data and fill gaps in knowledge of pollinators [gaps in what?
+taxonomic, geographic?
+habitat?] where possible.
+We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees.
+For the US and Europe there have been few regional bee extinctions [#@Ollerton; Bartomeus2013c] (#check this Bartomeus reference), but in disturbed ecosystems declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006].
 For most other pollinator taxa and regions throughout the world we know almost nothing.
@@
 IB was funded by FBBVA project.
+DW was funded through Landcare Research within the Characterising New Zealand’s Land Biota Portfolio.
 We thank E.
-Asensio for sharing his histrical data and knowledge.
+Asensio for sharing his historical data and knowledge.
 We thank Curro Molina, Patrick McQuinn, and Crona McMonagle for data entry.

### 2018-03-27 script/map global pollinator richness changes.R

@@ -8,8 +8,9 @@ declines <- read.csv(file = "data/richness_declines.csv", header = TRUE, sep = "
 
 #aggregate data
 declines.ave <- declines %>% 
-    group_by(country, taxa) %>% 
+    group_by(country, taxa, study) %>% 
     summarise(average = mean(percent_richness_change),
+              median = median(percent_richness_change),
               max = max(percent_richness_change),
               min = min(percent_richness_change))
 

### 2018-03-27 ToDo.Rmd
Make quick notes on whats left to do.
integrated comments/changes from Darren and myself. Have added segments to section on global pollinaotr declines (what we aready know)
Add reference notes to TODO

@@
 - Species declines in % for different taxa... what to do?
+- for NZ species abundance analysis - should data be split by exotic/native first?
 - think on gbif plot...
 - Spain.
+- Refs
 
@@
 
-@Ollerton2012 
-@Winfree2011 - >
+@Ollerton2012 -> Ollerton winfree and tarrant Oikos, I think 2012.
+@Buchman1996 -> http://www.worldcat.org/title/forgotten-pollinators/oclc/34077459
+@NAS -> https://www.nap.edu/catalog/11761/status-of-pollinators-in-north-america
+@Winfree2011 -> Winfree bartomeus Cariveau ARES
+@PickeringAscher -> Ascher J.S., Pickering J. (2015) Discover Life bee species guide and world checklist (Hymenoptera: Apoidea: Anthophila). www.discoverlife.org
+@Matrin et al2012: https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/110154?journalCode=fron
+@Archer: https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1600-0706.2013.00949.x
+@França: França, Filipe, et al.
+"Do space‐for‐time assessments underestimate the impacts of logging on tropical biodiversity?
+An Amazonian case study using dung beetles." Journal of Applied Ecology 53.4 (2016): 1098-1105.
+@DePalma2018: De Palma, Adriana, et al.
+"Challenges With Inferring How Land-Use Affects Terrestrial Biodiversity: Study Design, Time, Space and Synthesis." Next Generation Biomonitoring 58 (2018): 163.
+@DePalma2017 De Palma, Adriana, et al.
+"Dimensions of biodiversity loss: Spatial mismatch in land‐use impacts on species, functional and phylogenetic diversity of European bees." Diversity and Distributions 23.12 (2017): 1435-1446.
+@DePalma2016: De Palma, Adriana, et al. "Predicting bee community responses to land-use changes: Effects of geographic and taxonomic biases." Scientific Reports 6 (2016): 31153.
+@vanSwaay: van Swaay, Chris AM, et al. "Butterfly monitoring in Europe: methods, applications and perspectives." Biodiversity and Conservation 17.14 (2008): 3455-3469.
+@Scheper2014. PNAS Museum specimens reveal loss of pollen host plants as key factor driving wild bee decline in The Netherlands
+@young Are pollinating hawk moths declining in the Northeastern United States? An analysis of collection records
+@helminth helminthR: an R interface to the London Natural History Museum's Host–Parasite Database
+@keil Biodiversity change is scale‐dependent: an example from Dutch and UK hoverflies (Diptera, Syrphidae)
+P Keil, JC Biesmeijer, A Barendregt, M Reemer, WE Kunin
+Ecography 34 (3), 392-401
+@@taxonomy papers (2):
+https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1523-1739.2004.00233.x
+Huber, J.
+T.
+"The importance of voucher specimens, with practical guidelines for preserving specimens of the major invertebrate phyla for identification." Journal of Natural History 32.3 (1998): 367-385.
+@ref geonames, geoparser and opencage: thos are likns to software.
+@Hart Ten simple rules for digital data storage. DOI: 10.7287/peerj.preprints.1448v1
+@collado2018: https://www.biorxiv.org/content/early/2018/03/08/278812
+@Bartomeus_Winfree2013 F1000 https://f1000research.com/articles/2-146/v1
+@Aizen Morales, Carolina L., et al. "Rapid ecological replacement of a native bumble bee by invasive species." Frontiers in Ecology and the Environment 11.10 (2013): 529-534.
+@Pauw: Reconstruction of historical pollination rates reveals linked declines of pollinators and plants
+A Pauw, JA Hawkins - Oikos, 2011
+Cited by 53 Related articles All 11 versions
+@Bartomeus2011 PNAS phenology
+@Tylianakis2010ELE Ecology letters It's 2008 sorry
+@Kleijn raemakers: https://www.ncbi.nlm.nih.gov/pubmed/18705369
+@refAlaskaScience http://www.sciencemag.org/content/349/6255/1541
+@Oliveira http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0148983
+@R just cite R
+@rgbif https://cran.r-project.org/web/packages/rgbif/rgbif.pdf
+@cameron2013] PNAS paper on bumblebees
+@OllertonScience Science paper on uk bee extinctions
+
+#to add
+
+@Mikheyev Museum samples reveal rapid evolution by wild honey bees exposed to a novel parasite

### 2018-03-27 ms.Rmd
Add reference notes to TODO

@@
 
-Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [# ???] and food security [@Klein2007].
+Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007].
 As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions.
-The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@ref].
+The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@Buchman1996].
 However, early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time.
 These initial claims triggered the first efforts to assess the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines.
-The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@ref].
+The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@nas].
 Since then, studies on pollinator responses to various global change drivers have multiplied rapidly.
-Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @ Potts2010].
+Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010].
 Yet, the current status and population trends of most pollinator species worldwide remain unknown.
@@
 However, bees are also diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification.
-Furthermore, the uneven distribution of pollinator researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos].
+Furthermore, the uneven distribution of pollinator researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos, @DePalma2016].
 Another barrier to assessing pollinator declines is that pollinators are often highly mobile, short-lived and small.
@@
 One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients.
-Despite critiques on the robustness of this approach [@ref], this is the most extensive source of data on pollinator populations currently available.
-In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [DePalma2017].
+Despite critiques on the robustness of this approach [@Franca, @DePalma2018], this is the most extensive source of data on pollinator populations currently available.
+In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [@DePalma2017].
 A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists.
-This approach was inspired by successful butterfly monitoring programs in countries such as the UK and is currently being extended to other pollinator taxa [@ref].
+This approach was inspired by successful butterfly monitoring programs [@vanSwaay] and is currently being extended to other pollinator taxa.
 However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations.
@@
 The core aim of museums is to conserve and curate collections.
-Thus, they serve as a precious repository for specimens, and at the same time often ensure higher quality taxonomic identification [@primak].
+Thus, they serve as a precious repository for specimens, and at the same time often ensure higher quality taxonomic identification.
 Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization.
-Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014).
+Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014, @Young).
 This is despite the fact that museum collections often have a number of biases including: the unknown sampling effort that was employed; the personal interests and curatorial techniques of collectors (e.g.
@@
 
-Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@ref cory paper on taxonomy needs], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@ref on database management].
+Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@taxonomy], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@hart].
 Undertaking this process for tens or hundreds of thousands of museum collections specimens can be a daunting task and requires specialized personnel.
@@
 Outside of Europe and the US there are no studies on pollinator declines using historical museum collection records.
-South Africa: Pauw, maybe with plants… .
+South Africa: @Pauw, maybe with plants… .
 However, there are species-specific examples of historical (e.g., Bombus dalbhomi; #@Aizen).
@@
 
-GBIF [#@ref] is a central repository centralizing occurrence data from around the globe.
+GBIF (www.gbig.com?) is a central repository centralizing occurrence data from around the globe.
 Much of this data comes from museums and taxonomic collections [from Darren: Wider than just a ‘museum’ in the traditional sense; also a lot goes on at govt research institutes that have a collection (eg like NZAC), and also sometimes at Universities,], but it integrates several other sources.
-Alongside the popular statistical language R [#@ref], GBIF can be directly queried into your computer [#@rgbif] and data availability checked for your region of interest.
+Alongside the popular statistical language R [#@R], GBIF can be directly queried into your computer [#@rgbif] and data availability checked for your region of interest.
 In Figure 2 we show the number of modern and historical bee records already available per country.
@@
 Ecological datasets also do not fit well into museums databases unless there are voucher specimens from the ecology project] datasets can also be stored in different webpages or database repositories (e.g.
-Dryad), and can be retrieved and combined with other datasets using open science tools (#@ropensci), providing these tools are accessible [API e.g., # @heltmint].
+Dryad), and can be retrieved and combined with other datasets using open science tools (ropensci), providing these tools are accessible [API e.g., # @heltmint].
 
@@
 We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees.
-For the US and Europe there have been few regional bee extinctions [#@Ollerton; Bartomeus2013c] (#check this Bartomeus reference), but in disturbed ecosystems declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006].
+For the US and Europe there have been few regional bee extinctions [#@OllertonScience; Bartomeus2013c] (#Ref bartomeus should be PNAS historical declines), but in disturbed ecosystems declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006].
 For most other pollinator taxa and regions throughout the world we know almost nothing.

### 2018-04-01 ToDo.Rmd
Draft in quite good shape

@@
 @Scheper2014. PNAS Museum specimens reveal loss of pollen host plants as key factor driving wild bee decline in The Netherlands
+@pearce PEARCE, J. L. and BOYCE, M. S. (2006), Modelling distribution and abundance with presence‐only data. Journal of Applied Ecology, 43: 405-412. doi:10.1111/j.1365-2664.2005.01112.x
 @young Are pollinating hawk moths declining in the Northeastern United States? An analysis of collection records
@@
 Ecography 34 (3), 392-401
-@@taxonomy papers (2):
+@taxonomy1
 https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1523-1739.2004.00233.x
-Huber, J.
+@taxonomy2 Huber, J.
 T.
 "The importance of voucher specimens, with practical guidelines for preserving specimens of the major invertebrate phyla for identification." Journal of Natural History 32.3 (1998): 367-385.
-@ref geonames, geoparser and opencage: thos are likns to software.
+@ref geonames, geoparser and opencage: those are likns to software.
 @Hart Ten simple rules for digital data storage. DOI: 10.7287/peerj.preprints.1448v1
@@
 
+@greve M.
+Greve, A.M.
+Lykke, C.W.
+Fagg, R.E.
+Gereau, G.P.
+Lewis, R.
+Marchant, A.R.
+Marshall, J.
+Ndayishimiye, J.
+Bogaert and J.-C.
+Svenning, Realising the potential of herbarium records for conservation biology, South African Journal of Botany, 105, (317)
+@lavoie Claude Lavoie, Biological collections in an ever changing world: Herbaria as tools for biogeographical and environmental studies, Perspectives in Plant Ecology, Evolution and Systematics, 15, 1, (68)
+X@Thomson Thomson SA, Pyle RL, Ahyong ST, Alonso-Zarazaga M, Ammirati J, Araya JF, et al.
+(2018) Taxonomy based on science is necessary for global conservation.
+PLoS Biol 16(3): e2005075.
+https://doi.org/10.1371/journal.pbio.2005075
+X@brief1: https://biologicalsurvey.ca/public/Bsc/Controller/Page/briefs/importance.pdf THE IMPORTANCE OF RESEARCH COLLECTIONS OF TERRESTRIAL ARTHROPODS
+X@brief2: http://biologicalsurvey.ca/assets/file/43 The role of voucher specimens in validating faunistic and ecological research
+X@Packer Packer, L.
+, Monckton, S.
+K., Onuferko, T.
+M., Ferrari, R.
+R.
+and Didham, R.
+(2018), Validating taxonomic identifications in entomological research.
+Insect Conserv Divers, 11: 1-12.
+doi:10.1111/icad.12284
+
+@Cox Cox, Paul Alan.
+“Extinction of the Hawaiian Avifauna Resulted in a Change of Pollinators for the Ieie, Freycinetia Arborea.” Oikos, vol.
+41, no.
+2, 1983, pp.
+195–199.
+JSTOR, JSTOR, www.jstor.org/stable/3544263.
+@Robbirt Robbirt, Karen M., et al.
+"Validation of biological collections as a source of phenological data for use in climate change studies: a case study with the orchid Ophrys sphegodes." Journal of Ecology 99.1 (2011): 235-241.
+@Maebe Maebe, K., Meeus, I., Maharramov, J.
+et al.
+Apidologie (2013) 44: 188.
+https://doi.org/10.1007/s13592-012-0170-9 Microsatellite analysis in museum samples reveals inbreeding before the regression of Bombus veteranus.
 @Mikheyev Museum samples reveal rapid evolution by wild honey bees exposed to a novel parasite

### 2018-04-01 ms.Rmd
Draft in quite good shape

@@
 ---
-title: "Museum specimens as baseline for assessing a potential global pollinator crisis"
+title: "Historical collections as baseline for assessing the extent of the pollinator crisis"
 author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
@@
   html_document: default
-  word_document: default
 bibliography: data/references.xml
@@
 ```
+
+[Historical collections or museum collections?]()
+
+*Affiliations:*
+1 Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, E-41092 Sevilla, Spain 
+2 
+3
+4 Andrena Soluciones
+*Correspondence: nacho.bartomeus@gmail.com
+
+*Author contributions*: IB wrote the initial draft. DW and OA provided data. IB and JS analyzed the data. All authors contributed to writting the manuscript.
 
@@
 Yet, the current status and population trends of most pollinator species worldwide remain unknown.
-For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species check number fell into the “Data Deficient” category [@Nieto2014].
+For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, > 55% of bee species fell into the “Data Deficient” category [@Nieto2014].
 For countries outside of Europe and the US, data on pollinator populations is almost non-existent.
@@
 One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011].
+All this groups of pollinators are often highly mobile, short-lived and small, difficulting the monitoring of its poplations.
 Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017].
 However, bees are also diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification.
-Furthermore, the uneven distribution of pollinator researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos, @DePalma2016].
-Another barrier to assessing pollinator declines is that pollinators are often highly mobile, short-lived and small.
+Furthermore, the uneven distribution of researchers [#@Martinetal2012] has also resulted in geographical biases in bee decline research, as well as taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos, @DePalma2016].
 
@@
 Despite critiques on the robustness of this approach [@Franca, @DePalma2018], this is the most extensive source of data on pollinator populations currently available.
-In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [@DePalma2017].
+For example, in Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [@DePalma2017].
 A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists.
@@
 However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations.
-Finally, the most practical approach for assessing long-term pollinator population trends is to use museum specimen collection data.
-
-The core aim of museums is to conserve and curate collections.
-Thus, they serve as a precious repository for specimens, and at the same time often ensure higher quality taxonomic identification.
-Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization.
-Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014, @Young).
-This is despite the fact that museum collections often have a number of biases including: the unknown sampling effort that was employed; the personal interests and curatorial techniques of collectors (e.g.
-discarding damaged individuals, only accessioning a certain number of individuals, targeting rare or unusual over common taxa); the spatial biases where areas have been undersampled, or where samples are biased towards easily collected localities (e.g.
-near towns/cities and/or along roadsides);that information is often restricted only to the presence of a species (i.e.
-there is no information on where a species is absent);and the difficulty of getting information on other taxa from the same location (e.g.
-NHCs are organised taxonomically, not geographically [#@Keil; other @refs?].
-Nevertheless, in many cases, museum collections are the only data we have to reconstruct the past.
-
-In this review we first highlight the major technical difficulties involved in using museum collection data.
-We then assess current evidence for diversity declines and highlight the need to move beyond simple diversity descriptors [this is not very clear – are we talking about richness?].
-Finally we present a road map outlining a strategy for using museum data to fill current knowledge gaps.
-We demonstrate how to approach analysis of different types for museum collection data to identify long-term trends in pollinator populations.
-
-#The use of museum specimens
-
-Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@taxonomy], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@hart].
-Undertaking this process for tens or hundreds of thousands of museum collections specimens can be a daunting task and requires specialized personnel.
-While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process.
-High resolution photos of specimens and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers.
-In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage].
-[Explain zooniverse, NZ initiative and how data can be validated… Is there anything on AI to automate the process?
-e.g.
-there is something on wing venation from photos at Netherlands.
-Can be mentioned?]
+Finally, the most practical approach for assessing long-term pollinator population trends is to use the available historical information on species occurrences deposited in museum specimen collection data [e.g.
+@bartomeus2013].
+
+In this review we first assess current evidence for diversity declines and present a road map outlining a strategy for using museum data to fill current knowledge gaps.
+We highlight the major technical difficulties involved in using museum collection data and demonstrate how to approach analysis of different types for museum collection data to identify long-term trends in pollinator populations.
+Finaly, we highlight the need to move beyond simple diversity descriptors and unleash the power of historical data to assess also species interactions, ecosyetm functioning and evolutionary changes along time.
 
@@
 
-More detailed patterns for common bee species in the Netherlands have shown simultaneous plant and pollinator declines [#@Scheper2014].
+More detailed patterns for common bee species in the Netherlands, aslo using museum historical specimens, have shown simultaneous plant and pollinator declines [#@Scheper2014].
 Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction.
-[check this] [- EU: UK?
-Ollerton Science, Peter Keil flies?
-Moths?
-Butterflies!
-Add Ollerton paper!].
 Although studies of local pollinator communities often detect richness declines, regional richness may remain relatively stable.
 For example, regional estimates for species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012].
-These findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with remaining forest remnants.
+A pattern also found in the UK [@OllertonScience].
+The US findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with remaining forest remnants.
 However, it is important to note that this study focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (#@collado2018).
@@
 
-Outside of Europe and the US there are no studies on pollinator declines using historical museum collection records.
-South Africa: @Pauw, maybe with plants… .
-However, there are species-specific examples of historical (e.g., Bombus dalbhomi; #@Aizen).
-add % in text also
-Justify focus on bees?
-We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes to highlight the little coverage that we have globally.
+Outside of Europe and the US there are no studies on pollinator declines using historical records.
+In Fig 1 we show a comparision of bee species richness per area (a), bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes (d) to highlight the little coverage that we have globally.
+See raw data at Sup Mat 1.
+For other parts of the world there are only species-specific examples of historical losses (e.g., Bombus dalbhomi; #@Aizen) [South Africa: @Pauw, maybe with plants… .
+However,Peter Keil flies?
+Moths?
+Butterflies!]()
 
@@
 map <- read.csv("data/bee_richness.csv")
-#Jamie, if you want, we could complete this data set based on the papers.
-# If two estimates available for the same country, I will use two rows. 
-# We can compute the mean or the max using reshape here.
-# For studies reporting large regions within country this is ok, I think, but
-# for studies reporting locations (like Burkle Science paper) the % richness loss would be inflated when scaling up... not sure how to add those up. Add a column maybe saying coverage
-# country, region, individual sites.
-#Bee richness can be gathered from Discover life webpage
-#head(map)
-map2 <- joinCountryData2Map(map, joinCode="NAME", nameJoinColumn = "country",
+size <- read.csv("data/country_area.csv")
+colnames(size) <- c("country", "size")
+map0 <- merge(map, size, all.x = TRUE)
+scatter.smooth(log(map0$species) ~ log(map0$size+1))
+m <- lm(log(map0$species) ~ log(map0$size+1), na.action = na.exclude) 
+map0$res <- residuals(m)
+map2 <- joinCountryData2Map(map0, joinCode="NAME", nameJoinColumn = "country",
                             nameCountryColumn = "distribution")
 #need to use ISO's to make sure we don't carry over NA's due to country misspellings
-#have plotted countries with missing values as black (shows we have richness for pretty much everywhere)
-map3 <- mapCountryData(map2, nameColumnToPlot = "species", missingCountryCol = "light grey",
+#have plotted countries with missing values as grey (shows we have richness for pretty much everywhere)
+map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "light grey",
                mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)
-#change it to grey, blak was too contrasting!
-
 #add a modified legend using the same initial parameters as mapCountryData
@@
 #Nacho needs to correct for country size
-
+#Need to 1) match all country names (includinh US and UK!)
+#2) check subsaharian countries maybe omit?
+
+#need to add all the pannels.
 ```
 
-Figure 1: Global map showing bee species richness. Countries without available data are coloured in grey. Data were downloaded from the Discover Life website.
-
-```{r}
-#plot for species declines??
-
-#Plots look mainly white... I wonder if they are useful...
-
-```
-
-#Beyond species occurrences
-
-[need integrate this two paragraphs!]()
-need integrate this two paragraphs!
-Regardless of the method, it is important for researchers to go beyond aspects of biodiversity focused solely in species richness.
-A recent survey highlighted that more than 90% of the papers studying pollinator responses to land use change focused on richness and abundance descriptors.
-But in addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities [@Winfree2015].
-Environmental changes often results in a few species as “winners” and many species becoming “losers” [@Bartomeus2013b].
-Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018].
-These changes can have important flow on effects for pollination of native plant species and crops.
-Add something about traits of winners and losers?
-https://www.sciencedirect.com/science/article/pii/S0065250417300296
-
-Museum can provide valuable information besides species occurrence records, and such information is important to capture when digitizing collections.
-This is specifically relevant to identify mechanisms of decline and adaptation.
-For example, one particularly important source of information is the date the specimen was collected, which allows the tracking of phenological advances congruent with contemporary climate change [#@Bartomeus2011].
-In addition, museum labels on pollinator specimens often include information about the host plant on which the specimen was collected.
-This information is precious to understand past and present species interactions [#@Tylianakis2010ELE].
-Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [#@kleinj_raemakers].
-Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.
+Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org).
+This is calculated as the residuals of the log-log regresion between bee richness and country size to account for the species-area relationships.
+We show mean local bee (b), syrphid (c) and butterflies (d) richness percent change in the last century.
+Warmer colors indicate steeper declines observed.
+Countries without available data are coloured in grey.
+
+#How to use historical records vouchered in museum specimens to fill the gaps.
+
+As evidenced in Fig 1, we lack specific estimates for most of the world countries and histrical data vouchered in museums may be the only change to fill those gaps.
+
+The core aim of museums is to conserve and curate collections.
+Thus, they serve as a precious repository for specimens, and at the same time often ensure higher quality taxonomic identification.
+Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization.
+Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@taxonomy, @taxonomy2, @Thomson], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@hart].
+Undertaking this process for tens or hundreds of thousands of museum collections specimens can be a daunting task and requires specialized personnel.
+While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process.
+High resolution photos of specimens and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers (e.g.
+https://www.zooniverse.org/).
+In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage].
+However, achieving all this goals needs propoer funding [@brief1].
+
+Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014, @Young).
+This is despite the fact that museum collections often have a number of biases.
+This biases includes the unknown sampling effort that was employed or the personal interests and curatorial techniques of collectors.
+For example, collectors tend to target rare or unusual over common taxa, discard damaged individuals or only accession a certain number of individuals.
+In addition, the opportunistic nature of museum collections often imply an spatial bias where areas have been undersampled, or where samples are biased towards easily collected localities (e.g.
+near towns/cities and/or along roadsides) and that information is often restricted only to the presence of a species (i.e.
+there is no information on where a species is absent).
+Given adequate sample sizes, most of this biases can be accounted for with propoer statistical data anslysis [e.g.
+@bartomeus2013, #@Keil, @pearce].
+Nevertheless, in many cases, museum collections are the only data we have to reconstruct the past.
 
@@
 
-There is a paucity of countries for which museum data is available (Figure 1), and hence can be used as baseline for assessing pollinator population declines.
+As we have seen, there is a paucity of countries for which museum data is available (Fig 1), and hence can be used as baseline for assessing pollinator population declines.
 While ideally, one would aim to digitize all museum records in the world, this is unlikely to happen in the near future due mainly to funding constraints.
-[from Darren: Not just about lack of funding, it is also about time.
-It could happen faster IF other things were not done but museums have competing priorities].
 Here we illustrate how this can be prioritized and optimized to encourage independent researchers to assess declines in its own country.
 
-GBIF (www.gbig.com?) is a central repository centralizing occurrence data from around the globe.
-Much of this data comes from museums and taxonomic collections [from Darren: Wider than just a ‘museum’ in the traditional sense; also a lot goes on at govt research institutes that have a collection (eg like NZAC), and also sometimes at Universities,], but it integrates several other sources.
+GBIF (https://www.gbif.org/) is a central repository centralizing occurrence data from around the globe.
+Much of this data comes from museums, collections and goverment research institutes, but it integrates several other sources.
 Alongside the popular statistical language R [#@R], GBIF can be directly queried into your computer [#@rgbif] and data availability checked for your region of interest.
@@
 #@cameron2013].
-Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X… Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.
+Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X… Also note that historical records are not always in local museums, but many EU and US museums contain large collection of exotic pollinators.
+Similar exploratry analysis can be easely done for other taxa.
 
@@
 Upper right quadrat are well covered countries, upper left quadrat are countries that need resurvey of old sites.
-Lower right are sites which needs to dig museum records and lower left are countries hard to get data.
-
-[we can point out % of countries with > 1000 records in either axes, for example, etc...
-would be cool to ggplot this and point where NZ and SP fall, for example.
-As well as the green and red dots in the lower right cuadrat]()
+Lower right are sites which needs to dig museum records and lower left are countries with very little data available.
+We highlight a few representative examples.
 
 Once museum collection datasets are made available, researchers need to gauge any potential biases in the data.
-We explore this process with two contrasting examples from Spain and New Zealand.
+We explore this process with two contrasting and ilustrative examples from Spain and New Zealand.
 In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people.
@@
 We then resurveyed the same sites using the same protocols 20 years later.
-Conversely, the New Zealand dataset includes a wide suite of collectors and collection locations.
-For this dataset, we ensured that there is no clear differences to geographical I haven’t actually looked at this as there weren’t and GPS coordinates in the data and taxonomic collection patterns through time.
-For example, some collectors may have targeted particular taxa, at specific locations or during certain time periods, which would represent a bias in the data.
+Conversely, the New Zealand dataset includes a wide suite of collectors and collection locations, but luckely, tehere is no obvious bias in the geographical and taxonomic coverage through time.
 For these case studies, we provide annotated R scripts to analyse the different kinds of data (e.g.
@@
 
-Spain provides an interesting study system because its natural habitats have been transformed extensively by humans, but land-use is not as intensive as many other European countries.
-In addition, Spain is a bee diversity hotspot and maintains a relatively heterogeneous landscape.
-Spain has already digitalized a large amount of occurrence data for both historical and recent periods (Figure 2).
+Spain provides an interesting study system because its natural habitats have been transformed extensively by humans, but land-use is not as intensive as in many other European countries.
+In addition, Spain is a bee diversity hotspot (Fig 1) and maintains a relatively heterogeneous landscape.
+Spain has already digitalized a large amount of occurrence data for both historical and recent periods (Fig 2).
 However, visual inspection of the data reveals that the data is clustered around a few localities.
@@
 In brief, we check sampling completeness at both time periods, and compare the species richness level of each site before 1980 and after 1980 with a paired t-test.
-We show that overall there is no richness loss (p > 0.05) [Should it be “no difference in species richness”].
-However, raw abundance [is this what you mean by numbers?] show a steep decline… and groups like nomadas, tyreus, prosopis and teralonias are ocmpletely extirpated (appendix?).
+We show that overall there is no no difference in species richness (p > 0.05).
+However, ...
+show a steep decline… and groups like nomadas, tyreus, prosopis and teralonias are completely extirpated (appendix?).
 [from Darren: Any information on these groups.
@@
 
-In global terms, human colonisation of a-New Zealand was relatively recent (c.
+In global terms, human colonisation of New Zealand was relatively recent (c.
 740 y) [@Wilmshurst2008].
@@
 Collection records from the New Zealand Arthropod Collection (NZAC) are freely available online (https://scd.landcareresearch.co.nz/).
-Fly pollinator data was obtained from six participating New Zealand museums [need to list in the acknowledgments] and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators.
+Fly pollinator data was obtained from six participating New Zealand museums [need to list in the acknowledgments]() and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators.
 Collections for the bee and fly datasets span over 100 years (early 1900s to late 2000s).
 
-First, we filtered our original datasets so that data used for analyses only included unique collection events.
+We follow Bartomeus et al.
+2013a protocol to analysze this data.
+First, we filtered our original datasets so that data used for analyses only included independent collection events.
 To do this, we removed specimens collected at the same location, on the same date, and by the same collector.
 We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980.
-We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years [@Bartomeus2013a].
+We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years.
 We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin.
@@
 
-To determine if the probability of finding a species in the collection changed over time, we used a general linear model with a binomial distribution and a logit link.
-[should we analyse natives and exotics together or separately?].
+Second, to determine if the probability of finding a species in the collection changed over time, we used a general linear model with a binomial distribution and a logit link.
 For species that showed overdispersion, we used a quasi-binomial distribution.
@@
 
-##Changes in species relative abundances over time
-
-We found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Figure 3).
+However, at the species level, we found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Fig 3).
 Interestingly, the two exotic bee species that declined in relative abundance were both in the genus Bombus, intentionally introduced into New Zealand for the pollination of crops.
@@
 
-- Do we need some sort of separate discussion or integrate into the results section?
-i would say integrate!
-… Although this is surprising, it could simply indicate a change in the evenness of species collected over time as the method we used actually measures both richness and evenness.
-Yes.
-
-#The way forward
+#Beyond species occurrences
+
+A recent survey highlighted that more than 90% of the papers studying pollinator responses to land use change focused solely on richness and abundance descriptors [@winfree2011].
+But in addition to local (alpha) diversity and regional (gamma) diversity, we need to understand changes in turnover between sites (beta diversity).
+Environmental changes often results in a few species as “winners” and many species becoming “losers” [@Bartomeus2013b].
+Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018].
+These changes can have important effects for pollination of native plant species and crops.
+
+But museum specimen collection can provide much more information besides species occurrence records, and such information is important to capture when digitizing collections.
+This is specifically relevant to identify mechanisms of decline and adaptation.
+For example, one particularly important source of information is the date the specimen was collected, which allows the tracking of phenological advances congruent with contemporary climate change [#@Bartomeus2011].
+In addition, museum labels on pollinator specimens often include information about the host plant on which the specimen was collected.
+This information is precious to understand past and present species interactions [#@Tylianakis2010ELE].
+Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [#@kleinj_raemakers].
+Finally, museum specimens can be measured to track evolutionary changes by measuring specific specimen traits.
+This has been already explored to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.
+
+#Conclucions
 
@@
 Using two case studies, we show that collaboration with both museum curators and ecologists is key to adequately understanding the data and treating it appropriately.
-We envision that instead of a central museum collection record repository, researchers and curators should simply aim to make digitalized data readily available and easy to share.
-[from Darren: I don’t think this is a good statement.
-It can be deleted and the discussion is still good.
-Saying “instead of a central museum..” is an argument against the main part of this paper (early in introduction) and also is against the discussion about collaboration].
-Ultimately, centralization of regional and national museum data in existing global platforms such as GBIF would facilitate free and widespread access.
-However, ecological [from Darren: Adding “ecological” means the data is different from the museum records in a central repository.
-Ecological datasets also do not fit well into museums databases unless there are voucher specimens from the ecology project] datasets can also be stored in different webpages or database repositories (e.g.
-Dryad), and can be retrieved and combined with other datasets using open science tools (ropensci), providing these tools are accessible [API e.g., # @heltmint].
+Researchers and curators should aim to make digitalized data readily available and easy to share.
+Centralization of regional and national museum data in existing global platforms such as GBIF would facilitate free and widespread access.
+However, specific datasets can also be stored in different webpages or database repositories (e.g.
+Universities and museum webpages, Dryad) as far as they are well documented and can be easely retrived and combined with other datasets using open science tools [e.g.
+ropensci, # @heltmint].
 
 We also need to revolutionize the way that researchers collaborate with museums, which should start with fostering healthy bidirectional relationships.
-For instance, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered, rendering them less useful for future research.
+For instance, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered (@Packer, @brief2), rendering them less useful for future research.
 To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is not adequate funding, and recognition that accurate data recording and long-term preservation are critical for research [@WardTEE].
@@
 Further, it is crucial that data are analysed appropriately.
-This requires researches to identify biases in the data and fill gaps in knowledge of pollinators [gaps in what?
-taxonomic, geographic?
-habitat?] where possible.
+This requires researches to identify biases in the data and fill  taxonomic and geographic gaps in knowledge of pollinators status where possible.
 We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees.
-For the US and Europe there have been few regional bee extinctions [#@OllertonScience; Bartomeus2013c] (#Ref bartomeus should be PNAS historical declines), but in disturbed ecosystems declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006].
+For the US and Europe there have been few regional bee extinctions [#@OllertonScience; @Bartomeus2013a] (#Ref bartomeus should be PNAS historical declines), but in disturbed ecosystems declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006].
 For most other pollinator taxa and regions throughout the world we know almost nothing.

### 2018-04-01 script/map global pollinator richness changes.R

@@ -5,6 +5,9 @@ library(reshape2)
 
 #read in data
 declines <- read.csv(file = "data/richness_declines.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+head(declines)
+
+#nothing outside EU- USA. Spain and NZ can be added.
 
 #aggregate data
 declines.ave <- declines %>% 

### 2018-04-03 ms.Rmd
silly edits
ms new version with gbif analysis!

@@
 bibliography: data/references.xml
+editor_options: 
+  chunk_output_type: console
 ---
@@
 map0 <- merge(map, size, all.x = TRUE)
-scatter.smooth(log(map0$species) ~ log(map0$size+1))
+#remove some countries with doubtful sampling
+#map0 <- subset(map0, !country %in% c("Kuwait", "Guinea-Bissau", "Chad", "Mali", "Mauritania"))
+#map0[which(is.na(map0$size)),] #some can be rescued?
+#scatter.smooth(log(map0$species) ~ log(map0$size+1))
 m <- lm(log(map0$species) ~ log(map0$size+1), na.action = na.exclude) 
 map0$res <- residuals(m)
+#subset(map0, res < -2)
 map2 <- joinCountryData2Map(map0, joinCode="NAME", nameJoinColumn = "country",
@@
 #have plotted countries with missing values as grey (shows we have richness for pretty much everywhere)
-map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "light grey",
+par(mfrow = c(2,2))
+map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "white",
                mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)
@@
 
-#Nacho needs to correct for country size
-#Need to 1) match all country names (includinh US and UK!)
-#2) check subsaharian countries maybe omit?
-
 #need to add all the pannels.
+library(plyr)
+library(dplyr)
+library(reshape2)
+
+#read in data
+declines <- read.csv(file = "data/richness_declines.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+#nothing outside EU- USA. Spain and NZ can be added.
+
+#aggregate data
+declines.ave <- declines %>% 
+    group_by(country, taxa, study) %>% 
+    summarise(average = mean(percent_richness_change),
+              median = median(percent_richness_change),
+              max = max(percent_richness_change),
+              min = min(percent_richness_change))
+
+####################################################
+#make syrphid map
+####################################################
+
+declines.syrphid <- as.data.frame(filter(declines.ave, taxa == "syrphid")) %>% droplevels()
+
+map.syrphid <- joinCountryData2Map(declines.syrphid, joinCode="NAME", nameJoinColumn = "country",
+                            nameCountryColumn = "distribution")
+
+#average richness change
+map.syrphid.ave <- mapCountryData(map.syrphid, nameColumnToPlot = "average", colourPalette="topo",
+                       mapTitle = "Average syrphid richness change", catMethod = "pretty", addLegend = F)
+do.call(addMapLegend, c(map.syrphid.ave, legendLabels="all", legendWidth=0.5))
+
+####################################################
+#make bee map
+####################################################
+
+declines.bee <- as.data.frame(filter(declines.ave, taxa == "bee")) %>% droplevels()
+
+map.bee <- joinCountryData2Map(declines.bee, joinCode="NAME", nameJoinColumn = "country",
+                                   nameCountryColumn = "distribution")
+
+#average richness change
+map.bee.ave <- mapCountryData(map.bee, nameColumnToPlot = "average", colourPalette="topo",
+                                  mapTitle = "Average bee richness change", catMethod = "pretty", addLegend = F)
+do.call(addMapLegend, c(map.bee.ave, legendLabels="all", legendWidth=0.5))
+
+####################################################
+#make butterfly map
+####################################################
+
+declines.butterfly <- as.data.frame(filter(declines.ave, taxa == "butterfly")) %>% droplevels()
+
+map.butterfly <- joinCountryData2Map(declines.butterfly, joinCode="NAME", nameJoinColumn = "country",
+                                     nameCountryColumn = "distribution")
+
+#average richness change
+map.butterfly.ave <- mapCountryData(map.butterfly, nameColumnToPlot = "average", colourPalette="topo",
+                                    mapTitle = "Average butterfly richness change", catMethod = "pretty", addLegend = F)
+do.call(addMapLegend, c(map.butterfly.ave, legendLabels="all", legendWidth=0.5))
+
+par(mfrow = c(1,1))
 ```
@@
 Warmer colors indicate steeper declines observed.
-Countries without available data are coloured in grey.
+Countries without available data are coloured in white.
+Note that some African countries may have incomplete faunas listed and that Alaska is included with US values.
 
@@
         lhf[i,1+j+7] <- as.numeric(occ_count(taxonKey= keys[j],
-                            from = 1850, to = 1980,
+                            from = 1850, to = 1979,
                             georeferenced=TRUE,
@@
 
+lhf <- lhf[-which(is.na(lhf$country)),]
 plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
-abline(h = 8, v = 8)
+abline(h = 7.5, v = 7.5)
+lhf$log_pre1980 <- log(lhf$pre1980)
+lhf$log_post1980 <- log(lhf$post1980)
+lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) < 7.5),]
+# poland switcherland need old records! > 10000 vs 300
+# sri lanka and french guayana, 
+# nicaragua, venezuela, Ecuador..>2000 vs >500.
+# Colombia & Panama, zimbawe > 3000 vs >1500
+nrow(lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),])
+lhf[which(lhf$country == "ES" | lhf$country == "NZ"),]
+col <- ifelse(lhf$country == "ES" | lhf$country == "NZ", "red", "black")
+col <- ifelse(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5, col, "grey")
+plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col)
+temp <- subset(lhf, country %in% c("ES", "NZ"))
+text(x = temp$log_post1980+1, y = temp$log_pre1980+1 , labels = temp$country)
+#36 countries!
+#exp(7.5) #1800 records.
+
+#Can we do rar richness per time period $ country and test? YES!! 
+#1) Select countries.
+countries <- lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),"country"]
+#2) Get all data
+occ_list <- list()
+for(i in 1:length(countries)){
+    taxon_list <- list()
+    count_list <- list()
+    for(j in 1:length(keys)){
+        taxon_list[[j]] <- occ_data(taxonKey= keys[j], 
+                            year = '1850,2018', 
+                            country = as.character(countries[i]),
+                            limit = 200000)
+        #count_list[[j]] <- occ_count(taxonKey= keys[j],
+                         #    from = 1850, to = 1979,
+                         #   georeferenced=TRUE,
+                         #   country = countries[i])
+    }
+    select <- c("name", "year")
+    out <- rbind(as.data.frame(taxon_list[[1]]$data[,select]), 
+                 as.data.frame(taxon_list[[2]]$data[,select]),
+                 as.data.frame(taxon_list[[3]]$data[,select]),
+                 as.data.frame(taxon_list[[4]]$data[,select]),
+                 as.data.frame(taxon_list[[5]]$data[,select]),
+                 as.data.frame(taxon_list[[6]]$data[,select]),
+                 as.data.frame(taxon_list[[7]]$data[,select]),
+                 as.data.frame(taxon_list[[8]]$data[,select]))
+    occ_list[[i]] <- out
+    print(i)
+}
+#save output
+save(occ_list, file = "data/gbif.RData")
+#3) split by year
+library(vegan)
+post1980 <- c()
+pre1980 <- c()
+n <- c()
+for(i in 1:length(countries)){
+    dat <- occ_list[[i]]
+    nrow(dat) 
+    dat1 <- dat[-which(is.na(dat$name)),]
+    nrow(dat1) # 
+    dat1$period <- ifelse(dat1$year < 1980, "pre1980", "post1980")
+    table(dat1$period) # :(
+#4) rar richness
+    comm <- dcast(dat1, period ~ name, fun.aggregate = length, value.var = "name")
+    rownames(comm) <- comm$period
+    comm <- comm[,-1]
+    temp <- rarefy(comm, sample = round(min(table(dat1$period))*0.75))
+    #str(temp)
+    post1980[i] <- temp[1]
+    pre1980[i] <- temp[2]
+    n[i] <- round(min(table(dat1$period))*0.75)
+}
+#5) t test!
+data <- data.frame(countries = as.character(countries), pre1980 = pre1980, post1980 = post1980, n = n)
+head(data)
+#maybe delete n < 500?
+data2 <- subset(data, n > 500)
+
+t.test(data2$pre1980, data2$post1980, paired = TRUE)
+
+boxplot(data2$pre1980, data2$post1980, names = c("pre1980", "post1980"), ylab = "rarefied species richness", 
+            las = 1, border = "light grey")
+points(x = c(rep(1,nrow(data2)), rep (2,nrow(data2))), y = c(data2$pre1980, data2$post1980), pch = 19)
+for(i in 1:nrow(data2)){
+    lines(x = 1:2, y = c(data2$pre1980[i], data2$post1980[i]))
+}
+
+#map
+data2$change <- ((data2$post1980 - data2$pre1980)*100 / data2$pre1980)
+
+map.gbif <- joinCountryData2Map(data2, joinCode="ISO2", nameJoinColumn = "countries",
+                            nameCountryColumn = "countries")
+
+#average richness change
+map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette="topo",
+                       mapTitle = "Average bee richness change", catMethod = "pretty", addLegend = F)
+do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))
 

### 2018-04-04 SP_data_cleaning.Rmd
minor
Rework spanish data

--- 2018-02-07 SP_data_cleaning.Rmd
@@
 #subset only the 7 localities.
+tesaurus <- read.csv("data/Locations.csv")
+head(tesaurus)
+levels(tesaurus$Localidad_Oscar) <- c("auditorio_Delibes",
+                                      "hornillos",
+                                      "hornillos",
+                                      "olmedo",
+                                      "olmedo",
+                                      "sanbernardo",
+                                      "toro",
+                                      "toro",
+                                      "vilalba",
+                                      "zamaduenas")
+#merge
+sp2 <- merge(sp, tesaurus, by.x = "LOCALIDAD", by.y = "Localidad_asensio")
+head(sp2)
 
@@
 #need to fix first locality
-comm <- table(sp2$LOCALIDAD, sp2$gen_sp)
+head(sp2)
+comm <- table(sp2$Localidad_Oscar, sp2$gen_sp)
 temp <- specaccum(comm, "random")
@@
 
+```{r}
+#make final dataset
+sp3 <- sp2[,c("Localidad_Oscar", "gen_sp", "NUM_EJEMPLARES")]
+colnames(sp3)[1] <- "LOCALIDAD"
+now3 <- now[, c("LOCALIDAD", "gen_sp")]
+now3$NUM_EJEMPLARES <- 1
+now3$when <- "now"
+sp3$when <- "asensio"
+
+library(lettercase)
+str_ucfirst(str_lower_case(sp3$gen_sp[1:10]))
+
+sp3$gen_sp <- str_ucfirst(str_lower_case(sp3$gen_sp))
+
+dat <- rbind(sp3, now3)
+#check stuff
+head(dat)
+unique(dat$gen_sp) #429
+#this can be cleaned, but looks decent!
+
+colnames(dat) <- c("locality", "gen_sp", "freq", "when")
+#freq should be melted.
+
+unique(dat$locality) #8
+str(dat)
+write.csv(dat, "data/spain_clean.csv", row.names = F)
+```
+
 1) only for 7 localities, join now and then and paired t-test on rarefied richness per site
 
+1.1) Show raref per site
+
 2) For common species we can do the same but with standardized frequency.
+
+```{r}
+dat <- read.csv("data/spain_clean.csv")
+#head(dat)
+library(reshape2)
+#dat2 <- melt(dat, id.vars = c("locality", "gen_sp", "when"), 
+ #            measure.vars = "freq")
+#head(dat2, 100)
+#sort(unique(dat2$gen_sp))
+#clean with clean_sp
+dat$gen_sp <- gsub(x = dat$gen_sp, pattern = " ", replacement = "", fixed = TRUE)
+#sort(unique(dat$gen_sp))
+dat$gen_sp <- gsub(x = dat$gen_sp, pattern = "_", replacement = " ", fixed = TRUE)
+#remove sp, NA,
+#NA_NA   
+#Hylaeus_Variegatus
+#source("clean_species.R") #sourced from Evenness project.
+library(taxize)
+spp <- clean_species(dat$gen_sp, rows = 1)
+species <- dat$gen_sp
+#table(dat2$gen_sp, dat2$when)
+#Andrena_pilipes = carbonaria
+#remove A. mellifera
+rich_tot <- dcast(dat, when ~ ., value.var = "gen_sp", 
+              fun.aggregate = function(x) length(unique(x)))
+rich_tot #wow.
+#extract genus
+pos <- regexpr(dat$gen_sp, pattern = " ", fixed = TRUE)
+dat$genus <- substr(dat$gen_sp, 
+                    start = 1,
+                    stop = pos-1)
+genus_tot <- dcast(dat, when ~ ., value.var = "genus", 
+              fun.aggregate = function(x) length(unique(x)))
+genus_tot2 <- dcast(dat,  genus ~ when, value.var = "genus", 
+              fun.aggregate = function(x) length(unique(x)))
+genus_tot #wow, reversed??
+genus_tot2
+subset(genus_tot2, now == 0 & asensio == 1)
+subset(genus_tot2, now == 1 & asensio == 1)
+subset(genus_tot2, now == 1 & asensio == 0)
+
+#remove non apis genus! check with Curro.
+
+#losts easely spotted: nomadas & Tyreus, prosopis, teralonias
+
+#calculate richness per site and time period
+#head(dat)
+rich <- dcast(dat, locality ~ when, value.var = "gen_sp", 
+              fun.aggregate = function(x) length(unique(x)))
+#rich
+#t.test(rich$asensio, rich$now, paired = TRUE)
+
+boxplot(rich$asensio[c(1,2,3,5,6,7)], rich$now[c(1,2,3,5,6,7)], names = c("1980's", "2016"), ylab = "species richness", 
+            las = 1, border = "light grey")
+points(x = c(rep(1,6), rep (2,6)), y = c(rich$asensio[c(1,2,3,5,6,7)],
+                                         rich$now[c(1,2,3,5,6,7)]), col = rep(c(1,2,3,5,6,7),2), pch = 19)
+for(i in c(1,2,3,5,6,7)){
+    lines(x = 1:2, y = c(rich$asensio[i], rich$now[i]), col = i)
+}
+
+library(vegan)
+temp <- subset(dat, when == "now")
+comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
+              fun.aggregate = sum, na.rm = TRUE)
+rownames(comm)[-8] <- as.character(comm$locality[-8])
+comm2 <- comm[,-1] 
+#rowSums(comm2) #toro is low, but same sampling effort. 
+temp <- specaccum(comm2, "random")
+#plot(temp)
+#rarecurve(comm2)
+now <- rarefy(x = comm2, sample = 120)
+
+temp <- subset(dat, when == "asensio")
+comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
+              fun.aggregate = sum, na.rm = TRUE)
+#comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
+ #             fun.aggregate = length, na.rm = TRUE)
+rownames(comm)[-8] <- as.character(comm$locality[-8])
+comm2 <- comm[,-1]
+#rowSums(comm2) #sanbernardo useless. Others ok! Delives super sampled!!
+temp <- specaccum(comm2, "random")
+#plot(temp)
+#rarecurve(comm2) #for delives quite complete!
+asensio <- rarefy(x = comm2, sample = 120)
+
+ttest <- t.test(asensio[c(1,2,3,5,6,7)], now[c(1,2,3,5,6,7)], paired = TRUE)
+
+boxplot(asensio[c(1,2,3,5,6,7)], now[c(1,2,3,5,6,7)], names = c("1980's", "2016"), ylab = "rarefied species richness", 
+            las = 1, border = "light grey")
+points(x = c(rep(1,6), rep (2,6)), y = c(asensio[c(1,2,3,5,6,7)], now[c(1,2,3,5,6,7)]), col = rep(c(1,2,3,5,6,7),2), pch = 19)
+for(i in c(1,2,3,5,6,7)){
+    lines(x = 1:2, y = c(asensio[i], now[i]), col = i)
+}
+
+#do by coverage?
+#clean sp.
+```

### 2018-04-04 ToDo.Rmd

--- 2018-03-26 ToDo.Rmd
@@
 - Species declines in % for different taxa... what to do?
+- for NZ species abundance analysis - should data be split by exotic/native first?
 - think on gbif plot...
 - Spain.
+- Refs -> e.g. not showing up in references in the final doc??
 
@@
 
-@Ollerton2012 
-@Winfree2011 - >
+@Ollerton2011 -> Ollerton winfree and tarrant Oikos, I think 2012.
+@Buchman1996 -> http://www.worldcat.org/title/forgotten-pollinators/oclc/34077459
+@NAS -> @National2007 https://www.nap.edu/catalog/11761/status-of-pollinators-in-north-america
+@Winfree2011 -> Winfree bartomeus Cariveau ARES
+@PickeringAscher -> @Ascher2015 Ascher J.S., Pickering J. (2015) Discover Life bee species guide and world checklist (Hymenoptera: Apoidea: Anthophila). www.discoverlife.org
+@Martinetal2012 -> @Martin2012: https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/110154?journalCode=fron
+@Archer -> @Archer2014: https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1600-0706.2013.00949.x
+@França: França, Filipe, et al.
+"Do space‐for‐time assessments underestimate the impacts of logging on tropical biodiversity?
+An Amazonian case study using dung beetles." Journal of Applied Ecology 53.4 (2016): 1098-1105.
+@DePalma2018: De Palma, Adriana, et al.
+"Challenges With Inferring How Land-Use Affects Terrestrial Biodiversity: Study Design, Time, Space and Synthesis." Next Generation Biomonitoring 58 (2018): 163.
+@DePalma2017 De Palma, Adriana, et al.
+"Dimensions of biodiversity loss: Spatial mismatch in land‐use impacts on species, functional and phylogenetic diversity of European bees." Diversity and Distributions 23.12 (2017): 1435-1446.
+@DePalma2016: De Palma, Adriana, et al. "Predicting bee community responses to land-use changes: Effects of geographic and taxonomic biases." Scientific Reports 6 (2016): 31153.
+@vanSwaay: -> @vanSwaay2008 van Swaay, Chris AM, et al. "Butterfly monitoring in Europe: methods, applications and perspectives." Biodiversity and Conservation 17.14 (2008): 3455-3469.
+@Scheper2014. PNAS Museum specimens reveal loss of pollen host plants as key factor driving wild bee decline in The Netherlands
+@pearce -> @Pearce2006 PEARCE, J.
+L.
+and BOYCE, M.
+S.
+(2006), Modelling distribution and abundance with presence‐only data.
+Journal of Applied Ecology, 43: 405-412.
+doi:10.1111/j.1365-2664.2005.01112.x
+@young -> @Young2017 Are pollinating hawk moths declining in the Northeastern United States? An analysis of collection records
+@helminth -> @Dallas2016 helminthR: an R interface to the London Natural History Museum's Host–Parasite Database
+@keil -> @Keil2011 Biodiversity change is scale‐dependent: an example from Dutch and UK hoverflies (Diptera, Syrphidae)
+P Keil, JC Biesmeijer, A Barendregt, M Reemer, WE Kunin
+Ecography 34 (3), 392-401
+@taxonomy1 -> @Meier2004
+https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1523-1739.2004.00233.x
+@taxonomy2 -> @Huber1998 Huber, J.
+T.
+"The importance of voucher specimens, with practical guidelines for preserving specimens of the major invertebrate phyla for identification." Journal of Natural History 32.3 (1998): 367-385.
+@ref geonames, geoparser and opencage: those are likns to software.
+@Hart -> @Hart2016 Ten simple rules for digital data storage. DOI: 10.7287/peerj.preprints.1448v1
+@Collado2018: https://www.biorxiv.org/content/early/2018/03/08/278812
+@Bartomeus_Winfree2013 -> @Bartomeus2013c F1000 https://f1000research.com/articles/2-146/v1
+@Aizen -> @Morales2013 Morales, Carolina L., et al. "Rapid ecological replacement of a native bumble bee by invasive species." Frontiers in Ecology and the Environment 11.10 (2013): 529-534.
+@Pauw: Reconstruction of historical pollination rates reveals linked declines of pollinators and plants
+A Pauw, JA Hawkins - Oikos, 2011
+Cited by 53 Related articles All 11 versions
+@Bartomeus2011 PNAS phenology
+@Tylianakis2010ELE Ecology letters It's 2008 sorry
+@Kleijn -> @Kleijn2008 raemakers: https://www.ncbi.nlm.nih.gov/pubmed/18705369
+@refAlaskaScience -> @Miller-Struttmann2015 http://www.sciencemag.org/content/349/6255/1541
+@Oliveira -> @Oliveira2016 http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0148983
+@R -> @RCoreTeam2017 just cite R
+@rgbif -> @Chamberlain2014 https://cran.r-project.org/web/packages/rgbif/rgbif.pdf
+@cameron2013 -> @Cameron2011 actually 2011 PNAS paper on bumblebees
+@OllertonScience -> @Ollerton2014 Science paper on uk bee extinctions
+
+#to add
+
+@Greve2016 M.
+Greve, A.M.
+Lykke, C.W.
+Fagg, R.E.
+Gereau, G.P.
+Lewis, R.
+Marchant, A.R.
+Marshall, J.
+Ndayishimiye, J.
+Bogaert and J.-C.
+Svenning, Realising the potential of herbarium records for conservation biology, South African Journal of Botany, 105, (317)
+@Lavoie2013 Claude Lavoie, Biological collections in an ever changing world: Herbaria as tools for biogeographical and environmental studies, Perspectives in Plant Ecology, Evolution and Systematics, 15, 1, (68)
+X@Thomson -> @Thomson2018 Thomson SA, Pyle RL, Ahyong ST, Alonso-Zarazaga M, Ammirati J, Araya JF, et al.
+(2018) Taxonomy based on science is necessary for global conservation.
+PLoS Biol 16(3): e2005075.
+https://doi.org/10.1371/journal.pbio.2005075
+X@brief1: -> @Wiggins1991 https://biologicalsurvey.ca/public/Bsc/Controller/Page/briefs/importance.pdf THE IMPORTANCE OF RESEARCH COLLECTIONS OF TERRESTRIAL ARTHROPODS
+X@brief2: -> @Wheeler2003 http://biologicalsurvey.ca/assets/file/43 The role of voucher specimens in validating faunistic and ecological research
+X@Packer2018 Packer, L.
+, Monckton, S.
+K., Onuferko, T.
+M., Ferrari, R.
+R.
+and Didham, R.
+(2018), Validating taxonomic identifications in entomological research.
+Insect Conserv Divers, 11: 1-12.
+doi:10.1111/icad.12284
+
+@Cox1983 Cox, Paul Alan.
+“Extinction of the Hawaiian Avifauna Resulted in a Change of Pollinators for the Ieie, Freycinetia Arborea.” Oikos, vol.
+41, no.
+2, 1983, pp.
+195–199.
+JSTOR, JSTOR, www.jstor.org/stable/3544263.
+@Robbirt2011 Robbirt, Karen M., et al.
+"Validation of biological collections as a source of phenological data for use in climate change studies: a case study with the orchid Ophrys sphegodes." Journal of Ecology 99.1 (2011): 235-241.
+@Maebe2013 Maebe, K., Meeus, I., Maharramov, J.
+et al.
+Apidologie (2013) 44: 188.
+https://doi.org/10.1007/s13592-012-0170-9 Microsatellite analysis in museum samples reveals inbreeding before the regression of Bombus veteranus.
+@Mikheyev2015 Museum samples reveal rapid evolution by wild honey bees exposed to a novel parasite

### 2018-04-04 ms.Rmd

@@
 title: "Historical collections as baseline for assessing the extent of the pollinator crisis"
-author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
+author: "I. Bartomeus1, J. Stavert2, D. Ward2,3, O. Aguado4"
 csl: data/philb_style.xml
@@
 1 Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, E-41092 Sevilla, Spain   
-2   
-3  
+2 Centre for Biodiversity and Biosecurity, School of Biological Sciences, The University of Auckland, Auckland, New Zealand
+3 Landcare Research, Auckland, New Zealand
 4 Andrena Soluciones  
@@
 
-*Author contributions*: IB wrote the initial draft. DW and OA provided data. IB and JS analyzed the data. All authors contributed to writting the manuscript.
+*Author contributions*: IB wrote the initial draft. DW and OA provided data. IB and JS analysed the data. All authors contributed to writing the manuscript.
 
@@
 These biases limit robust inference about any potential “pollinator crisis”.
-Analysis of historical time series data is the most direct approach for assessing species declines, but long term monitoring programs targeting pollinators are scarce.
+Analysis of historical time series data is the most direct approach for assessing species declines, but long term monitoring programs targeting pollinators are limited.
 However, non-structured and often opportunistic historical data collections are more frequently available.
@@
 Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits).
-Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and even pollination services.
-Despite recent advances, we show that information on the status and trends of most pollinators is absent, and that the lack of collection record data digitization is a key bottleneck, impeding progress in this field.
-We identify opportunities and limitations to progress in assessing pollinator declines at a global scale.
-Further, we demonstrate various approaches to analyzing museum collection data by providing two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
-We show that for both regions, winners and loser species emerge despite a lack of significant richness changes over time.
-There is immense potential for museum specimens to play a central role in assessing the global pollination crisis extent, but we must make these data feely accessible to researchers.
+Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and pollination services.
+Despite recent advances, we show that information on the status and trends of most pollinators is absent.
+Furthermore, we identify lack of collection record data digitization as a key bottleneck, impeding progress in this field.
+We highlight opportunities and limitations to progress assessment of pollinator declines globally.
+Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
+We show that for both regions, winner and loser species emerge despite significant changes in species richness over time.
+There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis, but to progress this field data must be made feely accessible to researchers.
 
@@
 
-Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007].
+Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [@Ollerton2011] and food security [@Klein2007].
 As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions.
@@
 These initial claims triggered the first efforts to assess the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines.
-The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@nas].
+The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@NationalResearchCouncil2007].
 Since then, studies on pollinator responses to various global change drivers have multiplied rapidly.
@@
 Yet, the current status and population trends of most pollinator species worldwide remain unknown.
-For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, > 55% of bee species fell into the “Data Deficient” category [@Nieto2014].
+For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, greater than 55% of bee species fell into the “Data Deficient” category [@Nieto2014].
 For countries outside of Europe and the US, data on pollinator populations is almost non-existent.
 
-One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011].
-All this groups of pollinators are often highly mobile, short-lived and small, difficulting the monitoring of its poplations.
+One of the main barriers to identifying pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [@Winfree2011].
+Additionally, many pollinators are highly mobile, short-lived and small, which makes monitoring their populations difficult.
 Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017].
-However, bees are also diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification.
-Furthermore, the uneven distribution of researchers [#@Martinetal2012] has also resulted in geographical biases in bee decline research, as well as taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos, @DePalma2016].
+However, bees are diverse, with more than 20,000 species currently described worldwide [@Archer2014], and often require expert taxonomists for identification.
+Furthermore, the uneven distribution of researchers [@Martin2012] has resulted in geographical biases in bee decline research, as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
 
 One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients.
-Despite critiques on the robustness of this approach [@Franca, @DePalma2018], this is the most extensive source of data on pollinator populations currently available.
-For example, in Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [@DePalma2017].
-A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists.
-This approach was inspired by successful butterfly monitoring programs [@vanSwaay] and is currently being extended to other pollinator taxa.
-However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations.
-Finally, the most practical approach for assessing long-term pollinator population trends is to use the available historical information on species occurrences deposited in museum specimen collection data [e.g.
-@bartomeus2013].
-
-In this review we first assess current evidence for diversity declines and present a road map outlining a strategy for using museum data to fill current knowledge gaps.
-We highlight the major technical difficulties involved in using museum collection data and demonstrate how to approach analysis of different types for museum collection data to identify long-term trends in pollinator populations.
-Finaly, we highlight the need to move beyond simple diversity descriptors and unleash the power of historical data to assess also species interactions, ecosyetm functioning and evolutionary changes along time.
-
-#Current evidence on pollinatior declines
-
-At a global scale, current evidence of pollinator declines highly limited with most data restricted to the US and countries in Europe.
-It is unsurprising that most evidence of pollinator declines is biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance.
-For example, in the UK and the Netherlands, a citizen science based study using both observations and museum collections detected strong declines for bee, hoverfly and flowering plant richness [@Biesmeijer2006].
-However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and data is lacking for natural areas.
+Despite critiques on the robustness of this approach [@Franca2016; @DePalma2018], these studies currently provide the most extensive source of data on pollinator populations.
+For example, researchers have recently estimated bee richness declines for every country in Europe using predictions from models of pollination associations with different land-use types [@DePalma2017].
+A second promising is the use of data collected from pollinator monitoring programs, which are often driven by citizen scientists.
+This approach was inspired by successful butterfly monitoring programs [@vanSwaay2008] and is currently being extended to other pollinator taxa.
+However, these programs require significant time to generate long-term datasets and cannot be used to assess historic pollinator populations.
+Finally, the most practical approach for assessing long-term pollinator population trends is to use the historical information on species occurrences, which is archived in museum collections [e.g., @bartomeus2013a].
+
+In this review, we first assess current evidence for pollinator richness declines and present a road map outlining a strategy for using museum collection data to fill current knowledge gaps.
+We highlight the major technical difficulties involved in using museum collection data and demonstrate several approaches for analysing different types of museum collect data to assess long-term trends in pollinator populations.
+Finally, we highlight the need to move beyond simple biological diversity descriptors and unleash the power of historical data to assess changes in species interactions, ecosystem functioning and evolutionary changes through time.
+
+#Current evidence on pollinator declines
+
+At a global scale, current evidence of pollinator declines is highly limited with most data restricted to the US and Europe.
+It is unsurprising that studies on pollinator declines are biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance.
+For example, in the UK and the Netherlands, a citizen science based study using both observations and museum collection data detected strong richness declines for bees, hoverflies and flowering plants [@Biesmeijer2006].
+However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus data is lacking for natural areas.
 Further exploration of this dataset revealed that for declining pollinator taxa, the trend has attenuated in recent decades [@Carvalheiro2013].
 
-More detailed patterns for common bee species in the Netherlands, aslo using museum historical specimens, have shown simultaneous plant and pollinator declines [#@Scheper2014].
+In the Netherlands, museum data have revealed simultaneous plant and pollinator declines [@Scheper2014].
 Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction.
@@
 For example, regional estimates for species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012].
-A pattern also found in the UK [@OllertonScience].
-The US findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with remaining forest remnants.
-However, it is important to note that this study focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (#@collado2018).
-In any case, there is strong concordance between local extinctions and regional declines [#@Bartomeus_Winfree2013], suggesting that local extinctions are indicators of regional population declines.
-
-As with bees, studies on changes in species richness for other pollinator taxa is scarce and geographically restricted.
-Generally speaking, reported declines for bumblebees have been more severe.
+This is a pattern also detected in the UK [@Ollerton2014].
+However, the US findings above are in stark contrast with the widespread local extinctions reported by Burkle [-@Burkle2013], who compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants.
+It is important to note that Burkle [-@Burkle2013] focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (@Collado2018).
+In any case, there is strong concordance between local extinctions and regional declines [@Bartomeus2013c], suggesting that local extinctions are indicators of regional population declines.
+
+As with bees, studies on species richness changes for other pollinator taxa are both scarce and geographically restricted.
+Generally speaking, reported declines for bumblebees are the most severe of all pollinator taxa.
 For example, declines of up to 18% in local bumble richness have been reported for Belgium and the Netherlands [@Carvalheiro2013].
-In other parts of Europe local richness declines range from 5% for Great Britain [@Carvalheiro2013] to 42% for Denmark [@Dupont].
-In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti] and 30% [@Bartomeus2013a].
-For butterflies, the only evidence of richness declines come from Europe.
-For example, butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although in Great Britain declines have been less severe [@Carvalheiro].
-Another study in Belgium showed that richness declines have been more severe (approximately 30%), although this study exampled changes in richness over a longer time period (early 1900’s to 2000) [@Maes].
-In parts of Germany, declines in local butterfly richness have been even more severe, with declines of up to 70% reported [@Wenzel].
-There are very few studies on changes in hoverfly species richness and are once again restricted to Europe.
-In Belgium, Great Britain and the Netherlands, changes in hoverfly richness have been modest [@Carvalheiro].
-In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer].
-Furthermore, directionality (richness increases or decreases) varies depending on the time period that richness changes have been assessed.
-For example, hoverfly richness decreased in Belgium decreased by approximately 6% from 1950-69 compared with 1970-80, but increased by approximately 10% between 1970-89 and 1990-2009 [@Carvalheiro2013].
-
-Outside of Europe and the US there are no studies on pollinator declines using historical records.
-In Fig 1 we show a comparision of bee species richness per area (a), bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes (d) to highlight the little coverage that we have globally.
+In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011].
+In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a].
+
+For butterflies, the only evidence of richness declines comes from Europe.
+For example, butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although declines in Great Britain have been less severe [@Carvalheiro2013].
+In Belgium, another study [@Maes2001] found that richness declines have been severe (approximately 30%), although this study assessed richness changes over a longer time period (early 1900’s to 2000) compared to Carvalheiro [-@Carvalheiro2013].
+In parts of Germany, declines in local butterfly richness of up to 70% have been reported [@Wenzel2006].
+
+There are very few studies on hoverfly species richness changes, those of which are restricted to Europe.
+In Belgium, Great Britain and the Netherlands, hoverfly richness changes have been modest [@Carvalheiro2013].
+In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer2006].
+Furthermore, directionality (richness increases or decreases) varies depending on the time period assessed.
+For example, hoverfly richness decreased in Belgium by approximately 6% from 1950-69 to 1970-80, but increased by approximately 10% between 1970-89 and 1990-2009 [@Carvalheiro2013].
+
+Outside of Europe and the US, there are no studies on pollinator declines using historical records.
+In Figure 1, we show a comparison of bee species richness per area (a), bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes (d) to highlight the little coverage that we have globally.
 See raw data at Sup Mat 1.
-For other parts of the world there are only species-specific examples of historical losses (e.g., Bombus dalbhomi; #@Aizen) [South Africa: @Pauw, maybe with plants… .
-However,Peter Keil flies?
+For other parts of the world, there are only species-specific examples of historical losses (e.g., Bombus dalbhomi; @Morales2013) [South Africa: @Pauw, maybe with plants… .
+However, Peter Keil flies?
 Moths?
@@
 
-Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org).
-This is calculated as the residuals of the log-log regresion between bee richness and country size to account for the species-area relationships.
-We show mean local bee (b), syrphid (c) and butterflies (d) richness percent change in the last century.
-Warmer colors indicate steeper declines observed.
-Countries without available data are coloured in white.
-Note that some African countries may have incomplete faunas listed and that Alaska is included with US values.
-
-#How to use historical records vouchered in museum specimens to fill the gaps.
-
-As evidenced in Fig 1, we lack specific estimates for most of the world countries and histrical data vouchered in museums may be the only change to fill those gaps.
+Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness and country size, which accounts for the species-area relationship; (b) syrphid (c) and butterflies (d) richness percent change in the last century.
+Warmer colours indicate steeper declines.
+Countries without data are coloured in white.
+Note that some African countries may have incomplete listed faunas and that Alaska is included with US values.
+
+#Using historical museum specimen records to fill knowledge gaps
+
+As shown in Figure 1, specific estimates are lacking for countries worldwide and use of historical museum collection data may be the most effective tool for filling those gaps.
 
 The core aim of museums is to conserve and curate collections.
-Thus, they serve as a precious repository for specimens, and at the same time often ensure higher quality taxonomic identification.
+Thus, they serve as a precious repository for specimens, and at the same time, often ensure higher quality taxonomic identification.
 Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization.
-Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@taxonomy, @taxonomy2, @Thomson], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@hart].
-Undertaking this process for tens or hundreds of thousands of museum collections specimens can be a daunting task and requires specialized personnel.
+Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@Meier2004; @Huber1998; @Thomson2018], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@Hart2016].
+Undertaking this process for tens or hundreds of thousands of museum collection specimens can be a daunting task and requires specialized personnel.
 While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process.
-High resolution photos of specimens and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers (e.g.
-https://www.zooniverse.org/).
+High resolution photos of specimens and associated labels can be uploaded to the internet, where the task of image transcription can be distributed across hundreds or thousands of volunteers (e.g., https://www.zooniverse.org/).
 In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage].
-However, achieving all this goals needs propoer funding [@brief1].
-
-Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014, @Young).
-This is despite the fact that museum collections often have a number of biases.
-This biases includes the unknown sampling effort that was employed or the personal interests and curatorial techniques of collectors.
+However, achieving this requires adequate funding [@Wiggins1991].
+
+Where digitization has been completed, the data provide a rich source of information, allowing assessment of the current status and long-term trends in pollinator populations [e.g., @Bartomeus2013a; @Scheper2014; @Young2017].
+This is despite the fact that museum collections often have a number of biases, including unknown sampling effort, personal interests of collectors and the curatorial techniques used.
 For example, collectors tend to target rare or unusual over common taxa, discard damaged individuals or only accession a certain number of individuals.
-In addition, the opportunistic nature of museum collections often imply an spatial bias where areas have been undersampled, or where samples are biased towards easily collected localities (e.g.
-near towns/cities and/or along roadsides) and that information is often restricted only to the presence of a species (i.e.
-there is no information on where a species is absent).
-Given adequate sample sizes, most of this biases can be accounted for with propoer statistical data anslysis [e.g.
-@bartomeus2013, #@Keil, @pearce].
-Nevertheless, in many cases, museum collections are the only data we have to reconstruct the past.
+In addition, collections are often made opportunistically, leading to a spatial bias where difficult to access areas are under-sampled or conversely, where samples are biased towards easily accessed locations (e.g., towns/cities and/or along roadsides).
+Furthermore, this approach can only determine where species are present and not where they are absent.
+However, given adequate sample sizes and appropriate statistical analyses, most of these biases can be accounted for [e.g., @Bartomeus2013a; @Keil2011; @Pearce2006].
 
@@
 
-As we have seen, there is a paucity of countries for which museum data is available (Fig 1), and hence can be used as baseline for assessing pollinator population declines.
-While ideally, one would aim to digitize all museum records in the world, this is unlikely to happen in the near future due mainly to funding constraints.
-Here we illustrate how this can be prioritized and optimized to encourage independent researchers to assess declines in its own country.
-
-GBIF (https://www.gbif.org/) is a central repository centralizing occurrence data from around the globe.
-Much of this data comes from museums, collections and goverment research institutes, but it integrates several other sources.
-Alongside the popular statistical language R [#@R], GBIF can be directly queried into your computer [#@rgbif] and data availability checked for your region of interest.
-In Figure 2 we show the number of modern and historical bee records already available per country.
-While for some countries like UK and US the numbers are high in both axes, and hence, data can be potentially analyzed without further data collection efforts, most countries fall short in one or both of the axes.
-Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis.
-On the contrary, for countries with large number of historical records, re-surveys should be organized [e.g.
-#@cameron2013].
-Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X… Also note that historical records are not always in local museums, but many EU and US museums contain large collection of exotic pollinators.
-Similar exploratry analysis can be easely done for other taxa.
+As we have shown, there is a paucity of countries for which museum data is available (Figure 1), and hence can be used as baseline for assessing pollinator population declines.
+While ideally one would aim to digitize all museum collection records, this is unlikely in the near future, mainly due to funding constraints.
+Here we show how researchers can optimize their use of museum collection data to assess long-term pollinator population changes.
+
+GBIF (https://www.gbif.org/) is a central repository for global species occurrence data.
+Much of these data come from museums, private collections and government research institutes, but it integrates several other sources [such as??].
+In combination with the popular statistical language R [@RCoreTeam2017], GBIF can be directly queried into your computer [@Chamberlain2014] and data availability checked for your region of interest.
+In Figure 2 we show the number of modern and historic bee records currently available for different countries.
+While values are high on both axes for countries such as the UK and the US so that data can potentially be analysed without further data collection effort, most countries fall short in one or both axes.
+Hence, for countries with high numbers of recent records, researchers should prioritize the digitalization of old material before embarking on data analyses.
+In contrast, for countries with large numbers of historic records, re-surveys are required before analysis [e.g., @Cameron2011].
+Interestingly, we show that greater than 100 countries have less than 1,000 records, making them poor candidates for analysing long-term pollinator population trends.
+However, countries like X or X… It is also important to note that historical records are not always vouchered in local museums, but many European and US museums contain large collections of exotic pollinators [do you mean exotic species collected in the US/EU or species collected in their native countries and bought back to and vouchered in US/EU museums?].
+Aside from bees, similar exploratory analysis can easily be done for other taxa.
 
@@
 Figure 2.
-The number of bee occurrences pre 1980 and post 1980 in GBIF.
-Upper right quadrat are well covered countries, upper left quadrat are countries that need resurvey of old sites.
-Lower right are sites which needs to dig museum records and lower left are countries with very little data available.
+The number of bee occurrences before 1980 and after 1980 in GBIF.
+The upper right quadrat contains well covered countries and the upper left quadrat contains countries that resurveys of old sites.
+The lower right quadrat contains countries that require museum record digitization and the lower left quadrat contains countries with very little available data.
 We highlight a few representative examples.
 
-Once museum collection datasets are made available, researchers need to gauge any potential biases in the data.
-We explore this process with two contrasting and ilustrative examples from Spain and New Zealand.
+Once museum collection datasets are made available, researchers must identify any potential biases in the data.
+We explore this process with two contrasting dataset examples (Spain and New Zealand).
 In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people.
 In this case, we contacted the original collectors to define their sampling protocols.
-We then resurveyed the same sites using the same protocols 20 years later.
-Conversely, the New Zealand dataset includes a wide suite of collectors and collection locations, but luckely, tehere is no obvious bias in the geographical and taxonomic coverage through time.
-For these case studies, we provide annotated R scripts to analyse the different kinds of data (e.g.
-for Spain and New Zealand, as described above).
-Essentially, these different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories.
-We hope this resource will encourage researchers to analyze data for regions where there is little current information on pollinator declines.
+We then resurveyed the same sites (20 years after the original surveys) using the same protocols.
+In contrast, the New Zealand dataset includes a wide suite of collectors and collection locations but shows no obvious biases in geographical and taxonomic coverage through time.
+For these case studies, we provide annotated R scripts for analysis of different dataset types (e.g., for Spain and New Zealand, as described above).
+These different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories.
+We hope this resource will encourage researchers to analyse data for regions where there is current information on pollinator declines is lacking.
 
@@
 
-Spain provides an interesting study system because its natural habitats have been transformed extensively by humans, but land-use is not as intensive as in many other European countries.
-In addition, Spain is a bee diversity hotspot (Fig 1) and maintains a relatively heterogeneous landscape.
-Spain has already digitalized a large amount of occurrence data for both historical and recent periods (Fig 2).
-However, visual inspection of the data reveals that the data is clustered around a few localities.
-Further, historic records do not spatially match recent records, making them difficult to compare.
+Spain provides an interesting study system because its natural habitats have been transformed extensively by humans over a long time period, but land-use is not as intensive compared with many other European countries.
+In addition, Spain is a bee diversity hotspot (Figure 1) and maintains a relatively heterogeneous landscape.
+Spain has already digitalized a large amount of pollinator occurrence data for both historical and recent periods (Figure 2).
+However, visual inspection of the data reveals clustering around a few localities.
+Further, historic records do not spatially match recent records, making comparisons difficult.
 For this dataset, most of the historical records are located around Valladolid and were collected by Enrique Asensio and his collaborators.
 There has been no recent sampling of bees in this area.
-However, we found that Enrique systematically sampled seven locations and that there was additional data available at the “Museo de Historia Nacional” and other minor collections.
-Digitization of these records, along with a re-survey of these locations provided an excellent dataset for a before and after comparison of bee communities at these locations.
+However, we found that Enrique systematically sampled seven locations and that additional data were available at the “Museo de Historia Nacional” and other minor collections.
+Digitization of these records, along with a re-survey of these locations provided an excellent dataset for a before and after comparison of bee communities at these specific locations.
 [from Darren: Has there been any changes in the land use between the two periods?]
 
-In brief, we check sampling completeness at both time periods, and compare the species richness level of each site before 1980 and after 1980 with a paired t-test.
-We show that overall there is no no difference in species richness (p > 0.05).
+In brief, we checked sampling completeness for both time periods, and compared species richness for each site before and after 1980 with a paired t-test.
+We found that overall there was no change in species richness (P > 0.05).
 However, ...
@@
 [from Darren: Any information on these groups.
-Are the specialists species, or more sensitive to land change?].
+Are they specialist species, or more sensitive to land change?].
 
@@
 
-However, at the species level, we found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Fig 3).
+However, at the species level, we found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Figure 3).
 Interestingly, the two exotic bee species that declined in relative abundance were both in the genus Bombus, intentionally introduced into New Zealand for the pollination of crops.
@@
 
-A recent survey highlighted that more than 90% of the papers studying pollinator responses to land use change focused solely on richness and abundance descriptors [@winfree2011].
-But in addition to local (alpha) diversity and regional (gamma) diversity, we need to understand changes in turnover between sites (beta diversity).
-Environmental changes often results in a few species as “winners” and many species becoming “losers” [@Bartomeus2013b].
-Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018].
+A recent study found that more than 90% of the papers studying pollinator responses to land-use change focused solely on richness and abundance descriptors [@winfree2011].
+But in addition to local (alpha) diversity and regional (gamma) diversity, we need to assess changes in turnover between sites (beta diversity).
+Environmental changes often result in a few “winner” species and many species becoming “losers” [@Bartomeus2013b].
+Identifying winners and losers is critical as the few winners are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2017].
 These changes can have important effects for pollination of native plant species and crops.
 
-But museum specimen collection can provide much more information besides species occurrence records, and such information is important to capture when digitizing collections.
-This is specifically relevant to identify mechanisms of decline and adaptation.
-For example, one particularly important source of information is the date the specimen was collected, which allows the tracking of phenological advances congruent with contemporary climate change [#@Bartomeus2011].
-In addition, museum labels on pollinator specimens often include information about the host plant on which the specimen was collected.
-This information is precious to understand past and present species interactions [#@Tylianakis2010ELE].
-Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [#@kleinj_raemakers].
+But museum specimen collections can provide much more information besides species occurrence records, providing such information is captured when digitizing collections.
+This is particularly important for identifying mechanisms of decline and adaptation.
+For example, recording the date of collection is particularly important for tracking of phenological advances congruent with contemporary climate change [@Bartomeus2011].
+In addition, pollinator specimen labels often include information about the host plant on which the specimen was collected.
+This information critical for understanding past and present species interactions [@Tylianakis2008].
+Regardless of this information, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [@Kleijn2008].
 Finally, museum specimens can be measured to track evolutionary changes by measuring specific specimen traits.
-This has been already explored to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.
+This has been already explored to show tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016] shrinkage with climate and land use change.
 
@@
 However, specific datasets can also be stored in different webpages or database repositories (e.g.
-Universities and museum webpages, Dryad) as far as they are well documented and can be easely retrived and combined with other datasets using open science tools [e.g.
-ropensci, # @heltmint].
+Universities and museum webpages, Dryad) as far as they are well documented and can be easely retrived and combined with other datasets using open science tools [e.g., @Dallas2016].
 
 We also need to revolutionize the way that researchers collaborate with museums, which should start with fostering healthy bidirectional relationships.
-For instance, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered (@Packer, @brief2), rendering them less useful for future research.
-To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is not adequate funding, and recognition that accurate data recording and long-term preservation are critical for research [@WardTEE].
+For instance, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered (@Packer2018; @Wheeler2003), rendering them less useful for future research.
+To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is not adequate funding, and recognition that accurate data recording and long-term preservation are critical for research [@Ward2015].
 
@@
 We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees.
-For the US and Europe there have been few regional bee extinctions [#@OllertonScience; @Bartomeus2013a] (#Ref bartomeus should be PNAS historical declines), but in disturbed ecosystems declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006].
+For the US and Europe there have been few regional bee extinctions [@Ollerton2014; @Bartomeus2013a], but in disturbed ecosystems declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006].
 For most other pollinator taxa and regions throughout the world we know almost nothing.

### 2018-04-04 ref_check.Rmd

--- 2018-04-01 ms.Rmd
@@
 1 Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, E-41092 Sevilla, Spain   
-2 
-3
+2 Centre for Biodiversity and Biosecurity, School of Biological Sciences, The University of Auckland, Auckland, New Zealand
+3 Landcare Research, Auckland, New Zealand
 4 Andrena Soluciones  
@@
 
-Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007].
+Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [@Ollerton2011] and food security [@Klein2007].
 As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions.
-The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@Buchman1996].
+The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@Buchmann1996].
 However, early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time.
 These initial claims triggered the first efforts to assess the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines.
-The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@nas].
+The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@NationalResearchCouncil2007].
 Since then, studies on pollinator responses to various global change drivers have multiplied rapidly.
@@
 
-One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011].
+One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [@Winfree2011].
 All this groups of pollinators are often highly mobile, short-lived and small, difficulting the monitoring of its poplations.
 Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017].
-However, bees are also diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification.
-Furthermore, the uneven distribution of researchers [#@Martinetal2012] has also resulted in geographical biases in bee decline research, as well as taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos, @DePalma2016].
+However, bees are also diverse, with more than 20,000 species currently described worldwide [@Archer2014], and often require expert taxonomists for identification.
+Furthermore, the uneven distribution of researchers [@Martin2012] has also resulted in geographical biases in bee decline research, as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
 
 One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients.
-Despite critiques on the robustness of this approach [@Franca, @DePalma2018], this is the most extensive source of data on pollinator populations currently available.
+Despite critiques on the robustness of this approach [@Franca2016; @DePalma2018], this is the most extensive source of data on pollinator populations currently available.
 For example, in Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [@DePalma2017].
 A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists.
-This approach was inspired by successful butterfly monitoring programs [@vanSwaay] and is currently being extended to other pollinator taxa.
+This approach was inspired by successful butterfly monitoring programs [@vanSwaay2008] and is currently being extended to other pollinator taxa.
 However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations.
-Finally, the most practical approach for assessing long-term pollinator population trends is to use the available historical information on species occurrences deposited in museum specimen collection data [e.g.
-@bartomeus2013].
+Finally, the most practical approach for assessing long-term pollinator population trends is to use the available historical information on species occurrences deposited in museum specimen collection data [e.g., @Bartomeus2013a].
 
@@
 
-More detailed patterns for common bee species in the Netherlands, aslo using museum historical specimens, have shown simultaneous plant and pollinator declines [#@Scheper2014].
+More detailed patterns for common bee species in the Netherlands, aslo using museum historical specimens, have shown simultaneous plant and pollinator declines [@Scheper2014].
 Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction.
@@
 For example, regional estimates for species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012].
-A pattern also found in the UK [@OllertonScience].
+A pattern also found in the UK [@Ollerton2014].
 The US findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with remaining forest remnants.
-However, it is important to note that this study focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (#@collado2018).
-In any case, there is strong concordance between local extinctions and regional declines [#@Bartomeus_Winfree2013], suggesting that local extinctions are indicators of regional population declines.
+However, it is important to note that this study focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (@Collado2018).
+In any case, there is strong concordance between local extinctions and regional declines [@Bartomeus2013c], suggesting that local extinctions are indicators of regional population declines.
 
@@
 For example, declines of up to 18% in local bumble richness have been reported for Belgium and the Netherlands [@Carvalheiro2013].
-In other parts of Europe local richness declines range from 5% for Great Britain [@Carvalheiro2013] to 42% for Denmark [@Dupont].
-In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti] and 30% [@Bartomeus2013a].
+In other parts of Europe local richness declines range from 5% for Great Britain [@Carvalheiro2013] to 42% for Denmark [@Dupont2011].
+In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a].
 For butterflies, the only evidence of richness declines come from Europe.
-For example, butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although in Great Britain declines have been less severe [@Carvalheiro].
-Another study in Belgium showed that richness declines have been more severe (approximately 30%), although this study exampled changes in richness over a longer time period (early 1900’s to 2000) [@Maes].
-In parts of Germany, declines in local butterfly richness have been even more severe, with declines of up to 70% reported [@Wenzel].
+For example, butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although in Great Britain declines have been less severe [@Carvalheiro2013].
+Another study in Belgium showed that richness declines have been more severe (approximately 30%), although this study exampled changes in richness over a longer time period (early 1900’s to 2000) [@Maes2001].
+In parts of Germany, declines in local butterfly richness have been even more severe, with declines of up to 70% reported [@Wenzel2006].
 There are very few studies on changes in hoverfly species richness and are once again restricted to Europe.
-In Belgium, Great Britain and the Netherlands, changes in hoverfly richness have been modest [@Carvalheiro].
-In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer].
+In Belgium, Great Britain and the Netherlands, changes in hoverfly richness have been modest [@Carvalheiro2013].
+In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer2006].
 Furthermore, directionality (richness increases or decreases) varies depending on the time period that richness changes have been assessed.
@@
 See raw data at Sup Mat 1.
-For other parts of the world there are only species-specific examples of historical losses (e.g., Bombus dalbhomi; #@Aizen) [South Africa: @Pauw, maybe with plants… .
+For other parts of the world there are only species-specific examples of historical losses (e.g., Bombus dalbhomi; @Morales2013) [South Africa: Pauw, maybe with plants… .
 However,Peter Keil flies?
@@
 Butterflies!]()
-
-```{r map, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
-library(rworldmap)
-map <- read.csv("data/bee_richness.csv")
-size <- read.csv("data/country_area.csv")
-colnames(size) <- c("country", "size")
-map0 <- merge(map, size, all.x = TRUE)
-scatter.smooth(log(map0$species) ~ log(map0$size+1))
-m <- lm(log(map0$species) ~ log(map0$size+1), na.action = na.exclude) 
-map0$res <- residuals(m)
-map2 <- joinCountryData2Map(map0, joinCode="NAME", nameJoinColumn = "country",
-                            nameCountryColumn = "distribution")
-#need to use ISO's to make sure we don't carry over NA's due to country misspellings
-#have plotted countries with missing values as grey (shows we have richness for pretty much everywhere)
-map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "light grey",
-               mapTitle = "Bee species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE)
-#add a modified legend using the same initial parameters as mapCountryData
-do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5))
-
-#Nacho needs to correct for country size
-#Need to 1) match all country names (includinh US and UK!)
-#2) check subsaharian countries maybe omit?
-
-#need to add all the pannels.
-```
 
@@
 Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization.
-Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@taxonomy, @taxonomy2, @Thomson], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@hart].
+Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@Meier2004; @Huber1998; @Thomson2018], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@Hart2016].
 Undertaking this process for tens or hundreds of thousands of museum collections specimens can be a daunting task and requires specialized personnel.
@@
 In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage].
-However, achieving all this goals needs propoer funding [@brief1].
+However, achieving all this goals needs propoer funding [@Wiggins1991].
 
-Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014, @Young).
+Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations [e.g., @Bartomeus2013a; @Scheper2014; @Young2017].
 This is despite the fact that museum collections often have a number of biases.
@@
 there is no information on where a species is absent).
-Given adequate sample sizes, most of this biases can be accounted for with propoer statistical data anslysis [e.g.
-@bartomeus2013, #@Keil, @pearce].
+Given adequate sample sizes, most of this biases can be accounted for with propoer statistical data anslysis [e.g., @Bartomeus2013a; @Keil2011; @Pearce2006].
 Nevertheless, in many cases, museum collections are the only data we have to reconstruct the past.
@@
 Much of this data comes from museums, collections and goverment research institutes, but it integrates several other sources.
-Alongside the popular statistical language R [#@R], GBIF can be directly queried into your computer [#@rgbif] and data availability checked for your region of interest.
+Alongside the popular statistical language R [@RCoreTeam2017], GBIF can be directly queried into your computer [@Chamberlain2014] and data availability checked for your region of interest.
 In Figure 2 we show the number of modern and historical bee records already available per country.
@@
 Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis.
-On the contrary, for countries with large number of historical records, re-surveys should be organized [e.g.
-#@cameron2013].
+On the contrary, for countries with large number of historical records, re-surveys should be organized [e.g., @Cameron2011].
 Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X… Also note that historical records are not always in local museums, but many EU and US museums contain large collection of exotic pollinators.
 Similar exploratry analysis can be easely done for other taxa.
-
-```{r low hanging fruits, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
-library(rgbif)
-apidae_key <- name_backbone(name="Apidae", rank = "family")$usageKey
-andrenidae_key <- name_backbone(name="Andrenidae", rank = "family")$usageKey
-halictidae_key <- name_backbone(name="Halictidae", rank = "family")$usageKey
-colletidae_key <- name_backbone(name="Colletidae", rank = "family")$usageKey
-megachilidae_key <- name_backbone(name="Megachilidae", rank = "family")$usageKey
-stenotritidae_key <- name_backbone(name="Stenotritidae", rank = "family")$usageKey
-melittidae_key <- name_backbone(name="Melittidae", rank = "family")$usageKey
-
-keys <- c(apidae_key, andrenidae_key, halictidae_key, halictidae_key, colletidae_key,
-          megachilidae_key, stenotritidae_key, melittidae_key)
-#loop through all families, through all countries and store in datarame
-country_codes <- isocodes$code 
-lhf <- matrix(ncol = 1+(7*2), nrow = length(country_codes))
-lhf <- as.data.frame(lhf)
-colnames(lhf) <- c("country", "apidaepre1980", "andrenidaepre1980", "halictidaepre1980", "colletidaepre1980", 
-                    "megachilidaepre1980", "stenotritidaepre1980", "melittidaepre1980",
-                   "apidaepost1980", "andrenidaepost1980", "halictidaepost1980", "colletidaepost1980", 
-                    "megachilidaepost1980", "stenotritidaepost1980", "melittidaepost1980")
-for(i in 1:length(country_codes)){
-    for(j in 1:length(keys)){
-        lhf[i,1] <- country_codes[i]
-        lhf[i,1+j] <- as.numeric(occ_count(taxonKey= keys[j],
-                            from = 1980, to = 2018,
-                            georeferenced=TRUE,
-                            country = country_codes[i]))
-        lhf[i,1+j+7] <- as.numeric(occ_count(taxonKey= keys[j],
-                            from = 1850, to = 1980,
-                            georeferenced=TRUE,
-                            country = country_codes[i]))
-    }
-}
-#very slow, we should save the data and load here when needed.
-lhf$post1980 <- rowSums(lhf[,2:8])
-lhf$pre1980 <- rowSums(lhf[,9:15])
-write.csv(lhf, "data/rgbif_output.csv")
-```
-
-```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
-lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
-#lhf
-#head(lhf)
-#str(lhf)
-lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
-#plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
-#abline(h = 150000, v = 150000)
-#plot needs legend by country, and to be ggplotted. Add some labels will be good, at least for ES and NZ, which we explore below.
-
-```
-
-```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
-lhf$percent_post1980 <- lhf$post1980 / (lhf$pre1980+lhf$post1980)
-lhf$percent_pre1980 <- lhf$pre1980 / (lhf$pre1980+lhf$post1980)
-
-#plot(sort(lhf$percent_post1980), xlab = "post1980", ylab = "%", col = as.factor(lhf$country))
-#Add country labels in x axes.
-
-plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
-abline(h = 8, v = 8)
-
-```
 
@@
 Are the specialists species, or more sensitive to land change?].
-
-```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
-dat <- read.csv("data/spain_clean.csv")
-#head(dat)
-library(reshape2)
-dat2 <- melt(dat, id.vars = c("locality", "gen_sp", "when"), 
-             measure.vars = "freq")
-#head(dat2, 100)
-#sort(unique(dat2$gen_sp))
-#clean with clean_sp
-#remove sp, NA, " ",
-#NA_NA   
-#Hylaeus_Variegatus
-
-#table(dat2$gen_sp, dat2$when)
-#Andrena_pilipes = carbonaria
-#remove A. mellifera
-
-#losts easely spotted: nomadas & Tyreus, prosopis, teralonias
-
-#calculate richness per site and time period
-#head(dat)
-rich <- dcast(dat, locality ~ when, value.var = "gen_sp", 
-              fun.aggregate = function(x) length(unique(x)))
-#rich
-#t.test(rich$asensio, rich$now, paired = TRUE)
-
-library(vegan)
-temp <- subset(dat, when == "now")
-comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
-              fun.aggregate = sum, na.rm = TRUE)
-rownames(comm)[-8] <- as.character(comm$locality[-8])
-comm2 <- comm[,-1]
-rowSums(comm2)
-temp <- specaccum(comm2, "random")
-#plot(temp)
-#rarecurve(comm2)
-now <- rarefy(x = comm2, sample = 120)
-
-temp <- subset(dat, when == "asensio")
-comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
-              fun.aggregate = sum, na.rm = TRUE)
-rownames(comm)[-8] <- as.character(comm$locality[-8])
-comm2 <- comm[,-1]
-rowSums(comm2)
-temp <- specaccum(comm2, "random")
-#plot(temp)
-#rarecurve(comm2)
-asensio <- rarefy(x = comm2, sample = 120)
-
-ttest <- t.test(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)], paired = TRUE)
-
-boxplot(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)], names = c("1980's", "2016"), ylab = "rarefies species richness", 
-            las = 1, border = "light grey")
-points(x = c(rep(1,5), rep (2,5)), y = c(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)]), col = rep(1:5,2), pch = 19)
-for(i in c(1,2,5,6,7)){
-    lines(x = 1:2, y = c(asensio[i], now[i]), col = i)
-}
-
-```
 
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
-library(plyr)
-library(dplyr)
-library(reshape2)
-library(ggplot2)
-library(maps)
-library(vegan)
-
-########################################
-#read in NZ bee data
-########################################
-
-d <- read.csv("data/NZBees_allrecords.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
-
-#select required variables
-d <- select(d, one_of(c("species", "date", "locality", "collector", "collection", "method", "exotic_native")))
-d <- na.omit(d) #remove NAs
-
-#remove honeybees
-d$species <- as.factor(d$species)
-d <- d[ ! d$species %in% c("Apis mellifera"), ]
-d$species <- droplevels(d$species)
-
-#need to correct space between genus and species
-#levels(d$species)
-#d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
-d$Gen_sp <- d$species
-#sort(table(d$Gen_sp))
-#sum(table(d$Gen_sp))
-#head(d)
-
-#standardise dates
-date2 <- strptime(d$date, "%d/%m/%Y")
-d$jday <- date2$yday
-d$year <- 1900 + date2$year
-d$month <- date2$mon + 1
-#head(d$month)
-
-#select fields to be independent
-#completeFun <- function(data, desiredCols) {
-#  completeVec <- complete.cases(data[, desiredCols])
-#  return(data[completeVec, ])
-#}
-
-#d <- completeFun(d, c("day", "month", "year"))
-uni <- as.data.frame(cbind(d$Gen_sp, d$locality, d$collector, d$jday, d$year))
-dup <- duplicated(uni)
-
-# remove duplicates
-d2 <- d[-which(dup == TRUE),]
-
-#retain only sweep netting and no data
-#keep <- c("sweep", "no_data") #method is missing for a number of records
-#d2 <- d2[d2$method %in% keep, ]
-
-native <- filter(d2, exotic_native == "native") %>% droplevels()
-exotic <- filter(d2, exotic_native == "exotic") %>% droplevels()
-
-#############################################
-#Resampling Analysis for natives species 
-#############################################
-
-par(mfrow = c(2,2), mar = c(4,4,2,1))
-
-#fin year = 0 quick and dirty
-#length(which(native$year < 1900))
-native <- subset(native, year > 1900)
-for (i in 3:10){
-    break1 <- quantile(native$year, probs = seq(0,1,1/i), na.rm = TRUE)
-    lab <- c(1:i)
-    for(j in 1:i){
-        lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
-    }
-    native$time_period_quant  <- cut(native$year, breaks = break1, labels = lab)
-    comm2 <- table(native$time_period_quant, factor(native$Gen_sp))
-    (x2 <- rarefy(comm2,150, se = TRUE)) #same result as my resampling...
-    #rar[[i-2+10]] <- x2
-    #str(x2)
-    time <- c(1:i)
-    for(j in 1:i){
-        time[j] <- (as.numeric(break1)[j] + as.numeric(break1)[j+1])/2
-    }
-    xx2 <- as.data.frame(x2)[1,]
-    se <- as.data.frame(x2)[2,]
-    #if(i == 3){
-        #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
-        #abline(v = 150)}
-    m <- lm(t(xx2)~ time)
-    summary(m)
-    corr <- cor(t(xx2), time, use = "complete.obs")
-    cor_dis <- c(1:1000)
-    for (k in 1:1000){
-        cor_dis[k] <- cor(sample(t(xx2),i), time, use = "complete.obs")
-    }
-    #hist(cor_dis)
-    #lines(c(corr, corr), c(0,200), col = "red")
-    (p <- pnorm(corr, mean = mean(cor_dis), sd = sd(cor_dis)))
-    #with lm
-    #if(summary(m)$coef[8]> 0.05 | is.na(summary(m)$coef[8]) == TRUE){
-    #abline(m, lty = 2)}
-    #else{
-    #abline(m, lty = 1)}    
-    #with corrs
-    #if(p > 0.05){
-        #abline(m, lty = 2)}
-    #else{
-        #abline(m, lty = 1)}    
-    #if(i == 10){
-        #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
-        #abline(v = 150)}    
-    #rarecurve(comm2)
-    #abline(v = 1000)
-
-}
-    plot(time,t(xx2), xlab = "", las = 2, ylab = "Number of native bee species", main = "", xaxt = "n", ylim = c(min(xx2, na.rm = TRUE)- max(se), max(xx2, na.rm = TRUE)+ max(se)))
-    axis(1, at = time ,labels = lab, las = 2, cex.axis = 0.7)
-    arrows(time,as.numeric(xx2)+as.numeric(se), time, as.numeric(xx2)-as.numeric(se), angle=90, length = 0)
-    abline(m, lty = 2)
-
-#############################################
-#Resampling Analysis for exotic species 
-#############################################
-
-#length(which(exotic$year < 1900))
-exotic <- subset(exotic, year > 1900)
-for (i in 3:6){ #error saying break are not unique - how to fix?
-    break1 <- quantile(exotic$year, probs = seq(0,1,1/i), na.rm = TRUE)
-    lab <- c(1:i)
-    for(j in 1:i){
-        lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
-    }
-    exotic$time_period_quant  <- cut(exotic$year, breaks = break1, labels = lab)
-    comm2 <- table(exotic$time_period_quant, factor(exotic$Gen_sp))
-    (x2 <- rarefy(comm2,25, se = TRUE)) #same result as my resampling...
-    #rar[[i-2+10]] <- x2
-    #str(x2)
-    time <- c(1:i)
-    for(j in 1:i){
-        time[j] <- (as.numeric(break1)[j] + as.numeric(break1)[j+1])/2
-    }
-    xx2 <- as.data.frame(x2)[1,]
-    se <- as.data.frame(x2)[2,]
-    #if(i == 3){
-        #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
-        #abline(v = 25)}
-    m <- lm(t(xx2)~ time)
-    summary(m)
-    corr <- cor(t(xx2), time, use = "complete.obs")
-    cor_dis <- c(1:1000)
-    for (k in 1:1000){
-        cor_dis[k] <- cor(sample(t(xx2),i), time, use = "complete.obs")
-    }
-    #hist(cor_dis)
-    #lines(c(corr, corr), c(0,200), col = "red")
-    (p <- pnorm(corr, mean = mean(cor_dis), sd = sd(cor_dis)))
-    #with lm
-    #if(summary(m)$coef[8]> 0.05 | is.na(summary(m)$coef[8]) == TRUE){
-    #abline(m, lty = 2)}
-    #else{
-    #abline(m, lty = 1)}    
-    #with corrs
-    #if(p > 0.05){
-      # abline(m, lty = 2)}
-   # else{
-       # abline(m, lty = 1)}    
-   # if(i == 10){
-        #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
-        #abline(v = 25)}    
-    #rarecurve(comm2)
-    #abline(v = 1000)
-}
-    plot(time,t(xx2), xlab = "", las = 2, ylab = "Number of exotic bee species", main = "", xaxt = "n", ylim = c(min(xx2, na.rm = TRUE)- max(se), max(xx2, na.rm = TRUE)+ max(se)))
-    axis(1, at = time ,labels = lab, las = 2, cex.axis = 0.7)
-    arrows(time,as.numeric(xx2)+as.numeric(se), time, as.numeric(xx2)-as.numeric(se), angle=90, length = 0)
-    abline(m, lty = 2)
-
-############################################
-#resampling analysis for flies
-############################################
-
-    #load data
-    d <- read.csv(file = "data/NZ_fly_pollinators.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
-
-    #remove unidentified species
-    completeFun <- function(data, desiredCols) {
-        completeVec <- complete.cases(data[, desiredCols])
-        return(data[completeVec, ])
-    }
-
-    d <- completeFun(d, "exotic_native")
-
-    d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
-
-    #date
-    date2 <- strptime(d$CollectionDateISOFrom, "%Y\\%m\\%d")
-    d$jday <- date2$yday
-    d$year <- 1900 + date2$year
-    d$month <- date2$mon + 1
-
-    #select fields to be independent
-    #two approaches lat = 0 and others
-    lat <- subset(d, Lat != 0 & is.na(year) == F)#remove all records with no latitude recording
-    #unique fields by lata, long, day of the year and year
-    uni <- as.data.frame(cbind(lat$Gen_sp, lat$Lat, lat$Long, lat$jday, lat$year))
-    dup <- duplicated(uni)
-    #d2 <- d[-which(dup == TRUE),]
-    # lat/long = 0
-    #lat0 <- subset(d, Dlat == 0 & is.na(year) == F)
-    #str(lat0)
-    #uni2 <- as.data.frame(cbind(lat0$Gen, lat0$sp, lat0$LocalityStr, lat0$SubDivStr, lat0$StateProv, lat0$Country, lat0$jday, lat0$year))
-    #str(uni2)
-    #dup2 <- duplicated(uni2)
-
-    # remove duplicates
-    lat <- lat[-which(dup == TRUE),]
-    #lat0 <- lat0[-which(dup2 == TRUE),]
-    #d2 <- rbind(lat,lat0)
-    d2 <- lat
-    native <- filter(d2, exotic_native == "native") %>% droplevels()
-    exotic <- filter(d2, exotic_native == "exotic") %>% droplevels()
-
-    #length(which(native$year < 1900))
-    native <- subset(native, year > 1900)
-    for (i in 3:10){
-        break1 <- quantile(native$year, probs = seq(0,1,1/i))
-        lab <- c(1:i)
-        for(j in 1:i){
-            lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
-        }
-        native$time_period_quant  <- cut(native$year, breaks = break1, labels = lab)
-        comm2 <- table(native$time_period_quant, factor(native$Gen_sp))
-        (x2 <- rarefy(comm2,100, se = TRUE)) #same result as my resampling...
-        #rar[[i-2+10]] <- x2
-        #str(x2)
-        time <- c(1:i)
-        for(j in 1:i){
-            time[j] <- (as.numeric(break1)[j] + as.numeric(break1)[j+1])/2
-        }
-        xx2 <- as.data.frame(x2)[1,]
-        se <- as.data.frame(x2)[2,]
-        #if(i == 3){
-            #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
-            #abline(v = 100)}
-        m <- lm(t(xx2)~ time)
-        summary(m)
-        corr <- cor(t(xx2), time, use = "complete.obs")
-        cor_dis <- c(1:1000)
-        for (k in 1:1000){
-            cor_dis[k] <- cor(sample(t(xx2),i), time, use = "complete.obs")
-        }
-        #hist(cor_dis)
-        #lines(c(corr, corr), c(0,200), col = "red")
-        (p <- pnorm(corr, mean = mean(cor_dis), sd = sd(cor_dis)))
-        #with lm
-        #if(summary(m)$coef[8]> 0.05 | is.na(summary(m)$coef[8]) == TRUE){
-        #abline(m, lty = 2)}
-        #else{
-        #abline(m, lty = 1)}    
-        #with corrs
-        #if(p > 0.05){
-           # abline(m, lty = 2)}
-        #else{
-            #abline(m, lty = 1)}    
-        #if(i == 10){
-            #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
-            #abline(v = 40)}	
-        #rarecurve(comm2)
-        #abline(v = 1000)
-    }
-
-    plot(time,t(xx2), xlab = "", las = 2, ylab = "Number of native fly species", main = "", xaxt = "n", ylim = c(min(xx2, na.rm = TRUE)- max(se), max(xx2, na.rm = TRUE)+ max(se)))
-    axis(1, at = time ,labels = lab, las = 2, cex.axis = 0.7)
-    arrows(time,as.numeric(xx2)+as.numeric(se), time, as.numeric(xx2)-as.numeric(se), angle=90, length = 0)
-    abline(m, lty = 2)
-
-    ##########################################################################
-    #Resampling Analysis for exotic flies
-    ##########################################################################
-
-    #fin year = 0 quick and dirty
-    #length(which(exotic$year < 1900))
-    exotic <- subset(exotic, year > 1900)
-    for (i in 3:10){
-        break1 <- quantile(exotic$year, probs = seq(0,1,1/i)) #breaks can be equal, un tie those
-        if(any(duplicated(break1))){
-            break1[duplicated(break1)] <- break1[duplicated(break1)]+1
-        }
-        if(any(duplicated(break1))){
-            break1[duplicated(break1)] <- break1[duplicated(break1)]+1
-        }
-        lab <- c(1:i)
-        for(j in 1:i){
-            lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
-        }
-        exotic$time_period_quant  <- cut(exotic$year, breaks = break1, labels = lab)
-        comm2 <- table(exotic$time_period_quant, factor(exotic$Gen_sp))
-        (x2 <- rarefy(comm2,20, se = TRUE)) #same result as my resampling...
-        #rar[[i-2+10]] <- x2
-        #str(x2)
-        time <- c(1:i)
-        for(j in 1:i){
-            time[j] <- (as.numeric(break1)[j] + as.numeric(break1)[j+1])/2
-        }
-        xx2 <- as.data.frame(x2)[1,]
-        se <- as.data.frame(x2)[2,]
-        #if(i == 3){
-            #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
-            #abline(v = 20)}
-        m <- lm(t(xx2)~ time)
-        summary(m)
-        corr <- cor(t(xx2), time, use = "complete.obs")
-        cor_dis <- c(1:1000)
-        for (k in 1:1000){
-            cor_dis[k] <- cor(sample(t(xx2),i), time, use = "complete.obs")
-        }
-        #hist(cor_dis)
-        #lines(c(corr, corr), c(0,200), col = "red")
-        (p <- pnorm(corr, mean = mean(cor_dis), sd = sd(cor_dis)))
-        #with lm
-        #if(summary(m)$coef[8]> 0.05 | is.na(summary(m)$coef[8]) == TRUE){
-        #abline(m, lty = 2)}
-        #else{
-        #abline(m, lty = 1)}    
-        #with corrs
-        #if(p > 0.05){
-            #abline(m, lty = 2)}
-        #else{
-            #abline(m, lty = 1)}    
-        #if(i == 10){
-            #rarecurve(comm2, label = TRUE, step = 20, cex = 0.1, las = 2)
-            #abline(v = 40)}	
-        #rarecurve(comm2)
-        #abline(v = 1000)
-    }
-
-    plot(time,t(xx2), xlab = "", las = 2, ylab = "Number of exotic fly species", main = "", xaxt = "n", ylim = c(min(xx2, na.rm = TRUE)- max(se), max(xx2, na.rm = TRUE)+ max(se)))
-    axis(1, at = time ,labels = lab, las = 2, cex.axis = 0.7)
-    arrows(time,as.numeric(xx2)+as.numeric(se), time, as.numeric(xx2)-as.numeric(se), angle=90, length = 0)
-    abline(m, lty = 2)
-
-    ##########################################################################
-    #END
-    ##########################################################################
-
-```
-
 Figure 4. Changes in rarefied species richness for different pollinator groups in New Zealand over time. All trends were non-significant (α = 0.05).
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
-
-############################################
-#analyse the bee dataset
-############################################
-
-d <- read.csv("data/NZBees_allrecords.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
-
-#select required variables
-d <- select(d, one_of(c("species", "date", "locality", "collector", "collection", "method", "exotic_native")))
-d <- na.omit(d) #remove NAs
-
-#remove honeybees
-d$species <- as.factor(d$species)
-d <- d[ ! d$species %in% c("Apis mellifera"), ]
-d$species <- droplevels(d$species)
-
-#need to correct space between genus and species
-#levels(d$species)
-#d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
-d$Gen_sp <- d$species
-#sort(table(d$Gen_sp))
-#sum(table(d$Gen_sp))
-#head(d)
-
-#standardise dates
-date2 <- strptime(d$date, "%d/%m/%Y")
-d$jday <- date2$yday
-d$year <- 1900 + date2$year
-d$month <- date2$mon + 1
-#head(d$month)
-
-#select fields to be independent
-#completeFun <- function(data, desiredCols) {
-#  completeVec <- complete.cases(data[, desiredCols])
-#  return(data[completeVec, ])
-#}
-
-#d <- completeFun(d, c("day", "month", "year"))
-uni <- as.data.frame(cbind(d$Gen_sp, d$locality, d$collector, d$jday, d$year))
-#str(uni)
-dup <- duplicated(uni)
-
-# remove duplicates
-d2 <- d[-which(dup == TRUE),]
-
-#retain only sweep netting and no data
-#keep <- c("sweep", "no_data") #method is missing for a number of records
-#d2 <- d2[d2$method %in% keep, ]
-
-#create dataframes to remove species with less that 30 records
-no.ind <- as.data.frame(sort(table(d2$Gen_sp))) #14 species with more than 30 datapoints
-species.remove <- filter(no.ind, Freq < 30)#create df with species that have less than 30 records
-#visualize the data: #not with bees
-#plot(d2$Long, d2$Lat)
-#library(maps)
-#map("world", col="red", add=TRUE)
-
-#plot(d2$Long, d2$Lat, ylim = c(-50,-32), xlim = c(163,183))
-#map("world", col="red", add=TRUE) #fix wrong lat/long
-
-#############################
-# species models
-#############################
-#collection history per year
-B <- d2
-frq <- table(factor(B$Gen_sp),factor(B$year))
-bin <- ifelse(frq > 0, 1, 0)
-
-#i=1
-#last_year2 <- c(1:length(bin[,1]))
-#for (i in 1:length(bin[,1])){
-#  bin2 <- bin[i,-which(bin[i,] == 0)]
-#  last_year2[i] <- names(tail(bin2,1))
-#}
-#l_yBB <- data.frame(Gen_sp = rownames(bin), last_year = last_year2)
-##l_yBB #broken at i = 5 check.
-
-#year vector
-y <- as.numeric(colnames(bin))
-
-#weights per year (specimens collected/max specimens collected)
-
-effort <- c(1:length(y))
-wei <- c(c(1:length(y)))
-for (i in 1:length(y)){effort[i] <- sum(frq[,i])}
-for (i in 1:length(y)){wei[i] <- sum(frq[,i])/max(effort)}
-
-#### estimates for all sp
-#mB <- frq[,c(1:11)]
-#colnames(mB) <- c("estimate_bin", "SE_bin", "p_bin","estimate_frq", "SE_frq", "p_frq", "pred1880", "pred 2010","pred1880SE", "pred 2010SE","Logit")
-mB <- frq[,c(1:10)]
-colnames(mB) <- c("estimate_bin", "SE_bin", "p_bin","estimate_frq", "SE_frq", "p_frq", "pred1880", "pred 2015","pred1880SE", "pred 2015SE")
-try(for (i in 1:length(frq[,1])){
-    mB[i,1] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[2]
-    mB[i,2] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[4]
-    mB[i,3] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[8]
-    if (deviance(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))/df.residual(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit)) > 1.5) {
-        mB[i,4] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[2]
-        mB[i,5] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[4]
-        mB[i,6] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[8]
-        mB[i,7] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[1]
-        mB[i,8] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[2]
-        mB[i,9] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[1]
-        mB[i,10] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[2]
-    }
-    #mB[i,11] <- plogis(summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[2])-0.5
-    else {
-        mB[i,4] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[2]
-        mB[i,5] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[4]
-        mB[i,6] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[8]
-        mB[i,7] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[1]
-        mB[i,8] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[2]
-        mB[i,9] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[1]
-        mB[i,10] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[2]
-    }
-})
-
-############################################
-#plot model estimates and predictions
-############################################
-
-#prepare dataframe
-bee.est <- as.data.frame(mB)
-bee.est.cast <- dcast(bee.est, Var1 ~ Var2, value.var="Freq")
-bee.est.cast <- bee.est.cast[!(bee.est.cast$Var1 %in% species.remove$Var1),]
-
-#add direction change based on estimate and p-value
-bee.est.cast$change_in_occurance <- with(bee.est.cast, ifelse(
-    estimate_frq < 0 & p_frq < 0.05, 'Negative', ifelse(
-        p_frq >= 0.05, 'Stable', ifelse(
-            estimate_frq > 0 & p_frq >= 0.05, 'whoops', 'Positive'))))
-bee.est.cast$taxa <- "Bee"
-
-############################################
-#analyse fly dataset
-############################################
-
-#load data
-d <- read.csv(file = "data/NZ_fly_pollinators.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
-
-#remove unidentified species
-completeFun <- function(data, desiredCols) {
-    completeVec <- complete.cases(data[, desiredCols])
-    return(data[completeVec, ])
-}
-
-d <- completeFun(d, "exotic_native")
-
-d$Gen_sp <- paste(d$NameGenus, d$NameSpecies, sep = "_")
-
-#date
-date2 <- strptime(d$CollectionDateISOFrom, "%Y\\%m\\%d")
-d$jday <- date2$yday
-d$year <- 1900 + date2$year
-d$month <- date2$mon + 1
-
-#select fields to be independent
-#two approaches lat = 0 and others
-lat <- subset(d, Lat != 0 & is.na(year) == F)#remove all records with no latitude recording
-#unique fields by lata, long, day of the year and year
-uni <- as.data.frame(cbind(lat$Gen_sp, lat$Lat, lat$Long, lat$jday, lat$year))
-dup <- duplicated(uni)
-#d2 <- d[-which(dup == TRUE),]
-# lat/long = 0
-#lat0 <- subset(d, Dlat == 0 & is.na(year) == F)
-#str(lat0)
-#uni2 <- as.data.frame(cbind(lat0$Gen, lat0$sp, lat0$LocalityStr, lat0$SubDivStr, lat0$StateProv, lat0$Country, lat0$jday, lat0$year))
-#str(uni2)
-#dup2 <- duplicated(uni2)
-
-# remove duplicates
-lat <- lat[-which(dup == TRUE),]
-#lat0 <- lat0[-which(dup2 == TRUE),]
-#d2 <- rbind(lat,lat0)
-d2 <- lat
-no.ind <- as.data.frame(sort(table(d2$Gen_sp))) #14 species with more than 30 datapoints
-species.remove <- filter(no.ind, Freq < 30)#create df with species that have less than 30 records
-
-no.ind.fly <- as.data.frame(sort(table(d2$Gen_sp))) #14 species with more than 30 datapoints
-species.remove.fly <- filter(no.ind.fly, Freq < 30)#create df with species that have less than 30 records
-colnames(species.remove.fly)[1] <- "Gen_sp"
-
-##########################################################################
-#species models
-##########################################################################
-
-#collection history per year
-B <- d2
-frq <- table(factor(B$Gen_sp),factor(B$year))
-bin <- ifelse(frq > 0, 1, 0)
-
-#i=1
-#last_year2 <- c(1:length(bin[,1]))
-#for (i in 1:length(bin[,1])){
-   #bin2 <- bin[i,-which(bin[i,] == 0)]
-    #last_year2[i] <- names(tail(bin2,1))
-#}
-#l_yBB <- data.frame(Gen_sp = rownames(bin), last_year = last_year2)
-#l_yBB #broken at i = 5 check.
-
-#year vector
-y <- as.numeric(colnames(bin))
-
-#weights per year (specimens collected/max specimens collected)
-
-effort <- c(1:length(y))
-wei <- c(c(1:length(y)))
-for (i in 1:length(y)){effort[i] <- sum(frq[,i])}
-for (i in 1:length(y)){wei[i] <- sum(frq[,i])/max(effort)}
-
-##########################################################################
-#estimates for all sp
-##########################################################################
-
-#mB <- frq[,c(1:11)]
-#colnames(mB) <- c("estimate_bin", "SE_bin", "p_bin","estimate_frq", "SE_frq", "p_frq", "pred1880", "pred 2010","pred1880SE", "pred 2010SE","Logit")
-mB <- frq[,c(1:10)]
-colnames(mB) <- c("estimate_bin", "SE_bin", "p_bin","estimate_frq", "SE_frq", "p_frq", "pred1880", "pred 2015","pred1880SE", "pred 2015SE")
-try(for (i in 1:length(frq[,1])){
-    mB[i,1] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[2]
-    mB[i,2] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[4]
-    mB[i,3] <- summary(glm(bin[i,] ~ y, family = binomial(link=logit), na.action=na.omit, weight = wei))$coefficients[8]
-    if (deviance(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))/df.residual(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit)) > 1.5) {
-        mB[i,4] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[2]
-        mB[i,5] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[4]
-        mB[i,6] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[8]
-        mB[i,7] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[1]
-        mB[i,8] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[2]
-        mB[i,9] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[1]
-        mB[i,10] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[2]
-    }
-    #mB[i,11] <- plogis(summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = quasibinomial(link=logit), na.action=na.omit))$coefficients[2])-0.5
-    else {
-        mB[i,4] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[2]
-        mB[i,5] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[4]
-        mB[i,6] <- summary(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit))$coefficients[8]
-        mB[i,7] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[1]
-        mB[i,8] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$fit[2]
-        mB[i,9] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[1]
-        mB[i,10] <- predict(glm(cbind(frq[i,], effort-frq[i,]) ~ y, family = binomial(link=logit), na.action=na.omit), newdata = data.frame(y = c(1880,2015)), type = "response", se.fit = TRUE)$se.fit[2]
-    }
-})
-
-##########################################################################
-#plot model estimates and predictions
-##########################################################################
-
-#prepare dataframe
-fly.est <- as.data.frame(mB)
-fly.est.cast <- dcast(fly.est, Var1 ~ Var2, value.var="Freq")
-fly.est.cast <- fly.est.cast[!(fly.est.cast$Var1 %in% species.remove$Var1),]
-
-#add direction change based on estimate and p-value
-fly.est.cast$change_in_occurance <- with(fly.est.cast, ifelse(
-    estimate_frq < 0 & p_frq < 0.05, 'Negative', ifelse(
-        p_frq >= 0.05, 'Stable', ifelse(
-            estimate_frq > 0 & p_frq >= 0.05, 'whoops', 'Positive'))))
-fly.est.cast$taxa <- "Fly"
-
-#merge bee and fly estimate datasets
-est.all <- rbind(fly.est.cast,bee.est.cast)
-est.all$Var1 <- gsub("_", " ", est.all$Var1)
-
-####################################################
-#plot frequency estimates
-####################################################
-
-p <- ggplot(est.all, aes(Var1, estimate_frq, group=taxa))
-p <- p + xlab(NULL) + ylab("Frequency estimate")
-p <- p + theme(text = element_text(size=16))
-p <- p + facet_wrap(~taxa,scales = "free_x")
-p <- p + geom_point(aes(colour=change_in_occurance), size = 2)
-p <- p + geom_errorbar(aes(ymin=estimate_frq-SE_frq, ymax=estimate_frq+SE_frq, colour=change_in_occurance), width = 0)
-p <- p + geom_hline(aes(yintercept=0), linetype="dashed")
-p <- p + theme(panel.grid.minor = element_blank(),
-               panel.background = element_blank(),
-               axis.line = element_line(colour = "black")) +
-    theme(panel.border=element_rect(colour = "black", fill = "NA", size = 1)) +
-    theme(axis.text.y=element_text(angle= 360, hjust = 0.5, vjust = 0.5, size =14),
-          axis.title.y=element_text(size=22, vjust = 1),
-          axis.text.x=element_text(face = "italic", angle= 90, hjust = 0, vjust = 0.5, size =7),
-          axis.title.x=element_text(size=22, vjust = 1),
-          axis.text=element_text(colour = "black"))+
-    theme(strip.background = element_rect(colour="NA", fill=NA),
-          strip.text = element_text(size=16))
-p <- p + theme(axis.title.y=element_text(margin=margin(0,20,0,0)))
-p <- p + scale_colour_brewer(palette = "Dark2")
-p$labels$colour <- "Change in occurance"
-p
-
-####################################################
-#END
-####################################################
-
-```
-
 Figure 5. Model estimated changes (± 1 SE) in the relative occurrence frequency for different New Zealand bee and fly species in museum collections over time.
@@
 
-A recent survey highlighted that more than 90% of the papers studying pollinator responses to land use change focused solely on richness and abundance descriptors [@winfree2011].
+A recent survey highlighted that more than 90% of the papers studying pollinator responses to land use change focused solely on richness and abundance descriptors [@Winfree2011].
 But in addition to local (alpha) diversity and regional (gamma) diversity, we need to understand changes in turnover between sites (beta diversity).
@@
 This is specifically relevant to identify mechanisms of decline and adaptation.
-For example, one particularly important source of information is the date the specimen was collected, which allows the tracking of phenological advances congruent with contemporary climate change [#@Bartomeus2011].
+For example, one particularly important source of information is the date the specimen was collected, which allows the tracking of phenological advances congruent with contemporary climate change [@Bartomeus2011].
 In addition, museum labels on pollinator specimens often include information about the host plant on which the specimen was collected.
-This information is precious to understand past and present species interactions [#@Tylianakis2010ELE].
-Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [#@kleinj_raemakers].
+This information is precious to understand past and present species interactions [@Tylianakis2008].
+Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [@Kleijn2008].
 Finally, museum specimens can be measured to track evolutionary changes by measuring specific specimen traits.
-This has been already explored to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.
+This has been already explored to show tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016] shrinkage with climate and land use change.
 

### 2018-04-04 SP_data_cleaning.Rmd
All data in, just need polishing!

@@
 library(taxize)
+#SLOW!!
 spp <- clean_species(dat$gen_sp, rows = 1) 
-species <- dat$gen_sp
-#table(dat2$gen_sp, dat2$when)
-#Andrena_pilipes = carbonaria
+spp[which(as.character(spp$species) != as.character(spp$final_names)),]
+spp$species
+temp <- spp[which(is.na(spp$final_names)),]
+str(temp)
+options(max.print = 4000)
+print(temp)
+spp[which(spp$species == "Andrena pilipes"),2] <- "Andrena carbonaria"
+spp[which(spp$species == "Megabombus laesus"),2] <- "Bombus laesus"
+spp[which(spp$species == "Colletes eousMorice,1904"),2] <- "Colletes eous"
+spp[which(spp$species == "Reanthidium nigricolle"),2] <- "Anthidium nigricolle"
+spp[which(spp$species == "Megabombus muscorum"),2] <- "Bombus muscorum"
+spp[which(spp$species == "Megabombus ruderatus"),2] <- "Bombus ruderatus"
+spp[which(spp$species == "Megabombus sylvarum"),2] <- "Bombus sylvarum"
+
+head(dat)
+head(spp)
+dat <- merge(dat, spp[,1:2], by.x = "gen_sp", by.y = "species", all.x = TRUE)
+head(dat, 100)
 #remove A. mellifera
-rich_tot <- dcast(dat, when ~ ., value.var = "gen_sp", 
+dat <- subset(dat, !matched_name2 %in% c("Apis mellifera"))
+# and genus only
+pos <- regexpr(dat$matched_name2, pattern = " ", fixed = TRUE)
+str(pos)
+dat2 <- dat[-which(pos < 1),]
+dat2$matched_name2
+#NA
+dat3<- dat2[-which(is.na(dat2$matched_name2)),]
+unique(dat3$matched_name2)
+dat <- dat3
+
+rich_tot <- dcast(dat, when ~ ., value.var = "matched_name2", 
               fun.aggregate = function(x) length(unique(x)))
@@
 #extract genus
-pos <- regexpr(dat$gen_sp, pattern = " ", fixed = TRUE)
-dat$genus <- substr(dat$gen_sp, 
+pos <- regexpr(dat$matched_name2, pattern = " ", fixed = TRUE)
+dat$genus <- substr(dat$matched_name2, 
                     start = 1,
@@
 subset(genus_tot2, now == 1 & asensio == 1)
-subset(genus_tot2, now == 1 & asensio == 0)
+(x <- subset(genus_tot2, now == 1 & asensio == 0))
 
 #remove non apis genus! check with Curro.
+#Anthocopa is Hoplitis
+#Chalicodoma
+#Pseudapis
+
+dat4 <- dat[which(!dat$genus %in% x$genus[c(1:5,7:25,27:34)]),]
+dat <- dat4
+#Heliophila <-  Anthophora
+#Metallinella <- Osmia
+#Reanthidium <-  Anthidium
+head(dat)
+dat[which(dat$genus == "Heliophila"),"matched_name2"] <- "Anthophora bimaculata"
+dat[which(dat$genus == "Metallinella"),"matched_name2"] <- "Osmia brevicornis"
+dat[which(dat$genus == "Reanthidium"),"matched_name2"] <- "Pseudoanthidium reticulatum"
+
+#resave data
+write.csv(dat, "data/spain_clean_used.csv", row.names = F)
+
+dat <- read.csv("data/spain_clean.csv")
+#Repeat richness and genus
+rich_tot <- dcast(dat, when ~ ., value.var = "matched_name2", 
+              fun.aggregate = function(x) length(unique(x)))
+rich_tot #wow.
+#extract genus
+pos <- regexpr(dat$matched_name2, pattern = " ", fixed = TRUE)
+dat$genus <- substr(dat$matched_name2, 
+                    start = 1,
+                    stop = pos-1)
+genus_tot <- dcast(dat, when ~ ., value.var = "genus", 
+              fun.aggregate = function(x) length(unique(x)))
+genus_tot2 <- dcast(dat,  genus ~ when, value.var = "genus", 
+              fun.aggregate = function(x) length(unique(x)))
+genus_tot #wow, lot's of genus lost!
+genus_tot2
+subset(genus_tot2, now == 0 & asensio == 1) #20 lost!
+subset(genus_tot2, now == 1 & asensio == 1) #27 stable
+subset(genus_tot2, now == 1 & asensio == 0) #2 added!
+genus_tot3 <- dcast(dat,  genus ~ when, value.var = "matched_name2", 
+              fun.aggregate = function(x) length(unique(x)))
+genus_tot3
+colSums(genus_tot3[,2:3])
 
@@
 #head(dat)
-rich <- dcast(dat, locality ~ when, value.var = "gen_sp", 
+rich <- dcast(dat, locality ~ when, value.var = "matched_name2", 
               fun.aggregate = function(x) length(unique(x)))
-#rich
-#t.test(rich$asensio, rich$now, paired = TRUE)
+(rich <- rich[-8,])
+t.test(rich$asensio[-4], rich$now[-4], paired = TRUE) #0.056!
 
@@
 library(vegan)
+str(dat)
+unique(dat$freq)
+dat[which(is.na(dat$freq)),"freq"] <- 1 
+dat$freq <- as.numeric(dat$freq)
 temp <- subset(dat, when == "now")
-comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
+comm <- dcast(temp, locality ~ matched_name2, value.var = "freq", 
               fun.aggregate = sum, na.rm = TRUE)
@@
 comm2 <- comm[,-1] 
-#rowSums(comm2) #toro is low, but same sampling effort. 
-temp <- specaccum(comm2, "random")
-#plot(temp)
-#rarecurve(comm2)
-now <- rarefy(x = comm2, sample = 120)
+comm2 <- comm2[-8,] 
+rowSums(comm2) #toro is low, but same sampling effort. 
+plot(specaccum(comm2, "random"))
+rarecurve(comm2) #Good!
+now <- rarefy(x = comm2, sample = 1000)
 
 temp <- subset(dat, when == "asensio")
-comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
+comm <- dcast(temp, locality ~ matched_name2, value.var = "freq", 
               fun.aggregate = sum, na.rm = TRUE)
-#comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
- #             fun.aggregate = length, na.rm = TRUE)
 rownames(comm)[-8] <- as.character(comm$locality[-8])
 comm2 <- comm[,-1]
-#rowSums(comm2) #sanbernardo useless. Others ok! Delives super sampled!!
-temp <- specaccum(comm2, "random")
-#plot(temp)
-#rarecurve(comm2) #for delives quite complete!
-asensio <- rarefy(x = comm2, sample = 120)
+rowSums(comm2) #sanbernardo useless. Others ok! Delives super sampled!!
+plot(specaccum(comm2, "random"))
+rarecurve(comm2) #for delives quite complete!
+head(comm2)
+rarecurve(comm2[-1,]) #for the rest is ok, but sanbernardo.
+abline(v = 1000)
+asensio <- rarefy(x = comm2, sample = 1000)
 
 ttest <- t.test(asensio[c(1,2,3,5,6,7)], now[c(1,2,3,5,6,7)], paired = TRUE)
+ttest #0.05828
 
@@
 }
-
-#do by coverage?
-#clean sp.
 ```

### 2018-04-04 ms.Rmd
All data in, just need polishing!

@@
 ---
-title: "Historical collections as baseline for assessing the extent of the pollinator crisis"
+title: Historical collections as baseline for assessing the extent of the pollinator
+  crisis
 author: "I. Bartomeus1, J. Stavert2, D. Ward2,3, O. Aguado4"
@@
 output:
+  word_document: default
   html_document: default
-bibliography: data/references.xml
 editor_options:
   chunk_output_type: console
+bibliography: data/references.xml
 ---
@@
 Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
-We show that for both regions, winner and loser species emerge despite significant changes in species richness over time.
+We show that for both regions, winner and loser species emerge despite non-significant changes in species richness over time.
 There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis, but to progress this field data must be made feely accessible to researchers.
@@
 For example, researchers have recently estimated bee richness declines for every country in Europe using predictions from models of pollination associations with different land-use types [@DePalma2017].
-A second promising is the use of data collected from pollinator monitoring programs, which are often driven by citizen scientists.
+A second important method is the use of data collected from pollinator monitoring programs, which are often driven by citizen scientists.
 This approach was inspired by successful butterfly monitoring programs [@vanSwaay2008] and is currently being extended to other pollinator taxa.
 However, these programs require significant time to generate long-term datasets and cannot be used to assess historic pollinator populations.
-Finally, the most practical approach for assessing long-term pollinator population trends is to use the historical information on species occurrences, which is archived in museum collections [e.g., @bartomeus2013a].
+Finally, the most practical approach for assessing long-term historical pollinator population trends is to use the historical information on species occurrences, which is archived in museum collections [e.g., @bartomeus2013a].
 
 In this review, we first assess current evidence for pollinator richness declines and present a road map outlining a strategy for using museum collection data to fill current knowledge gaps.
-We highlight the major technical difficulties involved in using museum collection data and demonstrate several approaches for analysing different types of museum collect data to assess long-term trends in pollinator populations.
+We highlight the major technical difficulties involved in using museum collection data and demonstrate several approaches for analysing different types of museum collection data to assess long-term trends in pollinator populations.
 Finally, we highlight the need to move beyond simple biological diversity descriptors and unleash the power of historical data to assess changes in species interactions, ecosystem functioning and evolutionary changes through time.
@@
 This is a pattern also detected in the UK [@Ollerton2014].
-However, the US findings above are in stark contrast with the widespread local extinctions reported by Burkle [-@Burkle2013], who compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants.
+However, the reginal US findings above are in stark contrast with the widespread local extinctions reported by Burkle [-@Burkle2013], who compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants.
 It is important to note that Burkle [-@Burkle2013] focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (@Collado2018).
@@
 Generally speaking, reported declines for bumblebees are the most severe of all pollinator taxa.
-For example, declines of up to 18% in local bumble richness have been reported for Belgium and the Netherlands [@Carvalheiro2013].
+For example, declines of up to 18% in local bumblebee richness have been reported for Belgium and the Netherlands [@Carvalheiro2013].
 In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011].
@@
 
-Outside of Europe and the US, there are no studies on pollinator declines using historical records.
+Outside of Europe and the US and for non-insect taxa, there are no studies on pollinator declines using historical records.
 In Figure 1, we show a comparison of bee species richness per area (a), bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes (d) to highlight the little coverage that we have globally.
@@
 
-Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness and country size, which accounts for the species-area relationship; (b) syrphid (c) and butterflies (d) richness percent change in the last century.
+Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness and country size, which accounts for the species-area relationship; Warmer colours indicate higher bee diversity; Note that some African countries may have incomplete listed faunas and that Alaska is included with US values.
+(b) Syrphid (c) bee and butterfly (d) richness percent change in the last century.
 Warmer colours indicate steeper declines.
 Countries without data are coloured in white.
-Note that some African countries may have incomplete listed faunas and that Alaska is included with US values.
 
@@
 
-As we have shown, there is a paucity of countries for which museum data is available (Figure 1), and hence can be used as baseline for assessing pollinator population declines.
+As we have shown, there is a paucity of countries for which historical data is available (Figure 1), and hence can be used as baseline for assessing pollinator population declines.
 While ideally one would aim to digitize all museum collection records, this is unlikely in the near future, mainly due to funding constraints.
@@
 GBIF (https://www.gbif.org/) is a central repository for global species occurrence data.
-Much of these data come from museums, private collections and government research institutes, but it integrates several other sources [such as??].
+Much of these data come from museums, private collections and government research institutes, but it integrates several other sources.
 In combination with the popular statistical language R [@RCoreTeam2017], GBIF can be directly queried into your computer [@Chamberlain2014] and data availability checked for your region of interest.
@@
 While values are high on both axes for countries such as the UK and the US so that data can potentially be analysed without further data collection effort, most countries fall short in one or both axes.
-Hence, for countries with high numbers of recent records, researchers should prioritize the digitalization of old material before embarking on data analyses.
+Hence, for countries with high numbers of recent records, researchers should prioritize the digitalization of old material before embarking on data analyses (e.g.
+Switcherland,...
+).
 In contrast, for countries with large numbers of historic records, re-surveys are required before analysis [e.g., @Cameron2011].
-Interestingly, we show that greater than 100 countries have less than 1,000 records, making them poor candidates for analysing long-term pollinator population trends.
-However, countries like X or X… It is also important to note that historical records are not always vouchered in local museums, but many European and US museums contain large collections of exotic pollinators [do you mean exotic species collected in the US/EU or species collected in their native countries and bought back to and vouchered in US/EU museums?].
+Interestingly, we show that greater than 100 [check]() countries have less than 1,000 records, making them poor candidates for analysing long-term pollinator population trends.
+However, countries like X or X… It is also important to note that historical records are not always vouchered in local museums, but many European and US museums contain large collections of pollinators collected elsewhere.
 Aside from bees, similar exploratory analysis can easily be done for other taxa.
@@
 
-```
-
-```{r alternative plot (bad), echo=FALSE, message=FALSE, warning=FALSE}
+#alternative plot
 lhf$percent_post1980 <- lhf$post1980 / (lhf$pre1980+lhf$post1980)
@@
 lhf <- lhf[-which(is.na(lhf$country)),]
-plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
-abline(h = 7.5, v = 7.5)
+#plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country))
+#abline(h = 7.5, v = 7.5)
 lhf$log_pre1980 <- log(lhf$pre1980)
 lhf$log_post1980 <- log(lhf$post1980)
+
+#table?
 lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) < 7.5),]
@@
 #36 countries!
-#exp(7.5) #1800 records.
-
+#exp(7.5) #> 1800 records in both axes!
+```
+
+```{r get rgbif, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
 #Can we do rar richness per time period $ country and test? YES!! 
@@
 save(occ_list, file = "data/gbif.RData")
+
+```
+
+```{r map rgbif, echo=FALSE, message=FALSE, warning=FALSE}
+countries <- lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),"country"]
+load("data/gbif.RData")
 #3) split by year
@@
 We highlight a few representative examples.
+b) boxplot showing rarefied number of species collected before and after 1980.
+c) same data plotted in a map.
+This data is likely to contain strong undetected biases, but indicates little change...
 
@@
 Digitization of these records, along with a re-survey of these locations provided an excellent dataset for a before and after comparison of bee communities at these specific locations.
-[from Darren: Has there been any changes in the land use between the two periods?]
-
-In brief, we checked sampling completeness for both time periods, and compared species richness for each site before and after 1980 with a paired t-test.
-We found that overall there was no change in species richness (P > 0.05).
-However, ...
-show a steep decline… and groups like nomadas, tyreus, prosopis and teralonias are completely extirpated (appendix?).
-[from Darren: Any information on these groups.
-Are they specialist species, or more sensitive to land change?].
+Interestingly, this localities have experienced all large agricultural intensification [expand]()
+
+In brief, we cleaned taxonomical names using taxize (@chamberlain), checked sampling completeness for both time periods and compared species richness for each site before and after 1980 with a paired t-test.
+We found that there is a trend to find a reduced number of species per site (P = 0.058).
+However, this is higlhly dependant on the site as 2 out of 3 sites show no richness declines.
+However, there are clear groups of species suffereing larger declines.
+Andrenids and its parasites being among the losers and Halictids among the winners.
+This is in accordance with findings elsewhere (e.g.
+@Bartomeusetal2013a) (appendix?).
 
 ```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
-dat <- read.csv("data/spain_clean.csv")
-#head(dat)
-library(reshape2)
-dat2 <- melt(dat, id.vars = c("locality", "gen_sp", "when"), 
-             measure.vars = "freq")
-#head(dat2, 100)
-#sort(unique(dat2$gen_sp))
-#clean with clean_sp
-#remove sp, NA, " ",
-#NA_NA   
-#Hylaeus_Variegatus
-
-#table(dat2$gen_sp, dat2$when)
-#Andrena_pilipes = carbonaria
-#remove A. mellifera
-
-#losts easely spotted: nomadas & Tyreus, prosopis, teralonias
+dat <- read.csv("data/spain_clean_used.csv")
+#Repeat richness and genus
+rich_tot <- dcast(dat, when ~ ., value.var = "matched_name2", 
+              fun.aggregate = function(x) length(unique(x)))
+rich_tot #wow.
+#extract genus
+pos <- regexpr(dat$matched_name2, pattern = " ", fixed = TRUE)
+dat$genus <- substr(dat$matched_name2, 
+                    start = 1,
+                    stop = pos-1)
+genus_tot <- dcast(dat, when ~ ., value.var = "genus", 
+              fun.aggregate = function(x) length(unique(x)))
+genus_tot2 <- dcast(dat,  genus ~ when, value.var = "genus", 
+              fun.aggregate = function(x) length(unique(x)))
+genus_tot #wow, lot's of genus lost!
+#genus_tot2
+subset(genus_tot2, now == 0 & asensio == 1) #20 lost!
+#subset(genus_tot2, now == 1 & asensio == 1) #27 stable
+#subset(genus_tot2, now == 1 & asensio == 0) #2 added!
+genus_tot3 <- dcast(dat,  genus ~ when, value.var = "matched_name2", 
+              fun.aggregate = function(x) length(unique(x)))
+genus_tot3
+#colSums(genus_tot3[,2:3])
+
+#losts easely spotted: Andrenas, Nomadas & Tyreus, prosopis, teralonias
 
@@
 #head(dat)
-rich <- dcast(dat, locality ~ when, value.var = "gen_sp", 
-              fun.aggregate = function(x) length(unique(x)))
-#rich
-#t.test(rich$asensio, rich$now, paired = TRUE)
+# rich <- dcast(dat, locality ~ when, value.var = "matched_name2", 
+#               fun.aggregate = function(x) length(unique(x)))
+# (rich <- rich[-8,])
+# t.test(rich$asensio[-4], rich$now[-4], paired = TRUE) #0.056!
+# 
+# boxplot(rich$asensio[c(1,2,3,5,6,7)], rich$now[c(1,2,3,5,6,7)], names = c("1980's", "2016"), ylab = "species richness", 
+#             las = 1, border = "light grey")
+# points(x = c(rep(1,6), rep (2,6)), y = c(rich$asensio[c(1,2,3,5,6,7)],
+#                                          rich$now[c(1,2,3,5,6,7)]), col = rep(c(1,2,3,5,6,7),2), pch = 19)
+# for(i in c(1,2,3,5,6,7)){
+#     lines(x = 1:2, y = c(rich$asensio[i], rich$now[i]), col = i)
+# }
 
 library(vegan)
+#str(dat)
+#unique(dat$freq)
+dat[which(is.na(dat$freq)),"freq"] <- 1 
+dat$freq <- as.numeric(dat$freq)
 temp <- subset(dat, when == "now")
-comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
+comm <- dcast(temp, locality ~ matched_name2, value.var = "freq", 
               fun.aggregate = sum, na.rm = TRUE)
@@
 comm2 <- comm[,-1] 
-rowSums(comm2)
-temp <- specaccum(comm2, "random")
-#plot(temp)
-#rarecurve(comm2)
-now <- rarefy(x = comm2, sample = 120)
+comm2 <- comm2[-8,] 
+#rowSums(comm2) #toro is low, but same sampling effort. 
+#plot(specaccum(comm2, "random"))
+#rarecurve(comm2) #Good!
+now <- rarefy(x = comm2, sample = 1000)
 
 temp <- subset(dat, when == "asensio")
-comm <- dcast(temp, locality ~ gen_sp, value.var = "freq", 
+comm <- dcast(temp, locality ~ matched_name2, value.var = "freq", 
               fun.aggregate = sum, na.rm = TRUE)
@@
 comm2 <- comm[,-1]
-rowSums(comm2)
-temp <- specaccum(comm2, "random")
-#plot(temp)
-#rarecurve(comm2)
-asensio <- rarefy(x = comm2, sample = 120)
-
-ttest <- t.test(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)], paired = TRUE)
-
-boxplot(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)], names = c("1980's", "2016"), ylab = "rarefies species richness", 
+#rowSums(comm2) #sanbernardo useless. Others ok! Delives super sampled!!
+#plot(specaccum(comm2, "random"))
+#rarecurve(comm2) #for delives quite complete!
+#head(comm2)
+#rarecurve(comm2[-1,]) #for the rest is ok, but sanbernardo.
+#abline(v = 1000)
+asensio <- rarefy(x = comm2, sample = 1000)
+
+ttest <- t.test(asensio[c(1,2,3,5,6,7)], now[c(1,2,3,5,6,7)], paired = TRUE)
+ttest #0.05828
+
+boxplot(asensio[c(1,2,3,5,6,7)], now[c(1,2,3,5,6,7)], names = c("1980's", "2016"), ylab = "rarefied species richness", 
             las = 1, border = "light grey")
-points(x = c(rep(1,5), rep (2,5)), y = c(asensio[c(1,2,5,6,7)], now[c(1,2,5,6,7)]), col = rep(1:5,2), pch = 19)
-for(i in c(1,2,5,6,7)){
+points(x = c(rep(1,6), rep (2,6)), y = c(asensio[c(1,2,3,5,6,7)], now[c(1,2,3,5,6,7)]), col = rep(c(1,2,3,5,6,7),2), pch = 19)
+for(i in c(1,2,3,5,6,7)){
     lines(x = 1:2, y = c(asensio[i], now[i]), col = i)
@@
 
-Figure 3. Rarefied richness at the 5 plots... need to work more on the raw data.
+Figure 3. Rarefied richness at the 6 plots resurveyed.
 

### 2018-04-05 ms.Rmd
altered plots and edited writing

@@
   crisis
-author: "I. Bartomeus1, J. Stavert2, D. Ward2,3, O. Aguado4"
+author: "Bartomeus^1^, I., Stavert^2^, J.R., Ward^2,3^, D., and Aguado^4^, O."
 csl: data/philb_style.xml
 output:
+  html_document: default
+  pdf_document: default
   word_document: default
-  html_document: default
 editor_options:
@@
 As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions.
-The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@Buchman1996].
+The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@Buchmann1996].
 However, early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time.
@@
 However, these programs require significant time to generate long-term datasets and cannot be used to assess historic pollinator populations.
-Finally, the most practical approach for assessing long-term historical pollinator population trends is to use the historical information on species occurrences, which is archived in museum collections [e.g., @bartomeus2013a].
+Finally, the most practical approach for assessing long-term historical pollinator population trends is to use the historical information on species occurrences, which is archived in museum collections [e.g., @Bartomeus2013a].
 
@@
 
-In brief, we cleaned taxonomical names using taxize (@chamberlain), checked sampling completeness for both time periods and compared species richness for each site before and after 1980 with a paired t-test.
+In brief, we cleaned taxonomical names using taxize [@chamberlain], checked sampling completeness for both time periods and compared species richness for each site before and after 1980 with a paired t-test.
 We found that there is a trend to find a reduced number of species per site (P = 0.058).
@@
 Andrenids and its parasites being among the losers and Halictids among the winners.
-This is in accordance with findings elsewhere (e.g.
-@Bartomeusetal2013a) (appendix?).
+This is in accordance with findings elsewhere [e.g.
+@Bartomeusetal2013a] (appendix?).
 
@@
 
-We follow Bartomeus et al.
-2013a protocol to analysze this data.
+We followed the protcols outlined in [@Bartomeus2013a] to analysze these data.
 First, we filtered our original datasets so that data used for analyses only included independent collection events.
@@
 #length(which(native$year < 1900))
-native <- subset(native, year > 1900)
-for (i in 3:10){
+native <- subset(native, year > 1800)
+for (i in 5){
     break1 <- quantile(native$year, probs = seq(0,1,1/i), na.rm = TRUE)
+    #x <- c(1909,1965,1970,1974,1976,1978,1981,1986,1994,2012)
     lab <- c(1:i)
@@
 bee.est.cast <- bee.est.cast[!(bee.est.cast$Var1 %in% species.remove$Var1),]
+bee.est.cast <- droplevels(bee.est.cast)
 
@@
 
+exotic.native.bee <- c("exotic","exotic","exotic","native","native","native","native","exotic","native","native","native","native","native",                        "native","native","native","native","native","native","native","native","native","native","native","native","native",                        "exotic")
+
+bee.est.cast <- cbind(bee.est.cast,exotic.native.bee)
+colnames(bee.est.cast)[14] <- "exotic_native"
 ############################################
@@
 fly.est.cast <- fly.est.cast[!(fly.est.cast$Var1 %in% species.remove$Var1),]
+fly.est.cast <- droplevels(fly.est.cast)
 
@@
 fly.est.cast$taxa <- "Fly"
+exotic.native.fly <- c("exotic","exotic","exotic","exotic","exotic","native","native","native","native","exotic","native","native","exotic","native")
+fly.est.cast <- cbind(fly.est.cast,exotic.native.fly)
+colnames(fly.est.cast)[14] <- "exotic_native"
 
@@
 
+est.all$Var1 <- factor(est.all$Var1, levels = est.all$Var1[order(est.all$exotic_native)])
+
 ####################################################
@@
 p <- ggplot(est.all, aes(Var1, estimate_frq, group=taxa))
-p <- p + xlab(NULL) + ylab("Frequency estimate")
-p <- p + theme(text = element_text(size=16))
-p <- p + facet_wrap(~taxa,scales = "free_x")
+p <- p + xlab(NULL) + ylab("Change in relative occurance")
+p <- p + facet_wrap(~taxa,nrow=2,ncol=1,scales = "free_x")
 p <- p + geom_point(aes(colour=change_in_occurance), size = 2)
@@
 p <- p + geom_hline(aes(yintercept=0), linetype="dashed")
+p <- p + geom_vline(data=filter(est.all, taxa=="Bee"), aes(xintercept=5.5), colour="grey")
+p <- p + geom_vline(data=filter(est.all, taxa=="Fly"), aes(xintercept=7.5), colour="grey")
+p <- p + geom_text(data=data.frame(Var1=3,estimate_frq=0.145,taxa="Bee",lab='Exotic'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
+p <- p + geom_text(data=data.frame(Var1=4,estimate_frq=0.145,taxa="Fly",lab='Exotic'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
+p <- p + geom_text(data=data.frame(Var1=16.5,estimate_frq=0.145,taxa="Bee",lab='Native'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
+p <- p + geom_text(data=data.frame(Var1=11,estimate_frq=0.145,taxa="Fly",lab='Native'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
 p <- p + theme(panel.grid.minor = element_blank(),
@@
     theme(panel.border=element_rect(colour = "black", fill = "NA", size = 1)) +
-    theme(axis.text.y=element_text(angle= 360, hjust = 0.5, vjust = 0.5, size =14),
-          axis.title.y=element_text(size=22, vjust = 1),
-          axis.text.x=element_text(face = "italic", angle= 90, hjust = 0, vjust = 0.5, size =7),
+    theme(axis.text.y=element_text(angle= 360, hjust = 0.5, vjust = 0.5, size =12),
+          axis.title.y=element_text(size=18, vjust = 1),
+          axis.text.x=element_text(face = "italic", angle= 45, hjust = 1, vjust = 1, size =7),
           axis.title.x=element_text(size=22, vjust = 1),
@@
     theme(strip.background = element_rect(colour="NA", fill=NA),
-          strip.text = element_text(size=16))
+          strip.text = element_text(size=14))
 p <- p + theme(axis.title.y=element_text(margin=margin(0,20,0,0)))
@@
 
-A recent study found that more than 90% of the papers studying pollinator responses to land-use change focused solely on richness and abundance descriptors [@winfree2011].
+A recent study found that more than 90% of the papers studying pollinator responses to land-use change focused solely on richness and abundance descriptors [@Winfree2011].
 But in addition to local (alpha) diversity and regional (gamma) diversity, we need to assess changes in turnover between sites (beta diversity).
@@
 
-#Conclucions
-
-As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors.
-Using two case studies, we show that collaboration with both museum curators and ecologists is key to adequately understanding the data and treating it appropriately.
-Researchers and curators should aim to make digitalized data readily available and easy to share.
-Centralization of regional and national museum data in existing global platforms such as GBIF would facilitate free and widespread access.
-However, specific datasets can also be stored in different webpages or database repositories (e.g.
-Universities and museum webpages, Dryad) as far as they are well documented and can be easely retrived and combined with other datasets using open science tools [e.g., @Dallas2016].
-
-We also need to revolutionize the way that researchers collaborate with museums, which should start with fostering healthy bidirectional relationships.
-For instance, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered (@Packer2018; @Wheeler2003), rendering them less useful for future research.
-To improve this, strong communication between museums and researchers is required, but this will only be achieved if there is not adequate funding, and recognition that accurate data recording and long-term preservation are critical for research [@Ward2015].
-
-Ultimately, to identify global trends in pollinator declines we require robust data, collected from diverse geographic regions.
-Further, it is crucial that data are analysed appropriately.
-This requires researches to identify biases in the data and fill  taxonomic and geographic gaps in knowledge of pollinators status where possible.
-We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees.
-For the US and Europe there have been few regional bee extinctions [@Ollerton2014; @Bartomeus2013a], but in disturbed ecosystems declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006].
+#Conclusions
+
+Unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors.
+Using two case studies, we show that collaboration with both museum curators and ecologists is key to understanding the data and treating it appropriately.
+To progress our understanding of the global pollination crisis, researchers and curators must aim to digitize museum collection data and make it readily available it a format that is widely accessible.
+Centralization of regional and national museum collection data in existing global platforms, such as GBIF, would facilitate free and widespread access.
+However, datasets could also be stored in alternative webpages or database repositories (e.g., University and museum webpages or Dryad) providing they are well documented, easily retrieved and can be combined with other datasets using open science tools [e.g., @Dallas2016].
+Further, we must revolutionize the way that researchers collaborate with museums, to foster healthy bidirectional relationships.
+For example, ecological researchers collect massive numbers of specimens, but these are often inappropriately vouchered [@Packer2018; @Wheeler2003], rendering them less useful for future research.
+To improve this, strong communication between museums and researchers is required, but can only be achieved with adequate funding and recognition that accurate data recording and long-term preservation are critical for research [@Ward2015].
+
+To identify global trends in pollinator declines we require robust data, collected from diverse geographic regions.
+It is also crucial that these data are analysed appropriately.
+This requires researches to identify biases and to any fill taxonomic and geographic gaps where possible.
+We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe, and for pollinator groups other than bees.
+For the US and Europe, there have been few regional bee extinctions [@Ollerton2014; @Bartomeus2013a] but in disturbed ecosystems, declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006].
 For most other pollinator taxa and regions throughout the world we know almost nothing.
 Moving forward, the first step for many taxa will be to identify and describe species.
-Only then can we begin to document declines.
+Only then can we begin to document pollinator declines.
 
@@
 
+We thank Curro Molina, Patrick McQuinn, and Crona McMonagle for data entry.
+Madrid museum of Historia Natural and ITACyL for access to historical collections.
+We thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.
 IB was funded by FBBVA project.
@@
 Asensio for sharing his historical data and knowledge.
-We thank Curro Molina, Patrick McQuinn, and Crona McMonagle for data entry.
-Madrid museum of Historia Natural and ITACyL for access to historical collections.
-We also thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.
 

### 2018-04-05 ms.Rmd
Small edits, too tired to finish it today :D

@@
 
-[Historical collections or museum collections?]()
-
 *Affiliations:*
@@
 Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
-We show that for both regions, winner and loser species emerge despite non-significant changes in species richness over time.
 There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis, but to progress this field data must be made feely accessible to researchers.
@@
 However, these programs require significant time to generate long-term datasets and cannot be used to assess historic pollinator populations.
-Finally, the most practical approach for assessing long-term historical pollinator population trends is to use the historical information on species occurrences, which is archived in museum collections [e.g., @Bartomeus2013a].
-
-In this review, we first assess current evidence for pollinator richness declines and present a road map outlining a strategy for using museum collection data to fill current knowledge gaps.
-We highlight the major technical difficulties involved in using museum collection data and demonstrate several approaches for analysing different types of museum collection data to assess long-term trends in pollinator populations.
+Finally, the most practical approach for assessing long-term historical pollinator population trends is to use the historical information on species occurrences, which is often archived in museum collections [e.g., @Bartomeus2013a].
+
+In this review, we first assess current evidence for pollinator richness declines and present a road map outlining a strategy for using historical collection data to fill current knowledge gaps.
+We highlight the major technical difficulties involved in using historical collection data and demonstrate several approaches for analysing different types of historical collection data to assess long-term trends in pollinator populations.
 Finally, we highlight the need to move beyond simple biological diversity descriptors and unleash the power of historical data to assess changes in species interactions, ecosystem functioning and evolutionary changes through time.
@@
 
-#Using historical museum specimen records to fill knowledge gaps
-
-As shown in Figure 1, specific estimates are lacking for countries worldwide and use of historical museum collection data may be the most effective tool for filling those gaps.
+#Using historical collection specimen records to fill knowledge gaps
+
+As shown in Figure 1, specific estimates are lacking for countries worldwide and use of historical  collection data deposited in museums may be the most effective tool for filling those gaps.
 
@@
 Once museum collection datasets are made available, researchers must identify any potential biases in the data.
-We explore this process with two contrasting dataset examples (Spain and New Zealand).
+[Talk about the caveats of teh above crude analysis!]() We explore this process with two contrasting dataset examples (Spain and New Zealand).
 In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people.

### 2018-04-06 ms.Rmd
altered richness change maps and editied writing

@@
 ---
-title: Historical collections as baseline for assessing the extent of the pollinator
-  crisis
-author: "Bartomeus^1^, I., Stavert^2^, J.R., Ward^2,3^, D., and Aguado^4^, O."
+title: Museum collections as baseline for assessing the global pollinator crisis
+author: "Bartomeus, I.^1^, Stavert, J.R.^2^, Ward, D.^2,3^, and Aguado, O.^4^"
 csl: data/philb_style.xml
@@
 However, despite reports that pollinator declines are widespread, data are scarce and often geographically and taxonomically biased.
-These biases limit robust inference about any potential “pollinator crisis”.
-Analysis of historical time series data is the most direct approach for assessing species declines, but long term monitoring programs targeting pollinators are limited.
-However, non-structured and often opportunistic historical data collections are more frequently available.
+These biases limit robust inference about any potential pollinator crisis.
+Analysis of historical time series data is the most direct approach for assessing species declines, but long-term monitoring programs targeting pollinators are limited.
+However, non-structured and often opportunistic specimen collection data are more frequently available.
 In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines.
 Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits).
-Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and pollination services.
+Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, morphology and pollination services.
 Despite recent advances, we show that information on the status and trends of most pollinators is absent.
-Furthermore, we identify lack of collection record data digitization as a key bottleneck, impeding progress in this field.
+Furthermore, we identify the lack of collection record data digitization as a key bottleneck impeding progress in this field.
 We highlight opportunities and limitations to progress assessment of pollinator declines globally.
 Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
-There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis, but to progress this field data must be made feely accessible to researchers.
+There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis.
+To progress this field, data must be made feely accessible to researchers and appropriate analytical treatment is imperative.
 
@@
 As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions.
-The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@Buchmann1996].
-However, early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time.
-These initial claims triggered the first efforts to assess the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines.
+The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “The forgotten pollinators” [@Buchmann1996].
+However, early accounts of pollinator declines were somewhat anecdotal, given the lack of pollinator population data at that time.
+These initial claims triggered the first efforts to assess this potential issue and included the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines.
 The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@NationalResearchCouncil2007].
@@
 
-One of the main barriers to identifying pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [@Winfree2011].
+One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and lizards [@Winfree2011].
 Additionally, many pollinators are highly mobile, short-lived and small, which makes monitoring their populations difficult.
@@
 However, bees are diverse, with more than 20,000 species currently described worldwide [@Archer2014], and often require expert taxonomists for identification.
-Furthermore, the uneven distribution of researchers [@Martin2012] has resulted in geographical biases in bee decline research, as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
-
-One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients.
-Despite critiques on the robustness of this approach [@Franca2016; @DePalma2018], these studies currently provide the most extensive source of data on pollinator populations.
-For example, researchers have recently estimated bee richness declines for every country in Europe using predictions from models of pollination associations with different land-use types [@DePalma2017].
+Furthermore, the uneven distribution of researchers has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
+
+One solution to overcoming these barriers is the use of space-for-time substitutions, where researchers compare pollinator populations across environmental gradients.
+Despite critiques on the robustness of this approach [@Franca2016; @DePalma2018], these studies currently provide the most extensive source of pollinator population data.
+For example, researchers have recently estimated bee richness declines for every country in Europe using predictions from models of pollinator associations with different land-use types [@DePalma2017].
 A second important method is the use of data collected from pollinator monitoring programs, which are often driven by citizen scientists.
@@
 However, these programs require significant time to generate long-term datasets and cannot be used to assess historic pollinator populations.
-Finally, the most practical approach for assessing long-term historical pollinator population trends is to use the historical information on species occurrences, which is often archived in museum collections [e.g., @Bartomeus2013a].
-
-In this review, we first assess current evidence for pollinator richness declines and present a road map outlining a strategy for using historical collection data to fill current knowledge gaps.
-We highlight the major technical difficulties involved in using historical collection data and demonstrate several approaches for analysing different types of historical collection data to assess long-term trends in pollinator populations.
+Finally, the most robust approach for assessing long-term historical pollinator population trends is to use historical information on species occurrences, which is often archived in museum collections [e.g., @Bartomeus2013a].
+
+In this review, we first assess current evidence for pollinator richness declines and present a roadmap outlining a strategy for using historical collection data to fill current knowledge gaps.
+We highlight the major technical difficulties involved in using historical collection data and demonstrate several approaches for analysing different types of collection data to assess long-term pollinator population trends.
 Finally, we highlight the need to move beyond simple biological diversity descriptors and unleash the power of historical data to assess changes in species interactions, ecosystem functioning and evolutionary changes through time.
@@
 For example, in the UK and the Netherlands, a citizen science based study using both observations and museum collection data detected strong richness declines for bees, hoverflies and flowering plants [@Biesmeijer2006].
-However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus data is lacking for natural areas.
+However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for natural areas.
 Further exploration of this dataset revealed that for declining pollinator taxa, the trend has attenuated in recent decades [@Carvalheiro2013].
@@
 This is a pattern also detected in the UK [@Ollerton2014].
-However, the reginal US findings above are in stark contrast with the widespread local extinctions reported by Burkle [-@Burkle2013], who compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants.
-It is important to note that Burkle [-@Burkle2013] focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (@Collado2018).
+However, these regional US findings are in stark contrast with the widespread local extinctions reported by [@Burkle2013], who compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants.
+It is important to note that [@Burkle2013] focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (@Collado2018).
 In any case, there is strong concordance between local extinctions and regional declines [@Bartomeus2013c], suggesting that local extinctions are indicators of regional population declines.
@@
 In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a].
-
 For butterflies, the only evidence of richness declines comes from Europe.
-For example, butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although declines in Great Britain have been less severe [@Carvalheiro2013].
-In Belgium, another study [@Maes2001] found that richness declines have been severe (approximately 30%), although this study assessed richness changes over a longer time period (early 1900’s to 2000) compared to Carvalheiro [-@Carvalheiro2013].
-In parts of Germany, declines in local butterfly richness of up to 70% have been reported [@Wenzel2006].
-
-There are very few studies on hoverfly species richness changes, those of which are restricted to Europe.
+Butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although declines in Great Britain have been less severe [@Carvalheiro2013].
+In Belgium, another study [@Maes2001] found that richness declines have been severe (approximately 30%), although this study assessed richness changes over a longer time period (early 1900’s to 2000) compared to [@Carvalheiro2013] (1950-69 vs.
+1970-80 and 1970-89 vs.
+1990-2009).
+In parts of Germany, up to 70% declines in local butterfly richness have been reported [@Wenzel2006].
+Compared with other insect pollinator taxa, there are very few studies on hoverfly species richness changes, which are all restricted to Europe.
 In Belgium, Great Britain and the Netherlands, hoverfly richness changes have been modest [@Carvalheiro2013].
 In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer2006].
-Furthermore, directionality (richness increases or decreases) varies depending on the time period assessed.
+Furthermore, directionality (richness increase or decrease) varies depending on the time period assessed.
 For example, hoverfly richness decreased in Belgium by approximately 6% from 1950-69 to 1970-80, but increased by approximately 10% between 1970-89 and 1990-2009 [@Carvalheiro2013].
 
-Outside of Europe and the US and for non-insect taxa, there are no studies on pollinator declines using historical records.
-In Figure 1, we show a comparison of bee species richness per area (a), bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes (d) to highlight the little coverage that we have globally.
-See raw data at Sup Mat 1.
+Figure one shows (a) bee species richness weighted by country area; (b) bee species richness changes; (c) syrphid species richness changes; and (d) butterfly species richness changes overtime from known studies, to highlight the little coverage that currently exists globally.
+See raw data (Sup Mat 1).
 For other parts of the world, there are only species-specific examples of historical losses (e.g., Bombus dalbhomi; @Morales2013) [South Africa: @Pauw, maybe with plants… .
@@
 Moths?
-Butterflies!]()
+Butterflies!]() Outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records.
 
@@
 #read in data
-declines <- read.csv(file = "data/richness_declines.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+declines <- read.csv(file = "data/richness_declines.V2.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
 #nothing outside EU- USA. Spain and NZ can be added.
@@
 declines.syrphid <- as.data.frame(filter(declines.ave, taxa == "syrphid")) %>% droplevels()
-
-map.syrphid <- joinCountryData2Map(declines.syrphid, joinCode="NAME", nameJoinColumn = "country",
+declines.syrphid.2 <- declines.syrphid %>% 
+    group_by(country) %>% 
+    summarise(average = mean(average),
+              median = median(median),
+              max = max(max),
+              min = min(min))
+
+map.syrphid <- joinCountryData2Map(declines.syrphid.2, joinCode="NAME", nameJoinColumn = "country",
                             nameCountryColumn = "distribution")
@@
 declines.bee <- as.data.frame(filter(declines.ave, taxa == "bee")) %>% droplevels()
-
-map.bee <- joinCountryData2Map(declines.bee, joinCode="NAME", nameJoinColumn = "country",
+declines.bee.2 <- declines.bee %>% 
+    group_by(country) %>% 
+    summarise(average = mean(average),
+              median = median(median),
+              max = max(max),
+              min = min(min))
+
+map.bee <- joinCountryData2Map(declines.bee.2, joinCode="NAME", nameJoinColumn = "country",
                                    nameCountryColumn = "distribution")
@@
 declines.butterfly <- as.data.frame(filter(declines.ave, taxa == "butterfly")) %>% droplevels()
-
-map.butterfly <- joinCountryData2Map(declines.butterfly, joinCode="NAME", nameJoinColumn = "country",
+declines.butterfly.2 <- declines.butterfly %>% 
+    group_by(country) %>% 
+    summarise(average = mean(average),
+              median = median(median),
+              max = max(max),
+              min = min(min))
+
+map.butterfly <- joinCountryData2Map(declines.butterfly.2, joinCode="NAME", nameJoinColumn = "country",
                                      nameCountryColumn = "distribution")
@@
 
-Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness and country size, which accounts for the species-area relationship; Warmer colours indicate higher bee diversity; Note that some African countries may have incomplete listed faunas and that Alaska is included with US values.
-(b) Syrphid (c) bee and butterfly (d) richness percent change in the last century.
+Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness per country and country size.
+This accounts for the species-area relationship.
+Warmer colours indicate higher bee diversity.
+Note that some African countries may have incomplete listed faunas and that Alaska is included with US values.
+Changes in (b) syrphid (c) bee and (d) butterfly richness within the last 100 years.
 Warmer colours indicate steeper declines.
@@
 
-As shown in Figure 1, specific estimates are lacking for countries worldwide and use of historical  collection data deposited in museums may be the most effective tool for filling those gaps.
-
+Estimates of pollinator declines are lacking for most countries worldwide (Figure 1).
+The use of historical museum collection data may be the most effective tool for filling these gaps.
 The core aim of museums is to conserve and curate collections.
@@
 
-Where digitization has been completed, the data provide a rich source of information, allowing assessment of the current status and long-term trends in pollinator populations [e.g., @Bartomeus2013a; @Scheper2014; @Young2017].
+Where digitization has been completed, the data provide a rich source of information, allowing assessment of the current status and long-term trends of pollinator populations [e.g., @Bartomeus2013a; @Scheper2014; @Young2017].
 This is despite the fact that museum collections often have a number of biases, including unknown sampling effort, personal interests of collectors and the curatorial techniques used.
 For example, collectors tend to target rare or unusual over common taxa, discard damaged individuals or only accession a certain number of individuals.
-In addition, collections are often made opportunistically, leading to a spatial bias where difficult to access areas are under-sampled or conversely, where samples are biased towards easily accessed locations (e.g., towns/cities and/or along roadsides).
-Furthermore, this approach can only determine where species are present and not where they are absent.
-However, given adequate sample sizes and appropriate statistical analyses, most of these biases can be accounted for [e.g., @Bartomeus2013a; @Keil2011; @Pearce2006].
+In addition, collections are often made opportunistically, leading to a spatial biases where difficult to access areas are under-sampled or conversely, where samples are biased towards easily accessed locations (e.g., towns/cities and/or roadsides).
+Further, museum collection data can only be used to determine where species are present and not where they are absent.
+However, given adequate sample sizes and appropriate statistical techniques, most biases can be accounted for [e.g., @Bartomeus2013a; @Keil2011; @Pearce2006].
 
@@
 As we have shown, there is a paucity of countries for which historical data is available (Figure 1), and hence can be used as baseline for assessing pollinator population declines.
-While ideally one would aim to digitize all museum collection records, this is unlikely in the near future, mainly due to funding constraints.
-Here we show how researchers can optimize their use of museum collection data to assess long-term pollinator population changes.
+While ideally one would aim to digitize all museum collection records, this is unlikely in the near future, predominantly due to funding constraints.
+Here we show how researchers can optimize the use of museum collection data to assess long-term pollinator population changes.
 
 GBIF (https://www.gbif.org/) is a central repository for global species occurrence data.
-Much of these data come from museums, private collections and government research institutes, but it integrates several other sources.
-In combination with the popular statistical language R [@RCoreTeam2017], GBIF can be directly queried into your computer [@Chamberlain2014] and data availability checked for your region of interest.
-In Figure 2 we show the number of modern and historic bee records currently available for different countries.
-While values are high on both axes for countries such as the UK and the US so that data can potentially be analysed without further data collection effort, most countries fall short in one or both axes.
-Hence, for countries with high numbers of recent records, researchers should prioritize the digitalization of old material before embarking on data analyses (e.g.
-Switcherland,...
+Much of these data come from museums, private collections and government research institutes, but several other sources are also integrated.
+In combination with the popular statistical language R [@RCoreTeam2017], GBIF can be directly queried into your computer [@Chamberlain2014] and data availability can be checked for the region of interest.
+We show the number of modern and historic bee records currently available for different countries (Figure 2).
+Although values are high on both axes for countries such as the UK and the US (these data can potentially be analysed without further data collection effort), most countries fall short in one or both axes.
+Hence, for countries with many recent records, researchers should prioritize the digitalization of old material before embarking on data analyses (e.g.
+Switzerland,...
 ).
@@
 Interestingly, we show that greater than 100 [check]() countries have less than 1,000 records, making them poor candidates for analysing long-term pollinator population trends.
-However, countries like X or X… It is also important to note that historical records are not always vouchered in local museums, but many European and US museums contain large collections of pollinators collected elsewhere.
-Aside from bees, similar exploratory analysis can easily be done for other taxa.
+However, countries like X or X… It is also important to note that historical records are not always vouchered in local museums (i.e., many European and US museums contain large collections of pollinators from other countries).
+Aside from bees, similar exploratory analyses can easily be conducted for other taxa.
+[expand on this?]
 
@@
 Figure 2.
-The number of bee occurrences before 1980 and after 1980 in GBIF.
-The upper right quadrat contains well covered countries and the upper left quadrat contains countries that resurveys of old sites.
+a) The number of bee occurrences before 1980 and after 1980 in GBIF.
+The upper right quadrat contains well covered countries and the upper left quadrat contains countries that require resurveys of old sites.
 The lower right quadrat contains countries that require museum record digitization and the lower left quadrat contains countries with very little available data.
-We highlight a few representative examples.
-b) boxplot showing rarefied number of species collected before and after 1980.
-c) same data plotted in a map.
-This data is likely to contain strong undetected biases, but indicates little change...
-
-Once museum collection datasets are made available, researchers must identify any potential biases in the data.
-[Talk about the caveats of teh above crude analysis!]() We explore this process with two contrasting dataset examples (Spain and New Zealand).
+We highlight a few representative examples; b) a boxplot showing the rarefied number of species collected before and after 1980 and; c) the same data plotted on a global map.
+This data are likely to contain strong undetected biases, but indicate little change...
+
+Once museum collection datasets are made available, researchers must identify any potential biases.
+[Talk about the caveats of the above crude analysis!]() We explore this process with two contrasting dataset examples (Spain and New Zealand).
 In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people.
 In this case, we contacted the original collectors to define their sampling protocols.
-We then resurveyed the same sites (20 years after the original surveys) using the same protocols.
+We then resurveyed the same sites (20 years after the original surveys) using the same sampling protocols.
 In contrast, the New Zealand dataset includes a wide suite of collectors and collection locations but shows no obvious biases in geographical and taxonomic coverage through time.
@@
 These different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories.
-We hope this resource will encourage researchers to analyse data for regions where there is current information on pollinator declines is lacking.
+We hope this resource will encourage researchers to analyse data for regions where current information on pollinator declines is lacking.
 
@@
 In addition, Spain is a bee diversity hotspot (Figure 1) and maintains a relatively heterogeneous landscape.
-Spain has already digitalized a large amount of pollinator occurrence data for both historical and recent periods (Figure 2).
-However, visual inspection of the data reveals clustering around a few localities.
-Further, historic records do not spatially match recent records, making comparisons difficult.
-For this dataset, most of the historical records are located around Valladolid and were collected by Enrique Asensio and his collaborators.
+Spain has already digitalized a large amount of pollinator occurrence data for both historic and recent periods (Figure 2).
+However, visual inspection of the data revealed clustering around a few localities.
+Further, historic records did not spatially match recent records, making comparisons difficult.
+For this dataset, most of the historic records were located around Valladolid and were collected by Enrique Asensio and collaborators.
 There has been no recent sampling of bees in this area.
 However, we found that Enrique systematically sampled seven locations and that additional data were available at the “Museo de Historia Nacional” and other minor collections.
-Digitization of these records, along with a re-survey of these locations provided an excellent dataset for a before and after comparison of bee communities at these specific locations.
-Interestingly, this localities have experienced all large agricultural intensification [expand]()
-
-In brief, we cleaned taxonomical names using taxize [@chamberlain], checked sampling completeness for both time periods and compared species richness for each site before and after 1980 with a paired t-test.
-We found that there is a trend to find a reduced number of species per site (P = 0.058).
-However, this is higlhly dependant on the site as 2 out of 3 sites show no richness declines.
-However, there are clear groups of species suffereing larger declines.
-Andrenids and its parasites being among the losers and Halictids among the winners.
+Digitization of these records, along with a re-survey of the original sampling locations provided an excellent dataset for a before and after comparison of bee communities.
+Interestingly, these localities have recently experienced substantial agricultural intensification [expand]()
+
+In brief, we cleaned taxonomic names using the taxize package [@Chamberlain2018], checked sampling completeness for both time periods and compared species richness for each site before and after 1980 with a paired t-test.
+We found that there were a reduced number of species at sites after 1980 (P = 0.058).
+However, this trend was highly dependent on site identity, as two out of three sites showed no richness declines.
+However, we found clear patterns in declines among species groups.
+For example, Andrenids and their parasites showed the strongest declines whereas Halictids increased.
 This is in accordance with findings elsewhere [e.g.
-@Bartomeusetal2013a] (appendix?).
+@Bartomeus2013a] (appendix?).
 
@@
 
-Figure 3. Rarefied richness at the 6 plots resurveyed.
+Figure 3. Rarefied richness at the six Spanish plots resurveyed.
 
@@
 Much of New Zealand’s pollinator fauna is also relatively depauperate.
-For example, New Zealand has only 27 species of native bees [@Donovan2007], which is a fraction of nearby Australia’s c.
-1600 species [#@ref].
-However, New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems [@Newstrom2005].
+For example, New Zealand has only 27 native bee species of [@Donovan2007], which is a fraction of nearby Australia’s c.
+1600 species [@Ascher2015].
+However, New Zealand has a surprisingly high diversity of flies (Diptera), which are important pollinators in many ecosystems [@Newstrom2005].
 Thus, New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.
@@
 Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 [@Perry2014].
-Therefore, human activity likely affected pollinator communities in New Zealand long before surveys and collections began.
+Therefore, human activity likely affected pollinator communities in New Zealand long before surveys and specimen collections began.
 Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.
@@
 Collection records from the New Zealand Arthropod Collection (NZAC) are freely available online (https://scd.landcareresearch.co.nz/).
-Fly pollinator data was obtained from six participating New Zealand museums [need to list in the acknowledgments]() and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators.
+Fly pollinator data was obtained from three participating New Zealand museums and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators.
 Collections for the bee and fly datasets span over 100 years (early 1900s to late 2000s).
 
-We followed the protcols outlined in [@Bartomeus2013a] to analysze these data.
+We followed protocols outlined in [@Bartomeus2013a] to analyse the New Zealand data.
 First, we filtered our original datasets so that data used for analyses only included independent collection events.
@@
 We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980.
+Further exploration of the New Zealand native bee data raised doubts on collection completeness in records prior to 1970, so we removed these records from further analyses.
 We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years.
@@
 Further, we only included species in this analysis for which we had 30 or more records.
-To account for differences in sampling effort between years, we weighted each year by the total number of samples collected in that year.
-
-We found that rarefied richness for both native and exotic bees increased overtime, but these trends were non-significant (P-values for both groups > 0.05).
+To account for differences in sampling effort between years, we weighted each year by the total number of samples collected that year.
+
+We found that rarefied richness for native bees was stable through time.
+Exotic bees showed an increase in rarefied richness, but this trend was non-significant (P-value for both natives and exotic bees > 0.05).
 In contrast, native fly richness declined, whereas exotic fly richness increased, although results for these groups were also non-significant (P-values for both groups > 0.05).
-Note that rarefied richness is sensitive to species evenness, so increases in rarefied richness over time may actually indicate increased species evenness.
+Note that rarefied richness is sensitive to species evenness, so increases in rarefied richness over time may actually indicate increased species evenness and vice-versa for decreased richness.
 
@@
 #length(which(native$year < 1900))
-native <- subset(native, year > 1800)
-for (i in 5){
+native <- subset(native, year > 1969)
+for (i in 10){
     break1 <- quantile(native$year, probs = seq(0,1,1/i), na.rm = TRUE)
-    #x <- c(1909,1965,1970,1974,1976,1978,1981,1986,1994,2012)
+    #x <- c(1909,1970,1971,1974,1976,1978,1981,1986,1994,2012)
     lab <- c(1:i)
@@
     comm2 <- table(native$time_period_quant, factor(native$Gen_sp))
-    (x2 <- rarefy(comm2,150, se = TRUE)) #same result as my resampling...
+    (x2 <- rarefy(comm2,88, se = TRUE)) #same result as my resampling...
     #rar[[i-2+10]] <- x2
@@
 #length(which(exotic$year < 1900))
-exotic <- subset(exotic, year > 1900)
-for (i in 3:6){ #error saying break are not unique - how to fix?
-    break1 <- quantile(exotic$year, probs = seq(0,1,1/i), na.rm = TRUE)
+exotic <- subset(exotic, year > 1800)
+for (i in 9){ #error saying break are not unique - how to fix?
+    break1 <- quantile(exotic$year, probs = seq(0,1,1/6), na.rm = TRUE)
+    x <- c(1916,1952,1953,1969,1971,1975,1978,1983,1991,2011)
     lab <- c(1:i)
     for(j in 1:i){
-        lab[j] <- paste(as.numeric(break1)[j],as.numeric(break1)[j+1], sep = "-")
-    }
-    exotic$time_period_quant  <- cut(exotic$year, breaks = break1, labels = lab)
+        lab[j] <- paste(as.numeric(x)[j],as.numeric(x)[j+1], sep = "-")
+    }
+    exotic$time_period_quant  <- cut(exotic$year, breaks = x, labels = lab)
     comm2 <- table(exotic$time_period_quant, factor(exotic$Gen_sp))
@@
     for(j in 1:i){
-        time[j] <- (as.numeric(break1)[j] + as.numeric(break1)[j+1])/2
+        time[j] <- (as.numeric(x)[j] + as.numeric(x)[j+1])/2
     }
@@
 
-However, at the species level, we found that 10 out of 27 bee species increased in relative abundance over time (nine native and one exotic) and three bee species declined in relative abundance (one native and two exotic) (Figure 3).
-Interestingly, the two exotic bee species that declined in relative abundance were both in the genus Bombus, intentionally introduced into New Zealand for the pollination of crops.
-Native bees that increased in relative abundance were mostly from the genus Leioproctus.
-Only two out of 14 fly species increased in relative abundance, both of which were exotic, whereas four species decreased in abundance (three native and one exotic).
-Native flies that decreased in relative abundance were all Syrphidae in the genus Helophilus.
+However, at the species level, we found that 11 out of 27 bee species increased in relative occurrence over time (10 native and one exotic) and three bee species declined in relative occurrence (one native and two exotic) (Figure 3).
+Interestingly, the two exotic bee species that declined in relative occurrence were both in the genus Bombus, which were intentionally introduced into New Zealand for the pollination of crops.
+Native bees that increased in relative occurrence were mostly from the genus Leioproctus, which are medium sized, ground nesting solitary bees.
+Only one out of 14 fly species increased in relative occurrence, which was exotic, whereas four species decreased in occurrence (three native and one exotic).
+Native flies that decreased in relative occurrence were all Syrphidae in the genus Helophilus.
 
@@
 
-Figure 5. Model estimated changes (± 1 SE) in the relative occurrence frequency for different New Zealand bee and fly species in museum collections over time.
+Figure 5. Model estimated changes (± 1 SE) in the relative occurrence frequency of different New Zealand bee and fly species in museum collections over time.
 
@@
 
-A recent study found that more than 90% of the papers studying pollinator responses to land-use change focused solely on richness and abundance descriptors [@Winfree2011].
-But in addition to local (alpha) diversity and regional (gamma) diversity, we need to assess changes in turnover between sites (beta diversity).
-Environmental changes often result in a few “winner” species and many species becoming “losers” [@Bartomeus2013b].
+A recent study found that more than 90% of the papers investigating pollinator responses to land-use change focused solely on richness and abundance descriptors [@Winfree2011].
+But in addition to local (alpha) diversity and regional (gamma) diversity, researchers need to assess changes in turnover between sites (beta diversity).
+Environmental changes often result in a few “winner” species and many “losers” species [@Bartomeus2013b].
 Identifying winners and losers is critical as the few winners are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2017].
-These changes can have important effects for pollination of native plant species and crops.
-
-But museum specimen collections can provide much more information besides species occurrence records, providing such information is captured when digitizing collections.
+These changes can have important effects for pollination of native plant species and crops [@Stavert2018].
+
+Museum specimen collections can provide much more information besides species occurrence records, given that such information is recorded when digitizing collections.
 This is particularly important for identifying mechanisms of decline and adaptation.
@@
 This information critical for understanding past and present species interactions [@Tylianakis2008].
-Regardless of this information, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [@Kleijn2008].
-Finally, museum specimens can be measured to track evolutionary changes by measuring specific specimen traits.
-This has been already explored to show tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016] shrinkage with climate and land use change.
+Aside from this information, bee specimens often contain pollen loads trapped on hairs, from which past visitation events can be identified [@Kleijn2008].
+Finally, museum specimens can be measured to track evolutionary changes by measuring the traits of specimen.
+This approach has been already used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016] in response to climate and land-use change.
 
@@
 
-Unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors.
-Using two case studies, we show that collaboration with both museum curators and ecologists is key to understanding the data and treating it appropriately.
+Unleashing the power of museum collection data to answer pressing ecological and evolutionary questions is at our hands, but requires the coordinated effort of many actors.
+Using two case studies, we show that strong collaboration with between museum curators and ecologists is key to understanding data and treating it appropriately.
 To progress our understanding of the global pollination crisis, researchers and curators must aim to digitize museum collection data and make it readily available it a format that is widely accessible.
 Centralization of regional and national museum collection data in existing global platforms, such as GBIF, would facilitate free and widespread access.
-However, datasets could also be stored in alternative webpages or database repositories (e.g., University and museum webpages or Dryad) providing they are well documented, easily retrieved and can be combined with other datasets using open science tools [e.g., @Dallas2016].
-Further, we must revolutionize the way that researchers collaborate with museums, to foster healthy bidirectional relationships.
-For example, ecological researchers collect massive numbers of specimens, but these are often inappropriately vouchered [@Packer2018; @Wheeler2003], rendering them less useful for future research.
-To improve this, strong communication between museums and researchers is required, but can only be achieved with adequate funding and recognition that accurate data recording and long-term preservation are critical for research [@Ward2015].
+However, datasets could also be stored in alternative webpages or database repositories (e.g., university and museum webpages or Dryad) providing they are thoroughly documented and easily retrieved and combined with other datasets using open science tools [e.g., @Dallas2016].
+We must revolutionize the way that researchers collaborate with museums, in order to foster healthy bidirectional relationships.
+For example, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered [@Packer2018; @Wheeler2003], rendering them less useful for future research.
+To improve this process, strong communication between museums and researchers is required.
+However, this can only be achieved with adequate funding and recognition that accurate data recording and long-term preservation are critical for research [@Ward2015].
 
@@
 We thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.
+We also thank Canterbury Museum, the New Zealand Arthropod Collection and the Museum of New Zealand Te Papa Tongarewa for providing the New Zealand fly pollinator data.
+We thank E.
+Asensio for sharing his historical data and knowledge.
 IB was funded by FBBVA project.
 DW was funded through Landcare Research within the Characterising New Zealand’s Land Biota Portfolio.
-We thank E.
-Asensio for sharing his historical data and knowledge.
 

### 2018-04-06 ms.Rmd
Main text ready

@@
 ---
-title: Museum collections as baseline for assessing the global pollinator crisis
+title: Historic collections as a tool for assessing the global pollination crisis
 author: "Bartomeus, I.^1^, Stavert, J.R.^2^, Ward, D.^2,3^, and Aguado, O.^4^"
@@
 *Affiliations:*
-1 Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, E-41092 Sevilla, Spain   
-2 Centre for Biodiversity and Biosecurity, School of Biological Sciences, The University of Auckland, Auckland, New Zealand
-3 Landcare Research, Auckland, New Zealand
-4 Andrena Soluciones  
+^1^ Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, E-41092 Sevilla, Spain   
+^2^ Centre for Biodiversity and Biosecurity, School of Biological Sciences, The University of Auckland, Auckland, New Zealand  
+^3^ Landcare Research, Auckland, New Zealand  
+^4^ Andrena Iniciativas y Estudios Medio Ambientales
 *Correspondence: nacho.bartomeus@gmail.com
@@
 These biases limit robust inference about any potential pollinator crisis.
-Analysis of historical time series data is the most direct approach for assessing species declines, but long-term monitoring programs targeting pollinators are limited.
-However, non-structured and often opportunistic specimen collection data are more frequently available.
-In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines.
-Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits).
-Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, morphology and pollination services.
-Despite recent advances, we show that information on the status and trends of most pollinators is absent.
-Furthermore, we identify the lack of collection record data digitization as a key bottleneck impeding progress in this field.
-We highlight opportunities and limitations to progress assessment of pollinator declines globally.
+Non-structured and opportunistic historical specimen collection data provide the only source of historical information which can serve as a baseline for identifying pollinator declines.
+Specimens historically collected and preserved in museums not only provide information on where and when species were collected, but also contain other ecological information such as species interactions and morphological traits.
+Here, we provide a synthesis of how researchers have used historical data to identify long-term changes in biodiversity, species abundances, morphology and pollination services.
+Despite recent advances, we show that information on the status and trends of most pollinators is absent, but we highlight opportunities and limitations to progress the assessment of pollinator declines globally.
 Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
 There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis.
-To progress this field, data must be made feely accessible to researchers and appropriate analytical treatment is imperative.
+
+*Keywords*: Museums, biodiversity, global change, bees, syrphid flies, butterflies.
 
@@
 
-One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and lizards [@Winfree2011].
+One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and lizards [@Ollerton2017].
 Additionally, many pollinators are highly mobile, short-lived and small, which makes monitoring their populations difficult.
-Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017].
-However, bees are diverse, with more than 20,000 species currently described worldwide [@Archer2014], and often require expert taxonomists for identification.
+Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Winfree2011].
+However, bees are diverse, with more than 20,000 species currently described worldwide, and often require expert taxonomists for identification.
 Furthermore, the uneven distribution of researchers has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
@@
 For example, in the UK and the Netherlands, a citizen science based study using both observations and museum collection data detected strong richness declines for bees, hoverflies and flowering plants [@Biesmeijer2006].
-However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for natural areas.
+In the Netherlands, museum data have also revealed simultaneous plant and pollinator declines [@Scheper2014].
+Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction.
+However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for well-preserved natural areas.
 Further exploration of this dataset revealed that for declining pollinator taxa, the trend has attenuated in recent decades [@Carvalheiro2013].
 
-In the Netherlands, museum data have revealed simultaneous plant and pollinator declines [@Scheper2014].
-Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction.
 Although studies of local pollinator communities often detect richness declines, regional richness may remain relatively stable.
-For example, regional estimates for species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012].
-This is a pattern also detected in the UK [@Ollerton2014].
-However, these regional US findings are in stark contrast with the widespread local extinctions reported by [@Burkle2013], who compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants.
-It is important to note that [@Burkle2013] focused on habitat remnants and did not take into account opportunistic bee species that use novel land-use types (@Collado2018).
-In any case, there is strong concordance between local extinctions and regional declines [@Bartomeus2013c], suggesting that local extinctions are indicators of regional population declines.
-
-As with bees, studies on species richness changes for other pollinator taxa are both scarce and geographically restricted.
-Generally speaking, reported declines for bumblebees are the most severe of all pollinator taxa.
+For example, regional estimates for bee species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012].
+This is a pattern also detected in the UK, where relatively few regional bee extinctions has been reported [@Ollerton2014].
+These regional findings are in stark contrast with the widespread local extinctions reported in local studies.
+For example, Burkle et al.
+[@Burkle2013] compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants and reports several local extinctions.
+However, it is important to note that there is strong concordance between local extinctions and regional declines [@Bartomeus2013c], suggesting that local extinctions are indicators of regional population declines.
+
+Reported declines for bumblebees are the most severe of all pollinator taxa.
 For example, declines of up to 18% in local bumblebee richness have been reported for Belgium and the Netherlands [@Carvalheiro2013].
@@
 In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a].
+However, studies on species richness changes for other pollinator taxa are both scarce and geographically restricted.
 For butterflies, the only evidence of richness declines comes from Europe.
@@
 
-Figure one shows (a) bee species richness weighted by country area; (b) bee species richness changes; (c) syrphid species richness changes; and (d) butterfly species richness changes overtime from known studies, to highlight the little coverage that currently exists globally.
-See raw data (Sup Mat 1).
-For other parts of the world, there are only species-specific examples of historical losses (e.g., Bombus dalbhomi; @Morales2013) [South Africa: @Pauw, maybe with plants… .
-However, Peter Keil flies?
-Moths?
-Butterflies!]() Outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records.
+For ilustrative purposes, we mapped the findings of this studies in Figure 1 to show the strong contrast between bee species richness worldwide, whith bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1).
+Despite outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records, there are species-specific examples of historical losses from different parts of the world (e.g., Bombus dalbhomi; @Morales2013).
 
@@
 #read in data
-declines <- read.csv(file = "data/richness_declines.V2.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+declines <- read.csv(file = "data/richness_declines.V2.csv", 
+                     header = TRUE, sep = ",", na.strings=c("","NA"))
 #nothing outside EU- USA. Spain and NZ can be added.
+#head(declines)
 
@@
 
-####################################################
+#make bee map
+declines.bee <- as.data.frame(filter(declines.ave, taxa == "bee")) %>% droplevels()
+declines.bee.2 <- declines.bee %>% 
+    group_by(country) %>% 
+    summarise(average = mean(average),
+              median = median(median),
+              max = max(max),
+              min = min(min))
+
+map.bee <- joinCountryData2Map(declines.bee.2, joinCode="NAME", nameJoinColumn = "country",
+                                   nameCountryColumn = "distribution")
+
+#average richness change
+map.bee.ave <- mapCountryData(map.bee, nameColumnToPlot = "average", colourPalette="topo",
+                                  mapTitle = "Average bee richness change", catMethod = "pretty", addLegend = F)
+do.call(addMapLegend, c(map.bee.ave, legendLabels="all", legendWidth=0.5))
+
 #make syrphid map
-####################################################
-
 declines.syrphid <- as.data.frame(filter(declines.ave, taxa == "syrphid")) %>% droplevels()
@@
 
-####################################################
-#make bee map
-####################################################
-
-declines.bee <- as.data.frame(filter(declines.ave, taxa == "bee")) %>% droplevels()
-declines.bee.2 <- declines.bee %>% 
-    group_by(country) %>% 
-    summarise(average = mean(average),
-              median = median(median),
-              max = max(max),
-              min = min(min))
-
-map.bee <- joinCountryData2Map(declines.bee.2, joinCode="NAME", nameJoinColumn = "country",
-                                   nameCountryColumn = "distribution")
-
-#average richness change
-map.bee.ave <- mapCountryData(map.bee, nameColumnToPlot = "average", colourPalette="topo",
-                                  mapTitle = "Average bee richness change", catMethod = "pretty", addLegend = F)
-do.call(addMapLegend, c(map.bee.ave, legendLabels="all", legendWidth=0.5))
-
-####################################################
 #make butterfly map
-####################################################
-
 declines.butterfly <- as.data.frame(filter(declines.ave, taxa == "butterfly")) %>% droplevels()
@@
 Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness per country and country size.
-This accounts for the species-area relationship.
+This correction accounts for the species-area relationship.
 Warmer colours indicate higher bee diversity.
 Note that some African countries may have incomplete listed faunas and that Alaska is included with US values.
-Changes in (b) syrphid (c) bee and (d) butterfly richness within the last 100 years.
-Warmer colours indicate steeper declines.
+Countries with available historical changes in (b) bee, (c) syrphid and (d) butterfly richness within the last 100 years.
+Warmer colours indicate steeper average declines.
 Countries without data are coloured in white.
@@
 Estimates of pollinator declines are lacking for most countries worldwide (Figure 1).
-The use of historical museum collection data may be the most effective tool for filling these gaps.
-The core aim of museums is to conserve and curate collections.
+The use of historic  collection data may be the most effective tool for filling these gaps.
+The core aim of museums is to conserve and curate historic collections.
 Thus, they serve as a precious repository for specimens, and at the same time, often ensure higher quality taxonomic identification.
 Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization.
-Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@Meier2004; @Huber1998; @Thomson2018], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@Hart2016].
+Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@Meier2004; @Huber1998; @Thomson2018], (ii) geo-locate the coordinates of collection events (e.g.
+http://www.geonames.org) and (iii) store the data in a properly curated database [@Hart2016].
 Undertaking this process for tens or hundreds of thousands of museum collection specimens can be a daunting task and requires specialized personnel.
@@
 High resolution photos of specimens and associated labels can be uploaded to the internet, where the task of image transcription can be distributed across hundreds or thousands of volunteers (e.g., https://www.zooniverse.org/).
-In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage].
+In addition, new algorithms have been created that allow location geo-referencing based on vernacular names (e.g.
+https://geoparser.io).
 However, achieving this requires adequate funding [@Wiggins1991].
@@
 While ideally one would aim to digitize all museum collection records, this is unlikely in the near future, predominantly due to funding constraints.
-Here we show how researchers can optimize the use of museum collection data to assess long-term pollinator population changes.
+Here we show how researchers can optimize the use of historical collection data to assess long-term pollinator population changes.
 
@@
 In combination with the popular statistical language R [@RCoreTeam2017], GBIF can be directly queried into your computer [@Chamberlain2014] and data availability can be checked for the region of interest.
-We show the number of modern and historic bee records currently available for different countries (Figure 2).
-Although values are high on both axes for countries such as the UK and the US (these data can potentially be analysed without further data collection effort), most countries fall short in one or both axes.
-Hence, for countries with many recent records, researchers should prioritize the digitalization of old material before embarking on data analyses (e.g.
-Switzerland,...
-).
-In contrast, for countries with large numbers of historic records, re-surveys are required before analysis [e.g., @Cameron2011].
-Interestingly, we show that greater than 100 [check]() countries have less than 1,000 records, making them poor candidates for analysing long-term pollinator population trends.
-However, countries like X or X… It is also important to note that historical records are not always vouchered in local museums (i.e., many European and US museums contain large collections of pollinators from other countries).
+Focusing on bee taxa, we show here the number of modern and historic bee records currently available for different countries (Figure 2a, Sup mat 2).
+Thirty-seven countries count with more than 1800 records in each time period, making these data potentially analyzable without further data collection effort (see Figure 2b and c for an initial exploration).
+However, a proper analysis of this dataset would need a careful look at the data as we detail later on with two examples (Spain and New Zealand).
+Most countries fall short in one or both axes.
+For example, a variety of countries located in different continents such as Switzerland, Sri Lanka, Nicaragua or Zimbawe have a decent number of recent records, but lack historical collections and researchers should prioritize the digitalization of old material before embarking on data analyses.
+It is also important to note that historical records are not always vouchered in local museums (i.e., many European and US museums contain large collections of pollinators from other countries).
+In contrast, more than 192 countries have less than 1000 records for both time periods, making them poor candidates for analysing long-term pollinator population trends.
 Aside from bees, similar exploratory analyses can easily be conducted for other taxa.
-[expand on this?]
 
@@
 lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
+write.csv(lhf, "data/SpMat2.csv")
 #plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
@@
 #table?
-lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) < 7.5),]
+#lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) < 7.5),]
+#lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),]
+#nrow(lhf[which(log(lhf$post1980) < 6.9 & log(lhf$pre1980) < 6.9),])
 # poland switcherland need old records! > 10000 vs 300
@@
 # Colombia & Panama, zimbawe > 3000 vs >1500
-nrow(lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),])
-lhf[which(lhf$country == "ES" | lhf$country == "NZ"),]
+#lhf[which(lhf$country == "ES" | lhf$country == "NZ"),]
 col <- ifelse(lhf$country == "ES" | lhf$country == "NZ", "red", "black")
 col <- ifelse(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5, col, "grey")
-plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col)
+pchs <- ifelse(lhf$country == "ES" | lhf$country == "NZ", 19, 1)
+plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col, pch = pchs)
 temp <- subset(lhf, country %in% c("ES", "NZ"))
-text(x = temp$log_post1980+1, y = temp$log_pre1980+1 , labels = temp$country)
+text(x = temp$log_post1980, y = temp$log_pre1980+2 , labels = temp$country, col = "red")
 #36 countries!
@@
     dat <- occ_list[[i]]
-    nrow(dat) 
-    dat1 <- dat[-which(is.na(dat$name)),]
-    nrow(dat1) # 
+    #nrow(dat) 
+    dat2 <- dat[-which(is.na(dat$name)),]
+    dat3 <- dat2[which(is.na(as.numeric(dat2$name))),]
+    dat4 <- subset(dat3, name != "Apis mellifera")
+    #nrow(dat1) 
+    dat1 <- dat4
+    #unique(dat1$name)
     dat1$period <- ifelse(dat1$year < 1980, "pre1980", "post1980")
-    table(dat1$period) # :(
+    #table(dat1$period) # :(
 #4) rar richness
@@
 data <- data.frame(countries = as.character(countries), pre1980 = pre1980, post1980 = post1980, n = n)
-head(data)
+#head(data)
 #maybe delete n < 500?
-data2 <- subset(data, n > 500)
-
-t.test(data2$pre1980, data2$post1980, paired = TRUE)
-
-boxplot(data2$pre1980, data2$post1980, names = c("pre1980", "post1980"), ylab = "rarefied species richness", 
+data2 <- subset(data, n > 500) #28 countries with enought data.
+
+#t.test(data2$pre1980, data2$post1980, paired = TRUE)
+
+boxplot(log(data2$pre1980), log(data2$post1980), names = c("pre1980", "post1980"), ylab = "log rarefied species richness", 
             las = 1, border = "light grey")
-points(x = c(rep(1,nrow(data2)), rep (2,nrow(data2))), y = c(data2$pre1980, data2$post1980), pch = 19)
+points(x = c(rep(1,nrow(data2)), rep (2,nrow(data2))), y = c(log(data2$pre1980), log(data2$post1980)), pch = 1)
 for(i in 1:nrow(data2)){
-    lines(x = 1:2, y = c(data2$pre1980[i], data2$post1980[i]))
+    lines(x = 1:2, y = c(log(data2$pre1980[i]), log(data2$post1980[i])), 
+          col = ifelse((data2$pre1980[i] - data2$post1980[i]) > 1, "red", "blue"))
 }
@@
 #average richness change
-map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette="topo",
-                       mapTitle = "Average bee richness change", catMethod = "pretty", addLegend = F)
+colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
+colorss <- (heat.colors(10))
+map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette= colorss,
+                       mapTitle = "Average bee richness change (%)", catMethod = "pretty", addLegend = F)
 do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))
@@
 Figure 2.
-a) The number of bee occurrences before 1980 and after 1980 in GBIF.
-The upper right quadrat contains well covered countries and the upper left quadrat contains countries that require resurveys of old sites.
-The lower right quadrat contains countries that require museum record digitization and the lower left quadrat contains countries with very little available data.
-We highlight a few representative examples; b) a boxplot showing the rarefied number of species collected before and after 1980 and; c) the same data plotted on a global map.
-This data are likely to contain strong undetected biases, but indicate little change...
-
-Once museum collection datasets are made available, researchers must identify any potential biases.
-[Talk about the caveats of the above crude analysis!]() We explore this process with two contrasting dataset examples (Spain and New Zealand).
-In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people.
-In this case, we contacted the original collectors to define their sampling protocols.
-We then resurveyed the same sites (20 years after the original surveys) using the same sampling protocols.
+Exploration of available data for bee records showing a) The number of bee occurrences before 1980 and after 1980 in GBIF for each country.
+The upper right quadrat (records in black) contains well covered countries with New Zealand (NZ) and Spain (ES) marked in red.
+For those countries, we show a temptaive comparision of the rarefied number of species in both time periods and show that for most countries the number of species recorded is similar in both time periods (b).
+A more careful analysis of this data would help complete the map of global declines (c).
+Note that this data are likely to contain strong undetected biases, as we explore below.
+
+As stated above, once historical collection datasets are made available, researchers must identify any potential biases.
+We explore this process with two contrasting dataset examples (Spain and New Zealand).
+In the Spanish dataset, most of the data comes from a few specific locations and was collected by a few specific teams.
+Hence the geographical coverage is not representative.
+Even worst, historical and modern collections do not overlap spatially, making any inference imposible to interpret.
+In this case, contacted the original collectors of the historical data to define their sampling protocols.
+We then resurveyed the same sites (30 years after the original surveys) using the same sampling protocols.
 In contrast, the New Zealand dataset includes a wide suite of collectors and collection locations but shows no obvious biases in geographical and taxonomic coverage through time.
-For these case studies, we provide annotated R scripts for analysis of different dataset types (e.g., for Spain and New Zealand, as described above).
+We complemented gbif data with further museum collections for bees and flies and analyze the regional richness changes through time.
+For these two case studies, we provide annotated R scripts as examples of analysis for different dataset types (Sup Mat 3).
 These different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories.
@@
 Spain provides an interesting study system because its natural habitats have been transformed extensively by humans over a long time period, but land-use is not as intensive compared with many other European countries.
-In addition, Spain is a bee diversity hotspot (Figure 1) and maintains a relatively heterogeneous landscape.
-Spain has already digitalized a large amount of pollinator occurrence data for both historic and recent periods (Figure 2).
+In addition, Spain is a bee diversity hotspot (Figure 1a) and maintains a relatively heterogeneous landscape.
+Spain has already digitalized a large amount of pollinator occurrence data for both historic and recent periods (Figure 2a).
 However, visual inspection of the data revealed clustering around a few localities.
@@
 There has been no recent sampling of bees in this area.
-However, we found that Enrique systematically sampled seven locations and that additional data were available at the “Museo de Historia Nacional” and other minor collections.
+However, we found that Enrique systematically sampled six independent locations and that additional historical data were available at the “Museo de Historia Nacional” and other minor collections.
 Digitization of these records, along with a re-survey of the original sampling locations provided an excellent dataset for a before and after comparison of bee communities.
-Interestingly, these localities have recently experienced substantial agricultural intensification [expand]()
-
-In brief, we cleaned taxonomic names using the taxize package [@Chamberlain2018], checked sampling completeness for both time periods and compared species richness for each site before and after 1980 with a paired t-test.
-We found that there were a reduced number of species at sites after 1980 (P = 0.058).
+
+In brief, after cleaning taxonomic names for erratas and synonims using the taxize package [@Chamberlain2018], we checked for sampling completeness for both time periods and compared rarefied species richness for each site before and after 1980 with a paired t-test (rarefication at 1000 specimens).
+We found that there were a reduced number of species at sites after 1980 (mean difference 20.27 species; 95 percent confidence interval: -1.03, 41.58; t = 2.44, df = 5, P = 0.06).
 However, this trend was highly dependent on site identity, as two out of three sites showed no richness declines.
-However, we found clear patterns in declines among species groups.
-For example, Andrenids and their parasites showed the strongest declines whereas Halictids increased.
-This is in accordance with findings elsewhere [e.g.
-@Bartomeus2013a] (appendix?).
+Interestingly, these two localities were the two that has experienced less land use changes (both are natural areas embeded into agro-ecosystemss).
+In contrast the other 4 localities suffered large urban or agricutural intensification.
+In addition, species not further recorded in the re-survey are not a random selection of species, but are clustered in a few genera.
+For example, Andrenids and their parasites (e.g.
+Nomada) showed the strongest declines whereas Halictids tend to be more stable (Sup mat 4).
+This pattern of winners and loosers of land use intensification is in accordance with findings elsewhere [e.g.
+@Bartomeus2013a], indicating that some clades are more sensitive to disturbance than others.
 
@@
               fun.aggregate = function(x) length(unique(x)))
-rich_tot #wow.
+#rich_tot #wow.
 #extract genus
@@
               fun.aggregate = function(x) length(unique(x)))
-genus_tot #wow, lot's of genus lost!
+#genus_tot #wow, lot's of genus lost!
 #genus_tot2
-subset(genus_tot2, now == 0 & asensio == 1) #20 lost!
+#subset(genus_tot2, now == 0 & asensio == 1) #20 lost!
 #subset(genus_tot2, now == 1 & asensio == 1) #27 stable
@@
               fun.aggregate = function(x) length(unique(x)))
-genus_tot3
+#genus_tot3
 #colSums(genus_tot3[,2:3])
+#write.csv(genus_tot3, "raw_data/spain_species_per_genus.csv")
 
@@
 ttest <- t.test(asensio[c(1,2,3,5,6,7)], now[c(1,2,3,5,6,7)], paired = TRUE)
-ttest #0.05828
+#ttest #0.05828
 
@@
 
-Figure 3. Rarefied richness at the six Spanish plots resurveyed.
+Figure 3. Comparision of historic collections (1980's) and modern re-surveys (2016) of the rarefied richness of bees at six Spanish localities.
 
@@
 Much of New Zealand’s pollinator fauna is also relatively depauperate.
-For example, New Zealand has only 27 native bee species of [@Donovan2007], which is a fraction of nearby Australia’s c.
+For example, New Zealand has only 27 native bee species [@Donovan2007], which is a fraction of nearby Australia’s c.
 1600 species [@Ascher2015].
@@
 
-We followed protocols outlined in [@Bartomeus2013a] to analyse the New Zealand data.
+We followed protocols outlined in [@Bartomeus2013a] to analyse the New Zealand data at the regional level.
 First, we filtered our original datasets so that data used for analyses only included independent collection events.
@@
 But in addition to local (alpha) diversity and regional (gamma) diversity, researchers need to assess changes in turnover between sites (beta diversity).
-Environmental changes often result in a few “winner” species and many “losers” species [@Bartomeus2013b].
+Environmental changes often result in a few “winner” species and many “losers” species [@Bartomeus2013a].
 Identifying winners and losers is critical as the few winners are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2017].
@@
 
-Museum specimen collections can provide much more information besides species occurrence records, given that such information is recorded when digitizing collections.
+In addition, museum specimen collections can provide much more information besides species occurrence records, given that such information is recorded when digitizing collections.
 This is particularly important for identifying mechanisms of decline and adaptation.
@@
 This approach has been already used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016] in response to climate and land-use change.
+Finally, plant herbariums can also contain indirect evidence of pollinator and pollination declines [60].
+A basic information for basic for linking pollinator declines with its consequences for ecosystem functioning.
 
@@
 Unleashing the power of museum collection data to answer pressing ecological and evolutionary questions is at our hands, but requires the coordinated effort of many actors.
-Using two case studies, we show that strong collaboration with between museum curators and ecologists is key to understanding data and treating it appropriately.
-To progress our understanding of the global pollination crisis, researchers and curators must aim to digitize museum collection data and make it readily available it a format that is widely accessible.
+Using two case studies, we show that strong collaboration between museum curators and ecologists is key to understanding data and treating it appropriately.
+To progress our understanding of the global pollination crisis, researchers and curators must aim to digitize museum collection data and make it readily available in a format that is widely accessible.
 Centralization of regional and national museum collection data in existing global platforms, such as GBIF, would facilitate free and widespread access.
 However, datasets could also be stored in alternative webpages or database repositories (e.g., university and museum webpages or Dryad) providing they are thoroughly documented and easily retrieved and combined with other datasets using open science tools [e.g., @Dallas2016].
+
 We must revolutionize the way that researchers collaborate with museums, in order to foster healthy bidirectional relationships.
@@
 We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe, and for pollinator groups other than bees.
-For the US and Europe, there have been few regional bee extinctions [@Ollerton2014; @Bartomeus2013a] but in disturbed ecosystems, declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006].
+For the US and Europe, there have been few regional bee extinctions [@Ollerton2014; @Bartomeus2013a] but in disturbed ecosystems, declines are widespread [@DePalma2017; @Biesmeijer2006].
 For most other pollinator taxa and regions throughout the world we know almost nothing.
@@
 
-We thank Curro Molina, Patrick McQuinn, and Crona McMonagle for data entry.
-Madrid museum of Historia Natural and ITACyL for access to historical collections.
-We thank Barry Donovan for providing New Zealand bee collection records and Carola Warner for entering the New Zealand data.
-We also thank Canterbury Museum, the New Zealand Arthropod Collection and the Museum of New Zealand Te Papa Tongarewa for providing the New Zealand fly pollinator data.
-We thank E.
+We thank Curro Molina, Carola Warner, Patrick McQuinn, and Crona McMonagle for data entry and Gregorio Aguado for carrying out the Spanish re-sampling.
+We thank Barry Donovan for providing New Zealand bee collection records and E.
 Asensio for sharing his historical data and knowledge.
-IB was funded by FBBVA project.
+We thank the "Museo Nacional de Ciencias Naturales", specially Mercedes Paris, ITACyL (Instituto Tecnológico Agrario de Castilla y León), Canterbury Museum, the New Zealand Arthropod Collection and the Museum of New Zealand Te Papa Tongarewa for access to historical collections.
+IB was funded by Fundación Banco Bilbao Vizcalla Argentaria (FBBVA) project.
 DW was funded through Landcare Research within the Characterising New Zealand’s Land Biota Portfolio.

### 2018-04-06 submitted/SupMat3.Rmd
Last touches

--- 2018-04-06 ms.Rmd
@@
 output:
+  word_document: default
   html_document: default
-  pdf_document: default
-  word_document: default
 editor_options:
@@
 For ilustrative purposes, we mapped the findings of this studies in Figure 1 to show the strong contrast between bee species richness worldwide, whith bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1).
-Despite outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records, there are species-specific examples of historical losses from different parts of the world (e.g., Bombus dalbhomi; @Morales2013).
+Despite outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records, there are species-specific examples of historical losses from different parts of the world (e.g., *Bombus dalbhomi*; @Morales2013).
 
@@
 
-Figure 1: Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness per country and country size.
+**Figure 1:** Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness per country and country size.
 This correction accounts for the species-area relationship.
@@
 #average richness change
-colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
+#colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
 colorss <- (heat.colors(10))
@@
 
-Figure 2.
-Exploration of available data for bee records showing a) The number of bee occurrences before 1980 and after 1980 in GBIF for each country.
+**Figure 2:** Exploration of available data for bee records showing (a) The number of bee occurrences before 1980 and after 1980 in GBIF for each country.
 The upper right quadrat (records in black) contains well covered countries with New Zealand (NZ) and Spain (ES) marked in red.
-For those countries, we show a temptaive comparision of the rarefied number of species in both time periods and show that for most countries the number of species recorded is similar in both time periods (b).
+For those countries, we show a temptaive comparision of the rarefied number of species in both time periods and show that for most countries the number of species recorded is slightly lower (red lines) in recent time periods (b).
 A more careful analysis of this data would help complete the map of global declines (c).
+In the map we plot % change is species recorded in gbif for the available countries to show the geographic coverage.
 Note that this data are likely to contain strong undetected biases, as we explore below.
@@
 #colSums(genus_tot3[,2:3])
-#write.csv(genus_tot3, "raw_data/spain_species_per_genus.csv")
+#write.csv(genus_tot3, "data/SupMat3.csv")
 
@@
 
-Figure 3. Comparision of historic collections (1980's) and modern re-surveys (2016) of the rarefied richness of bees at six Spanish localities.
+**Figure 3:** Comparision of historic collections (1980's) and modern re-surveys (2016) of the rarefied richness of bees at six Spanish localities.
 
@@
 
-Figure 4. Changes in rarefied species richness for different pollinator groups in New Zealand over time. All trends were non-significant (α = 0.05).
+**Figure 4:** Changes in rarefied species richness for different pollinator groups in New Zealand over time. All trends were non-significant (α = 0.05).
 
@@
 
-Figure 5. Model estimated changes (± 1 SE) in the relative occurrence frequency of different New Zealand bee and fly species in museum collections over time.
+**Figure 5** Model estimated changes (± 1 SE) in the relative occurrence frequency of different New Zealand bee and fly species in museum collections over time.
 
@@
 We thank the "Museo Nacional de Ciencias Naturales", specially Mercedes Paris, ITACyL (Instituto Tecnológico Agrario de Castilla y León), Canterbury Museum, the New Zealand Arthropod Collection and the Museum of New Zealand Te Papa Tongarewa for access to historical collections.
-IB was funded by Fundación Banco Bilbao Vizcalla Argentaria (FBBVA) project.
+IB was funded by Fundación Banco Bilbao Vizcaya Argentaria (FBBVA) project.
 DW was funded through Landcare Research within the Characterising New Zealand’s Land Biota Portfolio.

### 2018-04-06 ms.Rmd
Last touches 2

@@
 ^3^ Landcare Research, Auckland, New Zealand  
-^4^ Andrena Iniciativas y Estudios Medio Ambientales
+^4^ Andrena Iniciativas y Estudios Medio Ambientales, Valladolid, Spain  
 *Correspondence: nacho.bartomeus@gmail.com
@@
 However, these programs require significant time to generate long-term datasets and cannot be used to assess historic pollinator populations.
-Finally, the most robust approach for assessing long-term historical pollinator population trends is to use historical information on species occurrences, which is often archived in museum collections [e.g., @Bartomeus2013a].
+Finally, the most practical approach for assessing long-term historical pollinator population trends is to use historical information on species occurrences, which is often archived in museum collections [e.g., @Bartomeus2013a].
 
@@
 lhf <- lhf[order(lhf$pre1980),c(1,17,18)]
-write.csv(lhf, "data/SpMat2.csv")
+#write.csv(lhf, "data/SpMat2.csv")
 #plot(lhf$post1980, lhf$pre1980, xlab = "post1980", ylab = "pre1980", col = as.factor(lhf$country), xlim = c(0,300000), ylim = c(0, 300000))
@@
 pchs <- ifelse(lhf$country == "ES" | lhf$country == "NZ", 19, 1)
-plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col, pch = pchs)
+plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col, pch = pchs, las = 1)
 temp <- subset(lhf, country %in% c("ES", "NZ"))
 text(x = temp$log_post1980, y = temp$log_pre1980+2 , labels = temp$country, col = "red")
+abline(h = 7.5, v = 7.5)
 #36 countries!
@@
 #save output
-save(occ_list, file = "data/gbif.RData")
+#save(occ_list, file = "data/gbif.RData")
 
@@
 data2$change <- ((data2$post1980 - data2$pre1980)*100 / data2$pre1980)
-
+#mean(data2$change)
 map.gbif <- joinCountryData2Map(data2, joinCode="ISO2", nameJoinColumn = "countries",
@@
 #colSums(genus_tot3[,2:3])
-#write.csv(genus_tot3, "raw_data/spain_species_per_genus.csv")
+#write.csv(genus_tot3, "data/SupMat4.csv")
 

### 2018-06-12 ms.Rmd
new commit

--- 2018-04-06 ms.Rmd
@@
 #read in data
-declines <- read.csv(file = "data/richness_declines.V2.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+declines <- read.csv(file = "data/richness_declines.V2.csv", header = TRUE, sep = ",", na.strings=c("","NA"),fileEncoding="latin1")
 #nothing outside EU- USA. Spain and NZ can be added.

### 2018-06-12 ms.Rmd
new commit

--- 2018-04-06 ms.Rmd
@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
 editor_options:
@@
 Since then, studies on pollinator responses to various global change drivers have multiplied rapidly.
-Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010].
+Researchers have now developed strong consensus that disturbances such as habitat destruction [@jamie2016], land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010].
 Yet, the current status and population trends of most pollinator species worldwide remain unknown.
@@
 #read in data
-declines <- read.csv(file = "data/richness_declines.V2.csv", 
-                     header = TRUE, sep = ",", na.strings=c("","NA"))
+declines <- read.csv(file = "data/richness_declines.V2.csv", header = TRUE, sep = ",", stringsAsFactors=FALSE,na.strings=c("","NA"))
+declines$percent_richness_change <- as.numeric(declines$percent_richness_change)
+declines$proportion_richness_change <- as.numeric(declines$proportion_richness_change)
 #nothing outside EU- USA. Spain and NZ can be added.

### 2018-06-12 ms.Rmd
new commit

@@
 Since then, studies on pollinator responses to various global change drivers have multiplied rapidly.
-Researchers have now developed strong consensus that disturbances such as habitat destruction [@jamie2016], land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010].
+Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010].
 Yet, the current status and population trends of most pollinator species worldwide remain unknown.

### 2018-06-13 revision/RtoR.Rmd
Revision (95%)

---
title: "Response to reviewers"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Find included the revision of the manuscript "Historic collections as a tool for assessing the global pollination crisis". We want to thank the Editors and both reviewers for the encouraging words and the comments provided which have helped clarify the message. See a detailed response below.

## Editor

> Thank you for this contribution. Two reviewers who are very knowledgeable about pollinators and their decline with global change have assessed your work and find that it will be a valuable addition to the literature. I agree that this manuscript has been developed thoughtfully and only requires minor revision. I especially appreciated that the statistical methods demonstrate how researchers might alleviate the constraints of biases in specimen data. The only overarching comment I have is that you emphasize when you are introducing a method for alleviating biases to draw attention of the Phil Trans B readership.

Thanks for the positive comments. We have now emphasized the methodological parts that deal with biases through the text.

> Here are my minor comments.

> Line 104: How often are these changes in richness at the local scale mediated by introduced species at the regional scale? I suspect this paragraph is about native species exclusively, but it would be good to specify.

Yes, we specify now we are reporting data on native species trends at both scales. As there is not much information on exotic species displacing natives (other than by pathogen spillover) we do not discuss this issue here.

> Line 195: Can you give a citation for this statement that 1800 specimens is enough for analyses? And does this vary by spatial area?

Thanks for pointing this out. The 1800 specimens threshold is arbitrary and of course, it could vary with the spatial area. However, there is no magic number to put a threshold. For a global analysis, this represents a good compromise between data availability and number of countries that can be included. We clarified that in text.

> And are the data you use here just museum specimens or museum specimens and observations?

Gbif includes a variety of data sources. Within bees, most of them are vouchered in public collections, but a few may be vouchered elsewhere, but as far as I can tell observations are not included.

> Lines 228: Even ‘worse’ (not ‘worst’)

Fixed.

> Line 251: How did you find out that this was a systematic sample? From Enrique? From field notes? From the data structure?

We clarify in text that we contacted Enrique directly.

> Lines 264: Perhaps do not say “suffer” as it personifies the place and shows a bias toward more natural habitats. (This is tangential, but it is my understanding that often bees benefit from enhanced floral resources and openness of urban habitats.)

Thank you, we changed to "experienced".

> Lines 304-305: Can you be more specific about what led you to believe that these data were incomplete? All museum data are incomplete, but your specific assessment of what makes these data from prior to 1970 unanalyzable will be helpful to future researchers using museum specimens to assess decline.

We added the clues that we used for assessing incompleteness: "For example, old specimens lack particular genus, and geographically restricted."

> Line 318: Can you please insert a reference for only using species that have 30 or more records?

Again, very good point on thresholds. We clarified that "Preliminary analysis using random subsampling of specimens from species with large sample size indicated that below 30 records the trends become too variable. [JAMIE, can you look onto this?]()

> Lines 325-327: I know this is intuitive, but, if available, please insert a citation for this.

**Look for citation rarefied species confound evenness!

> Line 349: “Loser” not “losers”

Fixed.

> Lines 353-355: Here, I think you are referring to data you can retrieve from virtual collections. From physical collections, all label data are available . . .

Good point. We clarified this.

> Line 365: “herbaria” not “herbariums”

Fixed.

> Lines 395-396: It might be useful to note here that museums potentially contain vouchers of many (perhaps thousands? tens of thousands?) of undescribed pollinator species and are thus critical resources for the taxonomic/systematics studies required prior to ecological and evolutionary research on global change. (There should be a handy citation on this in the museum literature.)

** Add citation.

> Figure 2c: I realize that a graduated scale is the best way to show these data because they are continuous. However, perhaps there is a color palette that happens to shift enough at zero so that decline vs. no indication of decline can be differentiated visually. Right now, it is difficult to discern which regions have decline according to the data.

** Jamie?

> Figure 5. Please increase text size in y axes of both figures and the legend.

** Jamie?

# Referee: 1

> Comments to Author(s)
This is an excellent paper that reviews what we know and what we don’t know about bee declines from historical collections of bees around the world. I like the author’s introductory comments about the status of our understanding of bee declines. They comment that “early accounts of pollinator declines were somewhat anecdotal, given the lack of pollinator population data at that time.” I completely agree with that assessment. There are very few data on any group of bees other than bumble bees that can be used to assess whether bee populations are in decline or not. The use of museum specimen records provides an incredibly valuable tool, when analyzed correctly, for documenting and assessing the status of wild bees and this paper provides a nice overview of the limitations of these data and how best to use these data.

Thanks for the positive feed back!

> Minor, editorial comments: 
> Line 140-143 – this sentence is oddly worded. Rather than say “Despite outside of Europe…” I would say “Except for outside of Europe…”

Thanks, we reworded it as recommended.

> Line 208-209 – “Aside from bees, similar exploratory analyses can easily be conducted for other taxa” should be “We focus on bees, but similar exploratory analyses can easily be conducted for other taxa”
Line 220, 222 – “this data” should be “these data”
Line 224 – delete “collection”
Line 228 – “worst” should be “worse”
Line 263 – reword: “…the two that have experienced less dramatic change…”

We reworded all sentences as suggested, thank you for the edits.

> Line 266-268 – this statement matches my expectation for which groups are likely to be impacted by land use change – halictids are trash bees that hang on in even the most disturbed habitats.

Yes, this is also our experience with this group of species.

> Fig. 3 – can we be confident that collecting effort was similar between the data from the 1980’s and the more recent data (2016)? How many specimens does each collection represent?

**Comparisions of this kind are alwys hard to standardize. By rarefying species richness we minimize any bias due to sampling effort differences. We added the total number of spceimens collected by time period.

> Line 337 – “…ground-nesting, solitary bees.” [add hyphen and comma]

Added.

> Line 364 – another paper to cite that has some bearing on the issue of body size change in bees is: Renauld M, Hutchinson A, Loeb G, Poveda K, Connelly H. 2016. Landscape simplification constrains adult size in a native ground-nesting bee. PLoS One. 11(3):e0150946”

** Thanks for the reference, we added it. [JAMIE, can you do this?]()

> Line 364-367 – awkwardly worded sentence. Please rephrase.
Line 369-370 – awkwardly worded sentence. Please rephrase.

We reworded those sentences.

> Line 372 – change “progress” to “advance”

Changed.

> References 14 and 30 – titles of journal articles are capitalized; should be lower case for consistency with other citations.

Fixed
**[Jamie?]()

> Reference 57 – is this really the title of the article?

Fixed
**[Jamie?]()

> Overall, a much needed paper that will be widely cited as the use of museum specimen data is increasingly used to address the declines of bees and other pollinators.

# Referee: 2

> Comments to Author(s)
> In recent decades, reported declines in pollinators have raised global concerns about loss of pollinator biodiversity and delivery of pollination services. However, most data on pollinator trends originate from just a few well-developed countries (US, NW-Europe), and we actually know very little about the status of pollinator communities in other parts of the world. In addition, assessments of pollinator population trends have mostly been restricted to a limited set of taxonomic groups. In this paper, Bartomeus et al. make a case for better capitalizing the potential information stored in historic museum collections to improve our understanding of the geographical and taxonomic extent of pollinator declines. They review current evidence for pollinator declines and discuss the potential (and limitations) of entomological collections to address knowledge gaps and advance the assessment of the extent of pollinator decline. The potential of entomological collection data is nicely illustrated by presenting two case-studies.

> Overall, the paper is well-structured, well written and highly topical, and I really enjoyed reading it. I only have a few (minor) comments and tweaks to further improve the clarity and readability of the paper:

We are glad the reviewer enjoyed the manuscript.

> L27-30: Bit long, odd sentence (despite... but....). Suggest to make separate sentences.

We splited the sentence.

> L55: replace “greater” with “more”
> L59: I’d suggest to change into “diverse, including bees, etc.....”

> L65: I think I understand what you mean, but it is a bit vague what is meant by the phrase “the uneven distribution of researchers”.

We clarified this sentence.

> L97-99: Not quite. If I’m not mistaken this study showed that, regardless of diet breadth, the identity of preferred host plants explained decline of bee species. So bee species declined if their preferred host plant(s) declined.

Yes, you are right and we changed the wording to better reflect that result.

> L100-101: Could be, but also vice versa, when records from amateur insect collectors are biased towards just these well-preserved natural areas, that (still) contain “interesting species”.

Could be, but this bias would show even steeper declines. We clarified the sentence to refer to less human-developed areas.

> L102-103: For clarity, add that this concerns trends in richness, not population/species distribution trends of individual species. The distinction between trends in species richness and species population trends is also sometimes not entirely clear in other parts of the manuscript, please address.

Thanks for pointing this out as we agree is important to diferenciate richness from population trends data. We reworded it for clarity.

> L115-118: See also Nieto et al. 2014 European Red List of bees for Europa-wide assessment of bumblebee trends (24% of bumblebee species threatened with extinction; 46% of bumblebee species have declining population trend).

We added this reference.

> L136: replace “this” with “these”

Done.

> L139: Sup Mat 1: why are there sometimes lines with identical data for some studies in this file? Same data but for different grid cells?

Yes, those are different grid cells. We add the explanation to the caption.

> L140-142. This sentence is really confusing and hard to grasp. Please rephrase.

We split the sentence in two and re-worded it.

> L176: suggest to change into ”which may lead to spatial bias....”

Changed.

> L195, Figure 2a: I would like to see some justification for the used cut-off point of 1800 records in both periods. What’s the rationale behind this?

The 1800 specimens threshold is arbitrary. For a global analysis, this represents a good compromise between data availability and number of countries that can be included. We clarified that in text.

> L228: replace “ worst” with “worse”

Replaced.

> L284-291: I do not really see the point/relevance of this section. The same story (albeit over a longer timeframe) goes for e.g. NW Europe where human activity has also affected pollinator communities (either positively or negatively) before specimen collection began (e.g. through large-scale forestation, development of large areas of heathlands and extensive grasslands, etc.). I’d suggest to omit this.

We amoited this as suggested.

> L299: Please clarify what regional level.

We removed the reference to "region" to avoid confusions.

> L302-304: And what about spatial coverage?

Yes, we added a note on spacial coverage.

> L309: How was mean richness per bin calculated, over years within bins?

Yes, we reworded it for clarity.

> L361: see also ref [19] and [23]

We cite here the first study we are aware of using this technique.

> L389: switch “fill” and “any”

Switched.

### 2018-06-13 revision/ms.Rmd
Revision (95%)

--- 2018-06-12 ms.Rmd
@@
 Here, we provide a synthesis of how researchers have used historical data to identify long-term changes in biodiversity, species abundances, morphology and pollination services.
-Despite recent advances, we show that information on the status and trends of most pollinators is absent, but we highlight opportunities and limitations to progress the assessment of pollinator declines globally.
+Despite recent advances, we show that information on the status and trends of most pollinators is absent.
+We highlight opportunities and limitations to progress the assessment of pollinator declines globally.
 Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed.
@@
 Yet, the current status and population trends of most pollinator species worldwide remain unknown.
-For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, greater than 55% of bee species fell into the “Data Deficient” category [@Nieto2014].
+For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, more than 55% of bee species fell into the “Data Deficient” category [@Nieto2014].
 For countries outside of Europe and the US, data on pollinator populations is almost non-existent.
 
-One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and lizards [@Ollerton2017].
+One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse, including bees, flies, butterflies, beetles, birds, bats and lizards [@Ollerton2017].
 Additionally, many pollinators are highly mobile, short-lived and small, which makes monitoring their populations difficult.
@@
 However, bees are diverse, with more than 20,000 species currently described worldwide, and often require expert taxonomists for identification.
-Furthermore, the uneven distribution of researchers has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
+Furthermore, the worldwide uneven distribution of research studies has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
 
@@
 For example, in the UK and the Netherlands, a citizen science based study using both observations and museum collection data detected strong richness declines for bees, hoverflies and flowering plants [@Biesmeijer2006].
-In the Netherlands, museum data have also revealed simultaneous plant and pollinator declines [@Scheper2014].
+In the Netherlands, museum data have also revealed coupled plant and pollinator declines [@Scheper2014].
 Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction.
-However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for well-preserved natural areas.
-Further exploration of this dataset revealed that for declining pollinator taxa, the trend has attenuated in recent decades [@Carvalheiro2013].
-
-Although studies of local pollinator communities often detect richness declines, regional richness may remain relatively stable.
+However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for less human-developed areas [jamie check wording of the "human-developed"]().
+Further exploration of this dataset revealed that for declining pollinator taxa, the species richness declining trend has attenuated in recent decades [@Carvalheiro2013].
+
+Although studies of local native pollinator communities often detect richness declines, regional native richness may remain relatively stable.
 For example, regional estimates for bee species richness changes in the eastern US show moderate declines [@Bartomeus2013a] and very few regional extinctions [@Colla2012].
@@
 For example, declines of up to 18% in local bumblebee richness have been reported for Belgium and the Netherlands [@Carvalheiro2013].
-In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011].
+In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011] and 70 % of bumblebee species are clasified as threatened or with declining population trends [@Nieto2014].
 In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a].
@@
 
-For ilustrative purposes, we mapped the findings of this studies in Figure 1 to show the strong contrast between bee species richness worldwide, whith bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1).
-Despite outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records, there are species-specific examples of historical losses from different parts of the world (e.g., *Bombus dalbhomi*; @Morales2013).
+For ilustrative purposes, we mapped the findings of these studies in Figure 1 to show the strong contrast between bee species richness worldwide, whith bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1).
+Except for outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records.
+However, there are species-specific examples of historical losses from different parts of the world (e.g., *Bombus dalbhomi*; @Morales2013).
 
@@
 For example, collectors tend to target rare or unusual over common taxa, discard damaged individuals or only accession a certain number of individuals.
-In addition, collections are often made opportunistically, leading to a spatial biases where difficult to access areas are under-sampled or conversely, where samples are biased towards easily accessed locations (e.g., towns/cities and/or roadsides).
+In addition, collections are often made opportunistically, which may lead to spatial bias where difficult to access areas are under-sampled or conversely, where samples are biased towards easily accessed locations (e.g., towns/cities and/or roadsides).
 Further, museum collection data can only be used to determine where species are present and not where they are absent.
@@
 Focusing on bee taxa, we show here the number of modern and historic bee records currently available for different countries (Figure 2a, Sup mat 2).
-Thirty-seven countries count with more than 1800 records in each time period, making these data potentially analyzable without further data collection effort (see Figure 2b and c for an initial exploration).
-However, a proper analysis of this dataset would need a careful look at the data as we detail later on with two examples (Spain and New Zealand).
+We selected thirty-seven countries that count with more than 1800 records in each time period, making these data potentially analyzable without further data collection effort (see Figure 2b and c for an initial exploration).
+We selected this arbitrary threshold as a compromise between sample size and number of countries that can be included in a general analysis.
+However, a proper analysis of this dataset and the minimum sample size needed per country would need a careful look at the data characteristics as we detail later on with two examples (Spain and New Zealand).
 Most countries fall short in one or both axes.
@@
 In contrast, more than 192 countries have less than 1000 records for both time periods, making them poor candidates for analysing long-term pollinator population trends.
-Aside from bees, similar exploratory analyses can easily be conducted for other taxa.
+We focus on bees, but similar exploratory analyses can easily be conducted for other taxa.
 
@@
 For those countries, we show a temptaive comparision of the rarefied number of species in both time periods and show that for most countries the number of species recorded is slightly lower (red lines) in recent time periods (b).
-A more careful analysis of this data would help complete the map of global declines (c).
+A more careful analysis of these data would help complete the map of global declines (c).
 In the map we plot % change is species recorded in gbif for the available countries to show the geographic coverage.
@@
 
-As stated above, once historical collection datasets are made available, researchers must identify any potential biases.
+As stated above, once historical datasets are made available, researchers must identify any potential biases.
 We explore this process with two contrasting dataset examples (Spain and New Zealand).
@@
 Hence the geographical coverage is not representative.
-Even worst, historical and modern collections do not overlap spatially, making any inference imposible to interpret.
+Even worse, historical and modern collections do not overlap spatially, making any inference imposible to interpret.
 In this case, contacted the original collectors of the historical data to define their sampling protocols.
@@
 There has been no recent sampling of bees in this area.
-However, we found that Enrique systematically sampled six independent locations and that additional historical data were available at the “Museo de Historia Nacional” and other minor collections.
+However, we found that Enrique systematically sampled six independent locations and that additional historical data were available at the “Museo de Historia Nacional” and other minor collections (E.
+Asensio personal communication).
 Digitization of these records, along with a re-survey of the original sampling locations provided an excellent dataset for a before and after comparison of bee communities.
 
-In brief, after cleaning taxonomic names for erratas and synonims using the taxize package [@Chamberlain2018], we checked for sampling completeness for both time periods and compared rarefied species richness for each site before and after 1980 with a paired t-test (rarefication at 1000 specimens).
+In brief, after cleaning taxonomic names for erratas and synonims using the taxize package [@Chamberlain2018], we checked for sampling completeness for both time periods and compared rarefied species richness for each site before and after 1980 with a paired t-test (sample size pre-1980: X; sample size pre-1980: X; rarefication at 1000 specimens).
 We found that there were a reduced number of species at sites after 1980 (mean difference 20.27 species; 95 percent confidence interval: -1.03, 41.58; t = 2.44, df = 5, P = 0.06).
 However, this trend was highly dependent on site identity, as two out of three sites showed no richness declines.
-Interestingly, these two localities were the two that has experienced less land use changes (both are natural areas embeded into agro-ecosystemss).
-In contrast the other 4 localities suffered large urban or agricutural intensification.
+Interestingly, these two localities were the two that has experienced less dramatic land use changes (both are natural areas embeded into agro-ecosystemss).
+In contrast the other 4 localities experienced large urban or agricutural intensification.
 In addition, species not further recorded in the re-survey are not a random selection of species, but are clustered in a few genera.
@@
 Before human arrival, New Zealand was predominately forested, but has since been dramatically altered by people.
-Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 [@Perry2014].
-Therefore, human activity likely affected pollinator communities in New Zealand long before surveys and specimen collections began.
-Nevertheless, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.
+We can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.
 
@@
 
-We followed protocols outlined in [@Bartomeus2013a] to analyse the New Zealand data at the regional level.
+We followed protocols outlined in [@Bartomeus2013a] to analyse the New Zealand data.
 First, we filtered our original datasets so that data used for analyses only included independent collection events.
 To do this, we removed specimens collected at the same location, on the same date, and by the same collector.
-We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980.
-Further exploration of the New Zealand native bee data raised doubts on collection completeness in records prior to 1970, so we removed these records from further analyses.
+We found our data had reasonable coverage along time periods and across space, although there was a peak in collection occurrences from 1960-1980.
+Further exploration of the New Zealand native bee data raised doubts on collection completeness in records prior to 1970.
+For example, old specimens lack particular genus, and geographically restricted.
+Hence, we removed these records from further analyses.
 We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years.
-We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin.
+We then estimated species richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the rarefied species richness ±SE for each bin.
 Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness.
@@
 Further, we only included species in this analysis for which we had 30 or more records.
+Preliminary analysis using random subsampling of specimens from species with large sample size indicated that below 30 records the trends become too variable.
 To account for differences in sampling effort between years, we weighted each year by the total number of samples collected that year.
@@
 Interestingly, the two exotic bee species that declined in relative occurrence were both in the genus Bombus, which were intentionally introduced into New Zealand for the pollination of crops.
-Native bees that increased in relative occurrence were mostly from the genus Leioproctus, which are medium sized, ground nesting solitary bees.
+Native bees that increased in relative occurrence were mostly from the genus Leioproctus, which are medium sized, ground-nesting, solitary bees.
 Only one out of 14 fly species increased in relative occurrence, which was exotic, whereas four species decreased in occurrence (three native and one exotic).
@@
 But in addition to local (alpha) diversity and regional (gamma) diversity, researchers need to assess changes in turnover between sites (beta diversity).
-Environmental changes often result in a few “winner” species and many “losers” species [@Bartomeus2013a].
+Environmental changes often result in a few “winner” species and many “loser” species [@Bartomeus2013a].
 Identifying winners and losers is critical as the few winners are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2017].
@@
 
-In addition, museum specimen collections can provide much more information besides species occurrence records, given that such information is recorded when digitizing collections.
+In addition, digitalized museum specimen collections can provide much more information besides species occurrence records, given that such information is recorded when digitizing collections.
 This is particularly important for identifying mechanisms of decline and adaptation.
@@
 Finally, museum specimens can be measured to track evolutionary changes by measuring the traits of specimen.
-This approach has been already used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016] in response to climate and land-use change.
-Finally, plant herbariums can also contain indirect evidence of pollinator and pollination declines [60].
-A basic information for basic for linking pollinator declines with its consequences for ecosystem functioning.
+This approach has been already used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016, @Renauld2016] [jamie, ref to add]() in response to climate and land-use change.
+Finally, plant specimens stored in plant herbaria can also contain indirect evidence of pollinator and pollination declines [60] which can link pollinator declines with its consequences for ecosystem functioning.
 
@@
 Using two case studies, we show that strong collaboration between museum curators and ecologists is key to understanding data and treating it appropriately.
-To progress our understanding of the global pollination crisis, researchers and curators must aim to digitize museum collection data and make it readily available in a format that is widely accessible.
+To advance our understanding of the global pollination crisis, researchers and curators must aim to digitize museum collection data and make it readily available in a format that is widely accessible.
 Centralization of regional and national museum collection data in existing global platforms, such as GBIF, would facilitate free and widespread access.
@@
 It is also crucial that these data are analysed appropriately.
-This requires researches to identify biases and to any fill taxonomic and geographic gaps where possible.
+This requires researches to identify biases and to fill any taxonomic and geographic gaps where possible.
 We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe, and for pollinator groups other than bees.
@@
 For most other pollinator taxa and regions throughout the world we know almost nothing.
-Moving forward, the first step for many taxa will be to identify and describe species.
+Moving forward, the first step for many taxa will be to identify and describe species [REF].
 Only then can we begin to document pollinator declines.

### 2018-06-14 revision/ms.Rmd
have added a new reference list (references2) and made some minor edits on the manuscript (in revision/ms)

@@
   chunk_output_type: console
-bibliography: data/references.xml
+bibliography: data/references2.xml
 ---
@@
 
-One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse, including bees, flies, butterflies, beetles, birds, bats and lizards [@Ollerton2017].
-Additionally, many pollinators are highly mobile, short-lived and small, which makes monitoring their populations difficult.
+One of the main barriers to identifying long-term pollinator population trends is that pollinators are incredibly taxonomically diverse and include bees, flies, butterflies, beetles, birds, bats and lizards [@Ollerton2017].
+Additionally, many pollinators are highly mobile, short-lived and small in size, which makes monitoring their populations difficult.
 Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Winfree2011].
 However, bees are diverse, with more than 20,000 species currently described worldwide, and often require expert taxonomists for identification.
-Furthermore, the worldwide uneven distribution of research studies has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
+Furthermore, the uneven global distribution of research studies has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
 
@@
 Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction.
-However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for less human-developed areas [jamie check wording of the "human-developed"]().
-Further exploration of this dataset revealed that for declining pollinator taxa, the species richness declining trend has attenuated in recent decades [@Carvalheiro2013].
+However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for areas subject to less human activity.
+Further exploration of this dataset revealed that for declining pollinator taxa, the declining species richness trend has attenuated in recent decades [@Carvalheiro2013].
 
@@
 This is a pattern also detected in the UK, where relatively few regional bee extinctions has been reported [@Ollerton2014].
-These regional findings are in stark contrast with the widespread local extinctions reported in local studies.
+These regional findings contrast with the local extinctions reported in other studies.
 For example, Burkle et al.
-[@Burkle2013] compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants and reports several local extinctions.
+[@Burkle2013] compared historical observations of bee species’ occurrences in a large forested ecosystem with remaining forest remnants and reported several local extinctions.
 However, it is important to note that there is strong concordance between local extinctions and regional declines [@Bartomeus2013c], suggesting that local extinctions are indicators of regional population declines.
@@
 In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011] and 70 % of bumblebee species are clasified as threatened or with declining population trends [@Nieto2014].
-In the USA, reported bumble decline are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a].
+In the USA, reported bumble declines are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a].
 However, studies on species richness changes for other pollinator taxa are both scarce and geographically restricted.
@@
 Butterfly species richness has declined substantially in the Netherlands and Belgium since the 1950’s, although declines in Great Britain have been less severe [@Carvalheiro2013].
-In Belgium, another study [@Maes2001] found that richness declines have been severe (approximately 30%), although this study assessed richness changes over a longer time period (early 1900’s to 2000) compared to [@Carvalheiro2013] (1950-69 vs.
+In Belgium, another study [@Maes2001] found that butterfly richness declines have been severe (approximately 30%), although this study assessed richness changes over a longer time period (early 1900’s to 2000) compared to [@Carvalheiro2013] (1950-69 vs.
 1970-80 and 1970-89 vs.
@@
 In parts of Germany, up to 70% declines in local butterfly richness have been reported [@Wenzel2006].
-Compared with other insect pollinator taxa, there are very few studies on hoverfly species richness changes, which are all restricted to Europe.
-In Belgium, Great Britain and the Netherlands, hoverfly richness changes have been modest [@Carvalheiro2013].
+Compared with other insect pollinator taxa, there are very few studies on hoverfly species richness changes, of which are all restricted to Europe.
+Specifically, in Belgium, Great Britain and the Netherlands, hoverfly richness changes have been modest [@Carvalheiro2013].
 In the Netherlands, moderate increases in hoverfly species richness have been shown, whereas in Great Britain no significant directional changes were detected [@Biesmeijer2006].
@@
 
-For ilustrative purposes, we mapped the findings of these studies in Figure 1 to show the strong contrast between bee species richness worldwide, whith bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1).
+For illustrative purposes, we mapped the findings of these studies (Figure 1) to show the large differences in bee species richness worldwide, with bee diversity hotspots in mediterranean countries, against the paucity of countries for which we have any local or regional data on bee, syrphid or butterfly declines (raw data: Sup Mat 1).
 Except for outside of Europe and the US and for non-insect taxa, there are very few or no studies on pollinator declines using historical records.
@@
 For example, collectors tend to target rare or unusual over common taxa, discard damaged individuals or only accession a certain number of individuals.
-In addition, collections are often made opportunistically, which may lead to spatial bias where difficult to access areas are under-sampled or conversely, where samples are biased towards easily accessed locations (e.g., towns/cities and/or roadsides).
-Further, museum collection data can only be used to determine where species are present and not where they are absent.
+In addition, collections are often made opportunistically, which may lead to spatial bias where difficult to access areas are under-sampled or conversely, where samples are biased towards easily accessed locations (e.g., towns, cities and roadsides).
+Further, museum collection data can only be used to determine where species are present, not where they are absent.
 However, given adequate sample sizes and appropriate statistical techniques, most biases can be accounted for [e.g., @Bartomeus2013a; @Keil2011; @Pearce2006].
@@
 Focusing on bee taxa, we show here the number of modern and historic bee records currently available for different countries (Figure 2a, Sup mat 2).
-We selected thirty-seven countries that count with more than 1800 records in each time period, making these data potentially analyzable without further data collection effort (see Figure 2b and c for an initial exploration).
-We selected this arbitrary threshold as a compromise between sample size and number of countries that can be included in a general analysis.
-However, a proper analysis of this dataset and the minimum sample size needed per country would need a careful look at the data characteristics as we detail later on with two examples (Spain and New Zealand).
+We selected thirty-seven countries that have more than 1800 records in each time period, making these data potentially analyzable without further data collection effort (see Figure 2b and c for an initial exploration).
+We selected this arbitrary threshold as a compromise between sample size and the number of countries that can be included in a general analysis.
+However, proper analysis of this dataset and determining the minimum sample size needed per country would require careful inspection of the data, which we outline later with two examples (Spain and New Zealand).
 Most countries fall short in one or both axes.
-For example, a variety of countries located in different continents such as Switzerland, Sri Lanka, Nicaragua or Zimbawe have a decent number of recent records, but lack historical collections and researchers should prioritize the digitalization of old material before embarking on data analyses.
+For example, a variety of countries located in different continents such as Switzerland, Sri Lanka, Nicaragua or Zimbawe have a decent number of recent records, but lack historical collections.
+Thus, researchers should prioritize the digitalization of old material before embarking on data analyses for these countries.
 It is also important to note that historical records are not always vouchered in local museums (i.e., many European and US museums contain large collections of pollinators from other countries).
@@
 Hence the geographical coverage is not representative.
-Even worse, historical and modern collections do not overlap spatially, making any inference imposible to interpret.
-In this case, contacted the original collectors of the historical data to define their sampling protocols.
+Even worse, historical and modern collections do not overlap spatially, making any inference impossible to interpret.
+In this case, we contacted the original collectors of the historical data to define their sampling protocols.
 We then resurveyed the same sites (30 years after the original surveys) using the same sampling protocols.
@@
 Much of New Zealand’s pollinator fauna is also relatively depauperate.
-For example, New Zealand has only 27 native bee species [@Donovan2007], which is a fraction of nearby Australia’s c.
+For example, New Zealand has only 28 native bee species [@Donovan2007], which is a fraction of nearby Australia’s c.
 1600 species [@Ascher2015].
@@
 Before human arrival, New Zealand was predominately forested, but has since been dramatically altered by people.
-We can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.
+Therefore, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.
 
@@
 Collection records from the New Zealand Arthropod Collection (NZAC) are freely available online (https://scd.landcareresearch.co.nz/).
-Fly pollinator data was obtained from three participating New Zealand museums and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators.
+Fly pollinator data were obtained from three participating New Zealand museums and covers two families (Calliphoridae and Syrphidae) that contain important fly pollinators.
 Collections for the bee and fly datasets span over 100 years (early 1900s to late 2000s).
@@
 For species that showed overdispersion, we used a quasi-binomial distribution.
-Further, we only included species in this analysis for which we had 30 or more records.
+Further, we only included species in this analysis for which we had 30 or more independent records [@Bartomeus2013a].
 Preliminary analysis using random subsampling of specimens from species with large sample size indicated that below 30 records the trends become too variable.
@@
           axis.title.y=element_text(size=18, vjust = 1),
-          axis.text.x=element_text(face = "italic", angle= 45, hjust = 1, vjust = 1, size =7),
+          axis.text.x=element_text(face = "italic", angle= 45, hjust = 1, vjust = 1, size =8),
           axis.title.x=element_text(size=22, vjust = 1),
@@
 p <- p + theme(axis.title.y=element_text(margin=margin(0,20,0,0)))
+p <- p + theme(legend.text=element_text(size=12))
+p <- p + theme(legend.title = element_blank())
 p <- p + scale_colour_brewer(palette = "Dark2")
@@
 A recent study found that more than 90% of the papers investigating pollinator responses to land-use change focused solely on richness and abundance descriptors [@Winfree2011].
-But in addition to local (alpha) diversity and regional (gamma) diversity, researchers need to assess changes in turnover between sites (beta diversity).
+However, in addition to local (alpha) diversity and regional (gamma) diversity, researchers need to assess changes in turnover between sites (beta diversity).
 Environmental changes often result in a few “winner” species and many “loser” species [@Bartomeus2013a].
@@
 Aside from this information, bee specimens often contain pollen loads trapped on hairs, from which past visitation events can be identified [@Kleijn2008].
-Finally, museum specimens can be measured to track evolutionary changes by measuring the traits of specimen.
-This approach has been already used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016, @Renauld2016] [jamie, ref to add]() in response to climate and land-use change.
-Finally, plant specimens stored in plant herbaria can also contain indirect evidence of pollinator and pollination declines [60] which can link pollinator declines with its consequences for ecosystem functioning.
+Finally, museum specimens can be measured to track evolutionary changes by measuring the traits of specimens.
+This approach has already been used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016, @Renauld2016] in response to climate and land-use change.
+Finally, plant specimens stored in plant herbaria may contain indirect evidence of pollination declines [60], thus linking pollinator declines with consequences for ecosystem functioning.
 
@@
 Centralization of regional and national museum collection data in existing global platforms, such as GBIF, would facilitate free and widespread access.
-However, datasets could also be stored in alternative webpages or database repositories (e.g., university and museum webpages or Dryad) providing they are thoroughly documented and easily retrieved and combined with other datasets using open science tools [e.g., @Dallas2016].
-
-We must revolutionize the way that researchers collaborate with museums, in order to foster healthy bidirectional relationships.
-For example, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered [@Packer2018; @Wheeler2003], rendering them less useful for future research.
+However, datasets could also be stored in alternative webpages or database repositories (e.g., university and museum webpages or Dryad) providing they are thoroughly documented, and easy to retrieve and combined with other datasets using open science tools [e.g., @Dallas2016].
+
+We must revolutionize the way that researchers collaborate with museums in order to foster healthy bidirectional relationships.
+For example, ecological researchers collect massive amounts of specimens, but these are often inappropriately vouchered [@Packer2018; @Wheeler2003], rendering them less useless for future research.
 To improve this process, strong communication between museums and researchers is required.
@@
 For the US and Europe, there have been few regional bee extinctions [@Ollerton2014; @Bartomeus2013a] but in disturbed ecosystems, declines are widespread [@DePalma2017; @Biesmeijer2006].
-For most other pollinator taxa and regions throughout the world we know almost nothing.
+For most other pollinator taxa and regions throughout the world, we know almost nothing.
 Moving forward, the first step for many taxa will be to identify and describe species [REF].

### 2018-06-15 revision/RtoR.Rmd
have added a new reference list (references2) and made some minor edits on the manuscript (in revision/ms)
Almost there. Asher (now webpage discoverlife cited only), Pauw and Gottelli refs need to be fixed and Scale bar need new colors.

@@
 title: "Response to reviewers"
-output: html_document
+output:
+  word_document: default
+  html_document: default
 ---
@@
 
-Yes, we specify now we are reporting data on native species trends at both scales.
+Yes, we specify now that we are reporting data on native species trends at both scales.
 As there is not much information on exotic species displacing natives (other than by pathogen spillover) we do not discuss this issue here.
@@
 Thanks for pointing this out.
-The 1800 specimens threshold is arbitrary and of course, it could vary with the spatial area.
-However, there is no magic number to put a threshold.
-For a global analysis, this represents a good compromise between data availability and number of countries that can be included.
+The 1800 specimens threshold is arbitrary and of course, a the optimal threshold could vary with the country area.
+However, there is no magic number to set up a threshold.
+For a global analysis, 1800 specimens represents a good compromise between data availability and number of countries that can be included.
 We clarified that in text.
@@
 
-We added the clues that we used for assessing incompleteness: "For example, old specimens lack particular genus, and geographically restricted."
+We added the clues that we used for assessing incompleteness: "For example, some genus are not represented in old records."
 
@@
 
-Again, very good point on thresholds.
-We clarified that "Preliminary analysis using random subsampling of specimens from species with large sample size indicated that below 30 records the trends become too variable.
-[JAMIE, can you look onto this?]()
+Again, very good point on selecting thresholds.
+We clarified that "Preliminary analysis using random subsampling of specimens from species with large sample sizes indicated that below 30 records the trend become too variable.
 
@@
 
-**Look for citation rarefied species confound evenness!
+We added a reference.
 
@@
 
-** Add citation.
+We added a reference (Meier and Dikow 2004)
 
@@
 
-** Jamie?
+We updated the color scale to make this ditiction clearer.
 
@@
 
-** Jamie?
+Fixed.
 
@@
 
-**Comparisions of this kind are alwys hard to standardize.
+Comparisions of this kind are alwys hard to standardize.
 By rarefying species richness we minimize any bias due to sampling effort differences.
@@
 
-** Thanks for the reference, we added it. [JAMIE, can you do this?]()
+Thanks for the reference, we added it.
 
@@
 
-Fixed
-**[Jamie?]()
+Fixed.
 
@@
 
-Fixed
-**[Jamie?]()
+Yes, this is correct.
 
@@
 
+Both edits changed as suggested.
+
 > L65: I think I understand what you mean, but it is a bit vague what is meant by the phrase “the uneven distribution of researchers”.
@@
 
-Could be, but this bias would show even steeper declines. We clarified the sentence to refer to less human-developed areas.
+Could be, but this bias would show even steeper declines. We clarified the sentence to refer to "human-developed areas".
 
@@
 
-We amoited this as suggested.
+We omited this as suggested.
 

### 2018-06-15 revision/ms.Rmd
Almost there. Asher (now webpage discoverlife cited only), Pauw and Gottelli refs need to be fixed and Scale bar need new colors.

@@
 In the Netherlands, museum data have also revealed coupled plant and pollinator declines [@Scheper2014].
-Specifically, bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus, were most threatened with extinction.
+Specifically, bee species with the strongest host plant preferences (i.e., specialists) tend to display the strongest declines and thus, are most threatened with extinction.
 However, it is important to note that even for these two countries, local estimates of pollinator richness are biased toward large cities and regions dominated by agriculture, and thus lack data for areas subject to less human activity.
@@
 For example, declines of up to 18% in local bumblebee richness have been reported for Belgium and the Netherlands [@Carvalheiro2013].
-In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011] and 70 % of bumblebee species are clasified as threatened or with declining population trends [@Nieto2014].
+In other parts of Europe, local richness declines range from 5% in Great Britain [@Carvalheiro2013] to 42% in Denmark [@Dupont2011] and 70 % of bumblebee species are clasified as threatened or with declining population trends by IUCN [@Nieto2014].
 In the USA, reported bumble declines are also severe with estimates ranging between 25% [@Grixti2009] and 30% [@Bartomeus2013a].
@@
 We selected this arbitrary threshold as a compromise between sample size and the number of countries that can be included in a general analysis.
-However, proper analysis of this dataset and determining the minimum sample size needed per country would require careful inspection of the data, which we outline later with two examples (Spain and New Zealand).
-Most countries fall short in one or both axes.
+However, proper analysis of this dataset, including the determination of the minimum sample size needed per country, would require careful inspection of the data, We outline how this can be achived below using two examples (Spain and New Zealand).
+We show that most countries fall short in one or both axes.
 For example, a variety of countries located in different continents such as Switzerland, Sri Lanka, Nicaragua or Zimbawe have a decent number of recent records, but lack historical collections.
@@
 In contrast, more than 192 countries have less than 1000 records for both time periods, making them poor candidates for analysing long-term pollinator population trends.
-We focus on bees, but similar exploratory analyses can easily be conducted for other taxa.
+We focus here on bees, but similar exploratory analyses can easily be conducted for other taxa.
 
@@
 
-In brief, after cleaning taxonomic names for erratas and synonims using the taxize package [@Chamberlain2018], we checked for sampling completeness for both time periods and compared rarefied species richness for each site before and after 1980 with a paired t-test (sample size pre-1980: X; sample size pre-1980: X; rarefication at 1000 specimens).
+In brief, after cleaning taxonomic names for erratas and synonims using the taxize package [@Chamberlain2018], we checked for sampling completeness for both time periods and compared rarefied species richness for each site before and after 1980 with a paired t-test (mean sample size pre-1980 per site: 16376 specimens; mean sample size pre-1980 per site: 2006 specimens; rarefication at 1000 specimens per site).
 We found that there were a reduced number of species at sites after 1980 (mean difference 20.27 species; 95 percent confidence interval: -1.03, 41.58; t = 2.44, df = 5, P = 0.06).
@@
 ```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
-dat <- read.csv("data/spain_clean_used.csv")
+dat <- read.csv(file = "data/spain_clean_used.csv")
 #Repeat richness and genus
@@
 Therefore, we can use museum records to identify trends in pollinator communities during New Zealand’s more recent history.
-
 We used New Zealand bee collection records gathered from multiple sources, including university, research institute, museum and private collections.
@@
 Further exploration of the New Zealand native bee data raised doubts on collection completeness in records prior to 1970.
-For example, old specimens lack particular genus, and geographically restricted.
+For example, some genus are not represented in old records.
 Hence, we removed these records from further analyses.
@@
 In contrast, native fly richness declined, whereas exotic fly richness increased, although results for these groups were also non-significant (P-values for both groups > 0.05).
-Note that rarefied richness is sensitive to species evenness, so increases in rarefied richness over time may actually indicate increased species evenness and vice-versa for decreased richness.
+Note that rarefied richness is sensitive to species evenness, so increases in rarefied richness over time may actually indicate increased species evenness and vice-versa for decreased richness [Gotelli].
 
@@
 This approach has already been used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016, @Renauld2016] in response to climate and land-use change.
-Finally, plant specimens stored in plant herbaria may contain indirect evidence of pollination declines [60], thus linking pollinator declines with consequences for ecosystem functioning.
+Finally, plant specimens stored in plant herbaria may contain indirect evidence of pollination declines [Pauw], thus linking pollinator declines with consequences for ecosystem functioning.
 
@@
 For most other pollinator taxa and regions throughout the world, we know almost nothing.
-Moving forward, the first step for many taxa will be to identify and describe species [REF].
+Moving forward, the first step for many taxa will be to identify and describe species [@Meier2004].
 Only then can we begin to document pollinator declines.

### 2018-06-15 revision/ms.Rmd
Revised ms.

@@
 ---
-title: Historic collections as a tool for assessing the global pollination crisis
+title: "Historic collections as a tool for assessing the global pollination crisis"
 author: "Bartomeus, I.^1^, Stavert, J.R.^2^, Ward, D.^2,3^, and Aguado, O.^4^"
-csl: data/philb_style.xml
+csl: ../data/philb_style.xml
 output:
+  word_document: default
   html_document: default
-  word_document: default
 editor_options:
   chunk_output_type: console
-bibliography: data/references2.xml
+bibliography: ../data/references2.xml
 ---
@@
 Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Winfree2011].
-However, bees are diverse, with more than 20,000 species currently described worldwide, and often require expert taxonomists for identification.
+However, bees are diverse, with more than 20,000 species currently described worldwide [Asher], and often require expert taxonomists for identification.
 Furthermore, the uneven global distribution of research studies has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
@@
 library(rworldmap)
-map <- read.csv("data/bee_richness.csv")
-size <- read.csv("data/country_area.csv")
+map <- read.csv("../data/bee_richness.csv")
+size <- read.csv("../data/country_area.csv")
 colnames(size) <- c("country", "size")
@@
 #read in data
-declines <- read.csv(file = "data/richness_declines.V2.csv", header = TRUE, sep = ",", stringsAsFactors=FALSE,na.strings=c("","NA"))
+declines <- read.csv(file = "../data/richness_declines.V2.csv", header = TRUE, sep = ",", stringsAsFactors=FALSE,na.strings=c("","NA"))
 declines$percent_richness_change <- as.numeric(declines$percent_richness_change)
@@
 #average richness change
-map.bee.ave <- mapCountryData(map.bee, nameColumnToPlot = "average", colourPalette="topo",
+colorss <- c(heat.colors(9))
+map.bee.ave <- mapCountryData(map.bee, nameColumnToPlot = "average", colourPalette= colorss,
                                   mapTitle = "Average bee richness change", catMethod = "pretty", addLegend = F)
@@
 #average richness change
-map.syrphid.ave <- mapCountryData(map.syrphid, nameColumnToPlot = "average", colourPalette="topo",
+colorss <- c(heat.colors(5), c("#5e977c", "#3d88a5", "#1c79cf"))
+map.syrphid.ave <- mapCountryData(map.syrphid, nameColumnToPlot = "average", colourPalette= colorss,
                        mapTitle = "Average syrphid richness change", catMethod = "pretty", addLegend = F)
@@
 #average richness change
-map.butterfly.ave <- mapCountryData(map.butterfly, nameColumnToPlot = "average", colourPalette="topo",
+colorss <- c(heat.colors(6))
+map.butterfly.ave <- mapCountryData(map.butterfly, nameColumnToPlot = "average", colourPalette=colorss,
                                     mapTitle = "Average butterfly richness change", catMethod = "pretty", addLegend = F)
@@
 ```{r plot rgbif, echo=FALSE, message=FALSE, warning=FALSE}
-lhf <- read.csv("data/rgbif_output.csv", row.names = 1 )
+lhf <- read.csv("../data/rgbif_output.csv", row.names = 1 )
 #lhf
@@
 countries <- lhf[which(log(lhf$post1980) > 7.5 & log(lhf$pre1980) > 7.5),"country"]
-load("data/gbif.RData")
+load("../data/gbif.RData")
 #3) split by year
@@
 #colorss <- c("firebrick", "red", "orangered", "orange", "yellow", "wheat", "green", "blue", "darkblue")
-colorss <- (heat.colors(10))
+colorss <- c(heat.colors(5), c("#5e977c", "#3d88a5", "#1c79cf"))
 map.gbif.plot <- mapCountryData(map.gbif, nameColumnToPlot = "change", colourPalette= colorss,
@@
 ```{r  spain, echo=FALSE, message=FALSE, warning=FALSE}
-dat <- read.csv(file = "data/spain_clean_used.csv")
+dat <- read.csv(file = "../data/spain_clean_used.csv")
 #Repeat richness and genus
@@
 
-d <- read.csv("data/NZBees_allrecords.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+d <- read.csv("../data/NZBees_allrecords.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
 
@@
     #load data
-    d <- read.csv(file = "data/NZ_fly_pollinators.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+    d <- read.csv(file = "../data/NZ_fly_pollinators.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
 
@@
 #load data
-d <- read.csv(file = "data/NZ_fly_pollinators.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
+d <- read.csv(file = "../data/NZ_fly_pollinators.csv", header = TRUE, sep = ",", na.strings=c("","NA"))
 

### 2018-06-20 revision/ms.Rmd
added references

@@
 In contrast, native fly richness declined, whereas exotic fly richness increased, although results for these groups were also non-significant (P-values for both groups > 0.05).
-Note that rarefied richness is sensitive to species evenness, so increases in rarefied richness over time may actually indicate increased species evenness and vice-versa for decreased richness [Gotelli].
+Note that rarefied richness is sensitive to species evenness, so increases in rarefied richness over time may actually indicate increased species evenness and vice-versa for decreased richness [@Gotelli2001].
 
@@
 Finally, museum specimens can be measured to track evolutionary changes by measuring the traits of specimens.
-This approach has already been used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016, @Renauld2016] in response to climate and land-use change.
-Finally, plant specimens stored in plant herbaria may contain indirect evidence of pollination declines [Pauw], thus linking pollinator declines with consequences for ecosystem functioning.
+This approach has already been used to investigate tongue length [@Miller-Struttmann2015] and body size [@Oliveira2016; @Renauld2016] in response to climate and land-use change.
+Finally, plant specimens stored in plant herbaria may contain indirect evidence of pollination declines [@Pauw2011], thus linking pollinator declines with consequences for ecosystem functioning.
 

### 2018-09-11 revision/SupMat3.Rmd
I think this updates sup mat to match the ms.

--- 2018-06-20 revision/ms.Rmd
@@
 output:
+  word_document: default
   html_document: default
-  word_document: default
 editor_options:
@@
 However, there are species-specific examples of historical losses from different parts of the world (e.g., *Bombus dalbhomi*; @Morales2013).
+```{r, include=FALSE}
+#update joinCountryData2Map to avoid printing results
+
+joinCountryData2Map <- function (dF, joinCode = "ISO3", nameJoinColumn = "ISO3V10", 
+    nameCountryColumn = "Country", suggestForFailedCodes = FALSE, 
+    mapResolution = "coarse", projection = NA, verbose = FALSE) 
+{
+    mapWithData <- getMap(resolution = mapResolution)
+    if (!is.na(projection)) 
+        warning("the projection argument has been deprecated, returning Lat Lon, use spTransform from package rgdal as shown in help details or the FAQ")
+    listJoinCodesNew <- c("ISO_A2", "ISO_A3", "FIPS_10_", "ADMIN", 
+        "ISO_N3")
+    listJoinCodesOld <- c("ISO2", "ISO3", "FIPS", "NAME", "UN")
+    listJoinCodes <- c(listJoinCodesOld, listJoinCodesNew)
+    if (joinCode %in% listJoinCodes == FALSE) {
+        stop("your joinCode (", joinCode, ") in joinCountryData2Map() is not one of those supported. Options are :", 
+            paste(listJoinCodes, ""), "\\n")
+        return(FALSE)
+    }
+    joinCodeOld <- joinCode
+    if (joinCode %in% listJoinCodesOld) {
+        joinCode <- listJoinCodesNew[match(joinCode, listJoinCodesOld)]
+    }
+    if (is.na(match(nameJoinColumn, names(dF)))) {
+        stop("your chosen nameJoinColumn :'", nameJoinColumn, 
+            "' seems not to exist in your data, columns = ", 
+            paste(names(dF), ""))
+        return(FALSE)
+    }
+    dF[[joinCode]] <- as.character(dF[[nameJoinColumn]])
+    dF[[joinCode]] <- gsub("[[:space:]]*$", "", dF[[joinCode]])
+    if (joinCode == "ADMIN") {
+        dF$ISO3 <- NA
+        for (i in 1:nrow(dF)) dF$ISO3[i] = rwmGetISO3(dF[[joinCode]][i])
+        joinCode = "ISO3"
+        nameCountryColumn = nameJoinColumn
+    }
+    matchPosnsInLookup <- match(as.character(dF[[joinCode]]), 
+        as.character(mapWithData@data[[joinCode]]))
+    failedCodes <- dF[[joinCode]][is.na(matchPosnsInLookup)]
+    numFailedCodes <- length(failedCodes)
+    numMatchedCountries <- nrow(dF) - numFailedCodes
+    #cat(numMatchedCountries, "codes from your data successfully matched countries in the map\\n")
+    failedCountries <- dF[[nameCountryColumn]][is.na(matchPosnsInLookup)]
+    failedCountries <- cbind(failedCodes, failedCountries = as.character(failedCountries))
+    #cat(numFailedCodes, "codes from your data failed to match with a country code in the map\\n")
+    if (verbose) 
+        print(failedCountries)
+    matchPosnsInUserData <- match(as.character(mapWithData@data[[joinCode]]), 
+        as.character(dF[[joinCode]]))
+    codesMissingFromUserData <- as.character(mapWithData@data[[joinCode]][is.na(matchPosnsInUserData)])
+    countriesMissingFromUserData <- as.character(mapWithData@data[["NAME"]][is.na(matchPosnsInUserData)])
+    numMissingCodes <- length(codesMissingFromUserData)
+    #cat(numMissingCodes, "codes from the map weren't represented in your data\\n")
+    mapWithData@data <- cbind(mapWithData@data, dF[matchPosnsInUserData, 
+        ])
+    invisible(mapWithData)
+}
+```
 
@@
 library(rworldmap)
-map <- read.csv("../data/bee_richness.csv")
+map <- read.csv("../data/bee_richness.csv", sep = ";")
 size <- read.csv("../data/country_area.csv")
@@
 map2 <- joinCountryData2Map(map0, joinCode="NAME", nameJoinColumn = "country",
-                            nameCountryColumn = "distribution")
+                            nameCountryColumn = "distribution", verbose = FALSE)
 #need to use ISO's to make sure we don't carry over NA's due to country misspellings
 #have plotted countries with missing values as grey (shows we have richness for pretty much everywhere)
-par(mfrow = c(2,2))
+par(mfrow = c(3,2), mar = c(2.1, 3.1, 3.1, 2.1))
 map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "white",
@@
 #add a modified legend using the same initial parameters as mapCountryData
-do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5))
-
+do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5, labelFontSize=0.8, legendMar = 4))
+mtext("a)", side = 3, adj = 0.05, 
+    line = -1.3)
 #need to add all the pannels.
@@
 map.bee <- joinCountryData2Map(declines.bee.2, joinCode="NAME", nameJoinColumn = "country",
-                                   nameCountryColumn = "distribution")
+                                   nameCountryColumn = "distribution", verbose = FALSE)
 
@@
                                   mapTitle = "Average bee richness change", catMethod = "pretty", addLegend = F)
-do.call(addMapLegend, c(map.bee.ave, legendLabels="all", legendWidth=0.5))
+do.call(addMapLegend, c(map.bee.ave, legendLabels="all", legendWidth=0.5, labelFontSize=0.8, legendMar = 4))
+mtext("b)", side = 3, adj = 0.05, 
+    line = -1.3)
+
+#syrphid richness map
+map_s <- subset(map0, is.na(syrphid_complete))
+#scatter.smooth(log(map_s$syrphid) ~ log(map_s$size+1))
+m <- lm(log(map_s$syrphid) ~ log(map_s$size+1), na.action = na.exclude) 
+map_s$res <- residuals(m)
+#subset(map0, res < -2)
+map2 <- joinCountryData2Map(map_s, joinCode="NAME", nameJoinColumn = "country",
+                            nameCountryColumn = "distribution", verbose = FALSE)
+#need to use ISO's to make sure we don't carry over NA's due to country misspellings
+#have plotted countries with missing values as grey (shows we have richness for pretty much everywhere)
+map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "white",
+               mapTitle = "Syrphid species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE, mapRegion = "Europe")
+#add a modified legend using the same initial parameters as mapCountryData
+do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5, labelFontSize=0.8, legendMar = 4))
+mtext("c)", side = 3, adj = 0.05, 
+    line = -1.3)
 
@@
 map.syrphid <- joinCountryData2Map(declines.syrphid.2, joinCode="NAME", nameJoinColumn = "country",
-                            nameCountryColumn = "distribution")
+                            nameCountryColumn = "distribution", verbose = FALSE)
 
@@
 map.syrphid.ave <- mapCountryData(map.syrphid, nameColumnToPlot = "average", colourPalette= colorss,
-                       mapTitle = "Average syrphid richness change", catMethod = "pretty", addLegend = F)
-do.call(addMapLegend, c(map.syrphid.ave, legendLabels="all", legendWidth=0.5))
+                       mapTitle = "Average syrphid richness change", catMethod = "pretty", addLegend = F, mapRegion = "Europe")
+do.call(addMapLegend, c(map.syrphid.ave, legendLabels="all", legendWidth=0.5, labelFontSize=0.8, legendMar = 4))
+mtext("d)", side = 3, adj = 0.05, 
+    line = -1.3)
+
+#butterfly richness
+#scatter.smooth(log(map0$buterflies) ~ log(map0$size+1))
+m <- lm(log(map0$syrphid) ~ log(map0$size+1), na.action = na.exclude) 
+map0$res <- residuals(m)
+#subset(map0, res < -2)
+map2 <- joinCountryData2Map(map0, joinCode="NAME", nameJoinColumn = "country",
+                            nameCountryColumn = "distribution", verbose = FALSE)
+#need to use ISO's to make sure we don't carry over NA's due to country misspellings
+#have plotted countries with missing values as grey (shows we have richness for pretty much everywhere)
+map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "white",
+               mapTitle = "Butterfly species richness", numCats = 25, catMethod = "pretty", addLegend = FALSE, mapRegion = "Europe")
+#add a modified legend using the same initial parameters as mapCountryData
+do.call(addMapLegend, c(map3, legendLabels="all", legendWidth=0.5, labelFontSize=0.8, legendMar = 4))
+mtext("e)", side = 3, adj = 0.05, 
+    line = -1.3)
 
@@
 map.butterfly <- joinCountryData2Map(declines.butterfly.2, joinCode="NAME", nameJoinColumn = "country",
-                                     nameCountryColumn = "distribution")
+                                     nameCountryColumn = "distribution", verbose = FALSE)
 
@@
 map.butterfly.ave <- mapCountryData(map.butterfly, nameColumnToPlot = "average", colourPalette=colorss,
-                                    mapTitle = "Average butterfly richness change", catMethod = "pretty", addLegend = F)
-do.call(addMapLegend, c(map.butterfly.ave, legendLabels="all", legendWidth=0.5))
-
+                                    mapTitle = "Average butterfly richness change", catMethod = "pretty", addLegend = F, mapRegion = "Europe")
+do.call(addMapLegend, c(map.butterfly.ave, legendLabels="all", legendWidth=0.5, labelFontSize=0.8, legendMar = 4))
+mtext("f)", side = 3, adj = 0.05, 
+    line = -1.3)
 par(mfrow = c(1,1))
@@
 
-**Figure 1:** Global map showing a) bee species richness per area (Data from www.discoverlife.org) calculated as the residuals of the log-log regression between bee species richness per country and country size.
+**Figure 1:** Global map showing a), c) and e) bee, syrphid and butterfly species richness per area (Data from http://www.discoverlife.org, IUCN and http://www.syrphidae.com/).
+ALl richness levels are calculated as the residuals of the log-log regression between bee species richness per country and country size.
 This correction accounts for the species-area relationship.
@@
 Note that some African countries may have incomplete listed faunas and that Alaska is included with US values.
-Countries with available historical changes in (b) bee, (c) syrphid and (d) butterfly richness within the last 100 years.
+Countries with available historical changes in (b) bee, (d) syrphid and (f) butterfly richness within the last 100 years.
 Warmer colours indicate steeper average declines.
@@
 pchs <- ifelse(lhf$country == "ES" | lhf$country == "NZ", 19, 1)
-plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col, pch = pchs, las = 1)
-temp <- subset(lhf, country %in% c("ES", "NZ"))
-text(x = temp$log_post1980, y = temp$log_pre1980+2 , labels = temp$country, col = "red")
-abline(h = 7.5, v = 7.5)
+#par(mfrow = c(2,2))
+#plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col, pch = pchs, las = 1)
+#temp <- subset(lhf, country %in% c("ES", "NZ"))
+#text(x = temp$log_post1980, y = temp$log_pre1980+2 , labels = temp$country, col = "red")
+#abline(h = 7.5, v = 7.5)
 #36 countries!
@@
 
+#plot all
+par(mfrow = c(2,2),  mar = c(2.1, 3.1, 3.1, 2.1))
+plot(log(lhf$post1980+1), log(lhf$pre1980+1), xlab = "post1980", ylab = "pre1980", col = col, pch = pchs, las = 1)
+temp <- subset(lhf, country %in% c("ES", "NZ"))
+text(x = temp$log_post1980, y = temp$log_pre1980+2 , labels = temp$country, col = "red")
+abline(h = 7.5, v = 7.5)
+mtext("a)", side = 3, adj = 0.05, 
+    line = -1.3)
 boxplot(log(data2$pre1980), log(data2$post1980), names = c("pre1980", "post1980"), ylab = "log rarefied species richness", 
@@
 }
-
+mtext("b)", side = 3, adj = 0.05, 
+    line = -1.3)
 #map
@@
 map.gbif <- joinCountryData2Map(data2, joinCode="ISO2", nameJoinColumn = "countries",
-                            nameCountryColumn = "countries")
+                            nameCountryColumn = "countries", verbose = FALSE)
 
@@
 do.call(addMapLegend, c(map.gbif.plot, legendLabels="all", legendWidth=0.5))
+mtext("c)", side = 3, adj = 0.05, 
+    line = -1.3)
+par(mfrow = c(1,1))
 
@@
 For example, New Zealand has only 28 native bee species [@Donovan2007], which is a fraction of nearby Australia’s c.
-1600 species [@Ascher2015].
+1600 species [@Ascher2018].
 However, New Zealand has a surprisingly high diversity of flies (Diptera), which are important pollinators in many ecosystems [@Newstrom2005].
@@
     abline(m, lty = 2)
-
+mtext("a)", side = 3, adj = 0.05, 
+    line = -1.3)
 #############################################
@@
     abline(m, lty = 2)
-
+mtext("b)", side = 3, adj = 0.05, 
+    line = -1.3)
 ############################################
@@
     abline(m, lty = 2)
-
+mtext("c)", side = 3, adj = 0.05, 
+    line = -1.3)    
     ##########################################################################
@@
     abline(m, lty = 2)
-
+mtext("d)", side = 3, adj = 0.05, 
+    line = -1.3)    
     ##########################################################################
@@
 
+#dirty trick
+est.all$taxa <- ifelse(est.all$taxa == "Fly", "b)     Fly", 
+                                              "a)     Bee")
+
 p <- ggplot(est.all, aes(Var1, estimate_frq, group=taxa))
@@
 p <- p + geom_hline(aes(yintercept=0), linetype="dashed")
-p <- p + geom_vline(data=filter(est.all, taxa=="Bee"), aes(xintercept=5.5), colour="grey")
-p <- p + geom_vline(data=filter(est.all, taxa=="Fly"), aes(xintercept=7.5), colour="grey")
-p <- p + geom_text(data=data.frame(Var1=3,estimate_frq=0.145,taxa="Bee",lab='Exotic'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
-p <- p + geom_text(data=data.frame(Var1=4,estimate_frq=0.145,taxa="Fly",lab='Exotic'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
-p <- p + geom_text(data=data.frame(Var1=16.5,estimate_frq=0.145,taxa="Bee",lab='Native'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
-p <- p + geom_text(data=data.frame(Var1=11,estimate_frq=0.145,taxa="Fly",lab='Native'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
+p <- p + geom_vline(data=filter(est.all, taxa=="a)     Bee"), aes(xintercept=5.5), colour="grey")
+p <- p + geom_vline(data=filter(est.all, taxa=="b)     Fly"), aes(xintercept=7.5), colour="grey")
+p <- p + geom_text(data=data.frame(Var1=3,estimate_frq=0.145,taxa="a)     Bee",lab='Exotic'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
+p <- p + geom_text(data=data.frame(Var1=4,estimate_frq=0.145,taxa="b)     Fly",lab='Exotic'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
+p <- p + geom_text(data=data.frame(Var1=16.5,estimate_frq=0.145,taxa="a)     Bee",lab='Native'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
+p <- p + geom_text(data=data.frame(Var1=11,estimate_frq=0.145,taxa="b)     Fly",lab='Native'), aes(label=lab), parse=TRUE,colour="grey70", size=3)
 p <- p + theme(panel.grid.minor = element_blank(),

### 2018-10-10 revision/ms.Rmd
fix a reference

@@
 Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Winfree2011].
-However, bees are diverse, with more than 20,000 species currently described worldwide [Asher], and often require expert taxonomists for identification.
+However, bees are diverse, with more than 20,000 species currently described worldwide [@Ascher2018], and often require expert taxonomists for identification.
 Furthermore, the uneven global distribution of research studies has resulted in geographical biases in bee decline research [@Martin2012], as well as taxonomic biases toward species that are easier to identify, such as bumblebees [@Archer2014; @DePalma2016].
@@
 map0$res <- residuals(m)
-#subset(map0, res < -2)
+#map0 <- subset(map0, res > -2) #more contrasted plotting 
 map2 <- joinCountryData2Map(map0, joinCode="NAME", nameJoinColumn = "country",
-                            nameCountryColumn = "distribution")
+                            nameCountryColumn = "distribution", verbose = FALSE)
 #need to use ISO's to make sure we don't carry over NA's due to country misspellings
 #have plotted countries with missing values as grey (shows we have richness for pretty much everywhere)
-par(mfrow = c(2,2))
+par(mfrow = c(3,2), mar = c(2.1, 3.1, 3.1, 2.1))
 map3 <- mapCountryData(map2, nameColumnToPlot = "res", missingCountryCol = "white",
